### 2016-10-28 ParisEM.tex
Added .tex and .bib

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amssymb}
\usepackage{makeidx}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{amsfonts,amsmath,amssymb,amsthm}
\usepackage{oldgerm}
\usepackage{mathrsfs}
\usepackage[active]{srcltx}
\usepackage{verbatim}
\usepackage[toc,page]{appendix}
\usepackage{aliascnt}
\usepackage{array}
\usepackage{hyperref}
\usepackage[textwidth=4cm,textsize=footnotesize]{todonotes}
\usepackage{xargs}
\usepackage{cellspace}
\usepackage[Symbolsmallscale]{upgreek}
\usepackage{geometry}
\usepackage{array}
\geometry{top=3.5cm, bottom=3.5cm, left=3.5cm , right=3.5cm}
\usepackage{fancyhdr}
\pagestyle{fancy}

\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{xcolor}
\usepackage{algorithm,algorithmic}

\newcommand{\x}[2]{x_{#1}^{(#2)}}
\newcommand{\w}[2]{w_{#1}^{(#2)}}
\newcommand{\tw}[2]{\tilde{w}_{#1}^{(#2)}}
\newcommand{\p}[2]{\xi_{#1}^{(#2)}}
\newcommand{\tp}[2]{\tilde{\xi}_{#1}^{(#2)}}
\newcommand{\ta}[2]{\tau_{#1}^{(#2)}}
\newcommand{\rmd}{\mathrm{d}}
\newcommand{\eqsp}{\;}
\newcommand{\1}{\mathrm{1}}
\newcommand{\com}[1]{{\color{gray} // #1}}

\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}

\newcounter{hypA}
\newenvironment{hypA}{\refstepcounter{hypA}\begin{itemize}
\item[{\bf H\arabic{hypA}}]}{\end{itemize}}

\begin{document}

\author{Pierre Gloaguen\footnotemark[1] \and Marie-Pierre Etienne\footnotemark[1] \and Sylvain Le {C}orff\footnotemark[2]}

\footnotetext[1]{AgroParistech, UMR MIA 518, F-75231 Paris, France.}
\footnotetext[2]{Laboratoire de Math\'ematiques d'Orsay, Univ. Paris-Sud, CNRS, Universit\'e Paris-Saclay.}

\title{Efficient particle smoother for stochastic differential equations with linear complexity}

\lhead{Gloaguen et al.}
\rhead{Particle smoother for SDE}

\maketitle

\section{Introduction}
It is assumed that the dynamics of the states $(X_t)_{t\ge 0}$ depends on a parameter $\theta\in \Theta$ where $\Theta$ is a compact subspace of $\mathbb{R}^{\ell}$ and that $(X_t)_{t\ge 0}$ is a weak solution to the following Stochastic Differential Equation (SDE) in $\mathbb{R}^d$:
\begin{equation}
\label{eq:target:sde}
X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha_{\theta}(X_t)\rmd t + \rmd W_t\eqsp,
\end{equation}
where $(W_t)_{t\ge 0}$ is a standard Brownian motion. In the following, it is assumed that $\alpha_{\theta}$ is of the form $\alpha_{\theta}(x) = \nabla_x A_{\theta}(x)$ where $A_{\theta}: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function. A popular choice of $A_{\theta}$  is $A_{\theta} = \log \pi_{\theta}/2$ leading to a Langevin diffusion which may be used to approximate a given target distribution $\pi_{\theta}$. See for instance \cite{roberts:tweedie:1996} for conditions under which this diffusion converges exponentially quickly to $\pi_{\theta}$. The states $(X_t)_{t\ge 0}$ are only partially observed at discrete times $(t_k)_{0\le k \le n}$, with $t_0=0$, through the sequence $(Y_k)_{0\le k\le n}$ which is such that $(Y_k)_{0\le k\le n}$ are independent conditional on $(X_t)_{t\ge 0}$ and for all $0\le k\le n$, the conditional distribution of $Y_k$ given $(X_t)_{t\ge 0}$ depends only on $X_k = X_{t_k}$. This distribution has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g_{\theta}(X_k,\cdot)$. For all $1\le k \le n$, the distribution of $X_{k} $ conditional on $X_{k-1}$ has a density $q^{\Delta t_k}_{\theta}(X_k,\cdot)$ with respect to a reference measure $\mu$ on $\mathbb{R}^d$.

The aim of this paper is to estimate $\theta$ given the set of observations $(Y_0,\ldots,Y_n)$. The EM algorithm introduced in \cite{dempster:laird:rubin:1977} 
maximizes the incomplete data log-likelihood $\theta\mapsto \ell_{\theta}^{n}$ defined by
\begin{equation*}
\ell_{\theta}^{n}(Y_{0:n}) = \log\left(\int p_{\theta}(x_{0:n},Y_{0:n})\,\rmd x_{0:n}\right)\eqsp,
\end{equation*}  
where the complete data likelihood $p_{\theta}$ is given by
\begin{equation*}
p_{\theta}(x_{0:n},Y_{0:n}) = \chi_{\theta}(x_0)g_{\theta}(x_0;Y_0)\prod^{n}_{k=1}q^{\Delta t_k}_{\theta}(x_{k-1},x_k)g_{\theta}(x_k,Y_k)\eqsp.
\end{equation*}
Denote by $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{0:n}\right]$  the conditional expectation given $Y_{0:n}$ when the parameter value is $\theta$. The EM algorithm iteratively builds a sequence $\{\theta_{p}\}_{p\ge 0}$ of parameter estimates following the two steps:
\begin{enumerate}
	\item {\bf E-step}: compute $\theta \mapsto Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log p_{\theta}(X_{0:n},Y_{0:n})\middle|Y_{0:n}\right]$ ;
	\item {\bf M-step}: choose $\theta_{p+1}$ as a maximizer of $\theta \mapsto Q(\theta,\theta_{p})$.
\end{enumerate}
It can be shown that under weak assumptions on the model, the sequence $\{\theta_{p}\}_{p\ge 0}$ converges to the set of stationary points of $\theta \mapsto \ell_{\theta}^{n}(Y_{0:n})$. However, there are two main difficulties with applying the EM algorithm to partially observed SDE in general cases:
\begin{enumerate}[-]
\item The E-step requires to compute $q^{\Delta t_k}_{\theta}$ for which there is no analytic expression.
\item The computation of $Q(\theta,\theta_{p})$ relies on conditional expectations of additive functionals of the state sequence given the observations which are not available explicitly.
\end{enumerate}
To overcome these difficulties, \cite{olsson:strojby:2011} proposed to use General Poisson Estimators (GPE) introduced in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,fearnhead:papaspiliopoulos:roberts:2008}  to provide an unbiased estimation of $q^{\Delta t_k}_{\theta}$. These estimators are based on a representation of $q^{\Delta t_k}_{\theta}$ as an expectation under the probability measure of a Brownian bridge. The conditional expectation involved in the computation is approximated in  \cite{olsson:strojby:2011}  using fixed-lag Sequential Monte Carlo methods developed in \cite{olsson:cappe:douc:moulines:2008}. In this paper, we propose to improve this algorithm to obtain faster convergence of the sequence of Monte Carlo EM estimates. First, we introduce a new class off GPE based on the Ozaki discretization of the SDE  \eqref{eq:target:sde}, see \cite{ozaki:1992,shoji:ozaki:1998}. Instead of using fixed-lag techniques which introduce a systematic additional bias  in the approximation of $Q(\theta,\theta_{p})$, we use the new PaRIS algorithm which allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles.

\section{Random weight PaRIS-EM algorithm}
For a given value $\theta_p$ of the parameter, dropping the initial state, the E-step may be written:
\[
Q(\theta,\theta_{p})= \sum_{k=1}^n\mathbb{E}_{\theta_p}\left[\log \left\{q^{\Delta t_k}_{\theta}(X_{k-1},X_k)g_{\theta}(X_k,Y_k)\right\}\middle|Y_{0:n}\right]\eqsp.
\]
We first describe the Monte Carlo method used to approximate $Q(\theta,\theta_{p})$ when $q^{\Delta t_k}_{\theta}$ may be computed. Dropping $\theta$ from the notations, the E-step computes an expectation of the form
\begin{equation}
\label{def:addfunc}
\bar{H}_{n} =  \sum_{k=1}^n\mathbb{E}\left[h_k(X_{k-1},X_k)\middle|Y_{0:n}\right]\eqsp.
\end{equation}
Thus, for $0 \leq k \leq k' \leq n$ and a given measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, we need to compute an approximation of the smoothed expectation of the states $(X_k,\ldots,X_{k'})$ defined by:
\[
\phi_{k:k'|n}[h] := \mathbb{E}\left[h(X_k,\ldots,X_{k'})\middle|Y_{0:n}\right]\eqsp.
\]
The filtering distributions are defined, for $0\le k\le n$ by $\phi_{k} = \phi_{k:k|k}$. In this paper, we are interested in a particular class of Sequential Monte Carlo methods to approximate $\bar{H}_{n}$ known as the PaRIS algorithm introduced in \cite{olsson:westerborn:2016}.

The PaRIS algorithm is based on the first Sequential Monte Carlo algorithms introduced in  \cite{gordon:salmond:smith:1993} and \cite{kitagawa:1996}. 
They rely on importance sampling and importance resampling steps to approximate the filtering distribution $\phi_{k}$ (conditional distribution of $X_k$ given $Y_{0:k}$) using  a set of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$. At time $k = 0$, $N$ particles $\{\xi^{\ell}_0\}_{\ell=1}^N$ are initialized by setting $\xi^{\ell}_0=x_0$. Then, $\xi^{\ell}_0$ is associated with the importance weights $\omega_0^{\ell} = g_0 \left(\xi^{\ell}_0\right)$. Then, for any bounded and measurable function $h$ defined on $\mathbb{R}^d$, the expectation $\phi_{0}[h] $ is approximated by
\[
\phi^N_{0}[h] = \frac{1}{\Omega_0^N} \sum_{\ell=1}^N \omega_0^{\ell} h \left(\xi^{\ell}_0 \right)\eqsp, \quad \Omega_0^N:= \sum_{\ell=1}^N \omega_0^{\ell}\eqsp.
\]
Then, using $\{(\xi^{\ell}_{k-1},\omega^{\ell}_{k-1})\}_{\ell=1}^N$, pairs $\{(I^{\ell}_t,\xi^{\ell}_{t})\}_{\ell=1}^N$ of indices and particles are sampled and weighted using an instrumental transition density $p_k$ on $\mathbb{R}^d\times \mathbb{R}^d$ and an adjustment multiplier function $\vartheta_k$ on $\mathbb{R}^d$. Each new particle $\xi^{\ell}_{k}$ and weight $\omega^{\ell}_k$ at time $k$ are computing following these steps:
\begin{enumerate}[-]
\item choose a particle index $I^{\ell}_k$ at time $k-1$ in $\{1,\ldots,N\}$ with probabilities proportional to $\omega_{k-1}^{j} \vartheta_k (\xi^{j}_{k-1})$, for $j$ in $\{1,\ldots,N\}$ ;
\item sample  $\xi^{\ell}_{k}$ using this chosen particle according to $\xi^{\ell}_{k} \sim p_k(\xi^{I^{\ell}_k}_{k-1},\cdot)$.
\item $\xi^{\ell}_k$ is associated with:
\[
\omega^{\ell}_k := \frac{q^{\Delta t_{k}}(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_k (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
\]
\end{enumerate} 
The expectation $\phi_{k}[h]$ is approximated by
\[
\phi^N_{k}[h] := \frac{1}{\Omega_t^N} \sum_{\ell=1}^N \omega_k^{\ell} h \left(\xi^{\ell}_k \right)\eqsp,\quad\Omega_k^N:= \sum_{\ell=1}^N \omega_k^{\ell}\eqsp.
\]
The PaRIS algorithm uses the same decomposition as the Forward Filtering Backward Smoothing (FFBS) algorithm introduced in \cite{doucetgodsillandrieu:2000} and the Forward Filtering Backward Simulation (FFBSI) algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions. Both the FFBS and the FFBSi algorithms require first that a forward pass has produced a set of particles and importance weights $\{ (\xi^{\ell}_k, \omega^{\ell}_k)\}$, for $\ell \in \{1, \ldots, N\}$ and  $k \in \{0,\ldots,n\}$ to approximate $\phi_{k}$ for $k=0$ to $k=n$. The FFBS algorithm stores all particles and weights without taking into consideration the genealogy of the particles. Then, a backward pass is performed, keeping all the particles fixed but modifying all importance weights computed during the forward pass. This algorithm has a computational complexity of order $\mathcal{O}(N^{k'-k+1})$ to approximate joint smoothing distributions of the form $\phi_{k:k'|n}$ with $k<k'$ which makes it prohibitive. An important feature of the FFBS algorithm shown in \cite{mongillo:deneve:2008,cappe:2011,delmoral:doucet:singh:2010} is that it can be implemented using only a forward pass when approximating smoothed expectations of additive functions. 
%In more general case, \cite{dubarry:lecorff:2013} are interested in the FFBS algorithm for an additive function $h$ defined on $\mathsf{X}^{T+1}$,
%\[
%h(x_{0:T}) := \sum_{t=1}^{T} h_t (x_{t-1},x_t)
%\]
%where $\{h_t\}_{t=1}^{T}$ are bounded on $\mathsf{X}^2$. See \cite{dubarry:lecorff:2013} for their proposed FFBS algorithm in this case. As shown in \cite{delmoral:doucet:singh:2010} when applied to additive functionals, the computational complexity grows quadratically with number of particles. In the backward pass of the FFBS algorithm, the importance weights are updated while the particles are kept fixed. 
%\subsubsection*{FFBSi algorithm}
%As for the FFBS algorithm, the FFBSi algorithm uses a forward pass to produce $\{ (\xi^{\ell}_k, \omega^{\ell}_k)\}$, for $\ell \in \{1, \ldots, N\}$ and  $k \in \{0,\ldots,n\}$ to approximate $\phi_{k}$ for $k=0$ to $k=n$ and all these particles and weights are stored  without taking into consideration the genealogy of the particles. 
Instead of computing new weights in the backward pass, the FFBSi  samples backward new trajectories from time $n$ to time $0$ among all the $N^{n+1}$ trajectories which can be chosen in $\{\xi^{\ell}_k\}$, $1\le \ell\le N$ and $0\le k\le n$. 
%Each trajectory is sampled according to the following steps.
%\begin{enumerate}[-]
%\item At time $k = n$, sample $J_n\in\{1\ldots,N\}$ with probabilities proportional to $\omega^{\ell}_n/\Omega_n^N$, for $\ell\in\{1, \ldots, N\}$.
%\item For $k = n-1$ to $k=0$, $J_k$ is sampled in $\{1\ldots,N\}$ with probabilities 
%\[
%\Lambda_k^N(J_{k+1},i) = \frac{\omega^{i}_k q(\xi^{i}_k,\xi^{J_{k+1}}_{k+1})}{\sum_{\ell=1}^N\omega^{\ell}_k q(\xi^{\ell}_k,\xi^{J_{k+1}}_{k+1})}\eqsp,\quad 1\le i\le N\eqsp.
%\]
%\end{enumerate}
%These sampling steps are repeated $N$ times to produce $\{J_0^\ell,\ldots,J_n^{\ell}\}$, $1\le \ell \le N$. Then, the smoothed expectation of any measurable function $h$ on $(\mathbb{R}^d)^{n+1}$  $\mathbb{E}\left[h(X_{0:n})\middle|Y_{0:n}\right]$ is approximated by $N^{-1} \sum_{\ell = 1}^N  h(\tilde{\xi}^{J_0^{\ell}}_{0},\ldots, \tilde{\xi}^{J_n^{\ell}}_{n})$.
Contrary to the FFBS algorithm, the FFBSi algorithm samples trajectories and may be used to approximate joint smoothing distributions without increasing its complexity. Nevertheless, the computation of the transition probabilities to select particles in the backward pass makes it a $\mathcal{O}(N^2)$ algorithm. In the case where the transition probability $q$ is upper bounded, \cite{douc:garivier:moulines:olsson:2011} introduced an acceptance rejection mechanism to implement the FFBSi algorithm with a $\mathcal{O}(N)$ complexity, see also \cite{dubarry:lecorff:2011}.

The PaRIS algorithm combines both the forward only version of the FFBS algorithm with the sampling mechanism of the FFBSi algorithm. The algorithm does not produce an approximation of the smoothing distributions but of the smoothed expectation of a fixed additive functional and thus  may be used to approximate \eqref{def:addfunc}. Its crucial property is that it does not require a backward pass, the smoothed expectations is computed on-the-fly with the particle filter and no storage of the particles or weights is needed. The algorithm is based on sufficient statistics $\tau^i_k$, for $1\le i\le N$ and $0\le k \le n$ starting with $\tau^i_0 = 0$ for all $1\le i\le N$. Let $\tilde{N}\ge 1$. Then, at each time step $1\le k \le n$ these statistics are update according to the following steps.
\begin{enumerate}[(i)]
\item \label{it:PaRIS:filt} Run one step of a particle filter to produce $\{(\xi^{\ell}_k, \omega^{\ell}_k)\}$ for $1\le \ell \le N$.
\item \label{it:PaRIS:sampleindex} For all $1\le i \le N$, sample independently $J_k^{i,\ell}$ in $\{1,\ldots,N\}$ for $1\le \ell \le \widetilde N$ with probabilities $\Lambda_{k-1}^N(i,\cdot)$, given by
\[
\Lambda_{k-1}^N(i,\ell) = \frac{\omega^{\ell}_{k-1} q^{\Delta t_k}(\xi^{\ell}_{k-1},\xi_k^{i})}{\sum_{\ell=1}^N\omega^{\ell}_{k-1} q^{\Delta t_k}(\xi^{\ell}_{k-1},\xi_k^{i})}\eqsp,\quad 1\le \ell\le N\eqsp.
\]
\item \label{it:PaRIS:smooth} Set
\[
\tau^{i}_{k} := \frac{1}{\widetilde{N}} \sum^{\widetilde{N}}_{\ell=1} \left\{ \tau^{J_k^{i,\ell}}_{k-1} + h_{k} \left(\xi^{J_k^{i,\ell}}_{k-1}, \xi^{i}_{k}\right)  \right\}\eqsp.
\]
\end{enumerate}
Then, \eqref{def:addfunc} is approximated by
\[
H_n^N = \frac{1}{\Omega_n^N}\sum_{i=1}^N \omega^{i}_n \tau_n^i\eqsp.
\] 
As proved in \cite{olsson:westerborn:2016}, the algorithm is asymptotically consistent (as $N$ goes to infinity) for any fixed precision parameter $\tilde N$. However, there is a significant qualitative difference between the cases $\tilde{N} = 1$ and $\tilde{N} \geq 2$.

As for the FFBSi algorithm,  when there exists $0<q^{\Delta t_k}<\sigma_+$, the PaRIS algorithm may be implemented with $\mathcal{O}(N)$ complexity using the procedure of \cite{douc:garivier:moulines:olsson:2011}.
As $q^{\Delta t_k}(\xi^{\ell}_{k-1},\xi_k^{i})$ is not available explicitly for all $1\le i,\ell\le N$, the next section introduces GPE to replace it by $\widehat{q}^{\Delta t_k}(\xi^{\ell}_{k-1},\xi_k^{i};\Lambda)$, where $\Lambda$ is a random variable such that
\[
\mathbb{E}\left[\widehat{q}^{\Delta t_k}(\xi^{\ell}_{k-1},\xi_k^{i};\Lambda)\middle| \mathcal{F}_k^N\right] = q^{\Delta t_k}(\xi^{\ell}_{k-1},\xi_k^{i})\eqsp,
\]
where 
\[
\mathcal{F}_k^N = \sigma\left\{Y_{0:k};(\xi^{\ell}_{u},\omega^{\ell}_{u}), 1\le \ell\le N, 1\le u\le k\right\}\eqsp.
\]
In the case where $\widehat{q}^{\Delta t_k}$ may be upper bounded by $\sigma_+$, step~\eqref{it:PaRIS:sampleindex} is replaced by the acceptance-rejection mechanism given in Algorithm~\ref{alg:AR}. It is proved in Lemma~\ref{lem:AR:unbiased} that replacing $\widehat{q}^{\Delta t_k}$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.

\begin{algorithm}
\caption{Random weight Accept-reject-based backward sampling  (\cite{douc:garivier:moulines:olsson:2011})}
\begin{algorithmic}
\FORALL{$i \in 1,\dots,\widetilde N$}
\STATE Set $L =\{1,\ldots,N\}$. 
\WHILE{$L\neq\emptyset$}
\STATE $p = \#L$.
\STATE Sample $I = (I_1,\ldots,I_p)$ independently in $\{1,\ldots,N\}$ with probabilities proportional to $\omega_{k-1}^{\ell}$, $1\le \ell\le N$.
\STATE Sample $U = (U_1,\ldots,U_p)$ independently and independently of $I$ uniformly in $[0,1]$.
\STATE Sample $\Lambda = (\Lambda_1,\ldots,\Lambda_p)$ independently and independently of $I$ and $U$.
\STATE Set $L_p = \emptyset$.
\FORALL{$\ell \in 1,\dots,p$}
\IF{$U_\ell<\widehat{q}(\xi_{k-1}^{I_\ell},\xi_k^{L(\ell)};\Lambda_{\ell})/\sigma_+$}
\STATE $J_k^{L(\ell),i} = I_\ell$.
\ELSE
\STATE $L_p = L_p \cup \{L(\ell)\}$.
\ENDIF
\ENDFOR
\STATE $L = L_p$.
\ENDWHILE
\ENDFOR
\end{algorithmic}
\label{alg:AR}
\end{algorithm}

\section{Ozaki-PaRIS algorithm}
\label{sec:PaRIS:SDE}
In general situations, the PaRIS algorithm cannot be used for stochastic differential equations as $q^{\Delta t_k}_{\theta}$ is unknown. Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio in Algorithm~\ref{alg:AR} cannot be performed to apply the PaRIS algorithm. Following \cite{olsson:strojby:2011},
an extension of the PaRIS algorithm may be developed by replacing $q^{\Delta t_k}_{\theta}$  by an unbiased estimator. In \cite{olsson:strojby:2011},  GPE are used to estimate $q^{\Delta t_k}_{\theta}$ which relies on a representation of $q^{\Delta t_k}_{\theta}$ as an expectation under the probability measure of a Brownian bridge. In order to improve the Monte Carlo estimation of $q^{\Delta t_k}_{\theta}$,  other diffusion bridges may be used as long as they can be simulated efficiently. A simple idea is to consider a diffusion obtained as an approximation of the target SDE \eqref{eq:target:sde}. \cite{ozaki:1992,shoji:ozaki:1998} proposed a higher order approximation of the drift term $\alpha_{\theta}$ than the Euler scheme. The authors used a linear approximation of the diffusion drift $\alpha_{\theta}$, together with a constant approximation of the volatility over each time interval $(t_{k-1},t_k)$. Following \cite{ozaki:1992,shoji:ozaki:1998}, for all $1\le k\le n$, let $(Y^{k}_t)_{t\ge 0}$ be governed by the following stochastic differential equation for $t\in(t_{k-1},t_k)$:
\begin{equation}
\label{eq:SDE:Ozaki}
Y^k_{t_{k-1}} = X_{k-1}\quad\mbox{and}\quad\rmd Y^k_t = \widetilde{\alpha}^k_{\theta}(Y^k_t)\rmd t + \rmd W_t\eqsp,
\end{equation}
where
\[
\widetilde{\alpha}^k_{\theta}(x) = J_{\alpha_{\theta}}(X_{k-1})x + \alpha_{\theta}(X_{k-1}) - J_{\alpha_{\theta}}(X_{k-1})X_{k-1}
\]
and $J_{\alpha_{\theta}}$ is the Jacobian of $\alpha_{\theta}$. The probability density function of the transition kernel associated with $Y^k$ is written $\widetilde{q}^{\Delta t}_{\theta}$ and is given in Section~\ref{sec:ornstein:bridge}.

Writing $\mathbb{Q}^k_{\theta}$ (resp. $\mathbb{W}^{k}_{\theta}$) the probability measure induced by $(X_s)_{t_{k-1}\le s\le t_k}$  (resp. $(Y_s)_{t_{k-1}\le s\le t_k}$) on $(\mathsf{C},\mathcal{C})$, Girsanov theorem yields:
\[
\frac{\rmd \mathbb{Q}^k_{\theta}}{\rmd \mathbb{W}^{k}_{\theta}}(w) = \mathrm{exp}\left\{H_{\theta}(\omega_{t_k}) - H_{\theta}(\omega_{t_{k-1}}) - \frac{1}{2}\int_{t_{k-1}}^{t_k}\left\{\|\alpha_{\theta}(\omega_s)-\widetilde{\alpha}_{\theta}(\omega_s)\|^2+\Delta H_{\theta}(\omega_s)\right\}\rmd s\right\}\eqsp,
\]
where 
\[
H_{\theta}(x) = A_{\theta}(x) - x'J_{\alpha_{\theta}}(X_{k-1})x/2- (\alpha_{\theta}(X_{k-1}) - J_{\alpha_{\theta}}(X_{k-1})X_{k-1})'x
\]
and $\Delta$ is the Laplace operator. Then, the probability measures conditioned on hitting $X_{k}$ at time $t_k$ satisfy:
\begin{multline}
\label{eq:Girsanov}
\frac{\rmd \mathbb{Q}^{k,X_{k}}_{\theta}}{\rmd \mathbb{W}^{k,X_{k}}_{\theta}}(w) = \frac{\widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_{k})}{q^{\Delta t_k}_{\theta}(X_{k-1},X_{k})}\mathrm{exp}\left\{H_{\theta}(X_k) - H_{\theta}(X_{k-1})\right\} \\
\times \mathrm{exp}\left\{- \frac{1}{2}\int_{t_{k-1}}^{t_k}\left\{\|\alpha_{\theta}(\omega_s)-\widetilde{\alpha}_{\theta}(\omega_s)\|^2+\Delta H_{\theta}(\omega_s)\right\}\rmd s\right\}\eqsp.
\end{multline}
Consider the following asumptions on the model:
\begin{enumerate}[(a)]
\item For all $\theta\in\Theta$ and all $0\le k \le n-1$, $x\mapsto \alpha_{\theta}(x) - \alpha_{\theta}(X_k) - J_{\alpha_{\theta}}(X_{k})(x-X_k)$ is bounded.
\item For all $\theta\in\Theta$ and all $0\le k \le n-1$, $x\mapsto \mathrm{Tr}\left(J_{\alpha_{\theta}}(x) - J_{\alpha_{\theta}}(X_k)\right)$ is bounded, where $\mathrm{Tr}$ is the trace operator.
\end{enumerate}
Under these two assumptions, for all $\theta$, there exists $\ell(\theta)$ such that:
\[
\mathrm{inf}_{u\in \mathbb{R}^d} \left\{\frac{1}{2}\left(\|\alpha_{\theta}(u)-\widetilde{\alpha}_{\theta}(u)\|^2+\Delta H_{\theta}(u)\right)\right\}\ge \ell(\theta)\eqsp.
\]
Define
\[
\phi_{\theta}(x) = \frac{1}{2}\left\{\|\alpha_{\theta}(x)-\widetilde{\alpha}_{\theta}(x)\|^2+\Delta H_{\theta}(x)\right\}-\ell(\theta)\eqsp.
\]
The assumptions on the model also ensure that for all $\theta\in\Theta$, there exists $\Lambda(\theta)$ such that  $\mathrm{sup}_x\eqsp \phi_{\theta}(x)\le \Lambda_{\theta}$. Then,
\begin{multline*}
q^{\Delta t_k}_{\theta}(X_{k-1},X_k) = \widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k) \eqsp\mathrm{exp}\left\{H_{\theta}(X_k) - H_{\theta}(X_{k-1})-\ell(\theta)\Delta t_k\right\}\\
\times\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\right]\eqsp,
\end{multline*}
where $\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}$ is the expectation under the law of the bridge process defined by \eqref{eq:SDE:Ozaki}, starting at $X_{k-1}$ at time $0$ and ending at $X_{k}$ at time $\Delta t_k$.
An unbiased estimator of $q^{\Delta t_k}_{\theta}(X_{k-1},X_k)$ is then obtained by estimating $\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\right]$. 
Following \cite{}, if $\omega \sim\mathbb{W}_{\theta}^{k,X_{k}}$, $\kappa$ is a random variables taking values in $\mathbb{N}$ with distribution $\mu$ and if $(U_j)_{1\le j\le \kappa}$ are independent uniform random variables on $[0,t]$, 
\begin{align*}
\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\right] &= \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\right]\eqsp,\\
&= \mathrm{exp}\left\{-\Lambda_{\theta}\Delta t_k\right\}\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathrm{exp}\left\{\int_{0}^{\Delta t_k}(\Lambda_{\theta}-\phi_{\theta}(\omega_s))\rmd s\right\}\right]\eqsp,\\
&=\mathrm{exp}\left\{-\Lambda_{\theta}\Delta t_k\right\}\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\sum_{p\ge 0}\frac{1}{p!}\left(\int_{0}^{\Delta t_k}\Lambda_{\theta}-\phi_{\theta}(\omega_s)\rmd s\right)^p\right]\eqsp,\\
&=\mathrm{exp}\left\{-\Lambda_{\theta}\Delta t_k\right\}\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathbb{E}\left[\sum_{p\ge 0}\frac{(\Delta t_k)^p}{k!}\prod_{j=1}^p\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\right]\right]\eqsp,\\
&=\mathrm{exp}\left\{-\Lambda_{\theta}\Delta t_k\right\}\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\frac{(\Delta t_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\right]\eqsp.
\end{align*}
Interesting choices of $\mu$ are discussed in \cite{} and implemented in Section~\ref{sec:exp}. Writing
\[
\Lambda_k = \left\{\kappa,\omega,U_1,\ldots,U_\kappa\right\}\eqsp,
\]
Algorithm~\ref{alg:AR} is used with 
\begin{multline*}
\widehat{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k;\Lambda) = \widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k) \eqsp\mathrm{exp}\left\{H_{\theta}(X_k) - H_{\theta}(X_{k-1})-\ell(\theta)\Delta t_k\right\}\\ 
\times\mathrm{exp}\left\{-\Lambda_{\theta}\Delta t_k\right\}\frac{(\Delta t_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\eqsp.
\end{multline*}
The choice of $\mu$ is restricted to distributions such that the associated estimator is uper bounded by an eplicit constant $\sigma_+^k(\theta)$. Then, for all $1\le k \le n$, $1\le \ell\le N$, the importance weight $\omega_k^{\ell}$ is replaced by the unbiased estimator:
\begin{equation}
\label{eq:random:weight}
\widetilde{\omega}_k^{\ell} = \frac{M^{-1}\sum_{i=1}^M\widehat{q}_{\theta}^{\Delta t_{k}}(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k;\Lambda_k^i)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_k (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp,
\end{equation}
where $\{\Lambda_k^i\}_{i=1}^M$ are independent copies of $\Lambda_k$. The last step to approximate the intermediate quantity of the EM algorithm is to provide an unbiased estimate of $\log q^{\Delta t_k}_{\theta}(X_{k-1},X_k)$. By Girsanov theorem, this log density can be written as:
\begin{multline*}
\log q^{\Delta t_k}_{\theta}(X_{k-1},X_k) = \log \widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k) + H_{\theta}(X_k) - H_{\theta}(X_{k-1})-\ell(\theta)\Delta t_k \\
-\mathbb{E}_{\mathbb{Q}_{\theta}^{k,X_{k}}}\left[\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right]\eqsp.
\end{multline*}
If $U$ is uniformy distributed on $(0,\Delta t_k)$ and if $w \sim \mathbb{Q}_{\theta}^{k,X_{k}}$, $\log q^{\Delta t_k}_{\theta}(X_{k-1},X_k)$ may then be estimated by:
\begin{equation} 
\label{eq:logq}
\mathsf{q}_{\theta}^{\Delta t_k}(X_{k-1},X_k,\omega_U) = \log \widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k) + H_{\theta}(X_k) - H_{\theta}(X_{k-1})-\ell(\theta)\Delta t_k - \Delta t_k \phi_{\theta}(\omega_U)\eqsp.
\end{equation}
Sampling $w \sim \mathbb{Q}_{\theta}^{k,X_{k}}$ is possible as by \eqref{eq:Girsanov},
\[
\frac{\rmd \mathbb{Q}^{k,X_{k}}_{\theta}}{\rmd \mathbb{W}^{k,X_{k}}_{\theta}}(w) \propto 
\mathrm{exp}\left\{-\int_{t_{k-1}}^{t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\eqsp.
\]
As noted by \cite{}, this importance ratio is proportional to the probability that a marked Poisson process on $(0,\Delta t_k)\times (0,1)$ with intensity $\Lambda_{\theta}$ is below the graph $s \mapsto \phi_{\theta}(\omega_s)/\Lambda_{\theta}$. $\omega_U$ in \eqref{eq:logq} is then obtaine by first sampling $\tilde{\omega}\sim \mathbb{W}^{k,X_{k}}$ at times on $(0,\Delta t_k)$ defined by the marked Poison process (which is accepted if is is below $\phi_{\theta}(\omega_s)/\Lambda_{\theta}$) and then by interpolating the bridge at time $U$.

All these steps are summarized in Algorithm~\ref{alg:Ozaki:PaRIS} which provides an approximation of the E step of the EM algorithm.
\begin{algorithm}
\caption{Ozaki-PaRIS algorithm}
\begin{algorithmic}
\FORALL{$i \in 1,\dots, N$}
\STATE Set $\xi_0^i = X_0$, $\omega_0^i = 1/N$ and $\tau_0^i = 0$.
\ENDFOR
\FORALL{$k \in 1,\dots, n$}
\FORALL{$\ell \in 1,\dots, N$}
\STATE Choose $I_k^{\ell}$ in $\{1,\ldots,N\}$ with probabilities propotional to $\widetilde{\omega}_{k-1}^j\vartheta_k(\xi_{k-1}^j)$.
\STATE Sample $\xi_k^{\ell} \sim p_k(\xi_{k-1}^{I_k^{\ell}},\cdot)$.
\STATE Sample $\kappa \sim \mu$, $\omega\sim \mathbb{W}_{\theta}^{k,X_{k}}$ and $(U_j)_{1\le j\le \kappa}$ uniformly on $(0,\Delta t_k)$.
\STATE Compute $\widetilde{\omega}^{\ell}_k$ as in \eqref{eq:random:weight}.
\ENDFOR
\FORALL{$i \in 1,\dots,\widetilde N$}
\STATE Set $L =\{1,\ldots,N\}$. 
\WHILE{$L\neq\emptyset$}
\STATE $p = \#L$.
\STATE Sample $(I_1,\ldots,I_p)$ in $\{1,\ldots,N\}$ with probabilities prop. to $\omega_{k-1}^{\ell}$.
\STATE Sample $(V_1,\ldots,V_p)$ uniformly in $[0,1]$.
\STATE $L_p = \emptyset$
\FORALL{$\ell \in 1,\dots,p$}
\STATE Sample $\kappa_{\ell}\sim \mu$.
\STATE Sample $U^{\kappa_{\ell}}_j$, $1\le j\le \kappa_{\ell}$, independently and uniformly on $(0,\Delta t_k)$.
\STATE Sample $\omega^{\kappa_{\ell}}_{s}$ at times $U^{\kappa_{\ell}}_j$, $1\le j\le \kappa_{\ell}$, where $\omega^{\kappa_{\ell}}\sim \mathbb{W}_{\theta,\xi_{k-1}^{I_\ell}}^{k,\xi_k^{L(\ell)}}$.
\IF{$V_\ell<\widetilde{q}^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_\ell},\xi_k^{L(\ell)};\kappa_{\ell},\omega^{\kappa_{\ell}},U^{\kappa_{\ell}}_1,\ldots,U^{\kappa_{\ell}}_{\kappa_{\ell}})/\sigma_+^k(\theta)$}
\STATE $J_k^{L(\ell),i} = I_\ell$.
\ELSE
\STATE $L_p = L_p \cup \{L(\ell)\}$.
\ENDIF
\ENDFOR
\STATE $L = L_p$
\ENDWHILE
\ENDFOR
\FORALL{$i \in 1,\dots,N$}
\STATE Set $\tau_k^i = 0$.
\FORALL{$\ell \in 1,\dots,\widetilde{N}$}
\REPEAT
\STATE $\kappa \sim \mathcal{P}(\Lambda_{\theta}\Delta t_k)$, $(U_{m},\chi_m$) uniformly on $(0,\Delta t_k)\times(0,1)$, $1\le m \le \kappa$.
\STATE Sample $\widetilde{\omega}_s\sim \mathbb{W}_{\theta}^{k,X_{k}}$ at times $U_{m}$, $1\le m \le \kappa$.
\UNTIL{$\phi_{\theta}(\widetilde{\omega}_{U_m})/\Lambda_{\theta}<\chi_m$ for all $1\le m \le \kappa$}
\STATE Sample $U$ uniformly in $(0,\Delta t_k)$.
\STATE $\tau_k^i = \tau_k^i + \tau^{J_k^{i,\ell}}_{k-1} + \log g_{\theta}(\xi^i_k,Y_k) + \mathsf{q}_{\theta}^{\Delta t_k}(\xi^{J_k^{i,\ell}}_{k-1}, \xi^{i}_{k},\widetilde{\omega}_U)$.
\ENDFOR
\STATE Set $\tau^{i}_{k} = \tau^{i}_{k}/\widetilde{N}$.
\ENDFOR
\ENDFOR
\end{algorithmic}
\label{alg:Ozaki:PaRIS}
\end{algorithm}
%\log \left\{q^{\Delta t_k}_{\theta}(X_{k-1},X_k)g_{\theta}(X_k,Y_k)

\section{Convergence results}
Consider the following assumptions.
\begin{hypA}
\label{assum:boundmodel}
\begin{enumerate}[(i)]
\item For all $k \geq 0$ and  all $x\in \mathbb{R}^d$, $g_{k}(x) >0$.
\item $\underset{k\geq 0}{\sup}|g_{k}|_{\infty} < \infty$.
\end{enumerate}
\end{hypA}

\begin{hypA}
\label{assum:boundalgo}
$\underset{k\geq 1}{\sup}|\vartheta_k|_{\infty} < \infty$, $\underset{k\geq 1}{\sup}|p_k|_{\infty} < \infty$ and $\underset{k\geq 1}{\sup}|\widetilde{\omega}_{k}|_{\infty} < \infty$.
\end{hypA}
The following Lemma is proved in \cite[Lemma~11]{olsson:westerborn:2016}.
\begin{lemma}
\label{lem:smooth:rec}
Forward smoothing
\end{lemma}

\begin{lemma}
\label{lem:iid}
For all $0\le k \le n-1$, the random variables $\{\widetilde{\omega}_{k+1}^i\tau_{k+1}^i\}_{i=1}^N$ are independent conditionally on $\mathcal{F}_k^{N}$ and%For all bounded measurable function $h$ on $\mathbb{R}^d$,
\[
\mathbb{E}\left[\tilde{\omega}^1_{k+1}\tau^{1}_{k+1}\middle| \mathcal{F}_k^{N}\right] = \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\int q_{\theta}^{\Delta t_{k+1}}(\cdot,x)g_{k+1}(x)\left\{\tau_k(\cdot) + h_{k+1}(\cdot,x)\right\}\rmd x\right]\eqsp.
\]
\end{lemma}

\begin{proof}
By \eqref{eq:random:weight},
\[
\mathbb{E}\left[\tilde{\omega}^1_{k+1}\tau^{1}_{k+1}\middle| \mathcal{F}_k^{N}\right] = \mathbb{E}\left[\frac{M^{-1}\sum_{i=1}^M\widehat{q}_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{I^{1}_{k+1}},\xi^{1}_{k+1};\Lambda^i_{k+1})g_{k+1}(\xi^{1}_{k+1})}{\vartheta_{k+1}(\xi^{I^{1}_{k+1}}_{k}) p_{k+1}(\xi_{k}^{I^{1}_{k+1}},\xi^{\ell}_{k+1})}\tau^{1}_{k+1}\middle| \mathcal{F}_k^{N}\right]\eqsp.
%&= \mathbb{E}\left[\frac{g_{k+1}(\xi^{1}_{k+1})}{\vartheta_{k+1}(\xi^{I^{1}_{k+1}}_{k}) p_{k+1}(\xi_{k}^{I^{1}_{k+1}},\xi^{\ell}_{k+1})}\tau^{1}_{k+1}\widehat{q}_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{I^{1}_{k+1}},\xi^{1}_{k+1};\Lambda_{k+1})\middle| \mathcal{F}_k^{N}\right]\eqsp,
\]
Note that
\begin{align*}
&\mathbb{E}\left[\tau^{1}_{k+1}\middle|\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,\omega_{k+1}^i\}\right)\right]
 = \sum_{\ell=1}^N\frac{\omega_k^{\ell}q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{\ell},\xi^{1}_{k+1})\left(\tau^{\ell}_k + h_{k+1}(\xi_{k}^{\ell},\xi^{1}_{k+1})\right)}{\sum_{\ell'=1}^N\omega_k^{\ell'}q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{\ell'},\xi^{1}_{k+1})}\eqsp,\\
&\mathbb{E}\left[M^{-1}\sum_{i=1}^M\widehat{q}_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{I^{1}_{k+1}},\xi^{1}_{k+1};\Lambda^i_{k+1})\middle|\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,\omega_{k+1}^i\}\right)\right]
 = q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{I^{1}_{k+1}},\xi^{1}_{k+1})\eqsp.
\end{align*}
Since $\tau^{1}_{k+1}$ and $\{\Lambda^i_{k+1}\}_{i=1}^M$ are independent conditionally to $\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^j,I_{k+1}^j\}_{j=1}^N\right)$:
\begin{multline*}
\mathbb{E}\left[\tau^{1}_{k+1}M^{-1}\sum_{i=1}^M\widehat{q}_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{I^{1}_{k+1}},\xi^{1}_{k+1};\Lambda^i_{k+1})\middle|\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,\omega_{k+1}^i\}\right)\right]\\
 = q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{I^{1}_{k+1}},\xi^{1}_{k+1})\sum_{\ell=1}^N\frac{\omega_k^{\ell}q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{\ell},\xi^{1}_{k+1})\left(\tau^{\ell}_k + h_{k+1}(\xi_{k}^{\ell},\xi^{1}_{k+1})\right)}{\sum_{\ell'=1}^N\omega_k^{\ell'}q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{\ell'},\xi^{1}_{k+1})}\eqsp.
\end{multline*}
Using , this yields:
\begin{align*}
\mathbb{E}\left[\tilde{\omega}^1_{k+1}\tau^{1}_{k+1}\middle| \mathcal{F}_k^{N}\right]&= \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1} \sum_{i=1}^N\frac{\omega_k^i}{\Omega_k}\int\vartheta_{k+1}(\xi^{i}_{k})\frac{q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{i},x)g_{k+1}(x)}{\vartheta_{k+1}(\xi^{i}_{k}) p_{k+1}(\xi_{k}^{i},x)}\\
&\hspace{1cm}\times \sum_{\ell=1}^N\frac{\omega_k^{\ell}q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{\ell},x)\left(\tau^{\ell}_k + h_{k+1}(\xi_{k}^{\ell},x)\right)}{\sum_{\ell'=1}^N\omega_k^{\ell'}q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{\ell'},x)}p_{k+1}(\xi_{k}^{i},x)\rmd x\eqsp,\\
& =\left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\int q_{\theta}^{\Delta t_{k+1}}(\cdot,x)g_{k+1}(x)\left\{\tau_k(\cdot) + h_{k+1}(\cdot,x)\right\}\rmd x\right]\eqsp,
\end{align*}
which concludes the proof.
\end{proof}

\begin{proposition}
Assume that H\ref{assum:boundmodel} and H\ref{assum:boundalgo} hold and that for all $1\le k\le n$, $\mathrm{osc}(h_k)<+\infty$. For all $0\le k\le n$, there exist $b_k,c_k>0$ such that for all $N\ge 1$ and all $\varepsilon\in\mathbb{R}_+^\star$,
\[
\mathbb{P}\left(\left|\phi_k^N[\tau_k] - \phi_k\left[T_kh_k\right]\right|\ge \varepsilon\right)\le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\]
\end{proposition}

\begin{proof}
Write
\[
\phi_{k+1}^N[\tau_{k+1}] - \phi_{k+1}\left[T_{k+1}h_{k+1}\right] = \frac{a_N}{b_N}\eqsp,
\]
where $a_N = N^{-1}\sum_{i=1}^N \widetilde{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}h_{k+1}\right]\right)$ and $b_N =N^{-1}\sum_{i=1}^N \widetilde{\omega}_{k+1}^i$. By Lemma~\ref{lem:iid}, the random variables $\{\widetilde{\omega}_{k+1}^i\tau_{k+1}^i\}_{i=1}^N$ are independent conditionally on $\mathcal{F}_k^{N}$ and by H\ref{assum:boundalgo},
\[
\left|\widetilde{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}h_{k+1}\right]\right)\right| \le 2|\widetilde{\omega}_{k+1}|_{\infty}|H_{k+1}|_{\infty}\eqsp.
\]
Therefore, by Hoeffing inequality,
\[
\mathbb{P}\left(\left|a_N - \mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right) = \mathbb{E}\left[\mathbb{P}\left(\left|a_N - \mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\middle|\mathcal{F}_k^{N}\right)\right]\le 2\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\] 
On the other hand,
\[
\mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right] = \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\Upsilon_k\right] \eqsp,%\left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}
\]
where
\[
\Upsilon_k(x_k) = \int q_{\theta}^{\Delta t_{k+1}}(\cdot,x)g_{k+1}(x)\left(\tau_k(x_k) + h_{k+1}(x_k,x) - \phi_{k+1}\left[T_{k+1}h_{k+1}\right]\right)\rmd x\eqsp.
\]
By Lemma~\ref{lem:smooth:rec}, $\phi_{k}\left[\Upsilon_k\right] = 0$ which implies by Proposition~\ref{prop:filter} that 
\[
\mathbb{P}\left(\left|\mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right)\le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\]
Then,
\[
\mathbb{P}\left(\left|a_N\right|\ge \varepsilon\right) \le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\] 
Similarly, as $b_N \le |\widetilde{\omega}_k|_{\infty}$, by Hoeffding inequality,
\[
\mathbb{P}\left(\left|b_N - \mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right) = \mathbb{E}\left[\mathbb{P}\left(\left|b_N - \mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\middle|\mathcal{F}_k^{N}\right)\right]\le 2\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\] 
Note that
\[
\mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right] = \phi^N_{k}\left[\int q_{\theta}^{\Delta t_{k+1}}(\cdot,x)g_{k+1}(x)\rmd x\right]\eqsp.%\left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}
\]
By Proposition~\ref{prop:filter},
\[
\mathbb{P}\left(\left|\mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]-\phi_k\left[\int q_{\theta}^{\Delta t_{k+1}}(\cdot,x)g_{k+1}(x)\rmd x\right]\right|\ge \varepsilon\right)\le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
\]
The proof is completed using Lemma~\ref{lem:hoeffding:ratio}.
\end{proof}

\begin{lemma}\label{lem:hoeffding:ratio}
Assume that $a_N$, $b_N$, and $b$ are random variables defined on the same probability space such that there exist positive constants $\beta$, $B$, $C$, and $M$ satisfying
\begin{enumerate}[(i)]
    \item $|a_N/b_N|\leq M$, $\mathbb{P}$-a.s.\ and  $b \geq \beta$, $\mathbb{P}$-a.s.,
    \item For all $\epsilon>0$ and all $N\geq1$, $\mathbb{P}\left[|b_N-b|>\epsilon \right]\leq B \exp\left(-C N \epsilon^2\right)$,
    \item For all $\epsilon>0$ and all $N\geq1$, $\mathbb{P} \left[ |a_N|>\epsilon \right]\leq B \exp\left(-C N \left(\epsilon/M\right)^2\right)$.
\end{enumerate}
Then,
$$
    \mathbb{P}\left\{ \left| \frac{a_N}{b_N} \right| > \epsilon \right\} \leq B \exp{\left(-C N \left(\frac{\epsilon \beta}{2M} \right)^2 \right)} \eqsp.
$$
\end{lemma}
\begin{proof}
See \cite{}.
\end{proof}

\section{Numerical experiments}
\label{sec:exp}
This section investigates the performance of the proposed Ozaki-PaRIS EM algorithm. The different tuning parameters are:
\begin{enumerate}[i)]
\item the number of particles $N$ and the precision parameter $\widetilde{N}$;
\item the proposal transition density $p_k$ and the adjustment multiplier $\vartheta_k$ used to propagate and weight particles;
\item the discrete distribution $\mu$ used to define $\widetilde{q}_{\theta}^{\Delta t_k}$.
\end{enumerate}
The algorithm is compared to the fixed-lag particle smoother of \cite{}. To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated. Instead of the Ozaki discretization, the following bridges process are considered:
\begin{enumerate}[i)]
\item a bridge with constant drift as proposed in \cite{}.
\item a guided bridge process, see \cite{}.
\end{enumerate}
\paragraph{Choice of $\mu$}
A simple choice is to set $\mu$ as a Poisson distribution with mean $\Lambda_{\theta}t$ leading to the first estimator GPE-1:
\begin{multline*}
q^{t}_{\mathrm{GPE-1},\theta}(\kappa,\omega,(U_j)_{1\le j\le \kappa},x,y) =  \widetilde{q}^{t}_{\theta}(x,y)\eqsp\mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x)+\ell(\theta)t\right\}\Lambda_{\theta}^{-\kappa}\\
\times \prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\eqsp.
\end{multline*}
% An important feature of this estimator is that under the assumptions of this section it may be bounded uniformly in $x,y$, $\sup_{x,y\in\mathbb{R}^d}  q^{t}_{\mathrm{GPE-1},\theta}(\kappa,\omega,(U_j)_{1\le j\le \kappa},x,y) \le M_\theta$ where:
% \[
% M_{\theta} = \sup_{x,y\in\mathbb{R}^d} \left(\widetilde{q}^{t}_{\theta}(x,y)\mathrm{exp}\left\{ H_{\theta}(y) - H_{\theta}(x)\right\}\right)\mathrm{exp}\left\{\ell(\theta)t\right\}\eqsp.
% \]
\paragraph{Choice of bridge process}
The approximation of $q_{\theta}^{\Delta t_k}$ described in the paper may be compared to the approximation provided by other bridges with affine drift functions. Following \cite{}, let $(Y^{\beta}_t)_{t_{k-1}\le t\le t_k}$ be governed by the following stochastic differential equation for $t_{k-1}\le t\le t_k$:
\[
Y^{\beta}_{t_{k-1}} = X_{k-1}\quad\mbox{and}\quad\rmd Y^{\beta}_t = \beta\rmd t + \rmd W_t\eqsp.
\]
As suggestd by \cite{}, a guided process between $t_{k-1}$ and $t_k$ may alo be used:
\[
Y^{\mathsf{guid}}_{t_{k-1}} = X_{k-1}\quad\mbox{and}\quad\rmd Y^{\mathsf{guid}}_t = \frac{X_k - Y^{\mathsf{guid}}_t}{t_k-t}\rmd t + \rmd W_t\eqsp.
\]
Both processes lead to estimators of $q_{\theta}^{\Delta t_k}$ following the same steps as in Section~\ref{}.
% where $\beta\in\mathbb{R}^d$ is chosen by the user. Assume that $\alpha_{\theta}$ is of the form $\alpha_{\theta}(x) = \nabla_x A_{\theta}(x)$ where $A_{\theta}: \mathbb{R}^d \to \mathbb{R}$. Writing $\mathbb{Q}_{\theta}$ (resp. $\mathbb{W}^{\beta}_{\theta}$) the probability measure induced by $(X_s)_{0\le s\le t}$  (resp. $(Y_s)_{0\le s\le t}$) on $(\mathsf{C},\mathcal{C})$, Girsanov theorem yields:
% \[
% \frac{\rmd \mathbb{Q}_{\theta}}{\rmd \mathbb{W}^{\beta}_{\theta}}(w) = \mathrm{exp}\left\{H_{\theta}(\omega_t) - H_{\theta}(\omega_0) - \frac{1}{2}\int_{0}^{t}\left\{\|\alpha_{\theta}(\omega_s)-\beta\|^2+\Delta H_{\theta}(\omega_s)\right\}\rmd s\right\}\eqsp,
% \]
% where $H_{\theta}(x) = A_{\theta}(x) -\beta'x$. Then, the probability measures conditioned on hitting $y$ at time $t$ satisfy:
% \[
% \frac{\rmd \mathbb{Q}^{y}_{\theta}}{\rmd \mathbb{W}^{\beta,y}_{\theta}}(w) = \frac{\varphi_t(y-x-t\beta)}{q^{t}_{\theta}(x,y)} \mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x) - \frac{1}{2}\int_{0}^{t}\left\{\|\alpha_{\theta}(\omega_s)-\beta\|^2+\Delta H_{\theta}(\omega_s)\right\}\rmd s\right\}\eqsp.
% \]
% Assume also that for all $\theta$, there exists $\ell(\theta)$ such that:
% \[
% \mathrm{inf}_{u\in \mathbb{R}^d} \left\{\frac{1}{2}\left(\|\alpha_{\theta}(u)-\beta\|^2+\Delta H_{\theta}(u)\right)\right\}\ge \ell(\theta)\eqsp.
% \]
% Then
% \[
% q^{t}_{\theta}(x,y) = \varphi_t(y-x-t\beta)\eqsp\mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x)+\ell(\theta)t\right\}\mathbb{E}_{\mathbb{W}^{\beta,y}}\left[\mathrm{exp}\left\{-\int_{0}^{t}\phi_{\theta}(\omega_s)\rmd s\right\}\right]\eqsp,
% \]
% where 
% \[
% \phi_{\theta}(x) = \frac{1}{2}\left\{\|\alpha_{\theta}(x)-\beta\|^2+\Delta H_{\theta}(x)\right\}-\ell(\theta)\eqsp.
% \]
% An unbiased estimator of $q^{t}_{\theta}(x,y)$ is then obtained by estimating $\mathbb{E}_{\mathbb{W}^{\beta,y}}\left[\mathrm{exp}\left\{-\int_{0}^{t}\phi_{\theta}(\omega_s)\rmd s\right\}\right]$.

% Let $h$ be a function defined on $\{1,\ldots,N\}$,
% \begin{align*}
% \mathbb{E}\left[h(I_{\tau})\right] & = \sum_{m\ge 0}\mathbb{E}\left[h(I_m)\1_{\tau=m}\right]\eqsp,\\
% & = \sum_{m\ge 0}h(m)\left(\prod_{\ell=0}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\right]\right)\mathbb{E}\left[h(I_m)\1_{\mathcal{A}^k_{m}}\right]\eqsp,\\
% & = \sum_{m\ge}\left(\prod_{\ell=0}^{m-1}\mathbb{E}\left[1-\frac{q^{\Delta t_k}_{\mathrm{GPE-1},\theta}(\kappa_{\ell},\omega^{\kappa_{\ell}},(U^{\kappa_{\ell}}_j)_{1\le j\le \kappa_{\ell}},\xi_{k-1}^{I_\ell},\xi_k^{1})}{M_{\theta}}\right]\right)\\
% &\hspace{5cm}\times\mathbb{E}\left[h(I_m)\frac{q^{\Delta t_k}_{\mathrm{GPE-1},\theta}(\kappa_{m},\omega^{\kappa_{m}},(U^{\kappa_{m}}_j)_{1\le j\le \kappa_{m}},\xi_{k-1}^{I_m},\xi_k^{1})}{M_{\theta}}\right]\eqsp,\\
% & = \sum_{m\ge 0}\left(\mathbb{E}\left[1-\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{1})}{M_{\theta}}\right]\right)^m\mathbb{E}\left[h(I_m)\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{1})}{M_{\theta}}\right]\eqsp,\\
% & = \mathbb{E}\left[h(I_1)q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{1})\right]/\mathbb{E}\left[q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{1})\right]\eqsp,\\
% & = \sum_{\ell=1}^N \frac{h(\ell)\omega_{k-1}^{\ell}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{\ell},\xi_k^{1})}{\sum_{m=1}^N\omega_{k-1}^{m}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{m},\xi_k^{1})}\eqsp,
% \end{align*}
% which concludes the proof.

\appendix

\section{Proofs}

\begin{lemma}
\label{lem:AR:unbiased}
Let $1\le k\le n$.  For all $1\le i \le N$ and all $1\le j \le \widetilde{N}$, the conditional probability distribution given $\mathcal{F}_k^N$ of the random variable $J_k^{i,j}$ produced by Algorithm~\ref{alg:AR} is  $\Lambda_{k-1}^N(i,\cdot)$. 
\end{lemma}

\begin{proof}
Let $\tau$ be the first time  draws are accepted in the accept-reject mechanism. For all $\ell\ge 1$, write
\[
\mathcal{A}^k_{\ell} = \left\{U_\ell<\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_\ell},\xi_k^{i},\Lambda_{\ell})/\sigma_+\right\}\eqsp.
\]

Let $h$ be a function defined on $\{1,\ldots,N\}$,
\begin{align*}
\mathbb{E}\left[h(J^{i,j}_k)\middle| \mathcal{F}_k^N\right] & = \sum_{m\ge 1}\mathbb{E}\left[h(I_m)\1_{\tau=m}\middle| \mathcal{F}_k^N\right]\eqsp,\\
& = \sum_{m\ge 1}h(m)\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\middle| \mathcal{F}_k^N\right]\right)\mathbb{E}\left[h(I_m)\1_{\mathcal{A}^k_{m}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
& = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[1-\frac{\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_\ell},\xi_k^{i},\Lambda_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)\\
&\hspace{5cm}\times\mathbb{E}\left[h(I_m)\frac{\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_m},\xi_k^{i},\Lambda_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
& = \sum_{m\ge 1}\left(\mathbb{E}\left[1-\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)^{m-1}\mathbb{E}\left[h(I_m)\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
& = \mathbb{E}\left[h(I_1)q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{i})\middle| \mathcal{F}_k^N\right]/\mathbb{E}\left[q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{i})\middle| \mathcal{F}_k^N\right]\eqsp,\\
& = \sum_{\ell=1}^N \frac{h(\ell)\omega_{k-1}^{\ell}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{\ell},\xi_k^{i})}{\sum_{m=1}^N\omega_{k-1}^{m}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{m},\xi_k^{i})}\eqsp,\\
&= \sum_{\ell=1}^N \Lambda_{k-1}^N(i,\ell)h(\ell) \eqsp,
\end{align*}
which concludes the proof.
\end{proof}

\section{Multidimensional Ornstein Uhlenbeck bridge}
\label{sec:ornstein:bridge}
The GPE introduced in this paper requires to sample finite dimensional distributions of Ornstein Uhlenbeck bridges. These results may be found in \cite{} and are given here for completeness. Consider the following $d$-dimensional SDE:
\begin{equation}
\label{eq:sde:ornstein}
Z_0 = z_0\quad\mbox{and}\quad \rmd Z_t = (\Upsilon Z_t + \vartheta)\rmd t + \rmd W_t\eqsp,
\end{equation}
where $W$ is a standard Brownian motion on $\mathbb{R}^d$, $A$ is a $d\times d$ matrix and $b$ a $d$ dimensional vector. In the case of Section~\ref{sec:PaRIS:SDE}, for all $1\le k \le n$, the proposal stochastic differential equation governing $Y_t^k$, $t\in(t_{k-1},t_k)$, is of the form \eqref{eq:sde:ornstein} with
\[
\Upsilon = J_{\alpha_{\theta}}(X_{k-1})\;\;\mbox{and}\;\; \vartheta =\alpha_{\theta}(X_{k-1}) - J_{\alpha_{\theta}}(X_{k-1})X_{k-1}\eqsp,
\]
$J_{\alpha_{\theta}}$ being the Jacobian of $\alpha_{\theta}$. In this case, as $\alpha_{\theta}(x) = \nabla_x A_{\theta}(x)$, $A$ is symetric and \textcolor{red}{nonsingular, Syl : a voir}. By \cite[Section~5.6]{karatzas:shreve:1991}, the unique strong solution of \eqref{eq:sde:ornstein} is:
\[
Z_t = \mathrm{e}^{t\Upsilon}z_0 + \int_0^t\mathrm{e}^{(t-s)\Upsilon}\vartheta \rmd s + \int_0^t\mathrm{e}^{(t-s)\Upsilon} \rmd W_s\eqsp.
\]
Therefore, the probability density of the conditional distribution of $Z_t$ given $Z_0 = z_0$ is Gaussian with mean $\mu_t(z_0)$ and variance $\Sigma_{t}$ given by:
\begin{align*}
\Sigma_{t} &= \int_0^{t} \mathrm{e}^{(t -s)\Upsilon}\mathrm{e}^{(t -s)\Upsilon'}\rmd s\eqsp,\\
\mu_{t}(z) &= \mathrm{e}^{t \Upsilon}z_0 + \left(\int_0^t\mathrm{e}^{(t-s)\Upsilon} \rmd s\right)\vartheta\eqsp.
\end{align*}
As $\Upsilon$ is symetric and nonsingular,
\begin{align*}
\Sigma_{t} &= \int_0^{t} \mathrm{e}^{2(t -s)\Upsilon}\rmd s = \Upsilon^{-1}(\mathrm{e}^{2t\Upsilon}-I_d)/2\eqsp,\\
\mu_{t}(z) &= \mathrm{e}^{t \Upsilon}z + \left(\int_0^t\mathrm{e}^{(t-s)\Upsilon} \rmd s\right)\vartheta = \mathrm{e}^{t \Upsilon}z + \Upsilon^{-1}(\mathrm{e}^{t\Upsilon}-I_d)\vartheta \eqsp.
\end{align*}
For each interval $(t_{k-1},t_k)$, $1\le k \le n$, the GPE requires to sample a Ornstein-Uhlenbeck bridge at some ramdom time steps between $X_{k-1}$ at time $t_{k-1}$ and $X_k$ at time $t_k$. Conditional on $Z_s = x$ and $Z_T = y$, the probability density of $Z_t$ is Gaussian with mean $\mu$ and variance $\Sigma_t^{s,T}$ given by: 
\begin{align*}
\mu &=\Gamma(t,T)\Gamma(s,T)^{-1}\mu_- + \Gamma(s,t)'(\Gamma(s,T)')^{-1}\mu_+\eqsp,\\
\Sigma_t^{s,T} &=\Gamma(t,T)\Gamma(s,T)^{-1}\Gamma(s,t)\eqsp,\\
\Gamma(s,t)&=\int_s^{t} \mathrm{e}^{(s -u)\Upsilon}\mathrm{e}^{(t -u)\Upsilon'}\rmd u\eqsp = \Upsilon^{-1}\left(\mathrm{e}^{(t -s)\Upsilon}-\mathrm{e}^{(s -t)\Upsilon}\right)/2\eqsp,\\
\mu_- &= x - \Upsilon^{-1}(\mathrm{e}^{(s -t)\Upsilon}-I_d)\vartheta\eqsp,\\
\mu_+ &= y-\Upsilon^{-1}(\mathrm{e}^{(T -t)\Upsilon}-I_d)\vartheta\eqsp.
\end{align*}

\bibliographystyle{plain}
\bibliography{./ParisEM_bib}
\end{document}

### 2016-10-31 ParisEM.tex
Modified to remove EM parts

@@
 
-\title{Efficient particle smoother for stochastic differential equations with linear complexity}
+\title{Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations}
 
@@
 
-The aim of this paper is to estimate $\theta$ given the set of observations $(Y_0,\ldots,Y_n)$. The EM algorithm introduced in \cite{dempster:laird:rubin:1977} 
-maximizes the incomplete data log-likelihood $\theta\mapsto \ell_{\theta}^{n}$ defined by
+Statistical inference for partially observed stochastic differential equations usually involves the computation of conditional distributions of some unobserved states given the observations $(Y_0,\ldots,Y_n)$.
+These posterior distributions are crucial to compute maximum likelihood estimators defined as maximizers of the incomplete data log-likelihood $\theta\mapsto \ell_{\theta}^{n}$ given by
 \begin{equation*}
@@
 \end{equation*}
-Denote by $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{0:n}\right]$  the conditional expectation given $Y_{0:n}$ when the parameter value is $\theta$.
-The EM algorithm iteratively builds a sequence $\{\theta_{p}\}_{p\ge 0}$ of parameter estimates following the two steps:
+For instance, the EM algorithm introduced in \cite{dempster:laird:rubin:1977} iteratively builds a sequence $\{\theta_{p}\}_{p\ge 0}$ of parameter estimates following the two steps:
 \begin{enumerate}
-	\item {\bf E-step}: compute $\theta \mapsto Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log p_{\theta}(X_{0:n},Y_{0:n})\middle|Y_{0:n}\right]$ ;
+	\item {\bf E-step}: compute $\theta \mapsto Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log p_{\theta}(X_{0:n},Y_{0:n})\middle|Y_{0:n}\right]$, where $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{0:n}\right]$ is the conditional expectation given $Y_{0:n}$ when the parameter value is $\theta$.
 	\item {\bf M-step}: choose $\theta_{p+1}$ as a maximizer of $\theta \mapsto Q(\theta,\theta_{p})$.
 \end{enumerate}
-It can be shown that under weak assumptions on the model, the sequence $\{\theta_{p}\}_{p\ge 0}$ converges to the set of stationary points of $\theta \mapsto \ell_{\theta}^{n}(Y_{0:n})$.
-However, there are two main difficulties with applying the EM algorithm to partially observed SDE in general cases:
-\begin{enumerate}[-]
-\item The E-step requires to compute $q^{\Delta t_k}_{\theta}$ for which there is no analytic expression.
-\item The computation of $Q(\theta,\theta_{p})$ relies on conditional expectations of additive functionals of the state sequence given the observations which are not available explicitly.
-\end{enumerate}
+The E-step boils down to the computation of the smoothed additive function: 
+\begin{multline*}
+Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log \left(\chi_{\theta}(X_0)g_{\theta}(X_0;Y_0)\right)\middle|Y_{0:n}\right] \\
++ \sum_{k=1}^n\mathbb{E}_{\theta_p}\left[\log \left(q^{\Delta t_k}_{\theta}(X_{k-1},X_k)g_{\theta}(X_k,Y_k)\right)\middle|Y_{0:n}\right] \eqsp.
+\end{multline*}
+Similarly, by Fisher's identity, recursive maximum likelihood estimates may be computed using the gradient of the loglikelihood which can be written as the conditional expectation of an additive functional of the hidden states given all the observations.
+See \cite[Chapter $10$ and $11$]{cappe:moulines:ryden:2005}, \cite{kantas:doucet:signh:2015,doucet:poyiadjis:singh:2011,lecorff:fort:2013a,lecorff:fort:2013b}
+for further references on the use of these smoothed expectations of additive functionals  applied to maximum likelihood parameter inference in latent data models.
+
+The exact computation of these expectations is usually not possible for partially observed diffusions such as \eqref{eq:target:sde}.
+In this paper, we propose to use Sequential Monte Carlo methods to approximate smoothing distributions with random particles associated with importance weights.
+\cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers by combining importance sampling steps to propagate particles with importance resampling steps to duplicate or discard particles according to their importance weights.
+Approximations of the smoothing distributions may be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}.
+Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$.
+Then, the backward pass of the FFBS algorithm keeps all the particles sampled during the forward pass and computes new weights to take into account the information brought by all the observations from time $n$ to time $0$.
+Instead of computing new weights, the FFBSi  algorithm samples independently backward in time particle trajectories using the particles and weights produced by the forward pass.
+
+Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly, using the observations as they are received, smoothed expectations of additive functionals.
+Contrary to the FFBS algorithm the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.
+However, there is a major difficulty to apply SMC methods to partially observed SDE as the computation of the importance weights requires to evaluate $q^{\Delta t_k}_{\theta}$ for which there is no analytic expression.
+%the EM algorithm to partially observed SDE in general cases:
+%\begin{enumerate}[-]
+%\item The E-step requires to compute $q^{\Delta t_k}_{\theta}$ for which there is no analytic expression.
+%\item The computation of $Q(\theta,\theta_{p})$ relies on conditional expectations of additive functionals of the state sequence given the observations which are not available explicitly.
+%\end{enumerate}
 To overcome these difficulties, \cite{olsson:strojby:2011} proposed to use General Poisson Estimators (GPE) introduced in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,fearnhead:papaspiliopoulos:roberts:2008}  to provide an unbiased estimation of $q^{\Delta t_k}_{\theta}$.
-These estimators are based on a representation of $q^{\Delta t_k}_{\theta}$ as an expectation under the probability measure of a Brownian bridge.
-The conditional expectation involved in the computation is approximated in  \cite{olsson:strojby:2011}  using fixed-lag Sequential Monte Carlo methods developed in \cite{olsson:cappe:douc:moulines:2008}.
+These estimators are based on a representation of $q^{\Delta t_k}_{\theta}$ as an expectation under the probability measure of a Brownian bridge (the conditional expectation involved in the computation is approximated in  \cite{olsson:strojby:2011}  using fixed-lag Sequential Monte Carlo methods developed in \cite{olsson:cappe:douc:moulines:2008}).
+
 In this paper, we propose to improve this algorithm to obtain faster convergence of the sequence of Monte Carlo EM estimates.
 First, we introduce a new class off GPE based on the Ozaki discretization of the SDE  \eqref{eq:target:sde}, see \cite{ozaki:1992,shoji:ozaki:1998}.
-Instead of using fixed-lag techniques which introduce a systematic additional bias  in the approximation of $Q(\theta,\theta_{p})$, we use the new PaRIS algorithm which allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles.
+These new GPE estimators use a an order one Taylor expansion or the drift of \eqref{eq:target:sde} to obtain more efficient estimates than Euler discretization based GPE.
+Instead of using fixed-lag techniques which introduce a systematic additional bias  in the approximation of the conditional expectations, we use the new PaRIS algorithm which allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles.
+The striking result of the application of PaRIS algorithm to SDE is that the accept reject mechanism underlying the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
 
-\section{Random weight PaRIS-EM algorithm}
-For a given value $\theta_p$ of the parameter, dropping the initial state, the E-step may be written:
-\[
-Q(\theta,\theta_{p})= \sum_{k=1}^n\mathbb{E}_{\theta_p}\left[\log \left\{q^{\Delta t_k}_{\theta}(X_{k-1},X_k)g_{\theta}(X_k,Y_k)\right\}\middle|Y_{0:n}\right]\eqsp.
-\]
-We first describe the Monte Carlo method used to approximate $Q(\theta,\theta_{p})$ when $q^{\Delta t_k}_{\theta}$ may be computed.
+\section{Random weight PaRIS algorithm}
+%For a given value $\theta_p$ of the parameter, dropping the initial state, the E-step may be written:
+%\[
+%Q(\theta,\theta_{p})= \sum_{k=1}^n\mathbb{E}_{\theta_p}\left[\log \left\{q^{\Delta t_k}_{\theta}(X_{k-1},X_k)g_{\theta}(X_k,Y_k)\right\}\middle|Y_{0:n}\right]\eqsp.
+%\]
+%We first describe the Monte Carlo method used to approximate $Q(\theta,\theta_{p})$ when $q^{\Delta t_k}_{\theta}$ may be computed.
 Dropping $\theta$ from the notations, the E-step computes an expectation of the form
+The aim of this Section is to detail the extension of PaRIS algorithm to approximate expectations of the form
 \begin{equation}
 \label{def:addfunc}
-\bar{H}_{n} =  \sum_{k=1}^n\mathbb{E}\left[h_k(X_{k-1},X_k)\middle|Y_{0:n}\right]\eqsp.
+H_{n} =  \sum_{k=1}^n\mathbb{E}\left[h_k(X_{k-1},X_k)\middle|Y_{0:n}\right]\eqsp,
 \end{equation}
+when the transition density of the hidden states is not available explicitly where $\{h_k\}_{k=1}^n$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$.
 Thus, for $0 \leq k \leq k' \leq n$ and a given measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, we need to compute an approximation of the smoothed expectation of the states $(X_k,\ldots,X_{k'})$ defined by:
@@
 The filtering distributions are defined, for $0\le k\le n$ by $\phi_{k} = \phi_{k:k|k}$.
-In this paper, we are interested in a particular class of Sequential Monte Carlo methods to approximate $\bar{H}_{n}$ known as the PaRIS algorithm introduced in \cite{olsson:westerborn:2016}.
+%In this paper, we are interested in a particular class of Sequential Monte Carlo methods to approximate $\bar{H}_{n}$ known as the PaRIS algorithm introduced in \cite{olsson:westerborn:2016}.
 
-The PaRIS algorithm is based on the first Sequential Monte Carlo algorithms introduced in  \cite{gordon:salmond:smith:1993} and \cite{kitagawa:1996}. 
-They rely on importance sampling and importance resampling steps to approximate the filtering distribution $\phi_{k}$ (conditional distribution of $X_k$ given $Y_{0:k}$) using  a set of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$.
+%The PaRIS algorithm is based on the first Sequential Monte Carlo algorithms introduced in  \cite{gordon:salmond:smith:1993} and \cite{kitagawa:1996}. 
+Sequential Monte Carlo algorithms introduced in  \cite{gordon:salmond:smith:1993} and \cite{kitagawa:1996} rely on importance sampling and importance resampling steps to approximate the filtering distribution $\phi_{k}$ (conditional distribution of $X_k$ given $Y_{0:k}$) using  a set of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$.
 At time $k = 0$, $N$ particles $\{\xi^{\ell}_0\}_{\ell=1}^N$ are initialized by setting $\xi^{\ell}_0=x_0$.
@@
 \]
-The PaRIS algorithm uses the same decomposition as the Forward Filtering Backward Smoothing (FFBS) algorithm introduced in \cite{doucetgodsillandrieu:2000} and the Forward Filtering Backward Simulation (FFBSI) algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions.
-Both the FFBS and the FFBSi algorithms require first that a forward pass has produced a set of particles and importance weights $\{ (\xi^{\ell}_k, \omega^{\ell}_k)\}$, for $\ell \in \{1, \ldots, N\}$ and  $k \in \{0,\ldots,n\}$ to approximate $\phi_{k}$ for $k=0$ to $k=n$.
+The PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSI algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions.
+%Both the FFBS and the FFBSi algorithms require first that a forward pass has produced a set of particles and importance weights $\{ (\xi^{\ell}_k, \omega^{\ell}_k)\}$, for $\ell \in \{1, \ldots, N\}$ and  $k \in \{0,\ldots,n\}$ to approximate $\phi_{k}$ for $k=0$ to $k=n$.
 The FFBS algorithm stores all particles and weights without taking into consideration the genealogy of the particles.
 Then, a backward pass is performed, keeping all the particles fixed but modifying all importance weights computed during the forward pass.
-This algorithm has a computational complexity of order $\mathcal{O}(N^{k'-k+1})$ to approximate joint smoothing distributions of the form $\phi_{k:k'|n}$ with $k<k'$ which makes it prohibitive.
+The FFBS algorithm has a computational complexity of order $\mathcal{O}(N^{k'-k+1})$ to approximate joint smoothing distributions of the form $\phi_{k:k'|n}$ with $k<k'$ which makes it prohibitive.
 An important feature of the FFBS algorithm shown in \cite{mongillo:deneve:2008,cappe:2011,delmoral:doucet:singh:2010} is that it can be implemented using only a forward pass when approximating smoothed expectations of additive functions.
@@
 \[
-\mathcal{F}_k^N = \sigma\left\{Y_{0:k};(\xi^{\ell}_{u},\omega^{\ell}_{u}), 1\le \ell\le N, 1\le u\le k\right\}\eqsp.
+\mathcal{F}_k^N = \sigma\left\{Y_{0:k};(\xi^{\ell}_{u},\omega^{\ell}_{u},\tau^{\ell}_{u}), 1\le \ell\le N, 1\le u\le k\right\}\eqsp.
 \]
@@
 \end{algorithm}
+
+\begin{lemma}
+\label{lem:AR:unbiased}
+Let $1\le k\le n$.
+For all $1\le i \le N$ and all $1\le j \le \widetilde{N}$, the conditional probability distribution given $\mathcal{F}_k^N$ of the random variable $J_k^{i,j}$ produced by Algorithm~\ref{alg:AR} is  $\Lambda_{k-1}^N(i,\cdot)$.
+\end{lemma}
 
@@
 An unbiased estimator of $q^{\Delta t_k}_{\theta}(X_{k-1},X_k)$ is then obtained by estimating $\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\right]$.
-Following \cite{}, if $\omega \sim\mathbb{W}_{\theta}^{k,X_{k}}$, $\kappa$ is a random variables taking values in $\mathbb{N}$ with distribution $\mu$ and if $(U_j)_{1\le j\le \kappa}$ are independent uniform random variables on $[0,t]$,
+Following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, if $\omega \sim\mathbb{W}_{\theta}^{k,X_{k}}$, $\kappa$ is a random variables taking values in $\mathbb{N}$ with distribution $\mu$ and if $(U_j)_{1\le j\le \kappa}$ are independent uniform random variables on $[0,t]$,
 \begin{align*}
@@
 \end{align*}
-Interesting choices of $\mu$ are discussed in \cite{} and implemented in Section~\ref{sec:exp}. Writing
+Interesting choices of $\mu$ are discussed in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} and implemented in Section~\ref{sec:exp}. Writing
 \[
@@
 where $\{\Lambda_k^i\}_{i=1}^M$ are independent copies of $\Lambda_k$. 
-The last step to approximate the intermediate quantity of the EM algorithm is to provide an unbiased estimate of $\log q^{\Delta t_k}_{\theta}(X_{k-1},X_k)$.
+%The last step to approximate the intermediate quantity of the EM algorithm is to provide an unbiased estimate of $\log q^{\Delta t_k}_{\theta}(X_{k-1},X_k)$.
 By Girsanov theorem, this log density can be written as:
-\begin{multline*}
-\log q^{\Delta t_k}_{\theta}(X_{k-1},X_k) = \log \widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k) + H_{\theta}(X_k) - H_{\theta}(X_{k-1})-\ell(\theta)\Delta t_k \\
--\mathbb{E}_{\mathbb{Q}_{\theta}^{k,X_{k}}}\left[\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right]\eqsp.
-\end{multline*}
-If $U$ is uniformy distributed on $(0,\Delta t_k)$ and if $w \sim \mathbb{Q}_{\theta}^{k,X_{k}}$, $\log q^{\Delta t_k}_{\theta}(X_{k-1},X_k)$ may then be estimated by:
-\begin{equation} 
-\label{eq:logq}
-\mathsf{q}_{\theta}^{\Delta t_k}(X_{k-1},X_k,\omega_U) = \log \widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k) + H_{\theta}(X_k) - H_{\theta}(X_{k-1})-\ell(\theta)\Delta t_k - \Delta t_k \phi_{\theta}(\omega_U)\eqsp.
-\end{equation}
-Sampling $w \sim \mathbb{Q}_{\theta}^{k,X_{k}}$ is possible as by \eqref{eq:Girsanov},
-\[
-\frac{\rmd \mathbb{Q}^{k,X_{k}}_{\theta}}{\rmd \mathbb{W}^{k,X_{k}}_{\theta}}(w) \propto 
-\mathrm{exp}\left\{-\int_{t_{k-1}}^{t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\eqsp.
-\]
-As noted by \cite{}, this importance ratio is proportional to the probability that a marked Poisson process on $(0,\Delta t_k)\times (0,1)$ with intensity $\Lambda_{\theta}$ is below the graph $s \mapsto \phi_{\theta}(\omega_s)/\Lambda_{\theta}$.
+%\begin{multline*}
+%\log q^{\Delta t_k}_{\theta}(X_{k-1},X_k) = \log \widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k) + H_{\theta}(X_k) - H_{\theta}(X_{k-1})-\ell(\theta)\Delta t_k \\
+%-\mathbb{E}_{\mathbb{Q}_{\theta}^{k,X_{k}}}\left[\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right]\eqsp.
+%\end{multline*}
+%If $U$ is uniformy distributed on $(0,\Delta t_k)$ and if $w \sim \mathbb{Q}_{\theta}^{k,X_{k}}$, $\log q^{\Delta t_k}_{\theta}(X_{k-1},X_k)$ may then be estimated by:
+%\begin{equation} 
+%\label{eq:logq}
+%\mathsf{q}_{\theta}^{\Delta t_k}(X_{k-1},X_k,\omega_U) = \log \widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k) + H_{\theta}(X_k) - H_{\theta}(X_{k-1})-\ell(\theta)\Delta t_k - \Delta t_k \phi_{\theta}(\omega_U)\eqsp.
+%\end{equation}
+%Sampling $w \sim \mathbb{Q}_{\theta}^{k,X_{k}}$ is possible as by \eqref{eq:Girsanov},
+%\[
+%\frac{\rmd \mathbb{Q}^{k,X_{k}}_{\theta}}{\rmd \mathbb{W}^{k,X_{k}}_{\theta}}(w) \propto 
+%\mathrm{exp}\left\{-\int_{t_{k-1}}^{t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\eqsp.
+%\]
+%As noted by \cite{}, this importance ratio is proportional to the probability that a marked Poisson process on $(0,\Delta t_k)\times (0,1)$ with intensity $\Lambda_{\theta}$ is below the graph $s \mapsto \phi_{\theta}(\omega_s)/\Lambda_{\theta}$.
 $\omega_U$ in \eqref{eq:logq} is then obtaine by first sampling $\tilde{\omega}\sim \mathbb{W}^{k,X_{k}}$ at times on $(0,\Delta t_k)$ defined by the marked Poison process (which is accepted if is is below $\phi_{\theta}(\omega_s)/\Lambda_{\theta}$) and then by interpolating the bridge at time $U$.
-
-All these steps are summarized in Algorithm~\ref{alg:Ozaki:PaRIS} which provides an approximation of the E step of the EM algorithm.
+All these steps are summarized in Algorithm~\ref{alg:Ozaki:PaRIS} which provides an approximation of \eqref{def:addfunc}. %the E step of the EM algorithm.
 \begin{algorithm}
@@
 \FORALL{$\ell \in 1,\dots,\widetilde{N}$}
-\REPEAT
-\STATE $\kappa \sim \mathcal{P}(\Lambda_{\theta}\Delta t_k)$, $(U_{m},\chi_m$) uniformly on $(0,\Delta t_k)\times(0,1)$, $1\le m \le \kappa$.
-\STATE Sample $\widetilde{\omega}_s\sim \mathbb{W}_{\theta}^{k,X_{k}}$ at times $U_{m}$, $1\le m \le \kappa$.
-\UNTIL{$\phi_{\theta}(\widetilde{\omega}_{U_m})/\Lambda_{\theta}<\chi_m$ for all $1\le m \le \kappa$}
-\STATE Sample $U$ uniformly in $(0,\Delta t_k)$.
-\STATE $\tau_k^i = \tau_k^i + \tau^{J_k^{i,\ell}}_{k-1} + \log g_{\theta}(\xi^i_k,Y_k) + \mathsf{q}_{\theta}^{\Delta t_k}(\xi^{J_k^{i,\ell}}_{k-1}, \xi^{i}_{k},\widetilde{\omega}_U)$.
+%\REPEAT
+%\STATE $\kappa \sim \mathcal{P}(\Lambda_{\theta}\Delta t_k)$, $(U_{m},\chi_m$) uniformly on $(0,\Delta t_k)\times(0,1)$, $1\le m \le \kappa$.
+%\STATE Sample $\widetilde{\omega}_s\sim \mathbb{W}_{\theta}^{k,X_{k}}$ at times $U_{m}$, $1\le m \le \kappa$.
+%\UNTIL{$\phi_{\theta}(\widetilde{\omega}_{U_m})/\Lambda_{\theta}<\chi_m$ for all $1\le m \le \kappa$}
+%\STATE Sample $U$ uniformly in $(0,\Delta t_k)$.
+\STATE $\tau_k^i = \tau_k^i + \tau^{J_k^{i,\ell}}_{k-1} + h_k(\xi^{J_k^{i,\ell}}_{k-1},\xi^i_k)$.%\log g_{\theta}(\xi^i_k,Y_k) + \mathsf{q}_{\theta}^{\Delta t_k}(\xi^{J_k^{i,\ell}}_{k-1}, \xi^{i}_{k},\widetilde{\omega}_U)$.
 \ENDFOR
@@
 \end{hypA}
-The following Lemma is proved in \cite[Lemma~11]{olsson:westerborn:2016}.
-\begin{lemma}
-\label{lem:smooth:rec}
-Forward smoothing
-\end{lemma}
+%The following Lemma is proved in \cite[Lemma~11]{olsson:westerborn:2016}.
+%\begin{lemma}
+%\label{lem:smooth:rec}
+%Forward smoothing
+%\end{lemma}
 
@@
 \]
-By Lemma~\ref{lem:smooth:rec}, $\phi_{k}\left[\Upsilon_k\right] = 0$ which implies by Proposition~\ref{prop:filter} that 
+By \cite[Lemma~11]{olsson:westerborn:2016}., $\phi_{k}\left[\Upsilon_k\right] = 0$ which implies by Proposition~\ref{prop:filter} that 
 \[
@@
 \begin{proof}
-See \cite{}.
+See \cite{douc:garivier:moulines:olsson:2011}.
 \end{proof}
@@
 \section{Proofs}
-
-\begin{lemma}
-\label{lem:AR:unbiased}
-Let $1\le k\le n$.
-For all $1\le i \le N$ and all $1\le j \le \widetilde{N}$, the conditional probability distribution given $\mathcal{F}_k^N$ of the random variable $J_k^{i,j}$ produced by Algorithm~\ref{alg:AR} is  $\Lambda_{k-1}^N(i,\cdot)$.
-\end{lemma}
-
-\begin{proof}
+\begin{proof}[Proof of Lemma~\ref{lem:AR:unbiased}]
 Let $\tau$ be the first time  draws are accepted in the accept-reject mechanism. For all $\ell\ge 1$, write
@@
 $J_{\alpha_{\theta}}$ being the Jacobian of $\alpha_{\theta}$.
-In this case, as $\alpha_{\theta}(x) = \nabla_x A_{\theta}(x)$, $A$ is symetric and \textcolor{red}{nonsingular, Syl : a voir}.
+In this case, as $\alpha_{\theta}(x) = \nabla_x A_{\theta}(x)$, $A$ is symmetric and nonsingular.
 By \cite[Section~5.6]{karatzas:shreve:1991}, the unique strong solution of \eqref{eq:sde:ornstein} is:

### 2016-11-01 ParisEM.tex
Minor modifications

@@
 
-\title{Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations}
+\title{Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations: the Oziris algorithm}
 
@@
 \section{Introduction}
-It is assumed that the dynamics of the states $(X_t)_{t\ge 0}$ depends on a parameter $\theta\in \Theta$ where $\Theta$ is a compact subspace of $\mathbb{R}^{\ell}$ and that $(X_t)_{t\ge 0}$ is a weak solution to the following Stochastic Differential Equation (SDE) in $\mathbb{R}^d$:
+The dynamics of the states $(X_t)_{t\ge 0}$ depends on a parameter $\theta\in \Theta$ where $\Theta$ is a compact subspace of $\mathbb{R}^{\ell}$ and that $(X_t)_{t\ge 0}$ is a weak solution to the following Stochastic Differential Equation (SDE) in $\mathbb{R}^d$:
 \begin{equation}
@@
 \end{equation*}  
-where the complete data likelihood $p_{\theta}$ is given by
+where $p_{\theta}$ is the complete data likelihood:
 \begin{equation*}
@@
 Dropping $\theta$ from the notations, the E-step computes an expectation of the form
-The aim of this Section is to detail the extension of PaRIS algorithm to approximate expectations of the form
+Let $0 \leq k \leq k' \leq n$. The joint smoothing distributions of the hidden states are defined, for all measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, by:
+\[
+\phi_{k:k'|n}[h] := \mathbb{E}\left[h(X_k,\ldots,X_{k'})\middle|Y_{0:n}\right]\eqsp.
+\]
+For all $0\le k\le n$, $\phi_{k} = \phi_{k:k|k}$ denotes the filtering distributions. The aim of this section is to detail the extension of PaRIS algorithm to approximate expectations of the form
 \begin{equation}
@@
 \end{equation}
-when the transition density of the hidden states is not available explicitly where $\{h_k\}_{k=1}^n$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$.
-Thus, for $0 \leq k \leq k' \leq n$ and a given measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, we need to compute an approximation of the smoothed expectation of the states $(X_k,\ldots,X_{k'})$ defined by:
-\[
-\phi_{k:k'|n}[h] := \mathbb{E}\left[h(X_k,\ldots,X_{k'})\middle|Y_{0:n}\right]\eqsp.
-\]
-The filtering distributions are defined, for $0\le k\le n$ by $\phi_{k} = \phi_{k:k|k}$.
+when the transition density of the hidden states is not available explicitly and where $\{h_k\}_{k=1}^n$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$.
 %In this paper, we are interested in a particular class of Sequential Monte Carlo methods to approximate $\bar{H}_{n}$ known as the PaRIS algorithm introduced in \cite{olsson:westerborn:2016}.

### 2016-11-02 ParisEM.tex
Changing proof of Lemma 1

@@
 \newcommand{\com}[1]{{\color{gray} // #1}}
-
+\newcommand{\mP}{\mathbb{P}}
+\newcommand{\E}{\mathbb{E}}
 \newtheorem{lemma}{Lemma}
@@
 \begin{proof}[Proof of Lemma~\ref{lem:AR:unbiased}]
-Let $\tau$ be the first time  draws are accepted in the accept-reject mechanism. For all $\ell\ge 1$, write
-\[
-\mathcal{A}^k_{\ell} = \left\{U_\ell<\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_\ell},\xi_k^{i},\Lambda_{\ell})/\sigma_+\right\}\eqsp.
-\]
-
-Let $h$ be a function defined on $\{1,\ldots,N\}$,
+%Let $\tau$ be the first time  draws are accepted in the accept-reject mechanism. For all $\ell\ge 1$, write
+%\[
+%\mathcal{A}^k_{\ell} = \left\{U_\ell<\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_\ell},\xi_k^{i},\Lambda_{\ell})/\sigma_+\right\}\eqsp.
+%\]
+%
+%Let $h$ be a function defined on $\{1,\ldots,N\}$,
+%\begin{align*}
+%\mathbb{E}\left[h(J^{i,j}_k)\middle| \mathcal{F}_k^N\right] & = \sum_{m\ge 1}\mathbb{E}\left[h(I_m)\1_{\tau=m}\middle| \mathcal{F}_k^N\right]\eqsp,\\
+%& = \sum_{m\ge 1}h(m)\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\middle| \mathcal{F}_k^N\right]\right)\mathbb{E}\left[h(I_m)\1_{\mathcal{A}^k_{m}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
+%& = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[1-\frac{\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_\ell},\xi_k^{i},\Lambda_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)\\
+%&\hspace{5cm}\times\mathbb{E}\left[h(I_m)\frac{\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_m},\xi_k^{i},\Lambda_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
+%& = \sum_{m\ge 1}\left(\mathbb{E}\left[1-\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)^{m-1}\mathbb{E}\left[h(I_m)\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
+%& = \mathbb{E}\left[h(I_1)q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{i})\middle| \mathcal{F}_k^N\right]/\mathbb{E}\left[q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{i})\middle| \mathcal{F}_k^N\right]\eqsp,\\
+%& = \sum_{\ell=1}^N \frac{h(\ell)\omega_{k-1}^{\ell}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{\ell},\xi_k^{i})}{\sum_{m=1}^N\omega_{k-1}^{m}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{m},\xi_k^{i})}\eqsp,\\
+%&= \sum_{\ell=1}^N \Lambda_{k-1}^N(i,\ell)h(\ell) \eqsp,
+%\end{align*}
+%which concludes the proof.
+The goal is to simulate realisations of  a discrete random variable $X$ having the following target distribution:
+\begin{equation}
+\mP(X=l)= p_X(l)=\frac{\omega_k^l q_k(\xi_k^l,\xi_{k+1})}{\sum_{l=1}^N\omega_k^l q_k(\xi_k^l,\xi_{k+1})},~~l\in \{1,\dots,N\}
+\end{equation}
+for some $\xi_{k+1}$ that has no importance in this result.
+We assume $\omega_k$ are normalized weights, i.e.
+$\sum_l \omega_k^l=1$.
+The candidate will therefore be sampled from $Y$, a r.v.
+with the following distribution:
+\begin{equation}
+\mP(Y=l)=p_Y(l)= \omega_k^l,~~l\in \{1,\dots,N\}
+\end{equation}
+Unfortunately, the $q_k$ function is unknown.
+Howevere, we let's suppose we can define an unbiased estimator $\hat{q}_k(\xi_k^l,\xi_{k+1},\zeta)$, where $\zeta$ is a random variable having th p.d.f.
+$p_\zeta$.
+We have therefore:
+$$\E_{p_\zeta}\left[\hat{q}_k(\xi_k^l,\xi_{k+1},\zeta)\right]=q(\xi_k^l,\xi_{k+1})$$
+Moreover, we impose the following property over this estimator, we want that:
+$$\exists \sigma_+\in \mathbb{R}\text{ such that }\forall x,y,~~\hat{q}_k(x,y,\zeta)\leq \sigma_+$$
+To simulate a random variable $X\sim p_X$, we propose the following algorithm:
+\begin{enumerate}
+\item Sample independantly $y\sim p_Y$, $z\sim p_\zeta$ and $u \sim \mathcal{U}[0,1]$
+\item \textbf{If} 
+$$u \leq \frac{\hat{q}_k(\xi_k^l,\xi_{k+1},z)}{\sigma_+},$$
+set $x=y$. \textbf{Else}, return to 1).
+\end{enumerate}
+The lemma \ref{lem:AR:unbiased} states that $x$ is a realisation of a random variable $X\sim p_X$. Indeed
 \begin{align*}
-\mathbb{E}\left[h(J^{i,j}_k)\middle| \mathcal{F}_k^N\right] & = \sum_{m\ge 1}\mathbb{E}\left[h(I_m)\1_{\tau=m}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \sum_{m\ge 1}h(m)\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\middle| \mathcal{F}_k^N\right]\right)\mathbb{E}\left[h(I_m)\1_{\mathcal{A}^k_{m}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[1-\frac{\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_\ell},\xi_k^{i},\Lambda_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)\\
-&\hspace{5cm}\times\mathbb{E}\left[h(I_m)\frac{\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_m},\xi_k^{i},\Lambda_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \sum_{m\ge 1}\left(\mathbb{E}\left[1-\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)^{m-1}\mathbb{E}\left[h(I_m)\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \mathbb{E}\left[h(I_1)q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{i})\middle| \mathcal{F}_k^N\right]/\mathbb{E}\left[q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{i})\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \sum_{\ell=1}^N \frac{h(\ell)\omega_{k-1}^{\ell}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{\ell},\xi_k^{i})}{\sum_{m=1}^N\omega_{k-1}^{m}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{m},\xi_k^{i})}\eqsp,\\
-&= \sum_{\ell=1}^N \Lambda_{k-1}^N(i,\ell)h(\ell) \eqsp,
+\mP\left(Y=l\vert U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\vert Y=l \right) \mathbb{P}(Y=l)}{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)}\\
+&=\frac{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^l,\xi_{k+1},\zeta)}{\sigma_+} \right) \omega_k^l}{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)}
+\intertext{We now note that}
+\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\E_{p_\zeta}\left[\E\left[\E_{p_Y}\left[\E\left[\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right) \vert Y\right] \right] \vert \zeta\right]\right]\\
+&=\E_{p_\zeta}\left[\sum_{l=1}^N w_k^l\frac{\hat{q}_k(\xi_k^l,\xi_{k+1},\zeta)}{\sigma_+} \right]~~(\text{As }\frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\leq 1)\\
+&=\frac{1}{\sigma_+}\sum_{l=1}^N w_k^l q_k(\xi_k^l,\xi_{k+1},\zeta)
+\intertext{In the same way, we have:}
+\mP\left(U\leq \frac{\hat{q}_k(\xi_k^l,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{1}{\sigma_+} q_k(\xi_k^l,\xi_{k+1},\zeta)
+\intertext{Which gives overall the wanted result}
+\mP\left(Y=l\vert U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{\omega_k^l q_k(\xi_k^l,\xi_{k+1},\zeta)}{\sum_{l=1}^N w_k^l q_k(\xi_k^l,\xi_{k+1},\zeta)}=p_X(l),~~l\in \{1,\dots,N\}
 \end{align*}
-which concludes the proof.
+%\end{proof}
 \end{proof}

### 2016-11-03 ParisEM.tex
small typos

@@
 \newcommand{\E}{\mathbb{E}}
+\newcommand{\qk}{q_{k}}
+%\newcommand{\qk}{q^{\Delta t_k}_{\theta}}
 \newtheorem{lemma}{Lemma}
@@
 \section{Introduction}
+In this paper, we propose a new algorithm to approximate expectations of additive functionals with respect to the joint smoothing distribution, in the Hidden Markov Model (HMM) framework.\\
 The dynamics of the states $(X_t)_{t\ge 0}$ depends on a parameter $\theta\in \Theta$ where $\Theta$ is a compact subspace of $\mathbb{R}^{\ell}$ and that $(X_t)_{t\ge 0}$ is a weak solution to the following Stochastic Differential Equation (SDE) in $\mathbb{R}^d$:
@@
 The states $(X_t)_{t\ge 0}$ are only partially observed at discrete times $(t_k)_{0\le k \le n}$, with $t_0=0$, through the sequence $(Y_k)_{0\le k\le n}$ which is such that $(Y_k)_{0\le k\le n}$ are independent conditional on $(X_t)_{t\ge 0}$ and for all $0\le k\le n$, the conditional distribution of $Y_k$ given $(X_t)_{t\ge 0}$ depends only on $X_k = X_{t_k}$.
-This distribution has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g_{\theta}(X_k,\cdot)$.
-For all $1\le k \le n$, the distribution of $X_{k} $ conditional on $X_{k-1}$ has a density $q^{\Delta t_k}_{\theta}(X_k,\cdot)$ with respect to a reference measure $\mu$ on $\mathbb{R}^d$.
-
+This distribution has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g_k(X_k) = g(\theta,X_k,\cdot)$. 
+The distribution of $X_0$ has a density with respect to a reference measure $\mu_0$ on $\mathbb{R}^d$ given by $\chi(\cdot):=\chi(\theta,\cdot)$.
+For all $1\le k \le n$, the distribution of $X_{k} $ conditional on $X_{k-1}$ has a density $\qk(X_{k-1},\cdot) :=q(\theta,\Delta_k,X_{k-1},\cdot),~\Delta_k:=t_k-t_{k-1}$ with respect to a reference measure $\mu$ on $\mathbb{R}^d$.\\
+\\
 Statistical inference for partially observed stochastic differential equations usually involves the computation of conditional distributions of some unobserved states given the observations $(Y_0,\ldots,Y_n)$.
@@
 \begin{equation*}
-p_{\theta}(x_{0:n},Y_{0:n}) = \chi_{\theta}(x_0)g_{\theta}(x_0;Y_0)\prod^{n}_{k=1}q^{\Delta t_k}_{\theta}(x_{k-1},x_k)g_{\theta}(x_k,Y_k)\eqsp.
+p_{\theta}(x_{0:n},Y_{0:n}) = \chi(x_0)g_{0}(x_0)\prod^{n}_{k=1}\qk(x_{k-1},x_k)g_{k}(x_k)\eqsp.
 \end{equation*}
@@
 \begin{multline*}
-Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log \left(\chi_{\theta}(X_0)g_{\theta}(X_0;Y_0)\right)\middle|Y_{0:n}\right] \\
-+ \sum_{k=1}^n\mathbb{E}_{\theta_p}\left[\log \left(q^{\Delta t_k}_{\theta}(X_{k-1},X_k)g_{\theta}(X_k,Y_k)\right)\middle|Y_{0:n}\right] \eqsp.
+Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log \left(\chi(X_0)g_{0}(X_0)\right)\middle|Y_{0:n}\right] \\
++ \sum_{k=1}^n\mathbb{E}_{\theta_p}\left[\log \left(\qk(X_{k-1},X_k)g_k(X_k)\right)\middle|Y_{0:n}\right] \eqsp.
 \end{multline*}
@@
  See \cite[Chapter $10$ and $11$]{cappe:moulines:ryden:2005}, \cite{kantas:doucet:signh:2015,doucet:poyiadjis:singh:2011,lecorff:fort:2013a,lecorff:fort:2013b}
-for further references on the use of these smoothed expectations of additive functionals  applied to maximum likelihood parameter inference in latent data models.
-
+for further references on the use of these smoothed expectations of additive functionals  applied to maximum likelihood parameter inference in latent data models.\\
+\\
 The exact computation of these expectations is usually not possible for partially observed diffusions such as \eqref{eq:target:sde}. 
@@
  Then, the backward pass of the FFBS algorithm keeps all the particles sampled during the forward pass and computes new weights to take into account the information brought by all the observations from time $n$ to time $0$.
-Instead of computing new weights, the FFBSi  algorithm samples independently backward in time particle trajectories using the particles and weights produced by the forward pass.
-
-Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly, using the observations as they are received, smoothed expectations of additive functionals.
-Contrary to the FFBS algorithm the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.
-However, there is a major difficulty to apply SMC methods to partially observed SDE as the computation of the importance weights requires to evaluate $q^{\Delta t_k}_{\theta}$ for which there is no analytic expression.
+ Instead of computing new weights, the FFBSi  algorithm samples independently backward in time particle trajectories using the particles and weights produced by the forward pass.\\ 
+\\
+Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e.
+using the observations as they are received) smoothed expectations of additive functionals.
+ Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required. 
+ However, there is a major difficulty to apply SMC methods to partially observed SDE as the computation of the importance weights requires to evaluate $q^{\Delta t_k}_{\theta}$ for which there is no analytic expression in general case.
 %the EM algorithm to partially observed SDE in general cases:
@@
 %\end{enumerate}
-To overcome these difficulties, \cite{olsson:strojby:2011} proposed to use General Poisson Estimators (GPE) introduced in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,fearnhead:papaspiliopoulos:roberts:2008}  to provide an unbiased estimation of $q^{\Delta t_k}_{\theta}$.
-These estimators are based on a representation of $q^{\Delta t_k}_{\theta}$ as an expectation under the probability measure of a Brownian bridge (the conditional expectation involved in the computation is approximated in  \cite{olsson:strojby:2011}  using fixed-lag Sequential Monte Carlo methods developed in \cite{olsson:cappe:douc:moulines:2008}).
-
-In this paper, we propose to improve this algorithm to obtain faster convergence of the sequence of Monte Carlo EM estimates.
+To overcome these difficulties, \cite{olsson:strojby:2011} proposed to use General Poisson Estimators (GPE) introduced in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,fearnhead:papaspiliopoulos:roberts:2008}  to provide an unbiased estimation of $\qk$.
+These estimators are based on a representation of $\qk$ as an expectation under the probability measure of a Brownian bridge (the conditional expectation involved in the computation is approximated in  \cite{olsson:strojby:2011}  using fixed-lag Sequential Monte Carlo methods developed in \cite{olsson:cappe:douc:moulines:2008}).\\
+\\
+In this paper, we propose to improve this algorithm to obtain faster convergence of the sequence of Monte Carlo EM estimates, and to extend the range of diffusion processes for which SMC methods can be applied.\\
 First, we introduce a new class off GPE based on the Ozaki discretization of the SDE  \eqref{eq:target:sde}, see \cite{ozaki:1992,shoji:ozaki:1998}. 
-These new GPE estimators use a an order one Taylor expansion or the drift of \eqref{eq:target:sde} to obtain more efficient estimates than Euler discretization based GPE.
+These new GPE estimators use a an order one Taylor expansion of the drift of \eqref{eq:target:sde} to obtain more efficient estimates than Brownian Bridge based GPE.\\ 
 Instead of using fixed-lag techniques which introduce a systematic additional bias  in the approximation of the conditional expectations, we use the new PaRIS algorithm which allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles.
 The striking result of the application of PaRIS algorithm to SDE is that the accept reject mechanism underlying the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
-
 \section{Random weight PaRIS algorithm}
-%For a given value $\theta_p$ of the parameter, dropping the initial state, the E-step may be written:
-%\[
-%Q(\theta,\theta_{p})= \sum_{k=1}^n\mathbb{E}_{\theta_p}\left[\log \left\{q^{\Delta t_k}_{\theta}(X_{k-1},X_k)g_{\theta}(X_k,Y_k)\right\}\middle|Y_{0:n}\right]\eqsp.
-%\]
-%We first describe the Monte Carlo method used to approximate $Q(\theta,\theta_{p})$ when $q^{\Delta t_k}_{\theta}$ may be computed.
-Dropping $\theta$ from the notations, the E-step computes an expectation of the form
 Let $0 \leq k \leq k' \leq n$. The joint smoothing distributions of the hidden states are defined, for all measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, by:
@@
 \]
-For all $0\le k\le n$, $\phi_{k} = \phi_{k:k|k}$ denotes the filtering distributions. The aim of this section is to detail the extension of PaRIS algorithm to approximate expectations of the form
+For all $0\le k\le n$, $\phi_{k} = \phi_{k:k|k}$ denote the filtering distributions. The aim of this section is to detail the extension of PaRIS algorithm to approximate expectations of the form
 \begin{equation}
@@
 when the transition density of the hidden states is not available explicitly and where $\{h_k\}_{k=1}^n$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$.
-%In this paper, we are interested in a particular class of Sequential Monte Carlo methods to approximate $\bar{H}_{n}$ known as the PaRIS algorithm introduced in \cite{olsson:westerborn:2016}.
-
-%The PaRIS algorithm is based on the first Sequential Monte Carlo algorithms introduced in  \cite{gordon:salmond:smith:1993} and \cite{kitagawa:1996}. 
-Sequential Monte Carlo algorithms introduced in  \cite{gordon:salmond:smith:1993} and \cite{kitagawa:1996} rely on importance sampling and importance resampling steps to approximate the filtering distribution $\phi_{k}$ (conditional distribution of $X_k$ given $Y_{0:k}$) using  a set of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$.
+Sequential Monte Carlo algorithms introduced in  \cite{gordon:salmond:smith:1993} and \cite{kitagawa:1996} rely on importance sampling and importance resampling steps to approximate the filtering distribution $\phi_{k}$ (conditional distribution of $X_k$ given $Y_{0:k}$) using  a set of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$.\\
 At time $k = 0$, $N$ particles $\{\xi^{\ell}_0\}_{\ell=1}^N$ are initialized by setting $\xi^{\ell}_0=x_0$.
@@
 \]
-Then, using $\{(\xi^{\ell}_{k-1},\omega^{\ell}_{k-1})\}_{\ell=1}^N$, pairs $\{(I^{\ell}_t,\xi^{\ell}_{t})\}_{\ell=1}^N$ of indices and particles are sampled and weighted using an instrumental transition density $p_k$ on $\mathbb{R}^d\times \mathbb{R}^d$ and an adjustment multiplier function $\vartheta_k$ on $\mathbb{R}^d$.
+Then, using $\{(\xi^{\ell}_{k-1},\omega^{\ell}_{k-1})\}_{\ell=1}^N$, pairs $\{(I^{\ell}_k,\xi^{\ell}_{k})\}_{\ell=1}^N$ of indices and particles are sampled and weighted using an instrumental transition density $p_k$ on $\mathbb{R}^d\times \mathbb{R}^d$ and an adjustment multiplier function $\vartheta_k$ on $\mathbb{R}^d$.
 Each new particle $\xi^{\ell}_{k}$ and weight $\omega^{\ell}_k$ at time $k$ are computing following these steps:
-\begin{enumerate}[-]
+\begin{enumerate}
 \item choose a particle index $I^{\ell}_k$ at time $k-1$ in $\{1,\ldots,N\}$ with probabilities proportional to $\omega_{k-1}^{j} \vartheta_k (\xi^{j}_{k-1})$, for $j$ in $\{1,\ldots,N\}$ ;
@@
 \[
-\omega^{\ell}_k := \frac{q^{\Delta t_{k}}(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_k (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
+\omega^{\ell}_k := \frac{\qk(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_k (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
 \]
@@
  Nevertheless, the computation of the transition probabilities to select particles in the backward pass makes it a $\mathcal{O}(N^2)$ algorithm.
-In the case where the transition probability $q$ is upper bounded, \cite{douc:garivier:moulines:olsson:2011} introduced an acceptance rejection mechanism to implement the FFBSi algorithm with a $\mathcal{O}(N)$ complexity, see also \cite{dubarry:lecorff:2011}.
-
+In the case where the transition probability $q_k$ is upper bounded, \cite{douc:garivier:moulines:olsson:2011} introduced an acceptance rejection mechanism to implement the FFBSi algorithm with a $\mathcal{O}(N)$ complexity, see also \cite{dubarry:lecorff:2011}.\\
+ \\
 The PaRIS algorithm combines both the forward only version of the FFBS algorithm with the sampling mechanism of the FFBSi algorithm. 
 The algorithm does not produce an approximation of the smoothing distributions but of the smoothed expectation of a fixed additive functional and thus  may be used to approximate \eqref{def:addfunc}. 
-Its crucial property is that it does not require a backward pass, the smoothed expectations is computed on-the-fly with the particle filter and no storage of the particles or weights is needed.
+Its crucial property is that it does not require a backward pass, the smoothed expectations is computed on-the-fly with the particle filter and no storage of the particles or weights is needed.\\ 
 The algorithm is based on sufficient statistics $\tau^i_k$, for $1\le i\le N$ and $0\le k \le n$ starting with $\tau^i_0 = 0$ for all $1\le i\le N$.
@@
 \[
-\Lambda_{k-1}^N(i,\ell) = \frac{\omega^{\ell}_{k-1} q^{\Delta t_k}(\xi^{\ell}_{k-1},\xi_k^{i})}{\sum_{\ell=1}^N\omega^{\ell}_{k-1} q^{\Delta t_k}(\xi^{\ell}_{k-1},\xi_k^{i})}\eqsp,\quad 1\le \ell\le N\eqsp.
+\Lambda_{k-1}^N(i,\ell) = \frac{\omega^{\ell}_{k-1} \qk(\xi^{\ell}_{k-1},\xi_k^{i})}{\sum_{\ell=1}^N\omega^{\ell}_{k-1} \qk(\xi^{\ell}_{k-1},\xi_k^{i})}\eqsp,\quad 1\le \ell\le N\eqsp.
 \]

### 2016-11-04 ParisEM.tex
Small typos
Major changes in the algorithm. Changes in notations with zeta as the random variable for bridges. Attempt of harmonizing notations of indexes k, i, l. Major modifications in the proof of lemma 1. Putting proof of lemma 2 in appendix with minor modifications (adding a step that was not straightforward for me). Addition of intermetiate algorithm between the one of Paris and ours

@@
 \newcommand{\qk}{q_{k}}
+\newcommand{\acom}[1]{\textit{\color{gray} //#1}}
+\newcommand{\Oz}{Z}%Letter for the Ozaki approximation
+\newcommand{\Jk}{J_{\alpha_\theta}^k}%Command for Jacobian of alpha
+\newcommand{\mw}{\mathsf{w}}%For bridge realisations
 %\newcommand{\qk}{q^{\Delta t_k}_{\theta}}
@@
 The distribution of $X_0$ has a density with respect to a reference measure $\mu_0$ on $\mathbb{R}^d$ given by $\chi(\cdot):=\chi(\theta,\cdot)$.
-For all $1\le k \le n$, the distribution of $X_{k} $ conditional on $X_{k-1}$ has a density $\qk(X_{k-1},\cdot) :=q(\theta,\Delta_k,X_{k-1},\cdot),~\Delta_k:=t_k-t_{k-1}$ with respect to a reference measure $\mu$ on $\mathbb{R}^d$.\\
+For all $1\le k \le n$, the distribution of $X_{k+1} $ conditional on $X_{k}$ has a density $\qk(X_{k},\cdot) :=q(\theta,\Delta_k,X_{k-1},\cdot),~\Delta_k:=t_{k+1}-t_{k}$ with respect to a reference measure $\mu$ on $\mathbb{R}^d$.\\
 \\
@@
 \begin{equation*}
-p_{\theta}(x_{0:n},Y_{0:n}) = \chi(x_0)g_{0}(x_0)\prod^{n}_{k=1}\qk(x_{k-1},x_k)g_{k}(x_k)\eqsp.
+p_{\theta}(x_{0:n},Y_{0:n}) = \chi(x_0)g_{0}(x_0)\prod^{n-1}_{k=0}\qk(x_{k},x_{k+1})g_{k+1}(x_{k+1})\eqsp.
 \end{equation*}
@@
 Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log \left(\chi(X_0)g_{0}(X_0)\right)\middle|Y_{0:n}\right] \\
-+ \sum_{k=1}^n\mathbb{E}_{\theta_p}\left[\log \left(\qk(X_{k-1},X_k)g_k(X_k)\right)\middle|Y_{0:n}\right] \eqsp.
++ \sum_{k=0}^{n-1}\mathbb{E}_{\theta_p}\left[\log \left(\qk(X_{k},X_{k+1})g_{k+1}(X_{k+1})\right)\middle|Y_{0:n}\right] \eqsp.
 \end{multline*}
@@
 \label{def:addfunc}
-H_{n} =  \sum_{k=1}^n\mathbb{E}\left[h_k(X_{k-1},X_k)\middle|Y_{0:n}\right]\eqsp,
+H_{n} =  \sum_{k=0}^{n-1}\mathbb{E}\left[h_k(X_{k},X_{k+1})\middle|Y_{0:n}\right]\eqsp,
 \end{equation}
-when the transition density of the hidden states is not available explicitly and where $\{h_k\}_{k=1}^n$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$.
+when the transition density of the hidden states is not available explicitly and where $\{h_k\}_{k=0}^{n-1}$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$.
+The algorithm is based on the following link between the filtering and smoothing distributions:
+\begin{equation}
+\phi_{0:n|n}[h] = \phi_n[\E\left[h(X_{0:n})\vert X_n,~Y_{0:n}]\right]:= \phi_n[T_n(h)] \eqsp.\label{eq:FFbsm:equality}
+\end{equation}
+Therefore, using the additive property and the usual filtering recursion applied on the right hand side of equation \eqref{eq:FFbsm:equality}, an approximation of \eqref{def:addfunc} can be computed using Sequential Monte Carlo techniques.\\
 Sequential Monte Carlo algorithms introduced in  \cite{gordon:salmond:smith:1993} and \cite{kitagawa:1996} rely on importance sampling and importance resampling steps to approximate the filtering distribution $\phi_{k}$ (conditional distribution of $X_k$ given $Y_{0:k}$) using  a set of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$.\\
@@
 \]
-The PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSI algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions.
+The PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSi algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions.
 %Both the FFBS and the FFBSi algorithms require first that a forward pass has produced a set of particles and importance weights $\{ (\xi^{\ell}_k, \omega^{\ell}_k)\}$, for $\ell \in \{1, \ldots, N\}$ and  $k \in \{0,\ldots,n\}$ to approximate $\phi_{k}$ for $k=0$ to $k=n$.
@@
 Its crucial property is that it does not require a backward pass, the smoothed expectations is computed on-the-fly with the particle filter and no storage of the particles or weights is needed.\\ 
-The algorithm is based on sufficient statistics $\tau^i_k$, for $1\le i\le N$ and $0\le k \le n$ starting with $\tau^i_0 = 0$ for all $1\le i\le N$.
+The focus is based sufficient statistics $\tau^i_k$, approximations of $T_k(h)$, for $1\le i\le N$ and $0\le k \le n$ starting with $\tau^i_0 = 0$ for all $1\le i\le N$.
 Let $\tilde{N}\ge 1$.
-Then, at each time step $1\le k \le n$ these statistics are update according to the following steps.
+Then, at each time step $0\le k \le {n-1}$ these statistics are updated according to the following steps.
 \begin{enumerate}[(i)]
 \item \label{it:PaRIS:filt} Run one step of a particle filter to produce $\{(\xi^{\ell}_k, \omega^{\ell}_k)\}$ for $1\le \ell \le N$.
-\item \label{it:PaRIS:sampleindex} For all $1\le i \le N$, sample independently $J_k^{i,\ell}$ in $\{1,\ldots,N\}$ for $1\le \ell \le \widetilde N$ with probabilities $\Lambda_{k-1}^N(i,\cdot)$, given by
+\item \label{it:PaRIS:sampleindex} For all $1\le i \le N$, sample independently $J_{k}^{i,\ell}$ in $\{1,\ldots,N\}$ for $1\le \ell \le \widetilde N$ with probabilities $\Lambda_{k}^N(i,\cdot)$, given by
 \[
-\Lambda_{k-1}^N(i,\ell) = \frac{\omega^{\ell}_{k-1} \qk(\xi^{\ell}_{k-1},\xi_k^{i})}{\sum_{\ell=1}^N\omega^{\ell}_{k-1} \qk(\xi^{\ell}_{k-1},\xi_k^{i})}\eqsp,\quad 1\le \ell\le N\eqsp.
+\Lambda_{k}^N(i,\ell) = \frac{\omega^{\ell}_{k} \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})}{\sum_{\ell=1}^N\omega^{\ell}_{k} \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})}\eqsp,\quad 1\le \ell\le N\eqsp.
 \]
@@
 \[
-\tau^{i}_{k} := \frac{1}{\widetilde{N}} \sum^{\widetilde{N}}_{\ell=1} \left\{ \tau^{J_k^{i,\ell}}_{k-1} + h_{k} \left(\xi^{J_k^{i,\ell}}_{k-1}, \xi^{i}_{k}\right)  \right\}\eqsp.
+\tau^{i}_{k+1} := \frac{1}{\widetilde{N}} \sum^{\widetilde{N}}_{\ell=1} \left\{ \tau^{J_{k}^{i,\ell}}_{k} + h_{k} \left(\xi^{J_{k}^{i,\ell}}_{k}, \xi^{i}_{k+1}\right)  \right\}\eqsp.
 \]
@@
 As proved in \cite{olsson:westerborn:2016}, the algorithm is asymptotically consistent (as $N$ goes to infinity) for any fixed precision parameter $\tilde N$.
-However, there is a significant qualitative difference between the cases $\tilde{N} = 1$ and $\tilde{N} \geq 2$.
+However, there is a significant qualitative difference between the cases $\tilde{N} = 1$ and $\tilde{N} \geq 2$.\\
+As for the FFBSi algorithm,  when there exists $\sigma_+$ such that $0<\qk <\sigma_+$, the PaRIS algorithm may be implemented with $\mathcal{O}(N)$ complexity using the procedure of \cite{douc:garivier:moulines:olsson:2011} recalled in algorithm \ref{alg:AR:knownq}.
+\begin{algorithm}[hb]
+\caption{Random weight Accept-reject-based backward sampling  (\cite{douc:garivier:moulines:olsson:2011})}
+\begin{algorithmic}
+\FORALL{$i \in 1,\dots, N$}
+\FORALL{$\ell \in 1,\dots, \widetilde N$} 
+\STATE \textbf{\sc Sampling Step} Sample independently\\ $j\sim \{\omega_{k}^1,\dots,\omega_{k}^N\}$, $u\sim \mathcal{U}[0,1]$
+\IF{ $$u \leq \frac{\qk(\xi_{k}^j,\xi_{k+1}^i,z)}{\sigma_+},$$}
+\STATE Set $J_k^{i,\ell} = j$.
+\ELSE 
+\STATE Return to \textbf{\sc Sampling Step}
+\ENDIF
+\ENDFOR
+\ENDFOR
+\end{algorithmic}
+\label{alg:AR:knownq}
+\end{algorithm}
 
-As for the FFBSi algorithm,  when there exists $0<q^{\Delta t_k}<\sigma_+$, the PaRIS algorithm may be implemented with $\mathcal{O}(N)$ complexity using the procedure of \cite{douc:garivier:moulines:olsson:2011}.
-As $q^{\Delta t_k}(\xi^{\ell}_{k-1},\xi_k^{i})$ is not available explicitly for all $1\le i,\ell\le N$, the next section introduces GPE to replace it by $\widehat{q}^{\Delta t_k}(\xi^{\ell}_{k-1},\xi_k^{i};\Lambda)$, where $\Lambda$ is a random variable such that
+%\begin{algorithm}
+%\caption{Random weight Accept-reject-based backward sampling  (\cite{douc:garivier:moulines:olsson:2011})}
+%\begin{algorithmic}
+%\FORALL{$i \in 1,\dots,\widetilde N$}
+%\STATE Set $L =\{1,\ldots,N\}$. 
+%\WHILE{$L\neq\emptyset$}
+%\STATE $p = \#L$.
+%\STATE Sample $I = (I_1,\ldots,I_p)$ independently in $\{1,\ldots,N\}$ with probabilities proportional to $\omega_{k-1}^{\ell}$, $1\le \ell\le N$.
+%\STATE Sample $U = (U_1,\ldots,U_p)$ independently and independently of $I$ uniformly in $[0,1]$.
+%\STATE Sample $\Lambda = (\Lambda_1,\ldots,\Lambda_p)$ independently and independently of $I$ and $U$.
+%\STATE Set $L_p = \emptyset$.
+%\FORALL{$\ell \in 1,\dots,p$}
+%\IF{$U_\ell<\widehat{q}(\xi_{k-1}^{I_\ell},\xi_k^{L(\ell)};\Lambda_{\ell})/\sigma_+$}
+%\STATE $J_k^{L(\ell),i} = I_\ell$.
+%\ELSE
+%\STATE $L_p = L_p \cup \{L(\ell)\}$.
+%\ENDIF
+%\ENDFOR
+%\STATE $L = L_p$.
+%\ENDWHILE
+%\ENDFOR
+%\end{algorithmic}
+%\label{alg:AR}
+%\end{algorithm}
+\section{The OzIRis algorithm}
+\label{sec:PaRIS:SDE}
+In general situations, the PaRIS algorithm cannot be used for stochastic differential equations as $\qk$ is unknown.
+Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio in Algorithm~\ref{alg:AR:knownq} cannot be performed to apply the PaRIS algorithm.
+Following \cite{olsson:strojby:2011},
+an extension of the PaRIS algorithm may be developed by replacing $\qk$  by an  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable such that:
 \[
-\mathbb{E}\left[\widehat{q}^{\Delta t_k}(\xi^{\ell}_{k-1},\xi_k^{i};\Lambda)\middle| \mathcal{F}_k^N\right] = q^{\Delta t_k}(\xi^{\ell}_{k-1},\xi_k^{i})\eqsp,
+\mathbb{E}\left[\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta)\middle| \mathcal{F}_{k+1}^N\right] = \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})\eqsp,
 \]
@@
 \[
-\mathcal{F}_k^N = \sigma\left\{Y_{0:k};(\xi^{\ell}_{u},\omega^{\ell}_{u},\tau^{\ell}_{u}), 1\le \ell\le N, 1\le u\le k\right\}\eqsp.
+\mathcal{F}_{k+1}^N = \sigma\left\{Y_{0:k+1};(\xi^{\ell}_{u},\omega^{\ell}_{u},\tau^{\ell}_{u}),~1\le \ell\le N,~0\le u\le k+1\right\}\eqsp.
 \]
-In the case where $\widehat{q}^{\Delta t_k}$ may be upper bounded by $\sigma_+$, step~\eqref{it:PaRIS:sampleindex} is replaced by the acceptance-rejection mechanism given in Algorithm~\ref{alg:AR}.
-It is proved in Lemma~\ref{lem:AR:unbiased} that replacing $\widehat{q}^{\Delta t_k}$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.
-
-\begin{algorithm}
-\caption{Random weight Accept-reject-based backward sampling  (\cite{douc:garivier:moulines:olsson:2011})}
+In the case where $\widehat{q}_k$ may be upper bounded by a constant $\hat{\sigma}_+$, step~\eqref{it:PaRIS:sampleindex} is replaced by the acceptance-rejection mechanism given in Algorithm~\ref{alg:AR:unknownq}.
+The Lemma~\ref{lem:AR:unbiased} states that replacing  $\qk$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.
+\begin{lemma}
+\label{lem:AR:unbiased}
+Let $0\le k\le n-1$.
+For all $1\le i \le N$ and all $1\le \ell \le \widetilde{N}$, the conditional probability distribution given $\mathcal{F}_{k+1}^N$ of the random variable $J_{k}^{i,\ell}$ produced by Algorithm~\ref{alg:AR:unknownq} is  $\Lambda_{k}^N(i,\cdot)$.
+\end{lemma}
+\begin{proof}
+See Appendix \ref{sec:append:proofs}.
+\end{proof}
+\begin{algorithm}[hb]
+\caption{Double random weight Accept-reject-based backward sampling}
 \begin{algorithmic}
-\FORALL{$i \in 1,\dots,\widetilde N$}
-\STATE Set $L =\{1,\ldots,N\}$. 
-\WHILE{$L\neq\emptyset$}
-\STATE $p = \#L$.
-\STATE Sample $I = (I_1,\ldots,I_p)$ independently in $\{1,\ldots,N\}$ with probabilities proportional to $\omega_{k-1}^{\ell}$, $1\le \ell\le N$.
-\STATE Sample $U = (U_1,\ldots,U_p)$ independently and independently of $I$ uniformly in $[0,1]$.
-\STATE Sample $\Lambda = (\Lambda_1,\ldots,\Lambda_p)$ independently and independently of $I$ and $U$.
-\STATE Set $L_p = \emptyset$.
-\FORALL{$\ell \in 1,\dots,p$}
-\IF{$U_\ell<\widehat{q}(\xi_{k-1}^{I_\ell},\xi_k^{L(\ell)};\Lambda_{\ell})/\sigma_+$}
-\STATE $J_k^{L(\ell),i} = I_\ell$.
+\FORALL{$i \in 1,\dots, N$}
+\FORALL{$\ell \in 1,\dots, \widetilde N$} 
+\STATE \textbf{\sc Sampling Step} Sample independently\\ $j\sim \{\omega_{k}^1,\dots,\omega_{k}^N\}$, $z\sim\zeta_k$, $u\sim \mathcal{U}[0,1]$
+\IF{ $$u \leq \frac{\widehat{\qk}(\xi_{k}^j,\xi_{k+1}^i,z)}{\hat{\sigma}_+},$$}
+\STATE Set $J_k^{i,\ell} = j$.
 \ELSE 
-\STATE $L_p = L_p \cup \{L(\ell)\}$.
+\STATE Return to \textbf{\sc Sampling Step}
 \ENDIF
 \ENDFOR
-\STATE $L = L_p$.
-\ENDWHILE
 \ENDFOR
 \end{algorithmic}
-\label{alg:AR}
+\label{alg:AR:unknownq}
 \end{algorithm}
-
-\begin{lemma}
-\label{lem:AR:unbiased}
-Let $1\le k\le n$.
-For all $1\le i \le N$ and all $1\le j \le \widetilde{N}$, the conditional probability distribution given $\mathcal{F}_k^N$ of the random variable $J_k^{i,j}$ produced by Algorithm~\ref{alg:AR} is  $\Lambda_{k-1}^N(i,\cdot)$.
-\end{lemma}
-
-\section{Ozaki-PaRIS algorithm}
-\label{sec:PaRIS:SDE}
-In general situations, the PaRIS algorithm cannot be used for stochastic differential equations as $q^{\Delta t_k}_{\theta}$ is unknown.
-Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio in Algorithm~\ref{alg:AR} cannot be performed to apply the PaRIS algorithm.
-Following \cite{olsson:strojby:2011},
-an extension of the PaRIS algorithm may be developed by replacing $q^{\Delta t_k}_{\theta}$  by an unbiased estimator.
-In \cite{olsson:strojby:2011},  GPE are used to estimate $q^{\Delta t_k}_{\theta}$ which relies on a representation of $q^{\Delta t_k}_{\theta}$ as an expectation under the probability measure of a Brownian bridge.
-In order to improve the Monte Carlo estimation of $q^{\Delta t_k}_{\theta}$,  other diffusion bridges may be used as long as they can be simulated efficiently.
+In \cite{olsson:strojby:2011},  Global Poisson Estimator (GPE) are used to estimate $\qk$ which relies on a representation of $\qk$ as an expectation under the probability measure of a Brownian bridge.
+In order to improve the Monte Carlo estimation of $\qk$,  other diffusion bridges may be used as long as they can be simulated efficiently.\\ 
 A simple idea is to consider a diffusion obtained as an approximation of the target SDE \eqref{eq:target:sde}. 
 \cite{ozaki:1992,shoji:ozaki:1998} proposed a higher order approximation of the drift term $\alpha_{\theta}$ than the Euler scheme. 
-The authors used a linear approximation of the diffusion drift $\alpha_{\theta}$, together with a constant approximation of the volatility over each time interval $(t_{k-1},t_k)$.
-Following \cite{ozaki:1992,shoji:ozaki:1998}, for all $1\le k\le n$, let $(Y^{k}_t)_{t\ge 0}$ be governed by the following stochastic differential equation for $t\in(t_{k-1},t_k)$:
+The authors used a linear approximation of the diffusion drift $\alpha_{\theta}$, together with a constant approximation of the volatility over each time interval $(t_{k},t_{k+1})$.\\ 
+Formally, for all $0\le k\le n-1$, let $(Z^{k}_t)_{t\ge 0}$ be governed by the following stochastic differential equation for $t\in(t_{k},t_{k+1})$:
 \begin{equation}
 \label{eq:SDE:Ozaki}
-Y^k_{t_{k-1}} = X_{k-1}\quad\mbox{and}\quad\rmd Y^k_t = \widetilde{\alpha}^k_{\theta}(Y^k_t)\rmd t + \rmd W_t\eqsp,
+\Oz^k_{t_{k}} = X_{k}\quad\mbox{and}\quad\rmd \Oz^k_t = \widetilde{\alpha}^k_{\theta}(\Oz^k_t)\rmd t + \rmd W_t\eqsp,
 \end{equation}
@@
 \[
-\widetilde{\alpha}^k_{\theta}(x) = J_{\alpha_{\theta}}(X_{k-1})x + \alpha_{\theta}(X_{k-1}) - J_{\alpha_{\theta}}(X_{k-1})X_{k-1}
+\widetilde{\alpha}^k_{\theta}(x) = \Jk x + \alpha_{\theta}(X_{k}) - \Jk X_{k}
 \]
-and $J_{\alpha_{\theta}}$ is the Jacobian of $\alpha_{\theta}$.
-The probability density function of the transition kernel associated with $Y^k$ is written $\widetilde{q}^{\Delta t}_{\theta}$ and is given in Section~\ref{sec:ornstein:bridge}.
-
+and $\Jk$ is the Jacobian of $\alpha_{\theta}$ evaluated in $X_k$.
+The probability density function of the transition kernel associated with $\Oz^k$ is written $\widetilde{q}_k$ and is given in Section~\ref{sec:ornstein:bridge}.
 Writing $\mathbb{Q}^k_{\theta}$ (resp.
-$\mathbb{W}^{k}_{\theta}$) the probability measure induced by $(X_s)_{t_{k-1}\le s\le t_k}$  (resp.
-$(Y_s)_{t_{k-1}\le s\le t_k}$) on $(\mathsf{C},\mathcal{C})$, Girsanov theorem yields:
+$\mathbb{W}^{k}_{\theta}$) the probability measure induced by $(X_s)_{t_{k}\le s\le t_{k+1}}$  (resp.
+$(\Oz_s)_{t_{k}\le s\le t_{k+1}}$) on $(\mathsf{C},\mathcal{C})$, Girsanov theorem yields:
 \[
-\frac{\rmd \mathbb{Q}^k_{\theta}}{\rmd \mathbb{W}^{k}_{\theta}}(w) = \mathrm{exp}\left\{H_{\theta}(\omega_{t_k}) - H_{\theta}(\omega_{t_{k-1}}) - \frac{1}{2}\int_{t_{k-1}}^{t_k}\left\{\|\alpha_{\theta}(\omega_s)-\widetilde{\alpha}_{\theta}(\omega_s)\|^2+\Delta H_{\theta}(\omega_s)\right\}\rmd s\right\}\eqsp,
+\frac{\rmd \mathbb{Q}^k_{\theta}}{\rmd \mathbb{W}^{k}_{\theta}}(w) = \mathrm{exp}\left\{H_{\theta}(\omega_{t_{k+1}}) - H_{\theta}(\omega_{t_{k}}) - \frac{1}{2}\int_{t_{k}}^{t_{k+1}}\left\{\|\alpha_{\theta}(\mw_s)-\widetilde{\alpha}_{\theta}(\mw_s)\|^2+\Delta H_{\theta}(\mw_s)\right\}\rmd s\right\}\eqsp,
 \]
@@
 \[
-H_{\theta}(x) = A_{\theta}(x) - x'J_{\alpha_{\theta}}(X_{k-1})x/2- (\alpha_{\theta}(X_{k-1}) - J_{\alpha_{\theta}}(X_{k-1})X_{k-1})'x
+H_{\theta}(x) = A_{\theta}(x) - \frac{1}{2}x'\Jk x- (\alpha_{\theta}(X_{k}) - \Jk X_{k-1})'x
 \]
-and $\Delta$ is the Laplace operator. Then, the probability measures conditioned on hitting $X_{k}$ at time $t_k$ satisfy:
+and $\Delta$ is the Laplace operator. Then, the probability measures conditioned on hitting $X_{k+1}$ at time $t_{k+1}$ satisfy:
 \begin{multline}
 \label{eq:Girsanov}
-\frac{\rmd \mathbb{Q}^{k,X_{k}}_{\theta}}{\rmd \mathbb{W}^{k,X_{k}}_{\theta}}(w) = \frac{\widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_{k})}{q^{\Delta t_k}_{\theta}(X_{k-1},X_{k})}\mathrm{exp}\left\{H_{\theta}(X_k) - H_{\theta}(X_{k-1})\right\} \\
-\times \mathrm{exp}\left\{- \frac{1}{2}\int_{t_{k-1}}^{t_k}\left\{\|\alpha_{\theta}(\omega_s)-\widetilde{\alpha}_{\theta}(\omega_s)\|^2+\Delta H_{\theta}(\omega_s)\right\}\rmd s\right\}\eqsp.
+\frac{\rmd \mathbb{Q}^{k,X_{k+1}}_{\theta}}{\rmd \mathbb{W}^{k,X_{k+1}}_{\theta}}(\mw) = \frac{\widetilde{\qk}(X_{k},X_{k+1})}{\qk (X_{k},X_{k+1}) }\mathrm{exp}\left\{H_{\theta}(X_{k+1}) - H_{\theta}(X_{k})\right\} \\
+\times \mathrm{exp}\left\{- \frac{1}{2}\int_{t_{k}}^{t_{k+1}}\left\{\|\alpha_{\theta}(\mw_s)-\widetilde{\alpha}_{\theta}(\mw_s)\|^2+\Delta H_{\theta}(\mw_s)\right\}\rmd s\right\}\eqsp.
 \end{multline}
@@
 \begin{enumerate}[(a)]
-\item For all $\theta\in\Theta$ and all $0\le k \le n-1$, $x\mapsto \alpha_{\theta}(x) - \alpha_{\theta}(X_k) - J_{\alpha_{\theta}}(X_{k})(x-X_k)$ is bounded.
-\item For all $\theta\in\Theta$ and all $0\le k \le n-1$, $x\mapsto \mathrm{Tr}\left(J_{\alpha_{\theta}}(x) - J_{\alpha_{\theta}}(X_k)\right)$ is bounded, where $\mathrm{Tr}$ is the trace operator.
+\item For all $\theta\in\Theta$ and all $0\le k \le n-1$, $x\mapsto \alpha_{\theta}(x) - \alpha_{\theta}(X_k) - \Jk(x-X_k)$ is bounded.
+\item For all $\theta\in\Theta$ and all $0\le k \le n-1$, $x\mapsto \mathrm{Tr}\left(J_{\alpha_{\theta}}(x) - \Jk\right)$ is bounded, where $\mathrm{Tr}$ is the trace operator.
 \end{enumerate}
@@
 \begin{multline*}
-q^{\Delta t_k}_{\theta}(X_{k-1},X_k) = \widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k) \eqsp\mathrm{exp}\left\{H_{\theta}(X_k) - H_{\theta}(X_{k-1})-\ell(\theta)\Delta t_k\right\}\\
-\times\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\right]\eqsp,
+\qk(X_{k},X_{k+1}) = \widetilde{\qk}(X_{k},X_{k+1}) \eqsp\mathrm{exp}\left\{H_{\theta}(X_{k+1}) - H_{\theta}(X_{k})-\ell(\theta)\Delta_k\right\}\\
+\times\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{ \Delta_k}\phi_{\theta}(\mw_s)\rmd s\right\}\right]\eqsp,
 \end{multline*}
-where $\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}$ is the expectation under the law of the bridge process defined by \eqref{eq:SDE:Ozaki}, starting at $X_{k-1}$ at time $0$ and ending at $X_{k}$ at time $\Delta t_k$.
-An unbiased estimator of $q^{\Delta t_k}_{\theta}(X_{k-1},X_k)$ is then obtained by estimating $\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\right]$.
-Following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, if $\omega \sim\mathbb{W}_{\theta}^{k,X_{k}}$, $\kappa$ is a random variables taking values in $\mathbb{N}$ with distribution $\mu$ and if $(U_j)_{1\le j\le \kappa}$ are independent uniform random variables on $[0,t]$,
+where $\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}$ is the expectation under the law of the bridge process defined by \eqref{eq:SDE:Ozaki}, starting at $X_{k}$ at time $0$ and ending at $X_{k+1}$ at time $\Delta_k$.
+An unbiased estimator of $\qk(X_{k},X_{k+1})$ is then obtained by estimating $\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{ \Delta_k}\phi_{\theta}(\mw_s)\rmd s\right\}\right]$.
+Following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, if $\omega \sim\mathbb{W}_{\theta}^{k,X_{k}}$, $\kappa$ is a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and if $(U_j)_{1\le j\le \kappa}$ are independent uniform random variables on $[0,t]$,
 \begin{align*}
-\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\right] &= \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\right]\eqsp,\\
-&= \mathrm{exp}\left\{-\Lambda_{\theta}\Delta t_k\right\}\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathrm{exp}\left\{\int_{0}^{\Delta t_k}(\Lambda_{\theta}-\phi_{\theta}(\omega_s))\rmd s\right\}\right]\eqsp,\\
-&=\mathrm{exp}\left\{-\Lambda_{\theta}\Delta t_k\right\}\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\sum_{p\ge 0}\frac{1}{p!}\left(\int_{0}^{\Delta t_k}\Lambda_{\theta}-\phi_{\theta}(\omega_s)\rmd s\right)^p\right]\eqsp,\\
-&=\mathrm{exp}\left\{-\Lambda_{\theta}\Delta t_k\right\}\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\mathbb{E}\left[\sum_{p\ge 0}\frac{(\Delta t_k)^p}{k!}\prod_{j=1}^p\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\right]\right]\eqsp,\\
-&=\mathrm{exp}\left\{-\Lambda_{\theta}\Delta t_k\right\}\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k}}}\left[\frac{(\Delta t_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\right]\eqsp.
+\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta_k}\phi_{\theta}(\mw_s)\rmd s\right\}\right] &= \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}} \left[\mathrm{exp}\left\{-\int_{0}^{\Delta_k}\phi_{\theta}(\mw_s)\rmd s\right\}\right]\eqsp,\\
+&= \mathrm{exp}\left\{-\Lambda_{\theta} \Delta_k\right\} \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}} \left[\mathrm{exp}\left\{ \int_{0}^{\Delta_k}(\Lambda_{\theta}-\phi_{\theta}(\mw_s))\rmd s\right\}\right]\eqsp,\\
+&=\mathrm{exp}\left\{-\Lambda_{\theta}\Delta_k\right\} \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}\left[\sum_{p\ge 0}\frac{1}{p!}\left(\int_{0}^{\Delta_k}\Lambda_{\theta}-\phi_{\theta}(\mw_s)\rmd s\right)^p\right]\eqsp,\\
+&=\mathrm{exp}\left\{-\Lambda_{\theta}\Delta_k \right\} \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}} \left[\mathbb{E}\left[\sum_{p\ge 0}\frac{( \Delta_k)^p}{k!} \prod_{j=1}^p\left(\Lambda_{\theta}-\phi_{\theta}(\mw_{U_j})\right)\right]\right]\eqsp,\\
+&=\mathrm{exp}\left\{-\Lambda_{\theta} \Delta_k\right\} \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}\left[ \frac{(\Delta_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\mw_{U_j})\right)\right]\eqsp.
 \end{align*}
@@
 \[
-\Lambda_k = \left\{\kappa,\omega,U_1,\ldots,U_\kappa\right\}\eqsp,
+\zeta_k = \left\{\kappa,\mw,U_1,\ldots,U_\kappa\right\}\eqsp,
 \]
-Algorithm~\ref{alg:AR} is used with 
-\begin{multline*}
-\widehat{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k;\Lambda) = \widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k) \eqsp\mathrm{exp}\left\{H_{\theta}(X_k) - H_{\theta}(X_{k-1})-\ell(\theta)\Delta t_k\right\}\\
-\times\mathrm{exp}\left\{-\Lambda_{\theta}\Delta t_k\right\}\frac{(\Delta t_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\eqsp.
-\end{multline*}
-The choice of $\mu$ is restricted to distributions such that the associated estimator is uper bounded by an eplicit constant $\sigma_+^k(\theta)$.
-Then, for all $1\le k \le n$, $1\le \ell\le N$, the importance weight $\omega_k^{\ell}$ is replaced by the unbiased estimator:
+Algorithm~\ref{alg:AR:unknownq} is used with 
+\begin{multline}
+\widehat{\qk}(X_{k},X_{k+1};\zeta_k) = \widetilde{\qk}(X_{k},X_{k+1}) \eqsp\mathrm{exp}\left\{H_{\theta}(X_{k+1}) - H_{\theta}(X_{k})-\ell(\theta) \Delta_k\right\}\\ 
+\times\mathrm{exp}\left\{-\Lambda_{\theta}\Delta_k\right\}\frac{( \Delta_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\mw_{U_j})\right)\eqsp.\label{eq:unbiased:q}
+\end{multline}
+The choice of $\mu$ is restricted to distributions such that the associated estimator is upper bounded by an explicit constant $\hat{\sigma}_+^k(\theta)$.
+Then, for all $0\le k \le n-1$, $1\le i\le N$, the importance weight $\omega_k^{\ell}$ is replaced by the unbiased estimator:
 \begin{equation}
 \label{eq:random:weight}
-\widetilde{\omega}_k^{\ell} = \frac{M^{-1}\sum_{i=1}^M\widehat{q}_{\theta}^{\Delta t_{k}}(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k;\Lambda_k^i)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_k (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp,
+\widetilde{\omega}_{k+1}^{i} = \frac{M^{-1}\sum_{m=1}^M \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta_k^m)g_{k+1}(\xi^{i}_{k+1})}{\vartheta_{k+1}(\xi^{I^{i}_{k+1}}_{k}) p_{k} (\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})}\eqsp,
 \end{equation}
-where $\{\Lambda_k^i\}_{i=1}^M$ are independent copies of $\Lambda_k$. 
+where $\{\zeta_k^i\}_{i=1}^M$ are independent copies of $\zeta_k$.\\
 %The last step to approximate the intermediate quantity of the EM algorithm is to provide an unbiased estimate of $\log q^{\Delta t_k}_{\theta}(X_{k-1},X_k)$.
@@
 %\log q^{\Delta t_k}_{\theta}(X_{k-1},X_k) = \log \widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k) + H_{\theta}(X_k) - H_{\theta}(X_{k-1})-\ell(\theta)\Delta t_k \\
-%-\mathbb{E}_{\mathbb{Q}_{\theta}^{k,X_{k}}}\left[\int_{0}^{\Delta t_k}\phi_{\theta}(\omega_s)\rmd s\right]\eqsp.
+%-\mathbb{E}_{\mathbb{Q}_{\theta}^{k,X_{k}}}\left[\int_{0}^{\Delta t_k}\phi_{\theta}(\mw_s)\rmd s\right]\eqsp.
 %\end{multline*}
@@
 %\frac{\rmd \mathbb{Q}^{k,X_{k}}_{\theta}}{\rmd \mathbb{W}^{k,X_{k}}_{\theta}}(w) \propto 
-%\mathrm{exp}\left\{-\int_{t_{k-1}}^{t_k}\phi_{\theta}(\omega_s)\rmd s\right\}\eqsp.
+%\mathrm{exp}\left\{-\int_{t_{k-1}}^{t_k}\phi_{\theta}(\mw_s)\rmd s\right\}\eqsp.
 %\]
-%As noted by \cite{}, this importance ratio is proportional to the probability that a marked Poisson process on $(0,\Delta t_k)\times (0,1)$ with intensity $\Lambda_{\theta}$ is below the graph $s \mapsto \phi_{\theta}(\omega_s)/\Lambda_{\theta}$.
-$\omega_U$ in \eqref{eq:logq} is then obtaine by first sampling $\tilde{\omega}\sim \mathbb{W}^{k,X_{k}}$ at times on $(0,\Delta t_k)$ defined by the marked Poison process (which is accepted if is is below $\phi_{\theta}(\omega_s)/\Lambda_{\theta}$) and then by interpolating the bridge at time $U$.
-All these steps are summarized in Algorithm~\ref{alg:Ozaki:PaRIS} which provides an approximation of \eqref{def:addfunc}. %the E step of the EM algorithm.
+%As noted by \cite{}, this importance ratio is proportional to the probability that a marked Poisson process on $(0,\Delta t_k)\times (0,1)$ with intensity $\Lambda_{\theta}$ is below the graph $s \mapsto \phi_{\theta}(\mw_s)/\Lambda_{\theta}$.
+$\omega_U$ in \eqref{eq:logq} is then obtaine by first sampling $\tilde{\omega}\sim \mathbb{W}^{k,X_{k}}$ at times on $(0,\Delta t_k)$ defined by the marked Poison process (which is accepted if is is below $\phi_{\theta}(\mw_s)/\Lambda_{\theta}$) and then by interpolating the bridge at time $U$.
+The resulting Ozaki In PaRIS (OzIRis) algortihm, which provides an approximation of \eqref{def:addfunc}, is summarized in Algorithm~\ref{alg:Ozaki:PaRIS}.%the E step of the EM algorithm.
 \begin{algorithm}
-\caption{Ozaki-PaRIS algorithm}
+\caption{OzIRis algorithm}
 \begin{algorithmic}
@@
 \ENDFOR
-\FORALL{$k \in 1,\dots, n$}
-\FORALL{$\ell \in 1,\dots, N$}
-\STATE Choose $I_k^{\ell}$ in $\{1,\ldots,N\}$ with probabilities propotional to $\widetilde{\omega}_{k-1}^j\vartheta_k(\xi_{k-1}^j)$.
-\STATE Sample $\xi_k^{\ell} \sim p_k(\xi_{k-1}^{I_k^{\ell}},\cdot)$.
-\STATE Sample $\kappa \sim \mu$, $\omega\sim \mathbb{W}_{\theta}^{k,X_{k}}$ and $(U_j)_{1\le j\le \kappa}$ uniformly on $(0,\Delta t_k)$.
-\STATE Compute $\widetilde{\omega}^{\ell}_k$ as in \eqref{eq:random:weight}.
-\ENDFOR
-\FORALL{$i \in 1,\dots,\widetilde N$}
-\STATE Set $L =\{1,\ldots,N\}$. 
-\WHILE{$L\neq\emptyset$}
-\STATE $p = \#L$.
-\STATE Sample $(I_1,\ldots,I_p)$ in $\{1,\ldots,N\}$ with probabilities prop. to $\omega_{k-1}^{\ell}$.
-\STATE Sample $(V_1,\ldots,V_p)$ uniformly in $[0,1]$.
-\STATE $L_p = \emptyset$
-\FORALL{$\ell \in 1,\dots,p$}
-\STATE Sample $\kappa_{\ell}\sim \mu$.
-\STATE Sample $U^{\kappa_{\ell}}_j$, $1\le j\le \kappa_{\ell}$, independently and uniformly on $(0,\Delta t_k)$.
-\STATE Sample $\omega^{\kappa_{\ell}}_{s}$ at times $U^{\kappa_{\ell}}_j$, $1\le j\le \kappa_{\ell}$, where $\omega^{\kappa_{\ell}}\sim \mathbb{W}_{\theta,\xi_{k-1}^{I_\ell}}^{k,\xi_k^{L(\ell)}}$.
-\IF{$V_\ell<\widetilde{q}^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_\ell},\xi_k^{L(\ell)};\kappa_{\ell},\omega^{\kappa_{\ell}},U^{\kappa_{\ell}}_1,\ldots,U^{\kappa_{\ell}}_{\kappa_{\ell}})/\sigma_+^k(\theta)$}
-\STATE $J_k^{L(\ell),i} = I_\ell$.
+\FOR{$k \in 0,\dots, n-1$}
+\FORALL{$i \in 1,\dots, N$}
+\STATE Set $\tau_{k+1}^i=0$;
+\STATE Choose $I_{k+1}^{i}$ in $\{1,\ldots,N\}$ with probs $\propto \{\widetilde{\omega}_{k}^1\vartheta_{k+1}(\xi_{k}^1),\dots,\widetilde{\omega}_{k}^N\vartheta_{k+1}(\xi_{k}^N)\}$;
+\STATE Sample $\xi_{k+1}^{i} \sim p_k(\xi_{k}^{I_{k+1}^{i}},\cdot)$.
+\STATE Sample  $\{z_k^m=(\kappa_m, \mw_m, (U_j^m)_{1\leq j\leq \kappa_m})\}_{m=1}^M\overset{i.i.d}{\sim}  (\mu, \mathbb{W}_\theta^{k,X_{k+1}},\mathcal{U}[0,\Delta_k]^{\otimes \kappa_m})$.
+\STATE Compute $\widetilde{\omega}^{i}_{k+1}$ using equation \eqref{eq:random:weight}.
+\FORALL{$\ell \in 1,\dots, \widetilde N$} 
+\STATE \textbf{\sc Sampling Step} Sample independently\\ $j\sim \{\omega_{k}^1,\dots,\omega_{k}^N\}$, $z\sim\zeta_k$, $u\sim \mathcal{U}[0,1]$
+\IF{ $$u \leq \frac{\widehat{\qk}(\xi_{k}^j,\xi_{k+1}^i,z)}{\hat{\sigma}_+},$$}
+\STATE Set $J_k^{i,\ell} = j$.
+\STATE Update $\tau_{k+1}^i = \tau_{k+1}^i + \left(\tau^{J_k^{i,\ell}}_{k} + h_k(\xi^{J_k^{i,\ell}}_{k},\xi^i_{k+1})\right)/\tilde{N}$;
 \ELSE 
-\STATE $L_p = L_p \cup \{L(\ell)\}$.
+\STATE Return to \textbf{\sc Sampling Step}
 \ENDIF
 \ENDFOR
-\STATE $L = L_p$
-\ENDWHILE
-\ENDFOR
-\FORALL{$i \in 1,\dots,N$}
-\STATE Set $\tau_k^i = 0$.
-\FORALL{$\ell \in 1,\dots,\widetilde{N}$}
-%\REPEAT
-%\STATE $\kappa \sim \mathcal{P}(\Lambda_{\theta}\Delta t_k)$, $(U_{m},\chi_m$) uniformly on $(0,\Delta t_k)\times(0,1)$, $1\le m \le \kappa$.
-%\STATE Sample $\widetilde{\omega}_s\sim \mathbb{W}_{\theta}^{k,X_{k}}$ at times $U_{m}$, $1\le m \le \kappa$.
-%\UNTIL{$\phi_{\theta}(\widetilde{\omega}_{U_m})/\Lambda_{\theta}<\chi_m$ for all $1\le m \le \kappa$}
-%\STATE Sample $U$ uniformly in $(0,\Delta t_k)$.
-\STATE $\tau_k^i = \tau_k^i + \tau^{J_k^{i,\ell}}_{k-1} + h_k(\xi^{J_k^{i,\ell}}_{k-1},\xi^i_k)$.%\log g_{\theta}(\xi^i_k,Y_k) + \mathsf{q}_{\theta}^{\Delta t_k}(\xi^{J_k^{i,\ell}}_{k-1}, \xi^{i}_{k},\widetilde{\omega}_U)$.
-\ENDFOR
-\STATE Set $\tau^{i}_{k} = \tau^{i}_{k}/\widetilde{N}$.
 \ENDFOR
@@
 \begin{proof}
-By \eqref{eq:random:weight},
-\[
-\mathbb{E}\left[\tilde{\omega}^1_{k+1}\tau^{1}_{k+1}\middle| \mathcal{F}_k^{N}\right] = \mathbb{E}\left[\frac{M^{-1}\sum_{i=1}^M\widehat{q}_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{I^{1}_{k+1}},\xi^{1}_{k+1};\Lambda^i_{k+1})g_{k+1}(\xi^{1}_{k+1})}{\vartheta_{k+1}(\xi^{I^{1}_{k+1}}_{k}) p_{k+1}(\xi_{k}^{I^{1}_{k+1}},\xi^{\ell}_{k+1})}\tau^{1}_{k+1}\middle| \mathcal{F}_k^{N}\right]\eqsp.
-%&= \mathbb{E}\left[\frac{g_{k+1}(\xi^{1}_{k+1})}{\vartheta_{k+1}(\xi^{I^{1}_{k+1}}_{k}) p_{k+1}(\xi_{k}^{I^{1}_{k+1}},\xi^{\ell}_{k+1})}\tau^{1}_{k+1}\widehat{q}_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{I^{1}_{k+1}},\xi^{1}_{k+1};\Lambda_{k+1})\middle| \mathcal{F}_k^{N}\right]\eqsp,
-\]
-Note that
-\begin{align*}
-&\mathbb{E}\left[\tau^{1}_{k+1}\middle|\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,\omega_{k+1}^i\}\right)\right]
- = \sum_{\ell=1}^N\frac{\omega_k^{\ell}q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{\ell},\xi^{1}_{k+1})\left(\tau^{\ell}_k + h_{k+1}(\xi_{k}^{\ell},\xi^{1}_{k+1})\right)}{\sum_{\ell'=1}^N\omega_k^{\ell'}q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{\ell'},\xi^{1}_{k+1})}\eqsp,\\
-&\mathbb{E}\left[M^{-1}\sum_{i=1}^M\widehat{q}_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{I^{1}_{k+1}},\xi^{1}_{k+1};\Lambda^i_{k+1})\middle|\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,\omega_{k+1}^i\}\right)\right]
- = q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{I^{1}_{k+1}},\xi^{1}_{k+1})\eqsp.
-\end{align*}
-Since $\tau^{1}_{k+1}$ and $\{\Lambda^i_{k+1}\}_{i=1}^M$ are independent conditionally to $\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^j,I_{k+1}^j\}_{j=1}^N\right)$:
-\begin{multline*}
-\mathbb{E}\left[\tau^{1}_{k+1}M^{-1}\sum_{i=1}^M\widehat{q}_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{I^{1}_{k+1}},\xi^{1}_{k+1};\Lambda^i_{k+1})\middle|\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,\omega_{k+1}^i\}\right)\right]\\
- = q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{I^{1}_{k+1}},\xi^{1}_{k+1})\sum_{\ell=1}^N\frac{\omega_k^{\ell}q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{\ell},\xi^{1}_{k+1})\left(\tau^{\ell}_k + h_{k+1}(\xi_{k}^{\ell},\xi^{1}_{k+1})\right)}{\sum_{\ell'=1}^N\omega_k^{\ell'}q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{\ell'},\xi^{1}_{k+1})}\eqsp.
-\end{multline*}
-Using , this yields:
-\begin{align*}
-\mathbb{E}\left[\tilde{\omega}^1_{k+1}\tau^{1}_{k+1}\middle| \mathcal{F}_k^{N}\right]&= \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1} \sum_{i=1}^N\frac{\omega_k^i}{\Omega_k}\int\vartheta_{k+1}(\xi^{i}_{k})\frac{q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{i},x)g_{k+1}(x)}{\vartheta_{k+1}(\xi^{i}_{k}) p_{k+1}(\xi_{k}^{i},x)}\\
-&\hspace{1cm}\times \sum_{\ell=1}^N\frac{\omega_k^{\ell}q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{\ell},x)\left(\tau^{\ell}_k + h_{k+1}(\xi_{k}^{\ell},x)\right)}{\sum_{\ell'=1}^N\omega_k^{\ell'}q_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{\ell'},x)}p_{k+1}(\xi_{k}^{i},x)\rmd x\eqsp,\\
-& =\left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\int q_{\theta}^{\Delta t_{k+1}}(\cdot,x)g_{k+1}(x)\left\{\tau_k(\cdot) + h_{k+1}(\cdot,x)\right\}\rmd x\right]\eqsp,
-\end{align*}
-which concludes the proof.
+See appendix \ref{sec:append:proofs}
 \end{proof}
@@
 % \[
-% \frac{\rmd \mathbb{Q}_{\theta}}{\rmd \mathbb{W}^{\beta}_{\theta}}(w) = \mathrm{exp}\left\{H_{\theta}(\omega_t) - H_{\theta}(\omega_0) - \frac{1}{2}\int_{0}^{t}\left\{\|\alpha_{\theta}(\omega_s)-\beta\|^2+\Delta H_{\theta}(\omega_s)\right\}\rmd s\right\}\eqsp,
+% \frac{\rmd \mathbb{Q}_{\theta}}{\rmd \mathbb{W}^{\beta}_{\theta}}(w) = \mathrm{exp}\left\{H_{\theta}(\omega_t) - H_{\theta}(\omega_0) - \frac{1}{2}\int_{0}^{t}\left\{\|\alpha_{\theta}(\mw_s)-\beta\|^2+\Delta H_{\theta}(\mw_s)\right\}\rmd s\right\}\eqsp,
 % \]
@@
 % \[
-% \frac{\rmd \mathbb{Q}^{y}_{\theta}}{\rmd \mathbb{W}^{\beta,y}_{\theta}}(w) = \frac{\varphi_t(y-x-t\beta)}{q^{t}_{\theta}(x,y)} \mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x) - \frac{1}{2}\int_{0}^{t}\left\{\|\alpha_{\theta}(\omega_s)-\beta\|^2+\Delta H_{\theta}(\omega_s)\right\}\rmd s\right\}\eqsp.
+% \frac{\rmd \mathbb{Q}^{y}_{\theta}}{\rmd \mathbb{W}^{\beta,y}_{\theta}}(w) = \frac{\varphi_t(y-x-t\beta)}{q^{t}_{\theta}(x,y)} \mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x) - \frac{1}{2}\int_{0}^{t}\left\{\|\alpha_{\theta}(\mw_s)-\beta\|^2+\Delta H_{\theta}(\mw_s)\right\}\rmd s\right\}\eqsp.
 % \]
@@
 % \[
-% q^{t}_{\theta}(x,y) = \varphi_t(y-x-t\beta)\eqsp\mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x)+\ell(\theta)t\right\}\mathbb{E}_{\mathbb{W}^{\beta,y}}\left[\mathrm{exp}\left\{-\int_{0}^{t}\phi_{\theta}(\omega_s)\rmd s\right\}\right]\eqsp,
+% q^{t}_{\theta}(x,y) = \varphi_t(y-x-t\beta)\eqsp\mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x)+\ell(\theta)t\right\}\mathbb{E}_{\mathbb{W}^{\beta,y}}\left[\mathrm{exp}\left\{-\int_{0}^{t}\phi_{\theta}(\mw_s)\rmd s\right\}\right]\eqsp,
 % \]
@@
 % \]
-% An unbiased estimator of $q^{t}_{\theta}(x,y)$ is then obtained by estimating $\mathbb{E}_{\mathbb{W}^{\beta,y}}\left[\mathrm{exp}\left\{-\int_{0}^{t}\phi_{\theta}(\omega_s)\rmd s\right\}\right]$.
+% An unbiased estimator of $q^{t}_{\theta}(x,y)$ is then obtained by estimating $\mathbb{E}_{\mathbb{W}^{\beta,y}}\left[\mathrm{exp}\left\{-\int_{0}^{t}\phi_{\theta}(\mw_s)\rmd s\right\}\right]$.
 
@@
 \section{Proofs}
+\label{sec:append:proofs}
 \begin{proof}[Proof of Lemma~\ref{lem:AR:unbiased}]
@@
 \begin{equation}
-\mP(X=l)= p_X(l)=\frac{\omega_k^l q_k(\xi_k^l,\xi_{k+1})}{\sum_{l=1}^N\omega_k^l q_k(\xi_k^l,\xi_{k+1})},~~l\in \{1,\dots,N\}
+\mP(X=\ell)= p_X(\ell)=\frac{\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})}{\sum_{\ell=1}^N\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})},~~\ell\in \{1,\dots,N\}
 \end{equation}
@@
 We assume $\omega_k$ are normalized weights, i.e.
-$\sum_l \omega_k^l=1$.
+$\sum_\ell \omega_k^l=1$.
 The candidate will therefore be sampled from $Y$, a r.v.
@@
 \begin{equation}
-\mP(Y=l)=p_Y(l)= \omega_k^l,~~l\in \{1,\dots,N\}
+\mP(Y=\ell)=p_Y(\ell)= \omega_k^\ell,~~\ell\in \{1,\dots,N\}
 \end{equation}
 Unfortunately, the $q_k$ function is unknown.
-Howevere, we let's suppose we can define an unbiased estimator $\hat{q}_k(\xi_k^l,\xi_{k+1},\zeta)$, where $\zeta$ is a random variable having th p.d.f.
+Howevere, we let's suppose we can define an unbiased estimator $\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)$, where $\zeta$ is a random variable having th p.d.f.
 $p_\zeta$.
 We have therefore:
-$$\E_{p_\zeta}\left[\hat{q}_k(\xi_k^l,\xi_{k+1},\zeta)\right]=q(\xi_k^l,\xi_{k+1})$$
+$$\E_{p_\zeta}\left[\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)\right]=q(\xi_k^\ell,\xi_{k+1})$$
 Moreover, we impose the following property over this estimator, we want that:
@@
 \begin{enumerate}
-\item Sample independantly $y\sim p_Y$, $z\sim p_\zeta$ and $u \sim \mathcal{U}[0,1]$
+\item Sample independently $\ell\sim p_Y$, $z\sim p_\zeta$ and $u \sim \mathcal{U}[0,1]$
 \item \textbf{If} 
-$$u \leq \frac{\hat{q}_k(\xi_k^l,\xi_{k+1},z)}{\sigma_+},$$
-set $x=y$. \textbf{Else}, return to 1).
+$$u \leq \frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},z)}{\sigma_+},$$
+set $x=\ell$. \textbf{Else}, return to 1).
 \end{enumerate}
@@
 \begin{align*}
-\mP\left(Y=l\vert U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\vert Y=l \right) \mathbb{P}(Y=l)}{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)}\\
-&=\frac{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^l,\xi_{k+1},\zeta)}{\sigma_+} \right) \omega_k^l}{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)}
+\mP\left(Y=\ell\vert U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\vert Y=l \right) \mathbb{P}(Y=\ell)}{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)}\\
+&=\frac{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)}{\sigma_+} \right) \omega_k^\ell}{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)}
 \intertext{We now note that}
 \mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\E_{p_\zeta}\left[\E\left[\E_{p_Y}\left[\E\left[\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right) \vert Y\right] \right] \vert \zeta\right]\right]\\
-&=\E_{p_\zeta}\left[\sum_{l=1}^N w_k^l\frac{\hat{q}_k(\xi_k^l,\xi_{k+1},\zeta)}{\sigma_+} \right]~~(\text{As }\frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\leq 1)\\
-&=\frac{1}{\sigma_+}\sum_{l=1}^N w_k^l q_k(\xi_k^l,\xi_{k+1},\zeta)
+&=\E_{p_\zeta}\left[\sum_{\ell=1}^N w_k^l\frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)}{\sigma_+} \right]~~(\text{As }\frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\leq 1)\\
+&=\frac{1}{\sigma_+}\sum_{\ell=1}^N w_k^\ell q_k(\xi_k^\ell,\xi_{k+1})
 \intertext{In the same way, we have:}
-\mP\left(U\leq \frac{\hat{q}_k(\xi_k^l,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{1}{\sigma_+} q_k(\xi_k^l,\xi_{k+1},\zeta)
+\mP\left(U\leq \frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{1}{\sigma_+} q_k(\xi_k^\ell,\xi_{k+1})
 \intertext{Which gives overall the wanted result}
-\mP\left(Y=l\vert U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{\omega_k^l q_k(\xi_k^l,\xi_{k+1},\zeta)}{\sum_{l=1}^N w_k^l q_k(\xi_k^l,\xi_{k+1},\zeta)}=p_X(l),~~l\in \{1,\dots,N\}
+\mP\left(Y=\ell\vert U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})}{\sum_{l=1}^N w_k^\ell q_k(\xi_k^\ell,\xi_{k+1})}=p_X(\ell),~~\ell\in \{1,\dots,N\}
 \end{align*}
 %\end{proof}
+\end{proof}
+\begin{proof}[Proof of Lemma \ref{lem:iid}]
+The independence is ensured by the mechanism of SMC methods.\\
+By \eqref{eq:random:weight},
+\[
+\mathbb{E}\left[\tilde{\omega}^i_{k+1}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right] = \mathbb{E}\left[\frac{M^{-1}\sum_{m=1}^M \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}}, \xi^{i}_{k+1};\zeta^m_{k+1})g_{k+1}(\xi^{i}_{k+1})}{\vartheta_{k+1}(\xi^{I^{i}_{k+1}}_{k}) p_{k+1}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right]\eqsp.
+%&= \mathbb{E}\left[\frac{g_{k+1}(\xi^{1}_{k+1})}{\vartheta_{k+1}(\xi^{I^{1}_{k+1}}_{k}) p_{k+1}(\xi_{k}^{I^{1}_{k+1}},\xi^{\ell}_{k+1})}\tau^{1}_{k+1}\widehat{q}_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{I^{1}_{k+1}},\xi^{1}_{k+1};\Lambda_{k+1})\middle| \mathcal{F}_k^{N}\right]\eqsp,
+\]
+Note that
+\begin{align*}
+&\mathbb{E}\left[\tau^{i}_{k+1}\middle|\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,I_{k+1}^i\}\right)\right]
+ = \sum_{\ell=1}^N\frac{\omega_k^{\ell} \qk(\xi_{k}^{\ell}, \xi^{i}_{k+1}) \left(\tau^{\ell}_k + h_{k}(\xi_{k}^{\ell},\xi^{i}_{k+1})\right)}{\sum_{\ell'=1}^N\omega_k^{\ell'} \qk(\xi_{k}^{\ell'},\xi^{i}_{k+1})}\eqsp,\\
+&\mathbb{E} \left[ M^{-1}\sum_{m=1}^M \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta^m_{k+1}) \middle| \sigma \left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,I^{i}_{k+1}\}\right)\right]
+ = \qk(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})\eqsp.
+\end{align*}
+Since $\tau^{i}_{k+1}$ and $\{\zeta^m_{k+1}\}_{i=1}^M$ are independent conditionally to $\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^l, I_{k+1}^l\}_{l=1}^N\right)$:
+\begin{multline*}
+\mathbb{E}\left[\tau^{i}_{k+1}M^{-1}\sum_{m=1}^M \widehat{\qk} (\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta^m_{k+1})\middle|\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,I_{k+1}^i\} \right)\right]\\
+ = q_k(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})\sum_{\ell=1}^N\frac{\omega_k^{\ell} \qk (\xi_{k}^{\ell},\xi^{i}_{k+1})\left(\tau^{\ell}_k + h_{k}(\xi_{k}^{\ell},\xi^{i}_{k+1})\right)}{\sum_{\ell'=1}^N\omega_k^{\ell'} \qk (\xi_{k}^{\ell'},\xi^{i}_{k+1})}\eqsp.
+\end{multline*}
+Moreover, The p.d.f. of the couple $(\xi_{k+1}^i,I_{k+1}^i)\vert \mathcal{F}_k^N$ is given by
+\begin{align*}
+p(x,j) &= \frac{\omega_k^j\vartheta_{k+1}(\xi_k^j)p_k(\xi_k^j,x)}{\sum_{l=1}^N\int \omega_k^l\vartheta_{k+1}(\xi_k^l)p_k(\xi_k^l,y)\rmd y}=\frac{\omega_k^j\vartheta_{k+1}(\xi_k^j)p_k(\xi_k^j,x)}{\phi_k^N[\vartheta_{k+1}]}
+\end{align*}
+Therefore, this yields:
+\begin{align*}
+\mathbb{E}\left[\tilde{\omega}^i_{k+1}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right]&= \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1} \sum_{j=1}^N\omega_k^j \int \vartheta_{k+1}(\xi^{j}_{k})\frac{\qk(\xi_{k}^{j},x) g_{k+1}(x)}{\vartheta_{k+1}(\xi^{j}_{k}) p_{k}(\xi_{k}^{j},x)}\\
+&\hspace{1cm}\times \sum_{\ell=1}^N\frac{\omega_k^{\ell} \qk (\xi_{k}^{\ell},x)\left(\tau^{\ell}_k + h_{k}(\xi_{k}^{\ell},x)\right)}{\sum_{\ell'=1}^N\omega_k^{\ell'}\qk(\xi_{k}^{\ell'},x)}p_{k}(\xi_{k}^{j},x)\rmd x\eqsp,\\
+&= \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\\
+&~~~~\times\sum_{\ell=1}^N \omega_k^\ell\left[\int \frac{ \sum_{j=1}^N \omega_k^j\qk(\xi_k^j,x) }{ \sum_{\ell'=1}^N\omega_k^{\ell'}\qk(\xi_{k}^{\ell'},x) } g_{k+1}(x)\qk (\xi_{k}^{\ell},x)\left(\tau^{\ell}_k + h_{k}(\xi_{k}^{\ell},x)\right) \rmd x \right]\\
+& =\left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\int \qk(\cdot,x)g_{k+1}(x)\left\{\tau_k(\cdot) + h_{k}(\cdot,x)\right\}\rmd x\right]\eqsp,
+\end{align*}
+which concludes the proof.
 \end{proof}

### 2016-12-20 ParisEM.tex
Changing omega tilde to omega hat
Mistake in GFirsanov formula
New version - without experiments and appendices

@@
 \newcommand{\Oz}{Z}%Letter for the Ozaki approximation
-\newcommand{\Jk}{J_{\alpha_\theta}^k}%Command for Jacobian of alpha
+\newcommand{\Jk}{J_{\alpha}^k}%Command for Jacobian of alpha
 \newcommand{\mw}{\mathsf{w}}%For bridge realisations
@@
 
-\title{Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations: the Oziris algorithm}
+\title{Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations}
 
@@
 \section{Introduction}
-In this paper, we propose a new algorithm to approximate expectations of additive functionals with respect to the joint smoothing distribution, in the Hidden Markov Model (HMM) framework.\\
-The dynamics of the states $(X_t)_{t\ge 0}$ depends on a parameter $\theta\in \Theta$ where $\Theta$ is a compact subspace of $\mathbb{R}^{\ell}$ and that $(X_t)_{t\ge 0}$ is a weak solution to the following Stochastic Differential Equation (SDE) in $\mathbb{R}^d$:
-\begin{equation}
-\label{eq:target:sde}
-X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha_{\theta}(X_t)\rmd t + \rmd W_t\eqsp,
-\end{equation}
-where $(W_t)_{t\ge 0}$ is a standard Brownian motion.
-In the following, it is assumed that $\alpha_{\theta}$ is of the form $\alpha_{\theta}(x) = \nabla_x A_{\theta}(x)$ where $A_{\theta}: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function.
-A popular choice of $A_{\theta}$  is $A_{\theta} = \log \pi_{\theta}/2$ leading to a Langevin diffusion which may be used to approximate a given target distribution $\pi_{\theta}$.
-See for instance \cite{roberts:tweedie:1996} for conditions under which this diffusion converges exponentially quickly to $\pi_{\theta}$.
-The states $(X_t)_{t\ge 0}$ are only partially observed at discrete times $(t_k)_{0\le k \le n}$, with $t_0=0$, through the sequence $(Y_k)_{0\le k\le n}$ which is such that $(Y_k)_{0\le k\le n}$ are independent conditional on $(X_t)_{t\ge 0}$ and for all $0\le k\le n$, the conditional distribution of $Y_k$ given $(X_t)_{t\ge 0}$ depends only on $X_k = X_{t_k}$.
-This distribution has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g_k(X_k) = g(\theta,X_k,\cdot)$. 
-The distribution of $X_0$ has a density with respect to a reference measure $\mu_0$ on $\mathbb{R}^d$ given by $\chi(\cdot):=\chi(\theta,\cdot)$.
-For all $1\le k \le n$, the distribution of $X_{k+1} $ conditional on $X_{k}$ has a density $\qk(X_{k},\cdot) :=q(\theta,\Delta_k,X_{k-1},\cdot),~\Delta_k:=t_{k+1}-t_{k}$ with respect to a reference measure $\mu$ on $\mathbb{R}^d$.\\
-\\
-Statistical inference for partially observed stochastic differential equations usually involves the computation of conditional distributions of some unobserved states given the observations $(Y_0,\ldots,Y_n)$.
-These posterior distributions are crucial to compute maximum likelihood estimators defined as maximizers of the incomplete data log-likelihood $\theta\mapsto \ell_{\theta}^{n}$ given by
-\begin{equation*}
-\ell_{\theta}^{n}(Y_{0:n}) = \log\left(\int p_{\theta}(x_{0:n},Y_{0:n})\,\rmd x_{0:n}\right)\eqsp,
-\end{equation*}  
-where $p_{\theta}$ is the complete data likelihood:
-\begin{equation*}
-p_{\theta}(x_{0:n},Y_{0:n}) = \chi(x_0)g_{0}(x_0)\prod^{n-1}_{k=0}\qk(x_{k},x_{k+1})g_{k+1}(x_{k+1})\eqsp.
-\end{equation*}
-For instance, the EM algorithm introduced in \cite{dempster:laird:rubin:1977} iteratively builds a sequence $\{\theta_{p}\}_{p\ge 0}$ of parameter estimates following the two steps:
-\begin{enumerate}
-	\item {\bf E-step}: compute $\theta \mapsto Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log p_{\theta}(X_{0:n},Y_{0:n})\middle|Y_{0:n}\right]$, where $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{0:n}\right]$ is the conditional expectation given $Y_{0:n}$ when the parameter value is $\theta$.
-	\item {\bf M-step}: choose $\theta_{p+1}$ as a maximizer of $\theta \mapsto Q(\theta,\theta_{p})$.
-\end{enumerate}
-The E-step boils down to the computation of the smoothed additive function: 
-\begin{multline*}
-Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log \left(\chi(X_0)g_{0}(X_0)\right)\middle|Y_{0:n}\right] \\
-+ \sum_{k=0}^{n-1}\mathbb{E}_{\theta_p}\left[\log \left(\qk(X_{k},X_{k+1})g_{k+1}(X_{k+1})\right)\middle|Y_{0:n}\right] \eqsp.
-\end{multline*}
-Similarly, by Fisher's identity, recursive maximum likelihood estimates may be computed using the gradient of the loglikelihood which can be written as the conditional expectation of an additive functional of the hidden states given all the observations.
+This paper introduces a new algorithm to solve the smoothing problem for partially observed continuous time stochastic processes.
+In this setting, the hidden state process $(X_t)_{t\ge 0}$ is assumed to be a solution to a stochastic differential equation (SDE) and the only information available is given by noisy observations $(Y_{k})_{0\le k\le n}$ of the states $(X_k)_{0\le k\le  n}$ at some discrete time points $(t_k)_{0\le k\le n}$.
+The bivariate stochastic process $\{(X_{k},Y_{k})\}_{0\le k\le n}$ is a state space model such that conditional on the state sequence $(X_{k})_{0\le k\le n}$ the observations $(Y_{k})_{0\le k\le n}$ are independent and for all $0\le \ell\le n$ the conditional distribution of $Y_{\ell}$ given $\{X_{k}\}_{0\le k\le n}$ depends on $X_{\ell}$ only.
+
+%Statistical inference for partially observed stochastic differential equations is a challenging task since some elementary quantities, such as transition probabilities, are not available explicitly.
+Statistical inference for partially observed stochastic differential equations requires to solve bayesian filtering and smoothing problems, i.e.
+the computation of the posterior distributions of a sequence of hidden states given observations.
+Filtering refers to the estimation of the distributions of the hidden state $X_k$ given the observations $(Y_0,\ldots,Y_k)$ up to time $k$, while fixed-interval smoothing stands for the estimation of the distribution of sequence of states $(Z_{k},\ldots,Z_{p})$ given observations $(Y_{0},\ldots,Y_{\ell})$ with $0\le k\le p<\ell \le n$.
+These posterior distributions are crucial to compute maximum likelihood estimators of the unknown parameters defining the model using the observations $(Y_0,\ldots,Y_n)$ only.
+% $\theta\mapsto \ell_{\theta}^{n}$ given by
+%\begin{equation*}
+%\ell_{\theta}^{n}(Y_{0:n}) = \log\left(\int p_{\theta}(x_{0:n},Y_{0:n})\,\rmd x_{0:n}\right)\eqsp,
+%\end{equation*}  
+%where $p_{\theta}$ is the complete data likelihood:
+%\begin{equation*}
+%p_{\theta}(x_{0:n},Y_{0:n}) = \chi(x_0)g_{0}(x_0)\prod^{n-1}_{k=0}\qk(x_{k},x_{k+1})g_{k+1}(x_{k+1})\eqsp.
+%\end{equation*}
+For instance, the E-step of the EM algorithm introduced in \cite{dempster:laird:rubin:1977} %iteratively builds a sequence of parameter estimates following two steps.
+%following the two steps:
+%\begin{enumerate}
+%	\item {\bf E-step}: compute $\theta \mapsto Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log p_{\theta}(X_{0:n},Y_{0:n})\middle|Y_{0:n}\right]$, where $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{0:n}\right]$ is the conditional expectation given $Y_{0:n}$ when the parameter value is $\theta$.
+%	\item {\bf M-step}: choose $\theta_{p+1}$ as a maximizer of $\theta \mapsto Q(\theta,\theta_{p})$.
+%\end{enumerate}
+%step 
+boils down to the computation of a conditional expectation of an additive functional of the hidden states given all the observations up to time $n$. 
+%\begin{multline*}
+%Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log \left(\chi(X_0)g_{0}(X_0)\right)\middle|Y_{0:n}\right] \\
+%+ \sum_{k=0}^{n-1}\mathbb{E}_{\theta_p}\left[\log \left(\qk(X_{k},X_{k+1})g_{k+1}(X_{k+1})\right)\middle|Y_{0:n}\right] \eqsp.
+%\end{multline*}
+Similarly, by Fisher's identity, recursive maximum likelihood estimates may be computed using the gradient of the loglikelihood which can be written as the conditional expectation of an additive functional of the hidden states.
  See \cite[Chapter $10$ and $11$]{cappe:moulines:ryden:2005}, \cite{kantas:doucet:signh:2015,doucet:poyiadjis:singh:2011,lecorff:fort:2013a,lecorff:fort:2013b}
-for further references on the use of these smoothed expectations of additive functionals  applied to maximum likelihood parameter inference in latent data models.\\
-\\
-The exact computation of these expectations is usually not possible for partially observed diffusions such as \eqref{eq:target:sde}. 
+for further references on the use of these smoothed expectations of additive functionals  applied to maximum likelihood parameter inference in latent data models.
+
+The exact computation of these expectations is usually not possible for general partially observed diffusions.
 In this paper, we propose to use Sequential Monte Carlo methods to approximate smoothing distributions with random particles associated with importance weights.
  \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers by combining importance sampling steps to propagate particles with importance resampling steps to duplicate or discard particles according to their importance weights.
- Approximations of the smoothing distributions may be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}.
+Unfortunately, these methods cannot be applied directly to partially observed stochastic differential equations since some elementary quantities, such as transition densities of the hidden states, are not available explicitly.
+The transition density may be approximated using discretization procedures such as the Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998} or a Gaussian based approximation using Taylor expansions of the posterior mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}.
+Other approaches based on Hermite polynomials expansion were also introduced by \cite{ait-sahalia:1999,ait-sahalia:2002,ait-sahalia:2008} and extended in several directions recently, see \cite{li:2013} and all the references on the approximation of transition densities therein.
+However, even the most recent discretization based approximations of the transition densities induce a systematic bias of particle based approximations of posterior distributions, see \cite{delmoral:jacod:protter:2001}.
+To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of transition densities based on the generalized Poisson estimator (GPE) introduced in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.
+
+The only solution to solve the smoothing problem for partially observed SDE using SMC methods has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}.
+Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias that does not vanish as the number of particles grows to infinity.
+Approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}.
 Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$.
+In the case of partially observed SDE, this forward pass may be performed with the SMC steps combined with the GPE of \cite{fearnhead:papaspiliopoulos:roberts:2008}.
 Then, the backward pass of the FFBS algorithm keeps all the particles sampled during the forward pass and computes new weights to take into account the information brought by all the observations from time $n$ to time $0$.
- Instead of computing new weights, the FFBSi  algorithm samples independently backward in time particle trajectories using the particles and weights produced by the forward pass.\\ 
-\\
+Instead of computing new weights, the FFBSi  algorithm samples independently backward in time particle trajectories using the particles and weights produced by the forward pass.
 Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e.
@@
 Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.
- However, there is a major difficulty to apply SMC methods to partially observed SDE as the computation of the importance weights requires to evaluate $q^{\Delta t_k}_{\theta}$ for which there is no analytic expression in general case.
-%the EM algorithm to partially observed SDE in general cases:
-%\begin{enumerate}[-]
-%\item The E-step requires to compute $q^{\Delta t_k}_{\theta}$ for which there is no analytic expression.
-%\item The computation of $Q(\theta,\theta_{p})$ relies on conditional expectations of additive functionals of the state sequence given the observations which are not available explicitly.
-%\end{enumerate}
-To overcome these difficulties, \cite{olsson:strojby:2011} proposed to use General Poisson Estimators (GPE) introduced in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,fearnhead:papaspiliopoulos:roberts:2008}  to provide an unbiased estimation of $\qk$.
-These estimators are based on a representation of $\qk$ as an expectation under the probability measure of a Brownian bridge (the conditional expectation involved in the computation is approximated in  \cite{olsson:strojby:2011}  using fixed-lag Sequential Monte Carlo methods developed in \cite{olsson:cappe:douc:moulines:2008}).\\
-\\
-In this paper, we propose to improve this algorithm to obtain faster convergence of the sequence of Monte Carlo EM estimates, and to extend the range of diffusion processes for which SMC methods can be applied.\\
-First, we introduce a new class off GPE based on the Ozaki discretization of the SDE  \eqref{eq:target:sde}, see \cite{ozaki:1992,shoji:ozaki:1998}. 
-These new GPE estimators use a an order one Taylor expansion of the drift of \eqref{eq:target:sde} to obtain more efficient estimates than Brownian Bridge based GPE.\\ 
-Instead of using fixed-lag techniques which introduce a systematic additional bias  in the approximation of the conditional expectations, we use the new PaRIS algorithm which allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles.
-The striking result of the application of PaRIS algorithm to SDE is that the accept reject mechanism underlying the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
+
+In this paper, instead of using fixed-lag techniques which introduce a systematic additional bias in the approximation of conditional expectations, we extend the use of  PaRIS algorithm to partially observed SDE.
+The proposed algorithm allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles.
+The crucial result of the application of PaRIS algorithm to SDE is that the accept reject mechanism ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by bounded unbiased estimates.
+We also propose to improve this algorithm to obtain faster convergence of the sequence of Monte Carlo EM estimates, and to extend the range of diffusion processes for which SMC methods can be applied.To do so, we introduce a new class of GPE based on the Ozaki discretization of the target SDE, see \cite{ozaki:1992,shoji:ozaki:1998}.
+These new GPE estimators use a an order one Taylor expansion of the drift of the SDE to obtain more efficient estimates than Brownian Bridge based GPE.
+
 \section{Random weight PaRIS algorithm}
+In this paper, $(X_t)_{t\ge 0}$ is a weak solution to the following SDE in $\mathbb{R}^d$:
+\begin{equation}
+\label{eq:target:sde}
+X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha(X_t)\rmd t + \rmd W_t\eqsp,
+\end{equation}
+where $(W_t)_{t\ge 0}$ is a standard Brownian motion.
+It is assumed that $\alpha$ is of the form $\alpha(x) = \nabla_x A(x)$ where $A: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function.
+A popular choice of $A$  is $A = \log \pi/2$ leading to a Langevin diffusion which may be used to approximate a given target distribution $\pi$ on $\mathbb{R}^d$.
+See for instance \cite{roberts:tweedie:1996} for conditions under which this diffusion converges exponentially quickly to $\pi$.
+SDE where the drift is given by the gradient of a potential function have been widely used in ecology for instance, see \cite{brillinger:et:al:2011,harris:blackwell:2013,preisler:et:al:2013} and the references therein.
+
+For all $0\le k \le n$, the distribution of $Y_k$ given $X_k$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,Y_k) = g_k(X_k)$. 
+The distribution of $X_0$ has a density with respect to $\lambda$ given by $\chi$.
+For all $1\le k \le n$, the distribution of $X_{k+1} $ conditional on $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to a reference measure $\mu$ on $\mathbb{R}^d$. 
 Let $0 \leq k \leq k' \leq n$. The joint smoothing distributions of the hidden states are defined, for all measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, by:
 \[
-\phi_{k:k'|n}[h] := \mathbb{E}\left[h(X_k,\ldots,X_{k'})\middle|Y_{0:n}\right]\eqsp.
+\phi_{k:k'|n}[h] = \mathbb{E}\left[h(X_k,\ldots,X_{k'})\middle|Y_{0:n}\right]\eqsp.
 \]
@@
 when the transition density of the hidden states is not available explicitly and where $\{h_k\}_{k=0}^{n-1}$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$.
-The algorithm is based on the following link between the filtering and smoothing distributions:
+The algorithm is based on the following link between the filtering and smoothing distributions for additive functionals, see \cite{olsson:westerborn:2016}:
 \begin{equation}
-\phi_{0:n|n}[h] = \phi_n[\E\left[h(X_{0:n})\vert X_n,~Y_{0:n}]\right]:= \phi_n[T_n(h)] \eqsp.\label{eq:FFbsm:equality}
+\phi_{0:n|n}[h] = \phi_n[T_n[h]]\eqsp,\;\mbox{where}\; T_n[h](X_n) = \E\left[h(X_{0:n})\vert X_n,~Y_{0:n}\right]\eqsp.\label{eq:FFbsm:equality}
 \end{equation}
-Therefore, using the additive property and the usual filtering recursion applied on the right hand side of equation \eqref{eq:FFbsm:equality}, an approximation of \eqref{def:addfunc} can be computed using Sequential Monte Carlo techniques.\\
-Sequential Monte Carlo algorithms introduced in  \cite{gordon:salmond:smith:1993} and \cite{kitagawa:1996} rely on importance sampling and importance resampling steps to approximate the filtering distribution $\phi_{k}$ (conditional distribution of $X_k$ given $Y_{0:k}$) using  a set of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$.\\
+The approximation of \eqref{eq:FFbsm:equality} requires first to approximate the sequence of filtering distributions.
+Sequential Monte Carlo methods provide an efficient and simple solution to obtain these approximations using sets of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$, $0\le k \le n$.
 At time $k = 0$, $N$ particles $\{\xi^{\ell}_0\}_{\ell=1}^N$ are initialized by setting $\xi^{\ell}_0=x_0$.
-Then, $\xi^{\ell}_0$ is associated with the importance weights $\omega_0^{\ell} = g_0 \left(\xi^{\ell}_0\right)$.
-Then, for any bounded and measurable function $h$ defined on $\mathbb{R}^d$, the expectation $\phi_{0}[h] $ is approximated by
+Then, $\xi^{\ell}_0$ is associated with the importance weights $\omega_0^{\ell} = g_0 (\xi^{\ell}_0)$.
+For any bounded and measurable function $h$ defined on $\mathbb{R}^d$, the expectation $\phi_{0}[h] $ is approximated by
 \[
@@
 \]
-Then, using $\{(\xi^{\ell}_{k-1},\omega^{\ell}_{k-1})\}_{\ell=1}^N$, pairs $\{(I^{\ell}_k,\xi^{\ell}_{k})\}_{\ell=1}^N$ of indices and particles are sampled and weighted using an instrumental transition density $p_k$ on $\mathbb{R}^d\times \mathbb{R}^d$ and an adjustment multiplier function $\vartheta_k$ on $\mathbb{R}^d$.
+Then, for $1\le k \le n$, using $\{(\xi^{\ell}_{k-1},\omega^{\ell}_{k-1})\}_{\ell=1}^N$, the auxiliary particle filter of \cite{pitt:shephard:1999} samples pairs $\{(I^{\ell}_k,\xi^{\ell}_{k})\}_{\ell=1}^N$ of indices and particles using an instrumental transition density $p_k$ on $\mathbb{R}^d\times \mathbb{R}^d$ and an adjustment multiplier function $\vartheta_k$ on $\mathbb{R}^d$.
 Each new particle $\xi^{\ell}_{k}$ and weight $\omega^{\ell}_k$ at time $k$ are computing following these steps:
-\begin{enumerate}
+\begin{enumerate}[-]
 \item choose a particle index $I^{\ell}_k$ at time $k-1$ in $\{1,\ldots,N\}$ with probabilities proportional to $\omega_{k-1}^{j} \vartheta_k (\xi^{j}_{k-1})$, for $j$ in $\{1,\ldots,N\}$ ;
-\item sample  $\xi^{\ell}_{k}$ using this chosen particle according to $\xi^{\ell}_{k} \sim p_k(\xi^{I^{\ell}_k}_{k-1},\cdot)$.
+\item sample  $\xi^{\ell}_{k}$ using this chosen particle according to $\xi^{\ell}_{k} \sim p_k(\xi^{I^{\ell}_k}_{k-1},\cdot)$ ; 
 \item $\xi^{\ell}_k$ is associated with:
@@
 The PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSi algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions.
-%Both the FFBS and the FFBSi algorithms require first that a forward pass has produced a set of particles and importance weights $\{ (\xi^{\ell}_k, \omega^{\ell}_k)\}$, for $\ell \in \{1, \ldots, N\}$ and  $k \in \{0,\ldots,n\}$ to approximate $\phi_{k}$ for $k=0$ to $k=n$.
-The FFBS algorithm stores all particles and weights without taking into consideration the genealogy of the particles.
-Then, a backward pass is performed, keeping all the particles fixed but modifying all importance weights computed during the forward pass.
 The FFBS algorithm has a computational complexity of order $\mathcal{O}(N^{k'-k+1})$ to approximate joint smoothing distributions of the form $\phi_{k:k'|n}$ with $k<k'$ which makes it prohibitive.
 An important feature of the FFBS algorithm shown in \cite{mongillo:deneve:2008,cappe:2011,delmoral:doucet:singh:2010} is that it can be implemented using only a forward pass when approximating smoothed expectations of additive functions.
-%In more general case, \cite{dubarry:lecorff:2013} are interested in the FFBS algorithm for an additive function $h$ defined on $\mathsf{X}^{T+1}$,
-%\[
-%h(x_{0:T}) := \sum_{t=1}^{T} h_t (x_{t-1},x_t)
-%\]
-%where $\{h_t\}_{t=1}^{T}$ are bounded on $\mathsf{X}^2$.
-See \cite{dubarry:lecorff:2013} for their proposed FFBS algorithm in this case.
-As shown in \cite{delmoral:doucet:singh:2010} when applied to additive functionals, the computational complexity grows quadratically with number of particles.
-In the backward pass of the FFBS algorithm, the importance weights are updated while the particles are kept fixed.
-%\subsubsection*{FFBSi algorithm}
-%As for the FFBS algorithm, the FFBSi algorithm uses a forward pass to produce $\{ (\xi^{\ell}_k, \omega^{\ell}_k)\}$, for $\ell \in \{1, \ldots, N\}$ and  $k \in \{0,\ldots,n\}$ to approximate $\phi_{k}$ for $k=0$ to $k=n$ and all these particles and weights are stored  without taking into consideration the genealogy of the particles.
 Instead of computing new weights in the backward pass, the FFBSi  samples backward new trajectories from time $n$ to time $0$ among all the $N^{n+1}$ trajectories which can be chosen in $\{\xi^{\ell}_k\}$, $1\le \ell\le N$ and $0\le k\le n$.
-%Each trajectory is sampled according to the following steps.
-%\begin{enumerate}[-]
-%\item At time $k = n$, sample $J_n\in\{1\ldots,N\}$ with probabilities proportional to $\omega^{\ell}_n/\Omega_n^N$, for $\ell\in\{1, \ldots, N\}$.
-%\item For $k = n-1$ to $k=0$, $J_k$ is sampled in $\{1\ldots,N\}$ with probabilities 
-%\[
-%\Lambda_k^N(J_{k+1},i) = \frac{\omega^{i}_k q(\xi^{i}_k,\xi^{J_{k+1}}_{k+1})}{\sum_{\ell=1}^N\omega^{\ell}_k q(\xi^{\ell}_k,\xi^{J_{k+1}}_{k+1})}\eqsp,\quad 1\le i\le N\eqsp.
-%\]
-%\end{enumerate}
-%These sampling steps are repeated $N$ times to produce $\{J_0^\ell,\ldots,J_n^{\ell}\}$, $1\le \ell \le N$.
-Then, the smoothed expectation of any measurable function $h$ on $(\mathbb{R}^d)^{n+1}$  $\mathbb{E}\left[h(X_{0:n})\middle|Y_{0:n}\right]$ is approximated by $N^{-1} \sum_{\ell = 1}^N  h(\tilde{\xi}^{J_0^{\ell}}_{0},\ldots, \tilde{\xi}^{J_n^{\ell}}_{n})$.
 Contrary to the FFBS algorithm, the FFBSi algorithm samples trajectories and may be used to approximate joint smoothing distributions without increasing its complexity.
  Nevertheless, the computation of the transition probabilities to select particles in the backward pass makes it a $\mathcal{O}(N^2)$ algorithm.
-In the case where the transition probability $q_k$ is upper bounded, \cite{douc:garivier:moulines:olsson:2011} introduced an acceptance rejection mechanism to implement the FFBSi algorithm with a $\mathcal{O}(N)$ complexity, see also \cite{dubarry:lecorff:2011}.\\
- \\
+In the case where the transition probability $q_k$ is upper bounded, \cite{douc:garivier:moulines:olsson:2011} introduced an acceptance rejection mechanism to implement the FFBSi algorithm with a $\mathcal{O}(N)$ complexity, see also \cite{dubarry:lecorff:2011}.
+
 The PaRIS algorithm combines both the forward only version of the FFBS algorithm with the sampling mechanism of the FFBSi algorithm. 
 The algorithm does not produce an approximation of the smoothing distributions but of the smoothed expectation of a fixed additive functional and thus  may be used to approximate \eqref{def:addfunc}. 
-Its crucial property is that it does not require a backward pass, the smoothed expectations is computed on-the-fly with the particle filter and no storage of the particles or weights is needed.\\ 
-The focus is based sufficient statistics $\tau^i_k$, approximations of $T_k(h)$, for $1\le i\le N$ and $0\le k \le n$ starting with $\tau^i_0 = 0$ for all $1\le i\le N$.
+Its crucial property is that it does not require a backward pass, the smoothed expectations is computed on-the-fly with the particle filter and no storage of the particles or weights is needed.
+The focus is based sufficient statistics $\tau^i_k$, approximations of $T_k[h](\xi^i_k)$, for $1\le i\le N$ and $0\le k \le n$ starting with $\tau^i_0 = 0$ for all $1\le i\le N$.
 Let $\tilde{N}\ge 1$.
@@
 As proved in \cite{olsson:westerborn:2016}, the algorithm is asymptotically consistent (as $N$ goes to infinity) for any fixed precision parameter $\tilde N$.
-However, there is a significant qualitative difference between the cases $\tilde{N} = 1$ and $\tilde{N} \geq 2$.\\
-As for the FFBSi algorithm,  when there exists $\sigma_+$ such that $0<\qk <\sigma_+$, the PaRIS algorithm may be implemented with $\mathcal{O}(N)$ complexity using the procedure of \cite{douc:garivier:moulines:olsson:2011} recalled in algorithm \ref{alg:AR:knownq}.
-\begin{algorithm}[hb]
-\caption{Random weight Accept-reject-based backward sampling  (\cite{douc:garivier:moulines:olsson:2011})}
-\begin{algorithmic}
-\FORALL{$i \in 1,\dots, N$}
-\FORALL{$\ell \in 1,\dots, \widetilde N$} 
-\STATE \textbf{\sc Sampling Step} Sample independently\\ $j\sim \{\omega_{k}^1,\dots,\omega_{k}^N\}$, $u\sim \mathcal{U}[0,1]$
-\IF{ $$u \leq \frac{\qk(\xi_{k}^j,\xi_{k+1}^i,z)}{\sigma_+},$$}
-\STATE Set $J_k^{i,\ell} = j$.
-\ELSE 
-\STATE Return to \textbf{\sc Sampling Step}
-\ENDIF
-\ENDFOR
-\ENDFOR
-\end{algorithmic}
-\label{alg:AR:knownq}
-\end{algorithm}
+However, there is a significant qualitative difference between the cases $\tilde{N} = 1$ and $\tilde{N} \geq 2$.
+As for the FFBSi algorithm,  when there exists $\sigma_+$ such that $0<\qk <\sigma_+$, the PaRIS algorithm may be implemented with $\mathcal{O}(N)$ complexity using the accept reject mechanism of \cite{douc:garivier:moulines:olsson:2011}.
 
-%\begin{algorithm}
-%\caption{Random weight Accept-reject-based backward sampling  (\cite{douc:garivier:moulines:olsson:2011})}
-%\begin{algorithmic}
-%\FORALL{$i \in 1,\dots,\widetilde N$}
-%\STATE Set $L =\{1,\ldots,N\}$. 
-%\WHILE{$L\neq\emptyset$}
-%\STATE $p = \#L$.
-%\STATE Sample $I = (I_1,\ldots,I_p)$ independently in $\{1,\ldots,N\}$ with probabilities proportional to $\omega_{k-1}^{\ell}$, $1\le \ell\le N$.
-%\STATE Sample $U = (U_1,\ldots,U_p)$ independently and independently of $I$ uniformly in $[0,1]$.
-%\STATE Sample $\Lambda = (\Lambda_1,\ldots,\Lambda_p)$ independently and independently of $I$ and $U$.
-%\STATE Set $L_p = \emptyset$.
-%\FORALL{$\ell \in 1,\dots,p$}
-%\IF{$U_\ell<\widehat{q}(\xi_{k-1}^{I_\ell},\xi_k^{L(\ell)};\Lambda_{\ell})/\sigma_+$}
-%\STATE $J_k^{L(\ell),i} = I_\ell$.
-%\ELSE
-%\STATE $L_p = L_p \cup \{L(\ell)\}$.
-%\ENDIF
-%\ENDFOR
-%\STATE $L = L_p$.
-%\ENDWHILE
-%\ENDFOR
-%\end{algorithmic}
-%\label{alg:AR}
-%\end{algorithm}
-\section{The OzIRis algorithm}
-\label{sec:PaRIS:SDE}
 In general situations, the PaRIS algorithm cannot be used for stochastic differential equations as $\qk$ is unknown.
-Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio in Algorithm~\ref{alg:AR:knownq} cannot be performed to apply the PaRIS algorithm.
-Following \cite{olsson:strojby:2011},
+Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio of \cite{douc:garivier:moulines:olsson:2011} is not tractable.
+Following \cite{fearnhead:papaspiliopoulos:roberts:2008,olsson:strojby:2011},
 an extension of the PaRIS algorithm may be developed by replacing $\qk$  by an  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable such that:
@@
 In the case where $\widehat{q}_k$ may be upper bounded by a constant $\hat{\sigma}_+$, step~\eqref{it:PaRIS:sampleindex} is replaced by the acceptance-rejection mechanism given in Algorithm~\ref{alg:AR:unknownq}.
-The Lemma~\ref{lem:AR:unbiased} states that replacing  $\qk$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.
+Lemma~\ref{lem:AR:unbiased} states that replacing  $\qk$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.
 \begin{lemma}
@@
 \end{proof}
-\begin{algorithm}[hb]
+\begin{algorithm}[H]
 \caption{Double random weight Accept-reject-based backward sampling}
@@
 \FORALL{$\ell \in 1,\dots, \widetilde N$} 
-\STATE \textbf{\sc Sampling Step} Sample independently\\ $j\sim \{\omega_{k}^1,\dots,\omega_{k}^N\}$, $z\sim\zeta_k$, $u\sim \mathcal{U}[0,1]$
-\IF{ $$u \leq \frac{\widehat{\qk}(\xi_{k}^j,\xi_{k+1}^i,z)}{\hat{\sigma}_+},$$}
-\STATE Set $J_k^{i,\ell} = j$.
+\STATE \textbf{\sc Sampling Step} \\
+Sample independently $
+\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\omega_{k}^1,\dots,\omega_{k}^N\}$.
+\IF{ $$U \leq \frac{\widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)}{\hat{\sigma}_+},$$}
+\STATE Set $J_k^{i,\ell} = J$.
 \ELSE 
@@
 \end{algorithm}
-In \cite{olsson:strojby:2011},  Global Poisson Estimator (GPE) are used to estimate $\qk$ which relies on a representation of $\qk$ as an expectation under the probability measure of a Brownian bridge.
-In order to improve the Monte Carlo estimation of $\qk$,  other diffusion bridges may be used as long as they can be simulated efficiently.\\ 
+
+\section{The OzIRis algorithm}
+\label{sec:PaRIS:SDE}
+In \cite{olsson:strojby:2011}, $\qk$ is estimated using GPE based on a representation of $\qk$ as an expectation under the probability measure of a Brownian bridge. 
+In order to improve the Monte Carlo estimation of $\qk$,  other diffusion bridges may be used as long as they can be simulated efficiently.
 A simple idea is to consider a diffusion obtained as an approximation of the target SDE \eqref{eq:target:sde}.
-\cite{ozaki:1992,shoji:ozaki:1998} proposed a higher order approximation of the drift term $\alpha_{\theta}$ than the Euler scheme. 
-The authors used a linear approximation of the diffusion drift $\alpha_{\theta}$, together with a constant approximation of the volatility over each time interval $(t_{k},t_{k+1})$.\\ 
-Formally, for all $0\le k\le n-1$, let $(Z^{k}_t)_{t\ge 0}$ be governed by the following stochastic differential equation for $t\in(t_{k},t_{k+1})$:
+In \cite{giesecke:schwenkler:2016}, the authors proposed to replace the Brownian bridge by a Brownian motion with constant drift between two consecutive observations.
+As an alternative, is is also possible to follow \cite{ozaki:1992,shoji:ozaki:1998} to propose a higher order approximation of the drift term $\alpha$.
+The authors used a linear approximation of the diffusion drift $\alpha$, together with a constant approximation of the volatility over each time interval $(t_{k},t_{k+1})$.
+
+For all $0\le k\le n-1$, let $(Z^{k}_t)_{t\ge 0}$ be governed by the following stochastic differential equation for $t\in(t_{k},t_{k+1})$:
 \begin{equation}
 \label{eq:SDE:Ozaki}
-\Oz^k_{t_{k}} = X_{k}\quad\mbox{and}\quad\rmd \Oz^k_t = \widetilde{\alpha}^k_{\theta}(\Oz^k_t)\rmd t + \rmd W_t\eqsp,
+\Oz^k_{t_{k}} = X_{k}\quad\mbox{and}\quad\rmd \Oz^k_t = \widetilde{\alpha}_k(\Oz^k_t)\rmd t + \rmd W_t\eqsp,
 \end{equation}
-where
-\[
-\widetilde{\alpha}^k_{\theta}(x) = \Jk x + \alpha_{\theta}(X_{k}) - \Jk X_{k}
-\]
-and $\Jk$ is the Jacobian of $\alpha_{\theta}$ evaluated in $X_k$.
-The probability density function of the transition kernel associated with $\Oz^k$ is written $\widetilde{q}_k$ and is given in Section~\ref{sec:ornstein:bridge}.
-Writing $\mathbb{Q}^k_{\theta}$ (resp.
-$\mathbb{W}^{k}_{\theta}$) the probability measure induced by $(X_s)_{t_{k}\le s\le t_{k+1}}$  (resp.
+where $\widetilde{\alpha}_k$ is an instrumental drift. The probability density function of the transition kernel associated with $\Oz^k$ is written $\widetilde{q}_k$.
+Writing $\mathbb{Q}_k$ (resp.
+$\mathbb{W}_{k}$) the probability measure induced by $(X_s)_{t_{k}\le s\le t_{k+1}}$  (resp.
 $(\Oz_s)_{t_{k}\le s\le t_{k+1}}$) on $(\mathsf{C},\mathcal{C})$, Girsanov theorem yields:
-\[
-\frac{\rmd \mathbb{Q}^k_{\theta}}{\rmd \mathbb{W}^{k}_{\theta}}(w) = \mathrm{exp}\left\{H_{\theta}(\omega_{t_{k+1}}) - H_{\theta}(\omega_{t_{k}}) - \frac{1}{2}\int_{t_{k}}^{t_{k+1}}\left\{\|\alpha_{\theta}(\mw_s)-\widetilde{\alpha}_{\theta}(\mw_s)\|^2+\Delta H_{\theta}(\mw_s)\right\}\rmd s\right\}\eqsp,
-\]
-where 
-\[
-H_{\theta}(x) = A_{\theta}(x) - \frac{1}{2}x'\Jk x- (\alpha_{\theta}(X_{k}) - \Jk X_{k-1})'x
-\]
-and $\Delta$ is the Laplace operator. Then, the probability measures conditioned on hitting $X_{k+1}$ at time $t_{k+1}$ satisfy:
+\begin{multline*}
+\frac{\rmd \mathbb{Q}_k}{\rmd \mathbb{W}_{k}}(w) = \mathrm{exp}\left\{H_{k}(\mw_{t_{k+1}}) - H_{k}(\mw_{t_{k}})\right\}\\
+\times \exp\left\{- \frac{1}{2}\int_{t_{k}}^{t_{k+1}}\left\{\| \alpha_{}(\mw_s)\|^2-\|\widetilde{\alpha}_k(\mw_s)\|^2+\Delta H_k(\mw_s)\right\}\rmd s\right\}\eqsp,
+\end{multline*}
+where $\nabla H_k = \alpha -\widetilde{\alpha}_k$ and $\Delta$ is the Laplace operator. Then, the probability measures conditioned on hitting $X_{k+1}$ at time $t_{k+1}$ satisfy:
 \begin{multline}
 \label{eq:Girsanov}
-\frac{\rmd \mathbb{Q}^{k,X_{k+1}}_{\theta}}{\rmd \mathbb{W}^{k,X_{k+1}}_{\theta}}(\mw) = \frac{\widetilde{\qk}(X_{k},X_{k+1})}{\qk (X_{k},X_{k+1}) }\mathrm{exp}\left\{H_{\theta}(X_{k+1}) - H_{\theta}(X_{k})\right\} \\
-\times \mathrm{exp}\left\{- \frac{1}{2}\int_{t_{k}}^{t_{k+1}}\left\{\|\alpha_{\theta}(\mw_s)-\widetilde{\alpha}_{\theta}(\mw_s)\|^2+\Delta H_{\theta}(\mw_s)\right\}\rmd s\right\}\eqsp.
+\frac{\rmd \mathbb{Q}_k^{X_{k+1}}}{\rmd \mathbb{W}_k^{X_{k+1}}}(\mw) = \frac{\widetilde{\qk}(X_{k},X_{k+1})}{\qk (X_{k},X_{k+1}) }\mathrm{exp}\left\{H_{k}(X_{k+1}) - H_{k}(X_{k})\right\} \\
+\times \mathrm{exp}\left\{- \frac{1}{2}\int_{t_{k}}^{t_{k+1}}\left\{\|\alpha_{}(\mw_s)\|^2-\|\widetilde{\alpha}_{k}(\mw_s)\|^2+\Delta H_{k}(\mw_s)\right\}\rmd s\right\}\eqsp.
 \end{multline}
-Consider the following asumptions on the model:
-\begin{enumerate}[(a)]
-\item For all $\theta\in\Theta$ and all $0\le k \le n-1$, $x\mapsto \alpha_{\theta}(x) - \alpha_{\theta}(X_k) - \Jk(x-X_k)$ is bounded.
-\item For all $\theta\in\Theta$ and all $0\le k \le n-1$, $x\mapsto \mathrm{Tr}\left(J_{\alpha_{\theta}}(x) - \Jk\right)$ is bounded, where $\mathrm{Tr}$ is the trace operator.
-\end{enumerate}
-Under these two assumptions, for all $\theta$, there exists $\ell(\theta)$ such that:
+\subsection{Linear approximation of the drift}
+In order to improve the performance of the GPE introduced in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,giesecke:schwenkler:2016}, the instrumental drift is chosen as the Ozaki  \cite{ozaki:1992,shoji:ozaki:1998} linear approximation of the target drift:
+\begin{equation}
+\label{eq:alpha:linear}
+\widetilde{\alpha}_k: x\mapsto \Jk x + \alpha(X_{k}) - \Jk X_{k}\eqsp,
+\end{equation}
+where $\Jk$ is the Jacobian of $\alpha$ evaluated at $X_k$.
+In this case, the transition kernel associated with $\Oz^k$ is given in Section~\ref{sec:ornstein:bridge} and the function $H_k$ may be chosen as
 \[
-\mathrm{inf}_{u\in \mathbb{R}^d} \left\{\frac{1}{2}\left(\|\alpha_{\theta}(u)-\widetilde{\alpha}_{\theta}(u)\|^2+\Delta H_{\theta}(u)\right)\right\}\ge \ell(\theta)\eqsp.
+H_{k}: x \mapsto A_{}(x) - \frac{1}{2}x'\Jk x- (\alpha(X_{k}) - \Jk X_{k-1})'x\eqsp.
+\]
+%Consider the following assumptions on the model:
+%\begin{enumerate}[(a)]
+%\item For all $0\le k \le n-1$, $x\mapsto \alpha_{}(x) - \alpha_{}(X_k) - \Jk(x-X_k)$ is bounded.
+%\item For all $0\le k \le n-1$, $x\mapsto \mathrm{Tr}\left(J_{\alpha_{}}(x) - \Jk\right)$ is bounded, where $\mathrm{Tr}$ is the trace operator.
+%\end{enumerate}
+Assume that there exists $\ell_k$ such that:
+\[
+\mathrm{inf}_{u\in \mathbb{R}^d} \left\{\| \alpha_{}(u)\|^2-\|\widetilde{\alpha}_{k}(u)\|^2+\Delta H_{k}(u)\right\}/2\ge \ell_k\eqsp.
 \]
@@
 \[
-\phi_{\theta}(x) = \frac{1}{2}\left\{\|\alpha_{\theta}(x)-\widetilde{\alpha}_{\theta}(x)\|^2+\Delta H_{\theta}(x)\right\}-\ell(\theta)\eqsp.
+\phi_{k}: x\mapsto \left\{\|\alpha_{}(x)\|^2-\|\widetilde{\alpha}_{k}(x)\|^2+\Delta H_{k}(x)\right\}/2-\ell_k\eqsp.
 \]
-The assumptions on the model also ensure that for all $\theta\in\Theta$, there exists $\Lambda(\theta)$ such that  $\mathrm{sup}_x\eqsp \phi_{\theta}(x)\le \Lambda_{\theta}$. Then,
+Assume also that the model is such that there exists $\Lambda_k$ satisfying  $\mathrm{sup}_x\eqsp \phi_{k}(x)\le \Lambda_{k}$.
+These assumptions might seem somehow restrictive but are necessary to define the transition density estimator in the linear case.
+The alternative given by \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,giesecke:schwenkler:2016} when the instrumental drift is constant does not assume any upper bound for $\phi_{k}$ but requires to sample skeletons of a Brownian bridge given its minimum/maximum between $t_k$ and $t_{k+1}$.
+To the best of our knowledge, there is no closed from formula to sample skeletons of an Ornstein-Uhlenbeck bridges given its extremal values on an interval.
+Then,
 \begin{multline*}
-\qk(X_{k},X_{k+1}) = \widetilde{\qk}(X_{k},X_{k+1}) \eqsp\mathrm{exp}\left\{H_{\theta}(X_{k+1}) - H_{\theta}(X_{k})-\ell(\theta)\Delta_k\right\}\\
-\times\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{ \Delta_k}\phi_{\theta}(\mw_s)\rmd s\right\}\right]\eqsp,
+\qk(X_{k},X_{k+1}) = \widetilde{\qk}(X_{k},X_{k+1}) \eqsp\mathrm{exp}\left\{H_{k}(X_{k+1}) - H_{k}(X_{k})-\ell\Delta_k\right\}\\
+\times\mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{ \Delta_k}\phi_{k}(\mw_s)\rmd s\right\}\right]\eqsp,
 \end{multline*}
-where $\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}$ is the expectation under the law of the bridge process defined by \eqref{eq:SDE:Ozaki}, starting at $X_{k}$ at time $0$ and ending at $X_{k+1}$ at time $\Delta_k$.
-An unbiased estimator of $\qk(X_{k},X_{k+1})$ is then obtained by estimating $\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{ \Delta_k}\phi_{\theta}(\mw_s)\rmd s\right\}\right]$.
-Following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, if $\omega \sim\mathbb{W}_{\theta}^{k,X_{k}}$, $\kappa$ is a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and if $(U_j)_{1\le j\le \kappa}$ are independent uniform random variables on $[0,t]$,
+where $\mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}$ is the expectation under the law of the bridge process defined by \eqref{eq:SDE:Ozaki} and \eqref{eq:alpha:linear}, starting at $X_{k}$ at time $0$ and ending at $X_{k+1}$ at time $\Delta_k$.
+An unbiased estimator of $\qk(X_{k},X_{k+1})$ is then obtained by estimating $\mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{ \Delta_k}\phi_{k}(\mw_s)\rmd s\right\}\right]$. 
+Following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, if $\omega \sim\mathbb{W}_{k}^{X_{k}}$, $\kappa$ is a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and if $(U_j)_{1\le j\le \kappa}$ are independent uniform random variables on $[0,t]$,
 \begin{align*}
-\mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta_k}\phi_{\theta}(\mw_s)\rmd s\right\}\right] &= \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}} \left[\mathrm{exp}\left\{-\int_{0}^{\Delta_k}\phi_{\theta}(\mw_s)\rmd s\right\}\right]\eqsp,\\
-&= \mathrm{exp}\left\{-\Lambda_{\theta} \Delta_k\right\} \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}} \left[\mathrm{exp}\left\{ \int_{0}^{\Delta_k}(\Lambda_{\theta}-\phi_{\theta}(\mw_s))\rmd s\right\}\right]\eqsp,\\
-&=\mathrm{exp}\left\{-\Lambda_{\theta}\Delta_k\right\} \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}\left[\sum_{p\ge 0}\frac{1}{p!}\left(\int_{0}^{\Delta_k}\Lambda_{\theta}-\phi_{\theta}(\mw_s)\rmd s\right)^p\right]\eqsp,\\
-&=\mathrm{exp}\left\{-\Lambda_{\theta}\Delta_k \right\} \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}} \left[\mathbb{E}\left[\sum_{p\ge 0}\frac{( \Delta_k)^p}{k!} \prod_{j=1}^p\left(\Lambda_{\theta}-\phi_{\theta}(\mw_{U_j})\right)\right]\right]\eqsp,\\
-&=\mathrm{exp}\left\{-\Lambda_{\theta} \Delta_k\right\} \mathbb{E}_{\mathbb{W}_{\theta}^{k,X_{k+1}}}\left[ \frac{(\Delta_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\mw_{U_j})\right)\right]\eqsp.
+\mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta_k}\phi_{k}(\mw_s)\rmd s\right\}\right] &\\
+&\hspace{-3cm}= \mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}} \left[\mathrm{exp}\left\{-\int_{0}^{\Delta_k}\phi_{k}(\mw_s)\rmd s\right\}\right]\eqsp,\\
+&\hspace{-3cm}= \mathrm{exp}\left\{-\Lambda_{k} \Delta_k\right\} \mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}} \left[\mathrm{exp}\left\{ \int_{0}^{\Delta_k}(\Lambda_{k}-\phi_{k}(\mw_s))\rmd s\right\}\right]\eqsp,\\
+&\hspace{-3cm}=\mathrm{exp}\left\{-\Lambda_{k}\Delta_k\right\} \mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}\left[\sum_{p\ge 0}\frac{1}{p!}\left(\int_{0}^{\Delta_k}\Lambda_{k}-\phi_{k}(\mw_s)\rmd s\right)^p\right]\eqsp,\\
+&\hspace{-3cm}=\mathrm{exp}\left\{-\Lambda_{k}\Delta_k \right\} \mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}} \left[\mathbb{E}\left[\sum_{p\ge 0}\frac{( \Delta_k)^p}{k!} \prod_{j=1}^p\left(\Lambda_{k}-\phi_{k}(\mw_{U_j})\right)\right]\right]\eqsp,\\
+&\hspace{-3cm}=\mathrm{exp}\left\{-\Lambda_{k} \Delta_k\right\} \mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}\left[ \frac{(\Delta_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\Lambda_{k}-\phi_{k}(\mw_{U_j})\right)\right]\eqsp.
 \end{align*}
@@
 \begin{multline}
-\widehat{\qk}(X_{k},X_{k+1};\zeta_k) = \widetilde{\qk}(X_{k},X_{k+1}) \eqsp\mathrm{exp}\left\{H_{\theta}(X_{k+1}) - H_{\theta}(X_{k})-\ell(\theta) \Delta_k\right\}\\ 
-\times\mathrm{exp}\left\{-\Lambda_{\theta}\Delta_k\right\}\frac{( \Delta_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\mw_{U_j})\right)\eqsp.\label{eq:unbiased:q}
+\widehat{\qk}(X_{k},X_{k+1};\zeta_k) = \widetilde{\qk}(X_{k},X_{k+1}) \eqsp\mathrm{exp}\left\{H_{k}(X_{k+1}) - H_{k}(X_{k})-\ell_k \Delta_k\right\}\\ 
+\times\mathrm{exp}\left\{-\Lambda_{k}\Delta_k\right\}\frac{( \Delta_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\Lambda_{k}-\phi_{k}(\mw_{U_j})\right)\eqsp.\label{eq:unbiased:q}
 \end{multline}
-The choice of $\mu$ is restricted to distributions such that the associated estimator is upper bounded by an explicit constant $\hat{\sigma}_+^k(\theta)$.
+The choice of $\mu$ is restricted to distributions such that the associated estimator is upper bounded by an explicit constant $\hat{\sigma}_+^k$.
 Then, for all $0\le k \le n-1$, $1\le i\le N$, the importance weight $\omega_k^{\ell}$ is replaced by the unbiased estimator:
@@
 \label{eq:random:weight}
-\widetilde{\omega}_{k+1}^{i} = \frac{M^{-1}\sum_{m=1}^M \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta_k^m)g_{k+1}(\xi^{i}_{k+1})}{\vartheta_{k+1}(\xi^{I^{i}_{k+1}}_{k}) p_{k} (\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})}\eqsp,
+\widehat{\omega}_{k+1}^{i} = \frac{M^{-1}\sum_{m=1}^M \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta_k^m)g_{k+1}(\xi^{i}_{k+1})}{\vartheta_{k+1}(\xi^{I^{i}_{k+1}}_{k}) p_{k} (\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})}\eqsp,
 \end{equation}
-where $\{\zeta_k^i\}_{i=1}^M$ are independent copies of $\zeta_k$.\\
-%The last step to approximate the intermediate quantity of the EM algorithm is to provide an unbiased estimate of $\log q^{\Delta t_k}_{\theta}(X_{k-1},X_k)$.
-By Girsanov theorem, this log density can be written as:
-%\begin{multline*}
-%\log q^{\Delta t_k}_{\theta}(X_{k-1},X_k) = \log \widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k) + H_{\theta}(X_k) - H_{\theta}(X_{k-1})-\ell(\theta)\Delta t_k \\
-%-\mathbb{E}_{\mathbb{Q}_{\theta}^{k,X_{k}}}\left[\int_{0}^{\Delta t_k}\phi_{\theta}(\mw_s)\rmd s\right]\eqsp.
-%\end{multline*}
-%If $U$ is uniformy distributed on $(0,\Delta t_k)$ and if $w \sim \mathbb{Q}_{\theta}^{k,X_{k}}$, $\log q^{\Delta t_k}_{\theta}(X_{k-1},X_k)$ may then be estimated by:
-%\begin{equation} 
-%\label{eq:logq}
-%\mathsf{q}_{\theta}^{\Delta t_k}(X_{k-1},X_k,\omega_U) = \log \widetilde{q}^{\Delta t_k}_{\theta}(X_{k-1},X_k) + H_{\theta}(X_k) - H_{\theta}(X_{k-1})-\ell(\theta)\Delta t_k - \Delta t_k \phi_{\theta}(\omega_U)\eqsp.
-%\end{equation}
-%Sampling $w \sim \mathbb{Q}_{\theta}^{k,X_{k}}$ is possible as by \eqref{eq:Girsanov},
-%\[
-%\frac{\rmd \mathbb{Q}^{k,X_{k}}_{\theta}}{\rmd \mathbb{W}^{k,X_{k}}_{\theta}}(w) \propto 
-%\mathrm{exp}\left\{-\int_{t_{k-1}}^{t_k}\phi_{\theta}(\mw_s)\rmd s\right\}\eqsp.
-%\]
-%As noted by \cite{}, this importance ratio is proportional to the probability that a marked Poisson process on $(0,\Delta t_k)\times (0,1)$ with intensity $\Lambda_{\theta}$ is below the graph $s \mapsto \phi_{\theta}(\mw_s)/\Lambda_{\theta}$.
-$\omega_U$ in \eqref{eq:logq} is then obtaine by first sampling $\tilde{\omega}\sim \mathbb{W}^{k,X_{k}}$ at times on $(0,\Delta t_k)$ defined by the marked Poison process (which is accepted if is is below $\phi_{\theta}(\mw_s)/\Lambda_{\theta}$) and then by interpolating the bridge at time $U$.
-The resulting Ozaki In PaRIS (OzIRis) algortihm, which provides an approximation of \eqref{def:addfunc}, is summarized in Algorithm~\ref{alg:Ozaki:PaRIS}.%the E step of the EM algorithm.
+where $\{\zeta_k^i\}_{i=1}^M$ are independent copies of $\zeta_k$.
+The resulting Ozaki In PaRIS (OzIRis) algortihm, which provides an approximation of \eqref{def:addfunc}, is summarized in Algorithm~\ref{alg:Ozaki:PaRIS}.
 \begin{algorithm}
@@
 \STATE Set $\tau_{k+1}^i=0$;
-\STATE Choose $I_{k+1}^{i}$ in $\{1,\ldots,N\}$ with probs $\propto \{\widetilde{\omega}_{k}^1\vartheta_{k+1}(\xi_{k}^1),\dots,\widetilde{\omega}_{k}^N\vartheta_{k+1}(\xi_{k}^N)\}$;
+\STATE Sample $I_{k+1}^{i}$ in $\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1\vartheta_{k+1}(\xi_{k}^1),\dots,\widehat{\omega}_{k}^N\vartheta_{k+1}(\xi_{k}^N)\}$.
 \STATE Sample $\xi_{k+1}^{i} \sim p_k(\xi_{k}^{I_{k+1}^{i}},\cdot)$.
-\STATE Sample  $\{z_k^m=(\kappa_m, \mw_m, (U_j^m)_{1\leq j\leq \kappa_m})\}_{m=1}^M\overset{i.i.d}{\sim}  (\mu, \mathbb{W}_\theta^{k,X_{k+1}},\mathcal{U}[0,\Delta_k]^{\otimes \kappa_m})$.
-\STATE Compute $\widetilde{\omega}^{i}_{k+1}$ using equation \eqref{eq:random:weight}.
+\STATE For all $1\le m\le M$, sample independently $\zeta_k^m=(\kappa_m,\mw_m, (U_j^m)_{1\leq j\leq \kappa_m})$ with $\kappa_m\sim \mu$, $\mw_m\sim \mathbb{W}_k^{X_{k+1}}$ and $(U_j^m)_{1\leq j\leq \kappa_m}\sim \mathcal{U}[0,\Delta_k]^{\otimes \kappa_m}$.
+\STATE Compute $\widehat{\omega}^{i}_{k+1}$ using equation \eqref{eq:random:weight}.
 \FORALL{$\ell \in 1,\dots, \widetilde N$} 
-\STATE \textbf{\sc Sampling Step} Sample independently\\ $j\sim \{\omega_{k}^1,\dots,\omega_{k}^N\}$, $z\sim\zeta_k$, $u\sim \mathcal{U}[0,1]$
-\IF{ $$u \leq \frac{\widehat{\qk}(\xi_{k}^j,\xi_{k+1}^i,z)}{\hat{\sigma}_+},$$}
+\STATE \textbf{\sc Sampling Step} \\
+Sample independently $\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to  $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
+\IF{ $$U \leq \frac{\widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)}{\hat{\sigma}^k_+},$$}
 \STATE Set $J_k^{i,\ell} = j$.
@@
 \end{algorithm}
-%\log \left\{q^{\Delta t_k}_{\theta}(X_{k-1},X_k)g_{\theta}(X_k,Y_k)
+
+\subsection{Constant approximation of the drift}
+\label{sec:constantdrift}
+In the case where the function $\phi_k$ cannot be upper bounded when $\widetilde{\alpha}_k$ is set as a linear approximation of $\alpha$, \cite{giesecke:schwenkler:2016} suggests to set $\widetilde{\alpha}_k$ as a constant value $\widetilde{\alpha}_k= \rho_k$.
+The GPE of \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} is a special case with $\rho_k = 0$.
+In this case, $H_k$ may be chosen as $H_k: x\mapsto A(x) - \rho_k x$.
+Then, following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,olsson:strojby:2011}, it is enough to sample a pair $(L_k,U_k)$ of random variables and a Brownian bridge $\mw\sim \mathbb{W}_k^{X_{k+1}}$ (or a Brownian bridge with a constant drift $\rho_k$) such that, for all $t_k\le t\le t_{k+1}$,
+\[
+L_k\le \phi_k(\mw_s)\le U_k\eqsp.
+\]
+Sampling the bridge conditionally on its minimum and maximum is possible using Bessel bridges as noted by \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.
+Then, the second term of \eqref{eq:unbiased:q} is replaced by:
+\[
+\mathrm{exp}\left\{-U_{k}\Delta_k\right\}\frac{( \Delta_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(U_{k}-\phi_{k}(\mw_{U_j})\right)\eqsp.
+\]
+Here again, the choice of $\mu$ is restricted to distributions such that the associated estimator is upper bounded by an explicit constant $\hat{\sigma}_+^k$, see Section~\ref{sec:exp}.
 
@@
 \label{assum:boundalgo}
-$\underset{k\geq 1}{\sup}|\vartheta_k|_{\infty} < \infty$, $\underset{k\geq 1}{\sup}|p_k|_{\infty} < \infty$ and $\underset{k\geq 1}{\sup}|\widetilde{\omega}_{k}|_{\infty} < \infty$.
+$\underset{k\geq 1}{\sup}|\vartheta_k|_{\infty} < \infty$, $\underset{k\geq 1}{\sup}|p_k|_{\infty} < \infty$ and $\underset{k\geq 1}{\sup}|\widehat{\omega}_{k}|_{\infty} < \infty$, where
+\[
+\widehat{\omega}_{k}(x,x';z) = \frac{\widehat{\qk}(x,x';z)g_{k+1}(x')}{\vartheta_{k+1}(x) p_{k} (x,x')}\eqsp.
+\]
 \end{hypA}
@@
 \label{lem:iid}
-For all $0\le k \le n-1$, the random variables $\{\widetilde{\omega}_{k+1}^i\tau_{k+1}^i\}_{i=1}^N$ are independent conditionally on $\mathcal{F}_k^{N}$ and%For all bounded measurable function $h$ on $\mathbb{R}^d$,
+For all $0\le k \le n-1$, the random variables $\{\widehat{\omega}_{k+1}^i\tau_{k+1}^i\}_{i=1}^N$ are independent conditionally on $\mathcal{F}_k^{N}$ and%For all bounded measurable function $h$ on $\mathbb{R}^d$,
 \[
@@
 \]
-where $a_N = N^{-1}\sum_{i=1}^N \widetilde{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}h_{k+1}\right]\right)$ and $b_N =N^{-1}\sum_{i=1}^N \widetilde{\omega}_{k+1}^i$.
-By Lemma~\ref{lem:iid}, the random variables $\{\widetilde{\omega}_{k+1}^i\tau_{k+1}^i\}_{i=1}^N$ are independent conditionally on $\mathcal{F}_k^{N}$ and by H\ref{assum:boundalgo},
+where $a_N = N^{-1}\sum_{i=1}^N \widehat{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}h_{k+1}\right]\right)$ and $b_N =N^{-1}\sum_{i=1}^N \widehat{\omega}_{k+1}^i$.
+By Lemma~\ref{lem:iid}, the random variables $\{\widehat{\omega}_{k+1}^i\tau_{k+1}^i\}_{i=1}^N$ are independent conditionally on $\mathcal{F}_k^{N}$ and by H\ref{assum:boundalgo},
 \[
-\left|\widetilde{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}h_{k+1}\right]\right)\right| \le 2|\widetilde{\omega}_{k+1}|_{\infty}|H_{k+1}|_{\infty}\eqsp.
+\left|\widehat{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}h_{k+1}\right]\right)\right| \le 2|\widehat{\omega}_{k+1}|_{\infty}|H_{k+1}|_{\infty}\eqsp.
 \]
@@
 \]
-By \cite[Lemma~11]{olsson:westerborn:2016}., $\phi_{k}\left[\Upsilon_k\right] = 0$ which implies by Proposition~\ref{prop:filter} that 
+By \cite[Lemma~11]{olsson:westerborn:2016}, $\phi_{k}\left[\Upsilon_k\right] = 0$ which implies by Proposition~\ref{prop:filter} that 
 \[
@@
 \] 
-Similarly, as $b_N \le |\widetilde{\omega}_k|_{\infty}$, by Hoeffding inequality,
+Similarly, as $b_N \le |\widehat{\omega}_k|_{\infty}$, by Hoeffding inequality,
 \[
@@
 \label{sec:exp}
-This section investigates the performance of the proposed Ozaki-PaRIS EM algorithm. The different tuning parameters are:
+This section investigates the performance of the proposed algorithm. The different tuning parameters are:
 \begin{enumerate}[i)]
@@
 \end{enumerate}
-The algorithm is compared to the fixed-lag particle smoother of \cite{}.
+The algorithm is compared to the fixed-lag particle smoother of \cite{olsson:strojby:2011}.
 To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated.
@@
 \begin{enumerate}[i)]
-\item a bridge with constant drift as proposed in \cite{}.
+\item a bridge with constant drift as proposed in \cite{giesecke:schwenkler:2016}.
 \item a guided bridge process, see \cite{}.

### 2016-12-22 ParisEM.tex
minor changes

@@
 %Statistical inference for partially observed stochastic differential equations is a challenging task since some elementary quantities, such as transition probabilities, are not available explicitly.
-Statistical inference for partially observed stochastic differential equations requires to solve bayesian filtering and smoothing problems, i.e.
-the computation of the posterior distributions of a sequence of hidden states given observations.
-Filtering refers to the estimation of the distributions of the hidden state $X_k$ given the observations $(Y_0,\ldots,Y_k)$ up to time $k$, while fixed-interval smoothing stands for the estimation of the distribution of sequence of states $(Z_{k},\ldots,Z_{p})$ given observations $(Y_{0},\ldots,Y_{\ell})$ with $0\le k\le p<\ell \le n$.
+Statistical inference for partially observed stochastic differential equations often requires to solve bayesian filtering and smoothing problems, i.e.
+the computation of the posterior distributions of sequences of hidden states given observations.
+Filtering refers to the estimation of the distributions of the hidden state $X_k$ given the observations $(Y_0,\ldots,Y_k)$, while smoothing stands for the estimation of the distribution of sequence of states $(Z_{k},\ldots,Z_{p})$ given observations $(Y_{0},\ldots,Y_{\ell})$ with $0\le k\le p<\ell \le n$.
 These posterior distributions are crucial to compute maximum likelihood estimators of the unknown parameters defining the model using the observations $(Y_0,\ldots,Y_n)$ only.
@@
 The exact computation of these expectations is usually not possible for general partially observed diffusions.
-In this paper, we propose to use Sequential Monte Carlo methods to approximate smoothing distributions with random particles associated with importance weights.
+In this paper, we propose to use Sequential Monte Carlo (SMC) methods to approximate smoothing distributions with random particles associated with importance weights.
  \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers by combining importance sampling steps to propagate particles with importance resampling steps to duplicate or discard particles according to their importance weights.
 Unfortunately, these methods cannot be applied directly to partially observed stochastic differential equations since some elementary quantities, such as transition densities of the hidden states, are not available explicitly.
-The transition density may be approximated using discretization procedures such as the Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998} or a Gaussian based approximation using Taylor expansions of the posterior mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}.
+Many discretization procedures may be used to approximate transition densities, for instance the Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998}, or a Gaussian based approximation using Taylor expansions of the posterior mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}.
 Other approaches based on Hermite polynomials expansion were also introduced by \cite{ait-sahalia:1999,ait-sahalia:2002,ait-sahalia:2008} and extended in several directions recently, see \cite{li:2013} and all the references on the approximation of transition densities therein.
@@
 To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of transition densities based on the generalized Poisson estimator (GPE) introduced in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.
+In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities.
 
@@
 The crucial result of the application of PaRIS algorithm to SDE is that the accept reject mechanism ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by bounded unbiased estimates.
-We also propose to improve this algorithm to obtain faster convergence of the sequence of Monte Carlo EM estimates, and to extend the range of diffusion processes for which SMC methods can be applied.To do so, we introduce a new class of GPE based on the Ozaki discretization of the target SDE, see \cite{ozaki:1992,shoji:ozaki:1998}.
+We also propose to improve this algorithm to obtain faster convergence of the sequence of Monte Carlo EM estimates, and to extend the range of diffusion processes for which SMC methods can be applied.
+To do so, we introduce a new class of GPE based on the Ozaki discretization of the target SDE, see \cite{ozaki:1992,shoji:ozaki:1998}.
 These new GPE estimators use a an order one Taylor expansion of the drift of the SDE to obtain more efficient estimates than Brownian Bridge based GPE.
@@
 The PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSi algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions.
-The FFBS algorithm has a computational complexity of order $\mathcal{O}(N^{k'-k+1})$ to approximate joint smoothing distributions of the form $\phi_{k:k'|n}$ with $k<k'$ which makes it prohibitive.
+%The FFBS algorithm has a computational complexity of order $\mathcal{O}(N^{k'-k+1})$ to approximate joint smoothing distributions of the form $\phi_{k:k'|n}$ with $k<k'$ which makes it prohibitive. 
 An important feature of the FFBS algorithm shown in \cite{mongillo:deneve:2008,cappe:2011,delmoral:doucet:singh:2010} is that it can be implemented using only a forward pass when approximating smoothed expectations of additive functions.
-Instead of computing new weights in the backward pass, the FFBSi  samples backward new trajectories from time $n$ to time $0$ among all the $N^{n+1}$ trajectories which can be chosen in $\{\xi^{\ell}_k\}$, $1\le \ell\le N$ and $0\le k\le n$.
+%Instead of computing new weights in the backward pass, the FFBSi  samples backward new trajectories from time $n$ to time $0$ among all the $N^{n+1}$ trajectories which can be chosen in $\{\xi^{\ell}_k\}$, $1\le \ell\le N$ and $0\le k\le n$.
 Contrary to the FFBS algorithm, the FFBSi algorithm samples trajectories and may be used to approximate joint smoothing distributions without increasing its complexity.
- Nevertheless, the computation of the transition probabilities to select particles in the backward pass makes it a $\mathcal{O}(N^2)$ algorithm.
+%Nevertheless, the computation of the transition probabilities to select particles in the backward pass makes it a $\mathcal{O}(N^2)$ algorithm.
 In the case where the transition probability $q_k$ is upper bounded, \cite{douc:garivier:moulines:olsson:2011} introduced an acceptance rejection mechanism to implement the FFBSi algorithm with a $\mathcal{O}(N)$ complexity, see also \cite{dubarry:lecorff:2011}.
+The PaRIS algorithm combines both the forward only version of the FFBS algorithm with the sampling mechanism of the FFBSi algorithm.
+It does not produce an approximation of the smoothing distributions but of the smoothed expectation of a fixed additive functional and thus  may be used to approximate \eqref{def:addfunc}.
+Its crucial property is that it does not require a backward pass, the smoothed expectations is computed on-the-fly with the particle filter and no storage of the particles or weights is needed.
 
-The PaRIS algorithm combines both the forward only version of the FFBS algorithm with the sampling mechanism of the FFBSi algorithm. 
-The algorithm does not produce an approximation of the smoothing distributions but of the smoothed expectation of a fixed additive functional and thus  may be used to approximate \eqref{def:addfunc}. 
-Its crucial property is that it does not require a backward pass, the smoothed expectations is computed on-the-fly with the particle filter and no storage of the particles or weights is needed.
-The focus is based sufficient statistics $\tau^i_k$, approximations of $T_k[h](\xi^i_k)$, for $1\le i\le N$ and $0\le k \le n$ starting with $\tau^i_0 = 0$ for all $1\le i\le N$.
-Let $\tilde{N}\ge 1$.
-Then, at each time step $0\le k \le {n-1}$ these statistics are updated according to the following steps.
+The algorithm introduces sufficient statistics $\tau^i_k$, approximating $T_k[h](\xi^i_k)$, for $1\le i\le N$ and $0\le k \le n$ starting with $\tau^i_0 = 0$ for all $1\le i\le N$.
+Choosing $\tilde{N}\ge 1$, at each time step $0\le k \le {n-1}$ these statistics are updated according to the following steps.
 \begin{enumerate}[(i)]
@@
 Following \cite{fearnhead:papaspiliopoulos:roberts:2008,olsson:strojby:2011},
-an extension of the PaRIS algorithm may be developed by replacing $\qk$  by an  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable such that:
+an extension of the PaRIS algorithm may be developed by replacing $\qk(\xi^{\ell}_{k},\xi_{k+1}^{i})$  by an  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable such that:
 \[
@@
 \begin{algorithm}[H]
-\caption{Double random weight Accept-reject-based backward sampling}
+\caption{Random weight accept-reject backward sampling}
 \begin{algorithmic}
@@
 In \cite{giesecke:schwenkler:2016}, the authors proposed to replace the Brownian bridge by a Brownian motion with constant drift between two consecutive observations.
-As an alternative, is is also possible to follow \cite{ozaki:1992,shoji:ozaki:1998} to propose a higher order approximation of the drift term $\alpha$.
+As an alternative, it is also possible to follow \cite{ozaki:1992,shoji:ozaki:1998} to propose a higher order approximation of the drift term $\alpha$.
 The authors used a linear approximation of the diffusion drift $\alpha$, together with a constant approximation of the volatility over each time interval $(t_{k},t_{k+1})$.
@@
 \[
-\mathrm{inf}_{u\in \mathbb{R}^d} \left\{\| \alpha_{}(u)\|^2-\|\widetilde{\alpha}_{k}(u)\|^2+\Delta H_{k}(u)\right\}/2\ge \ell_k\eqsp.
+\mathrm{inf}_{u\in \mathbb{R}^d} \left\{\| \alpha_{}(u)\|^2-\|\widetilde{\alpha}_{k}(u)\|^2+\Delta H_{k}(u)\right\}/2\ge \ell_k
 \]
-Define
+and
 \[

### 2017-01-10 ParisEM.tex
New minor corrections - All sections

@@
 
-The exact computation of these expectations is usually not possible for general partially observed diffusions.
+The exact computation of these expectations is usually not possible in the case of partially observed diffusions.
 In this paper, we propose to use Sequential Monte Carlo (SMC) methods to approximate smoothing distributions with random particles associated with importance weights.
- \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers by combining importance sampling steps to propagate particles with importance resampling steps to duplicate or discard particles according to their importance weights.
+ \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers for state space models by combining importance sampling steps to propagate particles with importance resampling steps to duplicate or discard particles according to their importance weights.
 Unfortunately, these methods cannot be applied directly to partially observed stochastic differential equations since some elementary quantities, such as transition densities of the hidden states, are not available explicitly.
-Many discretization procedures may be used to approximate transition densities, for instance the Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998}, or a Gaussian based approximation using Taylor expansions of the posterior mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}.
+Discretization procedures may be used to approximate transition densities, for instance the Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998}, or Gaussian based approximations using Taylor expansions of the posterior mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}.
 Other approaches based on Hermite polynomials expansion were also introduced by \cite{ait-sahalia:1999,ait-sahalia:2002,ait-sahalia:2008} and extended in several directions recently, see \cite{li:2013} and all the references on the approximation of transition densities therein.
-However, even the most recent discretization based approximations of the transition densities induce a systematic bias of particle based approximations of posterior distributions, see \cite{delmoral:jacod:protter:2001}.
-To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of transition densities based on the generalized Poisson estimator (GPE) introduced in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.
+However, even the most recent discretization based approximations of the transition densities induce a systematic bias of particle based approximations of posterior distributions, see for instance \cite{delmoral:jacod:protter:2001}.
+To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE) introduced in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.
 In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities.
@@
 Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias that does not vanish as the number of particles grows to infinity.
-Approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}.
+In the case of discrete time state space models, approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}.
 Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$.
@@
 
-In this paper, instead of using fixed-lag techniques which introduce a systematic additional bias in the approximation of conditional expectations, we extend the use of  PaRIS algorithm to partially observed SDE.
+In this paper, instead of using fixed-lag techniques which introduce a systematic additional bias in the approximation of conditional expectations, we extend the use of  the PaRIS algorithm to partially observed SDE.
 The proposed algorithm allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles.
 The crucial result of the application of PaRIS algorithm to SDE is that the accept reject mechanism ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by bounded unbiased estimates.
-We also propose to improve this algorithm to obtain faster convergence of the sequence of Monte Carlo EM estimates, and to extend the range of diffusion processes for which SMC methods can be applied.
+We also propose to improve the algorithm of \cite{fearnhead:papaspiliopoulos:roberts:2008} by extending the range of diffusion processes for which SMC methods can be applied.
 To do so, we introduce a new class of GPE based on the Ozaki discretization of the target SDE, see \cite{ozaki:1992,shoji:ozaki:1998}.
-These new GPE estimators use a an order one Taylor expansion of the drift of the SDE to obtain more efficient estimates than Brownian Bridge based GPE.
+These new GPE estimators use an order one Taylor expansion of the drift of the SDE to obtain more efficient estimates than Brownian Bridge based GPE.
+
+The paper is organized as follows.
+Section~\ref{sec:rwparis} describes the random weight PaRIS algorithm to approximate smoothed additive functionals using unbiased estimates of the transition density of the hidden states.
+Then, Section~\ref{sec:PaRIS:SDE} details the application of this algorithm when the transition density may be approximated using a GPE based on the Ozaki discretization or a GPE based on a constant approximation of the drift function.
+The latter is a straightforward generalization of \cite{fearnhead:papaspiliopoulos:roberts:2008}.
+In Section~\ref{sec:convergence}, classical convergence results for SMC smoothers are extended to the setting of this paper and illustrated with numerical experiments in Section~\ref{sec:exp}.
+All proofs are postponed to Appendix~\ref{sec:append:proofs}.
 
 \section{Random weight PaRIS algorithm}
-In this paper, $(X_t)_{t\ge 0}$ is a weak solution to the following SDE in $\mathbb{R}^d$:
+\label{sec:rwparis}
+$(X_t)_{t\ge 0}$ is defined as a weak solution to the following SDE in $\mathbb{R}^d$:
 \begin{equation}
@@
 For all $0\le k \le n$, the distribution of $Y_k$ given $X_k$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,Y_k) = g_k(X_k)$. 
-The distribution of $X_0$ has a density with respect to $\lambda$ given by $\chi$.
-For all $1\le k \le n$, the distribution of $X_{k+1} $ conditional on $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to a reference measure $\mu$ on $\mathbb{R}^d$. 
+The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
+For all $1\le k \le n$, the distribution of $X_{k+1} $ conditional on $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$. 
 Let $0 \leq k \leq k' \leq n$. The joint smoothing distributions of the hidden states are defined, for all measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, by:
@@
 %The FFBS algorithm has a computational complexity of order $\mathcal{O}(N^{k'-k+1})$ to approximate joint smoothing distributions of the form $\phi_{k:k'|n}$ with $k<k'$ which makes it prohibitive. 
-An important feature of the FFBS algorithm shown in \cite{mongillo:deneve:2008,cappe:2011,delmoral:doucet:singh:2010} is that it can be implemented using only a forward pass when approximating smoothed expectations of additive functions.
+%An important feature of the FFBS algorithm shown in \cite{mongillo:deneve:2008,cappe:2011,delmoral:doucet:singh:2010} is that it can be implemented using only a forward pass when approximating smoothed expectations of additive functions.
 %Instead of computing new weights in the backward pass, the FFBSi  samples backward new trajectories from time $n$ to time $0$ among all the $N^{n+1}$ trajectories which can be chosen in $\{\xi^{\ell}_k\}$, $1\le \ell\le N$ and $0\le k\le n$.
@@
 In the case where the transition probability $q_k$ is upper bounded, \cite{douc:garivier:moulines:olsson:2011} introduced an acceptance rejection mechanism to implement the FFBSi algorithm with a $\mathcal{O}(N)$ complexity, see also \cite{dubarry:lecorff:2011}.
-The PaRIS algorithm combines both the forward only version of the FFBS algorithm with the sampling mechanism of the FFBSi algorithm.
+It combines both the forward only version of the FFBS algorithm with the sampling mechanism of the FFBSi algorithm.
 It does not produce an approximation of the smoothing distributions but of the smoothed expectation of a fixed additive functional and thus  may be used to approximate \eqref{def:addfunc}.
@@
 The alternative given by \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,giesecke:schwenkler:2016} when the instrumental drift is constant does not assume any upper bound for $\phi_{k}$ but requires to sample skeletons of a Brownian bridge given its minimum/maximum between $t_k$ and $t_{k+1}$.
-To the best of our knowledge, there is no closed from formula to sample skeletons of an Ornstein-Uhlenbeck bridges given its extremal values on an interval.
+To the best of our knowledge, there is no closed from formula to sample skeletons of an Ornstein-Uhlenbeck bridge given its extremal values on an interval.
 Then,
@@
 where $\mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}$ is the expectation under the law of the bridge process defined by \eqref{eq:SDE:Ozaki} and \eqref{eq:alpha:linear}, starting at $X_{k}$ at time $0$ and ending at $X_{k+1}$ at time $\Delta_k$.
-An unbiased estimator of $\qk(X_{k},X_{k+1})$ is then obtained by estimating $\mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{ \Delta_k}\phi_{k}(\mw_s)\rmd s\right\}\right]$. 
+An unbiased estimator of $\qk(X_{k},X_{k+1})$ is then obtained by estimating the expectation in the previous equation. 
 Following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, if $\omega \sim\mathbb{W}_{k}^{X_{k}}$, $\kappa$ is a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and if $(U_j)_{1\le j\le \kappa}$ are independent uniform random variables on $[0,t]$,
@@
 \mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta_k}\phi_{k}(\mw_s)\rmd s\right\}\right] &\\
-&\hspace{-3cm}= \mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}} \left[\mathrm{exp}\left\{-\int_{0}^{\Delta_k}\phi_{k}(\mw_s)\rmd s\right\}\right]\eqsp,\\
 &\hspace{-3cm}= \mathrm{exp}\left\{-\Lambda_{k} \Delta_k\right\} \mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}} \left[\mathrm{exp}\left\{ \int_{0}^{\Delta_k}(\Lambda_{k}-\phi_{k}(\mw_s))\rmd s\right\}\right]\eqsp,\\
@@
 where $\{\zeta_k^i\}_{i=1}^M$ are independent copies of $\zeta_k$.
-The resulting Ozaki In PaRIS (OzIRis) algortihm, which provides an approximation of \eqref{def:addfunc}, is summarized in Algorithm~\ref{alg:Ozaki:PaRIS}.
+The resulting Ozaki in PaRIS (OzIRis) algortihm, which provides an approximation of \eqref{def:addfunc}, is summarized in Algorithm~\ref{alg:Ozaki:PaRIS}.
 \begin{algorithm}
@@
 \]
-Here again, the choice of $\mu$ is restricted to distributions such that the associated estimator is upper bounded by an explicit constant $\hat{\sigma}_+^k$, see Section~\ref{sec:exp}.
+Here again, the choice of $\mu$ is restricted to distributions such that the associated estimator is upper bounded by an explicit constant $\hat{\sigma}_+^k$.
 
 \section{Convergence results}
+\label{sec:convergence}
 Consider the following assumptions.
@@
 The proof is completed using Lemma~\ref{lem:hoeffding:ratio}.
+\end{proof}
+
+\begin{proposition}
+\label{prop:filter}
+Assume that H\ref{assum:boundmodel} and H\ref{assum:boundalgo} hold. Then, for all $0\le k\le n$,
+there exist $b_k, c_k>0$ such that for all $N\ge 1$, all $\varepsilon > 0$, and all measurable function $h$ on $\mathsf{X}$,
+\begin{equation*}
+\mathbb{P}\left( \left|\phi_k^N[h] - \phi_k\left[h\right]\right| \geq \varepsilon \right) \leq b_k \mathrm{exp}\left(-c_k N \epsilon^2\right)\eqsp.
+\end{equation*}
+\end{proposition}
+
+\begin{proof}
+\textcolor{red}{On l'ecrit peut-etre... }
+The proof is a straightforward extension of \cite{douc:garivier:moulines:olsson:2011}. 
 \end{proof}
@@
 \begin{proof}[Proof of Lemma~\ref{lem:AR:unbiased}]
-%Let $\tau$ be the first time  draws are accepted in the accept-reject mechanism. For all $\ell\ge 1$, write
-%\[
-%\mathcal{A}^k_{\ell} = \left\{U_\ell<\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_\ell},\xi_k^{i},\Lambda_{\ell})/\sigma_+\right\}\eqsp.
-%\]
-%
-%Let $h$ be a function defined on $\{1,\ldots,N\}$,
-%\begin{align*}
-%\mathbb{E}\left[h(J^{i,j}_k)\middle| \mathcal{F}_k^N\right] & = \sum_{m\ge 1}\mathbb{E}\left[h(I_m)\1_{\tau=m}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-%& = \sum_{m\ge 1}h(m)\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\middle| \mathcal{F}_k^N\right]\right)\mathbb{E}\left[h(I_m)\1_{\mathcal{A}^k_{m}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-%& = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[1-\frac{\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_\ell},\xi_k^{i},\Lambda_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)\\
-%&\hspace{5cm}\times\mathbb{E}\left[h(I_m)\frac{\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_m},\xi_k^{i},\Lambda_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-%& = \sum_{m\ge 1}\left(\mathbb{E}\left[1-\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)^{m-1}\mathbb{E}\left[h(I_m)\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-%& = \mathbb{E}\left[h(I_1)q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{i})\middle| \mathcal{F}_k^N\right]/\mathbb{E}\left[q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{i})\middle| \mathcal{F}_k^N\right]\eqsp,\\
-%& = \sum_{\ell=1}^N \frac{h(\ell)\omega_{k-1}^{\ell}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{\ell},\xi_k^{i})}{\sum_{m=1}^N\omega_{k-1}^{m}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{m},\xi_k^{i})}\eqsp,\\
-%&= \sum_{\ell=1}^N \Lambda_{k-1}^N(i,\ell)h(\ell) \eqsp,
-%\end{align*}
-%which concludes the proof.
-The goal is to simulate realisations of  a discrete random variable $X$ having the following target distribution:
-\begin{equation}
-\mP(X=\ell)= p_X(\ell)=\frac{\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})}{\sum_{\ell=1}^N\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})},~~\ell\in \{1,\dots,N\}
-\end{equation}
-for some $\xi_{k+1}$ that has no importance in this result.
+Let $\tau$ be the first time  draws are accepted in the accept-reject mechanism. For all $\ell\ge 1$, write
+\[
+\mathcal{A}^k_{\ell} = \left\{U_\ell<\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_\ell},\xi_k^{i},\Lambda_{\ell})/\sigma_+\right\}\eqsp.
+\]
+
+Let $h$ be a function defined on $\{1,\ldots,N\}$,
+\begin{align*}
+\mathbb{E}\left[h(J^{i,j}_k)\middle| \mathcal{F}_k^N\right] & = \sum_{m\ge 1}\mathbb{E}\left[h(I_m)\1_{\tau=m}\middle| \mathcal{F}_k^N\right]\eqsp,\\
+& = \sum_{m\ge 1}h(m)\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\middle| \mathcal{F}_k^N\right]\right)\mathbb{E}\left[h(I_m)\1_{\mathcal{A}^k_{m}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
+& = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[1-\frac{\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_\ell},\xi_k^{i},\Lambda_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)\\
+&\hspace{5cm}\times\mathbb{E}\left[h(I_m)\frac{\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_m},\xi_k^{i},\Lambda_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
+& = \sum_{m\ge 1}\left(\mathbb{E}\left[1-\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)^{m-1}\mathbb{E}\left[h(I_m)\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
+& = \mathbb{E}\left[h(I_1)q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{i})\middle| \mathcal{F}_k^N\right]/\mathbb{E}\left[q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{i})\middle| \mathcal{F}_k^N\right]\eqsp,\\
+& = \sum_{\ell=1}^N \frac{h(\ell)\omega_{k-1}^{\ell}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{\ell},\xi_k^{i})}{\sum_{m=1}^N\omega_{k-1}^{m}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{m},\xi_k^{i})}\eqsp,\\
+&= \sum_{\ell=1}^N \Lambda_{k-1}^N(i,\ell)h(\ell) \eqsp,
+\end{align*}
+which concludes the proof.
+
+%The goal is to simulate realisations of  a discrete random variable $X$ having the following target distribution:
+%\begin{equation}
+%\mP(X=\ell)= p_X(\ell)=\frac{\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})}{\sum_{\ell=1}^N\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})},~~\ell\in \{1,\dots,N\}
+%\end{equation}
+%for some $\xi_{k+1}$ that has no importance in this result.
 We assume $\omega_k$ are normalized weights, i.e.
@@
 with the following distribution:
-\begin{equation}
-\mP(Y=\ell)=p_Y(\ell)= \omega_k^\ell,~~\ell\in \{1,\dots,N\}
-\end{equation}
-Unfortunately, the $q_k$ function is unknown.
+%\begin{equation}
+%\mP(Y=\ell)=p_Y(\ell)= \omega_k^\ell,~~\ell\in \{1,\dots,N\}
+%\end{equation}
+%Unfortunately, the $q_k$ function is unknown.
 Howevere, we let's suppose we can define an unbiased estimator $\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)$, where $\zeta$ is a random variable having th p.d.f.
@@
 We have therefore:
-$$\E_{p_\zeta}\left[\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)\right]=q(\xi_k^\ell,\xi_{k+1})$$
-Moreover, we impose the following property over this estimator, we want that:
-$$\exists \sigma_+\in \mathbb{R}\text{ such that }\forall x,y,~~\hat{q}_k(x,y,\zeta)\leq \sigma_+$$
-To simulate a random variable $X\sim p_X$, we propose the following algorithm:
-\begin{enumerate}
-\item Sample independently $\ell\sim p_Y$, $z\sim p_\zeta$ and $u \sim \mathcal{U}[0,1]$
-\item \textbf{If} 
-$$u \leq \frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},z)}{\sigma_+},$$
-set $x=\ell$. \textbf{Else}, return to 1).
-\end{enumerate}
+%$$\E_{p_\zeta}\left[\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)\right]=q(\xi_k^\ell,\xi_{k+1})$$
+%Moreover, we impose the following property over this estimator, we want that:
+%$$\exists \sigma_+\in \mathbb{R}\text{ such that }\forall x,y,~~\hat{q}_k(x,y,\zeta)\leq \sigma_+$$
+%To simulate a random variable $X\sim p_X$, we propose the following algorithm:
+%\begin{enumerate}
+%\item Sample independently $\ell\sim p_Y$, $z\sim p_\zeta$ and $u \sim \mathcal{U}[0,1]$
+%\item \textbf{If} 
+%$$u \leq \frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},z)}{\sigma_+},$$
+%set $x=\ell$. \textbf{Else}, return to 1).
+%\end{enumerate}
 The lemma \ref{lem:AR:unbiased} states that $x$ is a realisation of a random variable $X\sim p_X$. Indeed
@@
 The GPE introduced in this paper requires to sample finite dimensional distributions of Ornstein Uhlenbeck bridges.
-These results may be found in \cite{} and are given here for completeness.
+These results are given here for completeness.
 Consider the following $d$-dimensional SDE:

### 2017-01-11 ParisEM.tex
completing log growth model section

@@
-\documentclass[12pt]{article}
+\documentclass[12pt,draft]{article}
 \usepackage[utf8]{inputenc}
@@
 \usepackage{graphicx}
+\graphicspath{{/home/gloaguen/Bureau/Travaux_agro/figures_article_oziris/}}
 \usepackage{enumerate}
@@
 \item the proposal transition density $p_k$ and the adjustment multiplier $\vartheta_k$ used to propagate and weight particles;
-\item the discrete distribution $\mu$ used to define $\widetilde{q}_{\theta}^{\Delta t_k}$.
+\item the discrete distribution $\mu$ used to define $\widetilde{q}_{\theta}^{\Delta t_k}$;
+\item the number of samples to compute $\widetilde{q}_{\theta}^{\Delta t_k}$.
 \end{enumerate}
+The performance of the algorithm is investigated on two known models based on diffusion process.
+\subsection*{The SINE model} 
 The algorithm is compared to the fixed-lag particle smoother of \cite{olsson:strojby:2011}.
@@
 Both processes lead to estimators of $q_{\theta}^{\Delta t_k}$ following the same steps as in Section~\ref{}.
+\subsection*{The Log-growth model}
+We now focus on a second well known model in population dynamics, the log-growth model, also used as example in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} and \cite{olsson:westerborn:2016}, defined by:
+\begin{equation}
+\rmd Z_t = \kappa Z_t\left(1-\frac{Z_t}{\gamma}\right)\rmd t + \sigma Z_t \rmd W_t,~~Z_0=Z_0. \label{eq:LG:SDE}
+\end{equation}
+In order to use Exact algorithms of \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} and GPE used in \cite{fearnhead:papaspiliopoulos:roberts:2008}, we focus on the version of \eqref{eq:LG:SDE} after the Lamperti transform, i.e., we define $X_t=\eta(Z_t)$, with $\eta(z) := -\log (z)/\sigma$, in this case, $(X_t)_{t\geq 0}$ satisfies the following SDE:
+\begin{equation}
+\rmd X_t = \overbrace{\left( \frac{\sigma}{2} -  \frac{\kappa}{\sigma} + \frac{\kappa}{\gamma\sigma}\exp\left(-\sigma X_t\right)\right)}^{:=\alpha(X_t)}\rmd t +\rmd W_t,~~X_0=X_0.\label{eq:Lamp:LG}
+\end{equation}
+In this case, the conditions of the EA2 defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} are satisfied, as:
+\begin{equation}
+\forall m \in \mathbb{R}, \exists~\mathsf{U}(m)~\vert~\forall x\geq m, \psi(x):=\alpha^2(x)+\alpha'(x) \leq \mathsf{U}_m.
+\end{equation}
+Moreover, there exists $\mathsf{L}$ such that $\forall x, \psi(x):=\alpha^2(x)+\alpha'(x)\geq \mathsf{L}$. 
+Then, GPE estimators could be computed by simulating a the minimum of a OU bridge, and simulating OU bridges conditionnaly to this minimum. 
+However, to the best of our knowledge, no general techniques exist to do so. 
+We therefore apply the proposed algorithm, but with a GPE estimator computed with Bessel bridges, as proposed by \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.\\
+The process solution of \eqref{eq:Lamp:LG} is observed regularly at times $t_0=0,\dots,t_{50}=100$ through the observation process $(Y_k)_{k= 0,\dots,50}$:
+\begin{equation}
+Y_k = X_k + \varepsilon_k,~~\varepsilon \overset{i.i.d}{\sim} \mathcal{N}(0,\sigma^2_{obs})\label{eq:obs:LG}
+\end{equation}
+In the example presented here, we fixed the set of  parameter $$\theta_{true} =(\kappa=0.1,\sigma=0.1,\gamma=1000,\sigma^2_{obs}=4).$$\\
+In order to evaluate the performance of the algorithm presented here, we compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50).
+We therefore compute the particle approximation of $\mathcal{Q}(\theta_{true},\theta_{true})$ for a set of observations presented on Figure \ref{fig:obs:LG}.\\ 
+This estimation is performed 300 times, using $N=400$ particles, with $\tilde{N}=2$, and 30 replications for the Monte Carlo approximation of each $q(\xi_k,\xi_{k+1})$. 
+Moreover, in the computation of the E step, the corresponding $h_k$ function of equation \eqref{def:addfunc} requires the computation of $\log\left(q(\xi_k,\xi_{k+1})\right)$. 
+This quantity is approximated independently from other approximations, using the unbiased estimator proposed in \cite{olsson:strojby:2011} (Appendix B). 
+Again, 30 Monte Carlo approximations are used to compute each estimation.\\ In order to obtain a reference value, the same estimation is performed 30 times using $N=10000$ particles (using the proposed algorithm), the reference value is then computed as the arithmetic mean of these 30 estimations.
+The resulting estimates are presented on Figure \ref{fig:res:LG}.\\
+The results for the fixed lag technique are in conformity with the ones presented in \cite{olsson:strojby:2011} (Figure 1). 
+Indeed, for small lags, athe variance of estimates is small, but showing a bias. 
+The bias rapidly decreases as the lag increases, together with a  rapidly increasing variance. 
+The PaRIS algorithm for diffusion process presented results in an unbiased estimator, as expected, and show a small variance (lower than the variance for the fixed lag when the lag is one, therefore showing the best results on the presented example.
+It is worth noting here that, as pointed out in \cite{olsson:strojby:2011}, results of the fixed lag technique indicates a quite strong forgetting in the model, and the choice of the lag should take into account this forgetting property.
+As the PaRIS algorithm is always unbiased, there is no need to consider the forgetting property of a given model in the approximation of the EM intermediate quantity. 
+\begin{figure}[p]
+\centering
+\includegraphics[width=0.8\textwidth]{obs}
+\caption{One realisation of the process $(X_t)_{t\geq 0}$, solution of the SDE \eqref{eq:Lamp:LG} is represented in white.
+The observations of $Y$ at times $t_0=0,\dots,t_{50}=100$, satisfying equation \eqref{eq:obs:LG},  are represented in black.}
+\label{fig:obs:LG}
+\end{figure}
+\begin{figure}[p]
+\centering
+\includegraphics[width=0.8\textwidth]{res_Estep_LG}
+\caption{Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$ using observations of Figure \ref{fig:obs:LG}, using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the PaRIS algorithm for diffusion process proposed in this work, using 300 replicates.
+The whiskers represent the extent of the 95\% central values.
+The dot represents the empirical mean over the 300 replicates.
+The dotted line shows the reference value, computed using the unbiased PaRIS algorithm with $N=10000$ particles.}
+\label{fig:res:LG}
+\end{figure}
 % where $\beta\in\mathbb{R}^d$ is chosen by the user.

### 2017-01-17 ParisEM.tex
Modifications in proofs and simulations

--- 2017-01-10 ParisEM.tex
@@
 Sequential Monte Carlo methods provide an efficient and simple solution to obtain these approximations using sets of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$, $0\le k \le n$.
-At time $k = 0$, $N$ particles $\{\xi^{\ell}_0\}_{\ell=1}^N$ are initialized by setting $\xi^{\ell}_0=x_0$.
-Then, $\xi^{\ell}_0$ is associated with the importance weights $\omega_0^{\ell} = g_0 (\xi^{\ell}_0)$.
+At time $k = 0$, $N$ particles $\{\xi^{\ell}_0\}_{\ell=1}^N$ are sampled independently according to  $\xi^{\ell}_0 \sim \eta_0$, where $\eta_0$ is a probability density with respect to $\mu$.
+Then, $\xi^{\ell}_0$ is associated with the importance weights $\omega_0^{\ell} = \chi(\xi^{\ell}_0)g_0 (\xi^{\ell}_0)/\eta_0(\xi^{\ell}_0)$.
 For any bounded and measurable function $h$ defined on $\mathbb{R}^d$, the expectation $\phi_{0}[h] $ is approximated by
@@
 \]
-where 
+where, for all $0\le k \le n$, 
 \[
-\mathcal{F}_{k+1}^N = \sigma\left\{Y_{0:k+1};(\xi^{\ell}_{u},\omega^{\ell}_{u},\tau^{\ell}_{u}),~1\le \ell\le N,~0\le u\le k+1\right\}\eqsp.
+\mathcal{F}_{k}^N = \sigma\left\{Y_{0:k};(\xi^{\ell}_{u},\omega^{\ell}_{u},\tau^{\ell}_{u}),~1\le \ell\le N,~0\le u\le k\right\}\eqsp.
 \]
-In the case where $\widehat{q}_k$ may be upper bounded by a constant $\hat{\sigma}_+$, step~\eqref{it:PaRIS:sampleindex} is replaced by the acceptance-rejection mechanism given in Algorithm~\ref{alg:AR:unknownq}.
+In the case where $\widehat{q}_k$ may be upper bounded by a constant $\hat{\sigma}_+$, step~\eqref{it:PaRIS:sampleindex} is replaced by the accept reject mechanism given in Algorithm~\ref{alg:AR:unknownq}.
 Lemma~\ref{lem:AR:unbiased} states that replacing  $\qk$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.
@@
 \[
-\widehat{\omega}_{k}(x,x';z) = \frac{\widehat{\qk}(x,x';z)g_{k+1}(x')}{\vartheta_{k+1}(x) p_{k} (x,x')}\eqsp.
+\widehat{\omega}_{0}(x) = \frac{\chi(x)g_0(x)}{\eta_0(x)} \quad\mbox{and for}\; k\ge1\quad\widehat{\omega}_{k}(x,x';z) = \frac{\widehat{\qk}(x,x';z)g_{k+1}(x')}{\vartheta_{k+1}(x) p_{k} (x,x')}\eqsp.
 \]
@@
 \[
-\mathbb{E}\left[\tilde{\omega}^1_{k+1}\tau^{1}_{k+1}\middle| \mathcal{F}_k^{N}\right] = \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\int q_{\theta}^{\Delta t_{k+1}}(\cdot,x)g_{k+1}(x)\left\{\tau_k(\cdot) + h_{k+1}(\cdot,x)\right\}\rmd x\right]\eqsp.
+\mathbb{E}\left[\widehat{\omega}^1_{k+1}\tau^{1}_{k+1}\middle| \mathcal{F}_k^{N}\right] = \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\int q_{k}(\cdot,x)g_{k+1}(x)\left\{\tau_k(\cdot) + h_{k+1}(\cdot,x)\right\}\rmd x\right]\eqsp.
 \]
@@
 Assume that H\ref{assum:boundmodel} and H\ref{assum:boundalgo} hold and that for all $1\le k\le n$, $\mathrm{osc}(h_k)<+\infty$.
-For all $0\le k\le n$, there exist $b_k,c_k>0$ such that for all $N\ge 1$ and all $\varepsilon\in\mathbb{R}_+^\star$,
+For all $0\le k\le n$ and all $\widetilde{N}\ge 1$, there exist $b_k,c_k>0$ such that for all $N\ge 1$ and all $\varepsilon\in\mathbb{R}_+^\star$,
 \[
@@
 \begin{proof}
-Write
+The results is proved by induction.
+At time $k=0$, the result holds using that for all $1\le i \le N$, $\rho_0^i = 0$ and the convention $T_0[h_0] =0$.
+In addition, $\phi_0^N$ is a standard importance sampler estimator of $\phi_0$ with $\omega_0^i\le |\widehat{\omega}_0|_{\infty}$ so that for any bounded function $h$ on $\mathsf{X}$,
 \[
-\phi_{k+1}^N[\tau_{k+1}] - \phi_{k+1}\left[T_{k+1}h_{k+1}\right] = \frac{a_N}{b_N}\eqsp,
+\mathbb{P}\left(\left|\phi_0^N[h] - \phi_0\left[h\right]\right|\ge \varepsilon\right)\le b_0\exp\left(-c_0N\varepsilon^2\right)\eqsp.
 \]
-where $a_N = N^{-1}\sum_{i=1}^N \widehat{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}h_{k+1}\right]\right)$ and $b_N =N^{-1}\sum_{i=1}^N \widehat{\omega}_{k+1}^i$.
+Assume the results holds for $k\ge 1$ and write
+\[
+\phi_{k+1}^N[\tau_{k+1}] - \phi_{k+1}\left[T_{k+1}[h_{k+1}]\right] = a_N/b_N\eqsp,
+\]
+where $a_N = N^{-1}\sum_{i=1}^N \widehat{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}[h_{k+1}]\right]\right)$ and $b_N =N^{-1}\sum_{i=1}^N \widehat{\omega}_{k+1}^i$.
 By Lemma~\ref{lem:iid}, the random variables $\{\widehat{\omega}_{k+1}^i\tau_{k+1}^i\}_{i=1}^N$ are independent conditionally on $\mathcal{F}_k^{N}$ and by H\ref{assum:boundalgo},
 \[
-\left|\widehat{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}h_{k+1}\right]\right)\right| \le 2|\widehat{\omega}_{k+1}|_{\infty}|H_{k+1}|_{\infty}\eqsp.
+\left|\widehat{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}[h_{k+1}]\right]\right)\right| \le 2|\widehat{\omega}_{k+1}|_{\infty}|H_{k+1}|_{\infty}\eqsp.
 \]
-Therefore, by Hoeffing inequality,
+Therefore, by Hoeffding inequality,
 \[
@@
 \[
-\mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right] = \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\Upsilon_k\right] \eqsp,%\left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}
+\mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right] = \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\Upsilon_k\right] \eqsp,
 \]
@@
 \[
-\Upsilon_k(x_k) = \int q_{\theta}^{\Delta t_{k+1}}(\cdot,x)g_{k+1}(x)\left(\tau_k(x_k) + h_{k+1}(x_k,x) - \phi_{k+1}\left[T_{k+1}h_{k+1}\right]\right)\rmd x\eqsp.
+\Upsilon_k(x_k) = \int q_{k}(\cdot,x)g_{k+1}(x)\left(\tau_k(x_k) + h_{k+1}(x_k,x) - \phi_{k+1}\left[T_{k+1}[h_{k+1}]\right]\right)\rmd x\eqsp.
 \]
-By \cite[Lemma~11]{olsson:westerborn:2016}, $\phi_{k}\left[\Upsilon_k\right] = 0$ which implies by Proposition~\ref{prop:filter} that 
+By \cite[Lemma~11]{olsson:westerborn:2016}, $\phi_{k}\left[\Upsilon_k\right] = 0$ which implies by the induction assumption that 
 \[
@@
 Similarly, as $b_N \le |\widehat{\omega}_k|_{\infty}$, by Hoeffding inequality,
-\[
-\mathbb{P}\left(\left|b_N - \mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right) = \mathbb{E}\left[\mathbb{P}\left(\left|b_N - \mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\middle|\mathcal{F}_k^{N}\right)\right]\le 2\exp\left(-c_kN\varepsilon^2\right)\eqsp.
-\] 
+\begin{multline*}
+\mathbb{P}\left(\left|b_N - \mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right) \\
+= \mathbb{E}\left[\mathbb{P}\left(\left|b_N - \mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\middle|\mathcal{F}_k^{N}\right)\right]\le 2\exp\left(-c_kN\varepsilon^2\right)\eqsp.
+\end{multline*}
 Note that
 \[
-\mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right] = \phi^N_{k}\left[\int q_{\theta}^{\Delta t_{k+1}}(\cdot,x)g_{k+1}(x)\rmd x\right]\eqsp.%\left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}
+\mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right] = \phi^N_{k}\left[\int q_{k}(\cdot,x)g_{k+1}(x)\rmd x\right]\eqsp.
 \]
-By Proposition~\ref{prop:filter},
+By  the induction assumption,
 \[
-\mathbb{P}\left(\left|\mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]-\phi_k\left[\int q_{\theta}^{\Delta t_{k+1}}(\cdot,x)g_{k+1}(x)\rmd x\right]\right|\ge \varepsilon\right)\le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
+\mathbb{P}\left(\left|\mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]-\phi_k\left[\int q_{k}(\cdot,x)g_{k+1}(x)\rmd x\right]\right|\ge \varepsilon\right)\le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
 \]
 The proof is completed using Lemma~\ref{lem:hoeffding:ratio}.
-\end{proof}
-
-\begin{proposition}
-\label{prop:filter}
-Assume that H\ref{assum:boundmodel} and H\ref{assum:boundalgo} hold. Then, for all $0\le k\le n$,
-there exist $b_k, c_k>0$ such that for all $N\ge 1$, all $\varepsilon > 0$, and all measurable function $h$ on $\mathsf{X}$,
-\begin{equation*}
-\mathbb{P}\left( \left|\phi_k^N[h] - \phi_k\left[h\right]\right| \geq \varepsilon \right) \leq b_k \mathrm{exp}\left(-c_k N \epsilon^2\right)\eqsp.
-\end{equation*}
-\end{proposition}
-
-\begin{proof}
-\textcolor{red}{On l'ecrit peut-etre... }
-The proof is a straightforward extension of \cite{douc:garivier:moulines:olsson:2011}. 
 \end{proof}
@@
 \label{sec:exp}
-This section investigates the performance of the proposed algorithm. The different tuning parameters are:
-\begin{enumerate}[i)]
-\item the number of particles $N$ and the precision parameter $\widetilde{N}$;
-\item the proposal transition density $p_k$ and the adjustment multiplier $\vartheta_k$ used to propagate and weight particles;
-\item the discrete distribution $\mu$ used to define $\widetilde{q}_{\theta}^{\Delta t_k}$.
-\end{enumerate}
+%This section investigates the performance of the proposed algorithm. The different tuning parameters are:
+%\begin{enumerate}[i)]
+%\item the number of particles $N$ and the precision parameter $\widetilde{N}$;
+%\item the proposal transition density $p_k$ and the adjustment multiplier $\vartheta_k$ used to propagate and weight particles;
+%\item the discrete distribution $\mu$ used to define $\widetilde{q}_{\theta}^{\Delta t_k}$;
+%\item the number of samples to compute $\widetilde{q}_{\theta}^{\Delta t_k}$.
+%\end{enumerate}
+%The performance of the algorithm is investigated on two known models based on diffusion process.
+
+\subsection*{Sine model} 
 The algorithm is compared to the fixed-lag particle smoother of \cite{olsson:strojby:2011}.
-To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated.
+
+%To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated.
 Instead of the Ozaki discretization, the following bridges process are considered:
-\begin{enumerate}[i)]
-\item a bridge with constant drift as proposed in \cite{giesecke:schwenkler:2016}.
-\item a guided bridge process, see \cite{}.
-\end{enumerate}
-\paragraph{Choice of $\mu$}
-A simple choice is to set $\mu$ as a Poisson distribution with mean $\Lambda_{\theta}t$ leading to the first estimator GPE-1:
-\begin{multline*}
-q^{t}_{\mathrm{GPE-1},\theta}(\kappa,\omega,(U_j)_{1\le j\le \kappa},x,y) =  \widetilde{q}^{t}_{\theta}(x,y)\eqsp\mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x)+\ell(\theta)t\right\}\Lambda_{\theta}^{-\kappa}\\
-\times \prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\eqsp.
-\end{multline*}
+%\begin{enumerate}[i)]
+%\item a bridge with constant drift as proposed in \cite{giesecke:schwenkler:2016}.
+%\item a guided bridge process, see \cite{}.
+%\end{enumerate}
+%\paragraph{Choice of $\mu$}
+%A simple choice is to set $\mu$ as a Poisson distribution with mean $\Lambda_{\theta}t$ leading to the first estimator GPE-1:
+%\begin{multline*}
+%q^{t}_{\mathrm{GPE-1},\theta}(\kappa,\omega,(U_j)_{1\le j\le \kappa},x,y) =  \widetilde{q}^{t}_{\theta}(x,y)\eqsp\mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x)+\ell(\theta)t\right\}\Lambda_{\theta}^{-\kappa}\\
+%\times \prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\eqsp.
+%\end{multline*}
+
 % An important feature of this estimator is that under the assumptions of this section it may be bounded uniformly in $x,y$, $\sup_{x,y\in\mathbb{R}^d}  q^{t}_{\mathrm{GPE-1},\theta}(\kappa,\omega,(U_j)_{1\le j\le \kappa},x,y) \le M_\theta$ where:
@@
 % \]
-\paragraph{Choice of bridge process}
-The approximation of $q_{\theta}^{\Delta t_k}$ described in the paper may be compared to the approximation provided by other bridges with affine drift functions.
+%
+%\paragraph{Choice of bridge process}
+%The approximation of $q_{\theta}^{\Delta t_k}$ described in the paper may be compared to the approximation provided by other bridges with affine drift functions.
 Following \cite{}, let $(Y^{\beta}_t)_{t_{k-1}\le t\le t_k}$ be governed by the following stochastic differential equation for $t_{k-1}\le t\le t_k$:
-\[
-Y^{\beta}_{t_{k-1}} = X_{k-1}\quad\mbox{and}\quad\rmd Y^{\beta}_t = \beta\rmd t + \rmd W_t\eqsp.
-\]
-As suggestd by \cite{}, a guided process between $t_{k-1}$ and $t_k$ may alo be used:
-\[
-Y^{\mathsf{guid}}_{t_{k-1}} = X_{k-1}\quad\mbox{and}\quad\rmd Y^{\mathsf{guid}}_t = \frac{X_k - Y^{\mathsf{guid}}_t}{t_k-t}\rmd t + \rmd W_t\eqsp.
-\]
-Both processes lead to estimators of $q_{\theta}^{\Delta t_k}$ following the same steps as in Section~\ref{}.
+%\[
+%Y^{\beta}_{t_{k-1}} = X_{k-1}\quad\mbox{and}\quad\rmd Y^{\beta}_t = \beta\rmd t + \rmd W_t\eqsp.
+%\]
+%As suggestd by \cite{}, a guided process between $t_{k-1}$ and $t_k$ may alo be used:
+%\[
+%Y^{\mathsf{guid}}_{t_{k-1}} = X_{k-1}\quad\mbox{and}\quad\rmd Y^{\mathsf{guid}}_t = \frac{X_k - Y^{\mathsf{guid}}_t}{t_k-t}\rmd t + \rmd W_t\eqsp.
+%\]
+%Both processes lead to estimators of $q_{\theta}^{\Delta t_k}$ following the same steps as in Section~\ref{}.
+\subsection*{Log-growth model}
+Following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} and \cite{olsson:westerborn:2016}, the performance of the proposed algorithm are also illustrated with the log-growth model defined by:
+\begin{equation}
+\rmd Z_t = \kappa Z_t\left(1-\frac{Z_t}{\gamma}\right)\rmd t + \sigma Z_t \rmd W_t,~~Z_0=z_0. \label{eq:LG:SDE}
+\end{equation}
+In order to use the exact algorithms of \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} and the GPE of \cite{fearnhead:papaspiliopoulos:roberts:2008}, we consider  \eqref{eq:LG:SDE} after the Lamperti transform, i.e., the process defined by $X_t=\eta(Z_t)$, with $\eta(z) := -\log (z)/\sigma$,  which satisfies the following SDE:
+\begin{equation}
+\rmd X_t = \overbrace{\left( \frac{\sigma}{2} -  \frac{\kappa}{\sigma} + \frac{\kappa}{\gamma\sigma}\exp\left(-\sigma X_t\right)\right)}^{:=\alpha(X_t)}\rmd t +\rmd W_t,~~X_0=x_0=\eta(z_0).\label{eq:Lamp:LG}
+\end{equation}
+In this case, the conditions of the Exact Algorithm~2 defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} are satisfied, as for any $m \in \mathbb{R}$ there exists $\mathsf{U}(m)$ such that for all $x\ge m$, $\psi(x):=\alpha^2(x)+\alpha'(x) \leq \mathsf{U}_m$.
+Moreover, $\psi$ is lower bounded uniformly by  $\mathsf{L}$.
+Then, GPE estimators may be computed by simulating a the minimum of a OU bridge, and simulating OU bridges conditionally to this minimum.
+However, to the best of our knowledge, no general techniques exist to do so.
+We therefore apply the proposed algorithm, but with a GPE estimator computed with Bessel bridges, as proposed by \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.
+
+The process solution to \eqref{eq:Lamp:LG} is observed regularly at times $t_0=0,\dots,t_{50}=100$ through the observation process $(Y_k)_{0\le k\le 50}$ defined as:
+\begin{equation}
+Y_k = X_k + \varepsilon_k\label{eq:obs:LG}\eqsp,
+\end{equation}
+where the $(\varepsilon_k)_{0\le k \le 50}$ are i.i.d. $ \mathcal{N}(0,\sigma^2_{obs})$.
+The parameters are given by $$\theta =(\kappa=0.1,\sigma=0.1,\gamma=1000,\sigma^2_{obs}=4)\eqsp.$$
+In order to evaluate the performance of our algorithm, we compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50).
+We therefore compute the particle approximation of $\mathcal{Q}(\theta,\theta)$ for a sampled set of observations displayed in Figure~\ref{fig:obs:LG}.
+This estimation is performed 300 times, using $N=400$ particles, with $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.
+Moreover, the E step boils down to the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.
+$\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite{olsson:strojby:2011} (Appendix B) based on 30 independent Monte Carlo simulations.
+
+The intermediate quantity of the EM algorithm is also estimated with our algorithm 30 times using $N=10000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, see Figure \ref{fig:res:LG}.
+The results for the fixed lag technique are similar to the ones presented in \cite{olsson:strojby:2011} (Figure 1).
+For small lags, the variance of the estimates is small, but the estimation is highly biased.
+The bias rapidly decreases as the lag increases, together with a  great variance increase.
+The PaRIS algorithm outperforms the fixed lag smoother as it shows  a similar (vanishing) bias as the fixed lag for the largest lag and a smaller variance than the variance of the fixed lag when the lag is one.
+%It is worth noting here that, as pointed out in \cite{olsson:strojby:2011}, results of the fixed lag technique indicates a quite strong forgetting in the model, and the choice of the lag should take into account this forgetting property.
+%As the PaRIS algorithm is always unbiased, there is no need to consider the forgetting property of a given model in the approximation of the EM intermediate quantity. 
+\begin{figure}[p]
+\centering
+\includegraphics[width=0.8\textwidth]{obs}
+\caption{One realisation of the process $(X_t)_{t\geq 0}$, solution of the SDE \eqref{eq:Lamp:LG} is represented in white.
+The observations of $Y$ at times $t_0=0,\dots,t_{50}=100$, satisfying equation \eqref{eq:obs:LG},  are represented in black.}
+\label{fig:obs:LG}
+\end{figure}
+\begin{figure}[p]
+\centering
+\includegraphics[width=0.8\textwidth]{res_Estep_LG}
+\caption{Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$ using observations of Figure \ref{fig:obs:LG}, using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the PaRIS algorithm for diffusion process proposed in this work, using 300 replicates.
+The whiskers represent the extent of the 95\% central values.
+The dot represents the empirical mean over the 300 replicates.
+The dotted line shows the reference value, computed using the unbiased PaRIS algorithm with $N=10000$ particles.}
+\label{fig:res:LG}
+\end{figure}
 % where $\beta\in\mathbb{R}^d$ is chosen by the user.
@@
 %\end{enumerate}
-The lemma \ref{lem:AR:unbiased} states that $x$ is a realisation of a random variable $X\sim p_X$. Indeed
-\begin{align*}
-\mP\left(Y=\ell\vert U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\vert Y=l \right) \mathbb{P}(Y=\ell)}{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)}\\
-&=\frac{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)}{\sigma_+} \right) \omega_k^\ell}{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)}
-\intertext{We now note that}
-\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\E_{p_\zeta}\left[\E\left[\E_{p_Y}\left[\E\left[\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right) \vert Y\right] \right] \vert \zeta\right]\right]\\
-&=\E_{p_\zeta}\left[\sum_{\ell=1}^N w_k^l\frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)}{\sigma_+} \right]~~(\text{As }\frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\leq 1)\\
-&=\frac{1}{\sigma_+}\sum_{\ell=1}^N w_k^\ell q_k(\xi_k^\ell,\xi_{k+1})
-\intertext{In the same way, we have:}
-\mP\left(U\leq \frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{1}{\sigma_+} q_k(\xi_k^\ell,\xi_{k+1})
-\intertext{Which gives overall the wanted result}
-\mP\left(Y=\ell\vert U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})}{\sum_{l=1}^N w_k^\ell q_k(\xi_k^\ell,\xi_{k+1})}=p_X(\ell),~~\ell\in \{1,\dots,N\}
-\end{align*}
+
+%The lemma \ref{lem:AR:unbiased} states that $x$ is a realisation of a random variable $X\sim p_X$. Indeed
+%\begin{align*}
+%\mP\left(Y=\ell\vert U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\vert Y=l \right) \mathbb{P}(Y=\ell)}{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)}\\
+%&=\frac{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)}{\sigma_+} \right) \omega_k^\ell}{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)}
+%\intertext{We now note that}
+%\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\E_{p_\zeta}\left[\E\left[\E_{p_Y}\left[\E\left[\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right) \vert Y\right] \right] \vert \zeta\right]\right]\\
+%&=\E_{p_\zeta}\left[\sum_{\ell=1}^N w_k^l\frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)}{\sigma_+} \right]~~(\text{As }\frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\leq 1)\\
+%&=\frac{1}{\sigma_+}\sum_{\ell=1}^N w_k^\ell q_k(\xi_k^\ell,\xi_{k+1})
+%\intertext{In the same way, we have:}
+%\mP\left(U\leq \frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{1}{\sigma_+} q_k(\xi_k^\ell,\xi_{k+1})
+%\intertext{Which gives overall the wanted result}
+%\mP\left(Y=\ell\vert U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})}{\sum_{l=1}^N w_k^\ell q_k(\xi_k^\ell,\xi_{k+1})}=p_X(\ell),~~\ell\in \{1,\dots,N\}
+%\end{align*}
 %\end{proof}

### 2017-02-12 ParisEM.tex
Getting rid of Oziris, make place to the GRand PaRIS algorithm, note on complexity, addition of results on SINE model

@@
 \usepackage{geometry}
+\usepackage{hyperref}
 \usepackage{array}
@@
 \newcommand{\mw}{\mathsf{w}}%For bridge realisations
+\newcommand{\U}{\mathsf{U}}
+\newcommand{\Lo}{\mathsf{L}}
 %\newcommand{\qk}{q^{\Delta t_k}_{\theta}}
@@
 However, even the most recent discretization based approximations of the transition densities induce a systematic bias of particle based approximations of posterior distributions, see for instance \cite{delmoral:jacod:protter:2001}.
-To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE) introduced in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.
-In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities.
-
+To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE) introduced in \cite{fearnhead:papaspiliopoulos:roberts:2008}.
+In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities.\\
  The only solution to solve the smoothing problem for partially observed SDE using SMC methods has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}.
@@
 Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$. 
-In the case of partially observed SDE, this forward pass may be performed with the SMC steps combined with the GPE of \cite{fearnhead:papaspiliopoulos:roberts:2008}.
+%In the case of partially observed SDE, this forward pass may be performed with the SMC steps combined with the GPE of \cite{fearnhead:papaspiliopoulos:roberts:2008}. 
 Then, the backward pass of the FFBS algorithm keeps all the particles sampled during the forward pass and computes new weights to take into account the information brought by all the observations from time $n$ to time $0$.
@@
 In this paper, instead of using fixed-lag techniques which introduce a systematic additional bias in the approximation of conditional expectations, we extend the use of  the PaRIS algorithm to partially observed SDE.
-The proposed algorithm allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles.
-The crucial result of the application of PaRIS algorithm to SDE is that the accept reject mechanism ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by bounded unbiased estimates.
-We also propose to improve the algorithm of \cite{fearnhead:papaspiliopoulos:roberts:2008} by extending the range of diffusion processes for which SMC methods can be applied.
+The proposed algorithm allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles in the best cases, and at worse with quadratic complexity.
+The crucial result of the application of PaRIS algorithm to SDE is that the accept reject mechanism ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by positive and bounded unbiased estimates.
+The proposed method results in a generalized random version of the PaRIS algorithm, hereafter named GRand PaRIS algorithm.
+The proposed algorithm apply to general hidden Markov models whose Markovian dynamics is ruled by a stochastic differential equation, belonging to one of the first two domains defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.
+%We also propose to improve the algorithm of \cite{fearnhead:papaspiliopoulos:roberts:2008} by extending the range of diffusion processes for which SMC methods can be applied.
 To do so, we introduce a new class of GPE based on the Ozaki discretization of the target SDE, see \cite{ozaki:1992,shoji:ozaki:1998}.
@@
 Section~\ref{sec:rwparis} describes the random weight PaRIS algorithm to approximate smoothed additive functionals using unbiased estimates of the transition density of the hidden states.
-Then, Section~\ref{sec:PaRIS:SDE} details the application of this algorithm when the transition density may be approximated using a GPE based on the Ozaki discretization or a GPE based on a constant approximation of the drift function.
-The latter is a straightforward generalization of \cite{fearnhead:papaspiliopoulos:roberts:2008}.
+Then, Section~\ref{sec:PaRIS:SDE} details the application of this algorithm when the transition density may be approximated using a GPE. 
 In Section~\ref{sec:convergence}, classical convergence results for SMC smoothers are extended to the setting of this paper and illustrated with numerical experiments in Section~\ref{sec:exp}. 
@@
 
-\section{Random weight PaRIS algorithm}
+\section{The generalized random PaRIS algorithm}
 \label{sec:rwparis}
@@
 See for instance \cite{roberts:tweedie:1996} for conditions under which this diffusion converges exponentially quickly to $\pi$.
-SDE where the drift is given by the gradient of a potential function have been widely used in ecology for instance, see \cite{brillinger:et:al:2011,harris:blackwell:2013,preisler:et:al:2013} and the references therein.
-
-For all $0\le k \le n$, the distribution of $Y_k$ given $X_k$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,Y_k) = g_k(X_k)$. 
+SDE where the drift is given by the gradient of a potential function have been widely used in ecology for instance, see \cite{brillinger:et:al:2011,harris:blackwell:2013,preisler:et:al:2013} and the references therein.\\
+The solution of \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{k=0,\dots,n}$ in $\mathbb{R}^m$. 
+For all $0\le k \le n$, the distribution of $Y_k$ given $X_k:= X_{t_k}$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,Y_k) = g_k(X_k)$. 
 The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
-For all $1\le k \le n$, the distribution of $X_{k+1} $ conditional on $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$. 
-Let $0 \leq k \leq k' \leq n$. The joint smoothing distributions of the hidden states are defined, for all measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, by:
+For all $1\le k \le n$, the distribution of $X_{k+1} $ conditional on $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$.\\ 
+Let $0 \leq k \leq k' \leq n$, the joint smoothing distributions of the hidden states are defined, for all measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, by:
 \[
@@
 \label{def:addfunc}
-H_{n} =  \sum_{k=0}^{n-1}\mathbb{E}\left[h_k(X_{k},X_{k+1})\middle|Y_{0:n}\right]\eqsp,
+\phi_{0:n\vert n}[H_{n}] = \mathbb{E}\left[H_n(X_{0:n})\middle|Y_{0:n}\right] \text{ where } H_n=\sum_{k=0}^{n-1}h_k(X_k,X_{k+1})\eqsp,
 \end{equation}
@@
 The approximation of \eqref{eq:FFbsm:equality} requires first to approximate the sequence of filtering distributions. 
-Sequential Monte Carlo methods provide an efficient and simple solution to obtain these approximations using sets of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$, $0\le k \le n$.
+Sequential Monte Carlo methods provide an efficient and simple solution to obtain these approximations using sets of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$, $0\le k \le n$.\\
 At time $k = 0$, $N$ particles $\{\xi^{\ell}_0\}_{\ell=1}^N$ are sampled independently according to  $\xi^{\ell}_0 \sim \eta_0$, where $\eta_0$ is a probability density with respect to $\mu$. 
@@
 \item $\xi^{\ell}_k$ is associated with:
-\[
+\begin{equation}
+\label{eq:importance:weights}
 \omega^{\ell}_k := \frac{\qk(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_k (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
-\]
+\end{equation}
 \end{enumerate} 
@@
 
-The algorithm introduces sufficient statistics $\tau^i_k$, approximating $T_k[h](\xi^i_k)$, for $1\le i\le N$ and $0\le k \le n$ starting with $\tau^i_0 = 0$ for all $1\le i\le N$.
+The PaRIS algorithm relies on a fundamental property of $T_k[H_k]$ when $H_k$ for additive functionals.
+Indeed, if $H_k$ satisfies the additive property as in \eqref{def:addfunc}, the following equality holds:
+\begin{equation}
+T_k[H_k](X_k)=\mathbb{E}\left[T_{k-1}[H_{k-1}](X_{k-1}) + h_{k-1}(X_{k-1},X_k) \right],~1\leq k\leq n.
+\end{equation}
+Therefore, \cite{olsson:westerborn:2016} introduces sufficient statistics $\tau^i_k$, approximating $T_k[h](\xi^i_k)$, for $1\le i\le N$ and $0\le k \le n$ starting with $\tau^i_0 = 0$ for all $1\le i\le N$.
 Choosing $\tilde{N}\ge 1$, at each time step $0\le k \le {n-1}$ these statistics are updated according to the following steps.
@@
 \[
-H_n^N = \frac{1}{\Omega_n^N}\sum_{i=1}^N \omega^{i}_n \tau_n^i\eqsp.
+\phi_{0:n\vert n}^N = \frac{1}{\Omega_n^N}\sum_{i=1}^N \omega^{i}_n \tau_n^i\eqsp.
 \] 
@@
 In general situations, the PaRIS algorithm cannot be used for stochastic differential equations as $\qk$ is unknown.
-Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio of \cite{douc:garivier:moulines:olsson:2011} is not tractable.
-Following \cite{fearnhead:papaspiliopoulos:roberts:2008,olsson:strojby:2011},
-an extension of the PaRIS algorithm may be developed by replacing $\qk(\xi^{\ell}_{k},\xi_{k+1}^{i})$  by an  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable such that:
-\[
-\mathbb{E}\left[\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta)\middle| \mathcal{F}_{k+1}^N\right] = \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})\eqsp,
-\]
+Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio of \cite{douc:garivier:moulines:olsson:2011} is not tractable.\\
+Following \cite{fearnhead:papaspiliopoulos:roberts:2008,olsson:strojby:2011}, filtering weights can be approximated by replacing $\qk(\xi^{\ell}_{k},\xi_{k+1}^{i})$  by an  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable such that:
+\begin{align*}
+\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)&> 0~~\text{a.s}\eqsp,\\
+\mathbb{E}\left[\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)\middle| \mathcal{F}_{k+1}^N\right] &= \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})\eqsp,
+\end{align*}
 where, for all $0\le k \le n$, 
@@
 \]
-In the case where $\widehat{q}_k$ may be upper bounded by a constant $\hat{\sigma}_+$, step~\eqref{it:PaRIS:sampleindex} is replaced by the accept reject mechanism given in Algorithm~\ref{alg:AR:unknownq}.
-Lemma~\ref{lem:AR:unbiased} states that replacing  $\qk$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.
+In the case where $\qk$ is unknown, the filtering weights in \eqref{eq:importance:weights} then become:
+\begin{equation}
+\label{eq:random:weight}
+\widehat{\omega}^{\ell}_k := \frac{\widehat{q}_k(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_k (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
+\end{equation}
+
+Therefore, to obtain a generalized random version of PaRIS, one need to be able to sample from weight $\Lambda_k^N(i,\ell)$ in the case of SDE based HMM.\\
+If $\widehat{q}_k$ may be upper bounded almost surely by a constant $\hat{\sigma}_+$, this task is performed using the accept reject mechanism given in Algorithm~\ref{alg:AR:unknownq}.
+Lemma~\ref{lem:AR:unbiased} states that replacing  $\qk$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.\\
 \begin{lemma}
@@
 \end{proof}
+Therefore, having an unbiased positive and bounded estimator is sufficient to perform the Generalized Random PaRIS (GRand PaRIS) algorithm, fully described in Algorithm \ref{alg:Ozaki:PaRIS}.\\
+One can notice that in Algorithm \ref{alg:AR:unknownq}, it is actually sufficient to have a constant $\hat{\sigma}_+^k$ such that, for all $j$, the random variable $\hat{q}_k(\xi_k^j,y,\zeta_k)< \hat{\sigma}_+^k$ almost surely.
+This improvement keeps the linear complexity property of PaRIS algorithm and Algorithm \ref{alg:Ozaki:PaRIS}, and allows to increase the acceptance rate of Algorithm \ref{alg:AR:unknownq}.
+Moreover, the AR algorithm still stands if one can define
+\begin{equation}
+\hat{\sigma}_{+}^{k,i}= \underset{j=1,\dots,N}{\text{sup}}\rho_\Delta(\xi_{k-1}^j,\xi_k^i)\eqsp, \label{eq:quadr:bound}
+\end{equation}
+which is a weaker assumption.
+However, in this case, the resulting algorithm has a quadratic complexity in $N$.
+Nevertheless, in practice, this choice can result in a much more efficient acceptance rate, which would reduce the computational time of the algorithm.
 \begin{algorithm}[H]
@@
 Sample independently $
-\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\omega_{k}^1,\dots,\omega_{k}^N\}$.
+\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
 \IF{ $$U \leq \frac{\widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)}{\hat{\sigma}_+},$$}
@@
 \end{algorithm}
+\subsection*{Bounded estimator of $q_k$}
+For $x, y \in \mathbb{R}^d$, using Girsanov adn Itô's formulas, one can show that the transition density $q_\Delta(x,y):=q(x,y,\Delta)$ related to \eqref{eq:target:sde} satisfies:
+\begin{align*}
+q_\Delta(x,y)=\varphi_\Delta(x,y)\exp\left\lbrace A(y)-A(x)\right\rbrace \mathbb{E}_{\mathbb{W}^{x,y,\Delta}}\left[ \exp \left\lbrace - \int_0^\Delta \phi(\mw_s)\rmd s \right\rbrace \right]
+\end{align*}
+where $\mathbb{W}^{x,y,\Delta}$ is the law of Brownian bridge starting at $x$ at 0 and hitting $y$ at $\Delta$, $(\mw_t)_{0\leq t \leq \Delta}$ is such a Brownian bridge, $\varphi_\Delta(x,y)$ is the p.d.f.
+of a normal distribution with mean $x$ and variance $\Delta$, evaluated in $y$ and $phi$ is a function from $\mathbb{R}^d$ to $\mathbb{R}$ such that
+\[
+\phi(x) =\parallel \alpha(x)^2 \parallel + \triangle A(x)\eqsp.
+\]
+Suppose there exist random variables $\Lo_\mw$ and $\U_\mw$ such that for all $0\leq s \leq \Delta$, $\Lo_\mw \leq \phi(\mw_s)\leq \U_\mw$.
+Let $\kappa$ be a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and $(U_j)_{1\le j\le \kappa}$ be independent uniform random variables on $[0,\Delta]$, and $
+\zeta_k = \left\{\kappa,\mw,U_1,\ldots,U_\kappa\right\}\eqsp$. 
+As shown in \cite{fearnhead:papaspiliopoulos:roberts:2008}, a positive unbiased estimator is given by 
+\begin{multline}
+\widehat{q}_\Delta(x,y;\zeta_k) = \varphi_\Delta(x,y) \exp \left\{A(y) - A(x)\right\}\\ 
+\times\mathrm{exp}\left\{-\U_\mw\Delta\right\}\frac{\Delta^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\U_\mw-\phi(\mw_{U_j})\right)\eqsp.\label{eq:unbiased:q}
+\end{multline}
+Interesting choices of $\mu$ are discussed in \cite{fearnhead:papaspiliopoulos:roberts:2008}n and we focus here on the so called GPE-1, when $\mu$ is the distribution of a Poisson law with intensity $(\U_\mw-\Lo_\mw)\Delta$.
+In that case, the estimator \eqref{eq:unbiased:q} becomes:
+\begin{equation}
+\widehat{q}_\Delta(x,y;\zeta_k) = \varphi_\Delta(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta \right\}\prod_{j=1}^{\kappa}\frac{\U_\mw-\phi(\mw_{U_j})}{\U_\mw-\Lo_\mw}\eqsp.\label{eq:GPE1}
+\end{equation}
+On the r.h.s. of \eqref{eq:GPE1}, the product over $\kappa$ elements is bounded by 1, therefore, a sufficient condition to obtain a bound over $\hat{q}_\Delta$ is that the function:
+\begin{align*}
+\rho_\Delta:\eqsp\mathbb{R}^d\times \mathbb{R}^d &\mapsto \mathbb{R}\\
+(x,y)&\mapsto \varphi_\Delta(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta \right\}
+\end{align*}
+is upper bounded almost surely by $\hat{\sigma}_+$.
+As said above, the GRand PaRIS algorithm only requires a bound $\hat{\sigma}_+^k$ for the function $\rho_\Delta(x,y)$ where $x$ is restricted to its particle support (i.e., at step $k+1$, $x \in \xi_{k}^1,\dots,\xi_{1}^N$.).\\
+In order to obtain a better acceptance rate (at a cost of a quadratic complexity in $N$), one can always compute the bound defined by \eqref{eq:quadr:bound} as long as $\Lo_\mw$ is bounded almost surely.
+Actually, this condition is always satisfied for models being in the domains $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, i.e.
+domains for which the exact algorithms 1 and 2 can be performed.
+\subsection*{Note on the complexity and practical implementation}
 
-\section{The OzIRis algorithm}
-\label{sec:PaRIS:SDE}
-In \cite{olsson:strojby:2011}, $\qk$ is estimated using GPE based on a representation of $\qk$ as an expectation under the probability measure of a Brownian bridge. 
-In order to improve the Monte Carlo estimation of $\qk$,  other diffusion bridges may be used as long as they can be simulated efficiently.
+As said above, as long as one can define a $\sigma_+^k$ such that
+$$ \forall\eqsp 0\leq k\leq n-1,\eqsp 1\leq i\leq N,\eqsp y\in\mathbb{R}^d,\eqsp \hat{q}_k(\xi_k^i,y,\zeta_k)<\hat{\sigma}_+^k \text{ a.s.,}$$ then the resulting GRand Paris algorithm has a linear complexity in $N$.\\
+However, it is worth noting here that for a reasonable amount of particles, it might be more efficient in terms of computation time to compute, for all $i$, the bound given by \eqref{eq:quadr:bound}.
+Indeed, this bound requires, at each time step $k$, $N^2$ computation of the function $\rho$ given above.
+However, this might increase significantly the acceptance rate of the AR algorithm, and therefore reduce the number of drawings for the random variable $\zeta_k$, which has a much higher cost than the computation of $\rho$, as it requires simulation of Brownian Bridges.
+It is worth noting here that this latter option can also avoid numerical optimization if no analytical version of $\hat{\sigma}_+^k$ is available.\\
+In the example of the SINE model presented below, for $N=400$ particles, the linear complexity algorithm has a 40\% higher computation time than the algorithm with a quadratic complexity.
+In practice, in addition to $n$ and $N$, the computation time will increase with
+\begin{itemize}
+\item $\tilde{N}$ the number of retrospective sampling;
+\item $M$ the number of Monte Carlo samples to obtain $\hat{q}$;
+\item $\Delta$ the acquisition time step;
+\item The expected value of $\U_\mw-\Lo_\mw$.
+\end{itemize}
+
+%\section{The OzIRis algorithm}
+%\label{sec:PaRIS:SDE}
+%In \cite{olsson:strojby:2011}, $\qk$ is estimated using GPE based on a representation of $\qk$ as an expectation under the probability measure of a Brownian bridge. 
+%In order to improve the Monte Carlo estimation of $\qk$,  other diffusion bridges may be used as long as they can be simulated efficiently.
 A simple idea is to consider a diffusion obtained as an approximation of the target SDE \eqref{eq:target:sde}.
 In \cite{giesecke:schwenkler:2016}, the authors proposed to replace the Brownian bridge by a Brownian motion with constant drift between two consecutive observations.
-As an alternative, it is also possible to follow \cite{ozaki:1992,shoji:ozaki:1998} to propose a higher order approximation of the drift term $\alpha$.
+%As an alternative, it is also possible to follow \cite{ozaki:1992,shoji:ozaki:1998} to propose a higher order approximation of the drift term $\alpha$.
 The authors used a linear approximation of the diffusion drift $\alpha$, together with a constant approximation of the volatility over each time interval $(t_{k},t_{k+1})$.
-
-For all $0\le k\le n-1$, let $(Z^{k}_t)_{t\ge 0}$ be governed by the following stochastic differential equation for $t\in(t_{k},t_{k+1})$:
-\begin{equation}
-\label{eq:SDE:Ozaki}
-\Oz^k_{t_{k}} = X_{k}\quad\mbox{and}\quad\rmd \Oz^k_t = \widetilde{\alpha}_k(\Oz^k_t)\rmd t + \rmd W_t\eqsp,
-\end{equation}
-where $\widetilde{\alpha}_k$ is an instrumental drift. The probability density function of the transition kernel associated with $\Oz^k$ is written $\widetilde{q}_k$.
-Writing $\mathbb{Q}_k$ (resp.
+%
+%For all $0\le k\le n-1$, let $(Z^{k}_t)_{t\ge 0}$ be governed by the following stochastic differential equation for $t\in(t_{k},t_{k+1})$:
+%\begin{equation}
+%\label{eq:SDE:Ozaki}
+%\Oz^k_{t_{k}} = X_{k}\quad\mbox{and}\quad\rmd \Oz^k_t = \widetilde{\alpha}_k(\Oz^k_t)\rmd t + \rmd W_t\eqsp,
+%\end{equation}
+%where $\widetilde{\alpha}_k$ is an instrumental drift. The probability density function of the transition kernel associated with $\Oz^k$ is written $\widetilde{q}_k$.
+%Writing $\mathbb{Q}_k$ (resp.
 $\mathbb{W}_{k}$) the probability measure induced by $(X_s)_{t_{k}\le s\le t_{k+1}}$  (resp.
 $(\Oz_s)_{t_{k}\le s\le t_{k+1}}$) on $(\mathsf{C},\mathcal{C})$, Girsanov theorem yields:
-\begin{multline*}
-\frac{\rmd \mathbb{Q}_k}{\rmd \mathbb{W}_{k}}(w) = \mathrm{exp}\left\{H_{k}(\mw_{t_{k+1}}) - H_{k}(\mw_{t_{k}})\right\}\\
-\times \exp\left\{- \frac{1}{2}\int_{t_{k}}^{t_{k+1}}\left\{\| \alpha_{}(\mw_s)\|^2-\|\widetilde{\alpha}_k(\mw_s)\|^2+\Delta H_k(\mw_s)\right\}\rmd s\right\}\eqsp,
-\end{multline*}
-where $\nabla H_k = \alpha -\widetilde{\alpha}_k$ and $\Delta$ is the Laplace operator. Then, the probability measures conditioned on hitting $X_{k+1}$ at time $t_{k+1}$ satisfy:
-\begin{multline}
-\label{eq:Girsanov}
-\frac{\rmd \mathbb{Q}_k^{X_{k+1}}}{\rmd \mathbb{W}_k^{X_{k+1}}}(\mw) = \frac{\widetilde{\qk}(X_{k},X_{k+1})}{\qk (X_{k},X_{k+1}) }\mathrm{exp}\left\{H_{k}(X_{k+1}) - H_{k}(X_{k})\right\} \\
-\times \mathrm{exp}\left\{- \frac{1}{2}\int_{t_{k}}^{t_{k+1}}\left\{\|\alpha_{}(\mw_s)\|^2-\|\widetilde{\alpha}_{k}(\mw_s)\|^2+\Delta H_{k}(\mw_s)\right\}\rmd s\right\}\eqsp.
-\end{multline}
-\subsection{Linear approximation of the drift}
-In order to improve the performance of the GPE introduced in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,giesecke:schwenkler:2016}, the instrumental drift is chosen as the Ozaki  \cite{ozaki:1992,shoji:ozaki:1998} linear approximation of the target drift:
-\begin{equation}
-\label{eq:alpha:linear}
-\widetilde{\alpha}_k: x\mapsto \Jk x + \alpha(X_{k}) - \Jk X_{k}\eqsp,
-\end{equation}
-where $\Jk$ is the Jacobian of $\alpha$ evaluated at $X_k$.
+%\begin{multline*}
+%\frac{\rmd \mathbb{Q}_k}{\rmd \mathbb{W}_{k}}(w) = \mathrm{exp}\left\{H_{k}(\mw_{t_{k+1}}) - H_{k}(\mw_{t_{k}})\right\}\\
+%\times \exp\left\{- \frac{1}{2}\int_{t_{k}}^{t_{k+1}}\left\{\| \alpha_{}(\mw_s)\|^2-\|\widetilde{\alpha}_k(\mw_s)\|^2+\Delta H_k(\mw_s)\right\}\rmd s\right\}\eqsp,
+%\end{multline*}
+%where $\nabla H_k = \alpha -\widetilde{\alpha}_k$ and $\Delta$ is the Laplace operator. Then, the probability measures conditioned on hitting $X_{k+1}$ at time $t_{k+1}$ satisfy:
+%\begin{multline}
+%\label{eq:Girsanov}
+%\frac{\rmd \mathbb{Q}_k^{X_{k+1}}}{\rmd \mathbb{W}_k^{X_{k+1}}}(\mw) = \frac{\widetilde{\qk}(X_{k},X_{k+1})}{\qk (X_{k},X_{k+1}) }\mathrm{exp}\left\{H_{k}(X_{k+1}) - H_{k}(X_{k})\right\} \\
+%\times \mathrm{exp}\left\{- \frac{1}{2}\int_{t_{k}}^{t_{k+1}}\left\{\|\alpha_{}(\mw_s)\|^2-\|\widetilde{\alpha}_{k}(\mw_s)\|^2+\Delta H_{k}(\mw_s)\right\}\rmd s\right\}\eqsp.
+%\end{multline}
+%\subsection{Linear approximation of the drift}
+%In order to improve the performance of the GPE introduced in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,giesecke:schwenkler:2016}, the instrumental drift is chosen as the Ozaki  \cite{ozaki:1992,shoji:ozaki:1998} linear approximation of the target drift:
+%\begin{equation}
+%\label{eq:alpha:linear}
+%\widetilde{\alpha}_k: x\mapsto \Jk x + \alpha(X_{k}) - \Jk X_{k}\eqsp,
+%\end{equation}
+%where $\Jk$ is the Jacobian of $\alpha$ evaluated at $X_k$.
 In this case, the transition kernel associated with $\Oz^k$ is given in Section~\ref{sec:ornstein:bridge} and the function $H_k$ may be chosen as
-\[
-H_{k}: x \mapsto A_{}(x) - \frac{1}{2}x'\Jk x- (\alpha(X_{k}) - \Jk X_{k-1})'x\eqsp.
-\]
-%Consider the following assumptions on the model:
-%\begin{enumerate}[(a)]
-%\item For all $0\le k \le n-1$, $x\mapsto \alpha_{}(x) - \alpha_{}(X_k) - \Jk(x-X_k)$ is bounded.
-%\item For all $0\le k \le n-1$, $x\mapsto \mathrm{Tr}\left(J_{\alpha_{}}(x) - \Jk\right)$ is bounded, where $\mathrm{Tr}$ is the trace operator.
-%\end{enumerate}
-Assume that there exists $\ell_k$ such that:
-\[
-\mathrm{inf}_{u\in \mathbb{R}^d} \left\{\| \alpha_{}(u)\|^2-\|\widetilde{\alpha}_{k}(u)\|^2+\Delta H_{k}(u)\right\}/2\ge \ell_k
-\]
-and
-\[
-\phi_{k}: x\mapsto \left\{\|\alpha_{}(x)\|^2-\|\widetilde{\alpha}_{k}(x)\|^2+\Delta H_{k}(x)\right\}/2-\ell_k\eqsp.
-\]
-Assume also that the model is such that there exists $\Lambda_k$ satisfying  $\mathrm{sup}_x\eqsp \phi_{k}(x)\le \Lambda_{k}$.
+%\[
+%H_{k}: x \mapsto A_{}(x) - \frac{1}{2}x'\Jk x- (\alpha(X_{k}) - \Jk X_{k-1})'x\eqsp.
+%\]
+%%Consider the following assumptions on the model:
+%%\begin{enumerate}[(a)]
+%%\item For all $0\le k \le n-1$, $x\mapsto \alpha_{}(x) - \alpha_{}(X_k) - \Jk(x-X_k)$ is bounded.
+%%\item For all $0\le k \le n-1$, $x\mapsto \mathrm{Tr}\left(J_{\alpha_{}}(x) - \Jk\right)$ is bounded, where $\mathrm{Tr}$ is the trace operator.
+%%\end{enumerate}
+%Assume that there exists $\ell_k$ such that:
+%\[
+%\mathrm{inf}_{u\in \mathbb{R}^d} \left\{\| \alpha_{}(u)\|^2-\|\widetilde{\alpha}_{k}(u)\|^2+\Delta H_{k}(u)\right\}/2\ge \ell_k
+%\]
+%and
+%\[
+%\phi_{k}: x\mapsto \left\{\|\alpha_{}(x)\|^2-\|\widetilde{\alpha}_{k}(x)\|^2+\Delta H_{k}(x)\right\}/2-\ell_k\eqsp.
+%\]
+%Assume also that the model is such that there exists $\Lambda_k$ satisfying  $\mathrm{sup}_x\eqsp \phi_{k}(x)\le \Lambda_{k}$.
 These assumptions might seem somehow restrictive but are necessary to define the transition density estimator in the linear case.
@@
 Then,
-\begin{multline*}
-\qk(X_{k},X_{k+1}) = \widetilde{\qk}(X_{k},X_{k+1}) \eqsp\mathrm{exp}\left\{H_{k}(X_{k+1}) - H_{k}(X_{k})-\ell\Delta_k\right\}\\
-\times\mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{ \Delta_k}\phi_{k}(\mw_s)\rmd s\right\}\right]\eqsp,
-\end{multline*}
-where $\mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}$ is the expectation under the law of the bridge process defined by \eqref{eq:SDE:Ozaki} and \eqref{eq:alpha:linear}, starting at $X_{k}$ at time $0$ and ending at $X_{k+1}$ at time $\Delta_k$.
-An unbiased estimator of $\qk(X_{k},X_{k+1})$ is then obtained by estimating the expectation in the previous equation. 
-Following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, if $\omega \sim\mathbb{W}_{k}^{X_{k}}$, $\kappa$ is a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and if $(U_j)_{1\le j\le \kappa}$ are independent uniform random variables on $[0,t]$,
-\begin{align*}
-\mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{\Delta_k}\phi_{k}(\mw_s)\rmd s\right\}\right] &\\
-&\hspace{-3cm}= \mathrm{exp}\left\{-\Lambda_{k} \Delta_k\right\} \mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}} \left[\mathrm{exp}\left\{ \int_{0}^{\Delta_k}(\Lambda_{k}-\phi_{k}(\mw_s))\rmd s\right\}\right]\eqsp,\\
-&\hspace{-3cm}=\mathrm{exp}\left\{-\Lambda_{k}\Delta_k\right\} \mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}\left[\sum_{p\ge 0}\frac{1}{p!}\left(\int_{0}^{\Delta_k}\Lambda_{k}-\phi_{k}(\mw_s)\rmd s\right)^p\right]\eqsp,\\
-&\hspace{-3cm}=\mathrm{exp}\left\{-\Lambda_{k}\Delta_k \right\} \mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}} \left[\mathbb{E}\left[\sum_{p\ge 0}\frac{( \Delta_k)^p}{k!} \prod_{j=1}^p\left(\Lambda_{k}-\phi_{k}(\mw_{U_j})\right)\right]\right]\eqsp,\\
-&\hspace{-3cm}=\mathrm{exp}\left\{-\Lambda_{k} \Delta_k\right\} \mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}\left[ \frac{(\Delta_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\Lambda_{k}-\phi_{k}(\mw_{U_j})\right)\right]\eqsp.
-\end{align*}
-Interesting choices of $\mu$ are discussed in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} and implemented in Section~\ref{sec:exp}. Writing
-\[
-\zeta_k = \left\{\kappa,\mw,U_1,\ldots,U_\kappa\right\}\eqsp,
-\]
-Algorithm~\ref{alg:AR:unknownq} is used with 
-\begin{multline}
-\widehat{\qk}(X_{k},X_{k+1};\zeta_k) = \widetilde{\qk}(X_{k},X_{k+1}) \eqsp\mathrm{exp}\left\{H_{k}(X_{k+1}) - H_{k}(X_{k})-\ell_k \Delta_k\right\}\\ 
-\times\mathrm{exp}\left\{-\Lambda_{k}\Delta_k\right\}\frac{( \Delta_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\Lambda_{k}-\phi_{k}(\mw_{U_j})\right)\eqsp.\label{eq:unbiased:q}
-\end{multline}
-The choice of $\mu$ is restricted to distributions such that the associated estimator is upper bounded by an explicit constant $\hat{\sigma}_+^k$.
-Then, for all $0\le k \le n-1$, $1\le i\le N$, the importance weight $\omega_k^{\ell}$ is replaced by the unbiased estimator:
-\begin{equation}
-\label{eq:random:weight}
-\widehat{\omega}_{k+1}^{i} = \frac{M^{-1}\sum_{m=1}^M \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta_k^m)g_{k+1}(\xi^{i}_{k+1})}{\vartheta_{k+1}(\xi^{I^{i}_{k+1}}_{k}) p_{k} (\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})}\eqsp,
-\end{equation}
-where $\{\zeta_k^i\}_{i=1}^M$ are independent copies of $\zeta_k$.
+%\begin{multline*}
+%\qk(X_{k},X_{k+1}) = \widetilde{\qk}(X_{k},X_{k+1}) \eqsp\mathrm{exp}\left\{H_{k}(X_{k+1}) - H_{k}(X_{k})-\ell\Delta_k\right\}\\
+%\times\mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{ \Delta_k}\phi_{k}(\mw_s)\rmd s\right\}\right]\eqsp,
+%\end{multline*}
+%where $\mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}$ is the expectation under the law of the bridge process defined by \eqref{eq:SDE:Ozaki} and \eqref{eq:alpha:linear}, starting at $X_{k}$ at time $0$ and ending at $X_{k+1}$ at time $\Delta_k$.
+%An unbiased estimator of $\qk(X_{k},X_{k+1})$ is then obtained by estimating the expectation in the previous equation. 
+%Following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, if $\omega \sim\mathbb{W}_{k}^{X_{k}}$, $\kappa$ is a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and if $(U_j)_{1\le j\le \kappa}$ are independent uniform random variables on $[0,t]$,
+%\begin{align*}
+
+%Then, for all $0\le k \le n-1$, $1\le i\le N$, the importance weight $\omega_k^{\ell}$ is replaced by the unbiased estimator:
+%\begin{equation}
+%\label{eq:random:weight}
+%\widehat{\omega}_{k+1}^{i} = \frac{M^{-1}\sum_{m=1}^M \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta_k^m)g_{k+1}(\xi^{i}_{k+1})}{\vartheta_{k+1}(\xi^{I^{i}_{k+1}}_{k}) p_{k} (\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})}\eqsp,
+%\end{equation}
+%where $\{\zeta_k^i\}_{i=1}^M$ are independent copies of $\zeta_k$.
 The resulting Ozaki in PaRIS (OzIRis) algortihm, which provides an approximation of \eqref{def:addfunc}, is summarized in Algorithm~\ref{alg:Ozaki:PaRIS}.
 \begin{algorithm}
-\caption{OzIRis algorithm}
+\caption{GRand PaRIS algorithm}
 \begin{algorithmic}
 \FORALL{$i \in 1,\dots, N$}
-\STATE Set $\xi_0^i = X_0$, $\omega_0^i = 1/N$ and $\tau_0^i = 0$.
+\STATE Set $\xi_0^i = X_0$, $\tau_0^i = 0$  and  $\omega_0^i = g_0(\xi_0^i)\chi_0(\xi_0^i)/\nu_0(\xi_0^i)$.
 \ENDFOR
@@
 
-\subsection{Constant approximation of the drift}
-\label{sec:constantdrift}
-In the case where the function $\phi_k$ cannot be upper bounded when $\widetilde{\alpha}_k$ is set as a linear approximation of $\alpha$, \cite{giesecke:schwenkler:2016} suggests to set $\widetilde{\alpha}_k$ as a constant value $\widetilde{\alpha}_k= \rho_k$.
+%\subsection{Constant approximation of the drift}
+%\label{sec:constantdrift}
+%In the case where the function $\phi_k$ cannot be upper bounded when $\widetilde{\alpha}_k$ is set as a linear approximation of $\alpha$, \cite{giesecke:schwenkler:2016} suggests to set $\widetilde{\alpha}_k$ as a constant value $\widetilde{\alpha}_k= \rho_k$.
 The GPE of \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} is a special case with $\rho_k = 0$.
@@
 Then, following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,olsson:strojby:2011}, it is enough to sample a pair $(L_k,U_k)$ of random variables and a Brownian bridge $\mw\sim \mathbb{W}_k^{X_{k+1}}$ (or a Brownian bridge with a constant drift $\rho_k$) such that, for all $t_k\le t\le t_{k+1}$,
-\[
-L_k\le \phi_k(\mw_s)\le U_k\eqsp.
-\]
-Sampling the bridge conditionally on its minimum and maximum is possible using Bessel bridges as noted by \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.
+%\[
+%L_k\le \phi_k(\mw_s)\le U_k\eqsp.
+%\]
+%Sampling the bridge conditionally on its minimum and maximum is possible using Bessel bridges as noted by \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.
 Then, the second term of \eqref{eq:unbiased:q} is replaced by:
-\[
-\mathrm{exp}\left\{-U_{k}\Delta_k\right\}\frac{( \Delta_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(U_{k}-\phi_{k}(\mw_{U_j})\right)\eqsp.
-\]
-Here again, the choice of $\mu$ is restricted to distributions such that the associated estimator is upper bounded by an explicit constant $\hat{\sigma}_+^k$.
+%\[
+%\mathrm{exp}\left\{-U_{k}\Delta_k\right\}\frac{( \Delta_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(U_{k}-\phi_{k}(\mw_{U_j})\right)\eqsp.
+%\]
+%Here again, the choice of $\mu$ is restricted to distributions such that the associated estimator is upper bounded by an explicit constant $\hat{\sigma}_+^k$.
 
@@
 
-\subsection*{Sine model} 
-The algorithm is compared to the fixed-lag particle smoother of \cite{olsson:strojby:2011}.
-
+\subsection*{The SINE model} 
+The performances of the GRand PaRIS algorithm is first presented on the SINE model, where $(X_t)_{t\geq 0}$ is supposed to be the solution of the SDE: \cite{olsson:strojby:2011}. 
+\begin{equation}
+\rmd X_t = \sin \left(X_t-\theta\right)\rmd t + \rmd W_t,~~X_0=X_0. \label{eq:SINE:SDE}
+\end{equation}
+This simple model has no explicit transition density $q$, however, GPE estimators may be computed by simulating Brownian bridges.
+The process solution to \eqref{eq:SINE:SDE} is observed regularly at times $t_0=0,t_200=200$ through the observation process $(Y_k)_{0\leq k \leq 200}$
+\begin{equation}
+Y_k = X_k + \varepsilon_k\label{eq:obs:SINE},~~\varepsilon\sim \mathcal{N}(0,1)\eqsp,
+\end{equation}
+In the example showed on Figure \ref{fig:obs:SINE}, we set $\theta=0$.\\
+\\
+In order to evaluate the performance of our algorithm, we compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50).
+We therefore compute the particle approximation of $\mathcal{Q}(\theta,\theta)$ for a sampled set of observations displayed in Figure~\ref{fig:obs:SINE}.
+This estimation is performed 300 times, using $N=400$ particles, with $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.
+Moreover, the E step boils down to the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.
+$\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite{olsson:strojby:2011} (Appendix B) based on 30 independent Monte Carlo simulations.
+The intermediate quantity of the EM algorithm is also estimated with our algorithm 30 times using $N=10000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, see Figure \ref{fig:res:SINE}.\\
+One can see that the GRand PaRIS algorithm outperforms the fixed lag methods (for any lag), as the empirical variance of the estimates is the smallest of all, and the bias is already neglictible for $N=400$, whereas small lags lead to strongly biased estimates for the Fixed lag methods, and unbiased estimates are at the cost of a large variance.
 %To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated.
@@
 Moreover, $\psi$ is lower bounded uniformly by  $\mathsf{L}$.
-Then, GPE estimators may be computed by simulating a the minimum of a OU bridge, and simulating OU bridges conditionally to this minimum.
-However, to the best of our knowledge, no general techniques exist to do so.
-We therefore apply the proposed algorithm, but with a GPE estimator computed with Bessel bridges, as proposed by \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.
+Then, GPE estimators may be computed by simulating the minimum of a Brownian bridge, and simulating Bessel bridges conditionally to this minimum, as proposed by \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.
 
@@
 The parameters are given by $$\theta =(\kappa=0.1,\sigma=0.1,\gamma=1000,\sigma^2_{obs}=4)\eqsp.$$
-In order to evaluate the performance of our algorithm, we compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50).
-We therefore compute the particle approximation of $\mathcal{Q}(\theta,\theta)$ for a sampled set of observations displayed in Figure~\ref{fig:obs:LG}.
-This estimation is performed 300 times, using $N=400$ particles, with $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.
-Moreover, the E step boils down to the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.
-$\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite{olsson:strojby:2011} (Appendix B) based on 30 independent Monte Carlo simulations.
-
-The intermediate quantity of the EM algorithm is also estimated with our algorithm 30 times using $N=10000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, see Figure \ref{fig:res:LG}.
-The results for the fixed lag technique are similar to the ones presented in \cite{olsson:strojby:2011} (Figure 1).
+Again, we evaluate the intermediate quantity of the EM algorithm using the same amount of samples as for the SINE model.\\
+The results for the fixed lag technique are similar to the ones presented in \cite{olsson:strojby:2011} on the same model (Figure 1 of their article).
 For small lags, the variance of the estimates is small, but the estimation is highly biased.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{obs}
+\includegraphics[width=0.8\textwidth]{obs_SINE}
+\caption{One realisation of the process $(X_t)_{t\geq 0}$, solution of the SDE \eqref{eq:Lamp:SINE} is represented in white.
+The observations of $Y$ at times $t_0=0,\dots,t_{200}=200$, satisfying equation \eqref{eq:obs:SINE},  are represented in black.}
+\label{fig:obs:SINE}
+\end{figure}
+\begin{figure}[p]
+\centering
+\includegraphics[width=0.8\textwidth]{obs_LG}
 \caption{One realisation of the process $(X_t)_{t\geq 0}$, solution of the SDE \eqref{eq:Lamp:LG} is represented in white.
-The observations of $Y$ at times $t_0=0,\dots,t_{50}=100$, satisfying equation \eqref{eq:obs:LG},  are represented in black.}
+The observations of $Y$ at times $t_0=0,\dots,t_{200}=100$, satisfying equation \eqref{eq:obs:LG},  are represented in black.}
 \label{fig:obs:LG}
@@
 \centering
-\includegraphics[width=0.8\textwidth]{res_Estep_LG}
-\caption{Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$ using observations of Figure \ref{fig:obs:LG}, using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the PaRIS algorithm for diffusion process proposed in this work, using 300 replicates.
+\includegraphics[width=0.8\textwidth]{res_Estep_SINE}
+\caption{Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$ using observations of Figure \ref{fig:obs:SINE}, using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm for diffusion process proposed in this work, using 300 replicates.
 The whiskers represent the extent of the 95\% central values.
 The dot represents the empirical mean over the 300 replicates.
-The dotted line shows the reference value, computed using the unbiased PaRIS algorithm with $N=10000$ particles.}
+The dotted line shows the reference value, computed using the unbiased GRand PaRIS algorithm with $N=10000$ particles.}
+\label{fig:res:SINE}
+\end{figure}
+\begin{figure}[p]
+\centering
+\includegraphics[width=0.8\textwidth]{res_Estep_LG}
+\caption{Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$ using observations of Figure \ref{fig:obs:LG}, using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm for diffusion process proposed in this work, using 300 replicates.
+The whiskers represent the extent of the 95\% central values.
+The dot represents the empirical mean over the 300 replicates.
+The dotted line shows the reference value, computed using the unbiased GRand PaRIS algorithm with $N=10000$ particles.}
 \label{fig:res:LG}
 \end{figure}
+\section{Conclusions}
+We presented a new online smoother for partially observed differential equations.
+This algorithm relies on an acceptance reject procedure inspired from the recent PaRIS algorithm.
+The main result of the article is that the mechanism of this procedure remains valid when the transition density is approximated by a an unbiased positive estimator.
+The resulting algorithm to approximate smoothing expectations outperforms the existing fixed laged smoother for SDEs of \cite{olsson:strojby:2011}, as it is unbiased, and moreover shows a lower variance on two simulated examples.
+The proposed algorithm can be implemented for the class of models $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} with a linear (or at worse quadratic) complexity in $N$.
 % where $\beta\in\mathbb{R}^d$ is chosen by the user.
@@
 
-\section{Multidimensional Ornstein Uhlenbeck bridge}
-\label{sec:ornstein:bridge}
-The GPE introduced in this paper requires to sample finite dimensional distributions of Ornstein Uhlenbeck bridges.
+%\section{Multidimensional Ornstein Uhlenbeck bridge}
+%\label{sec:ornstein:bridge}
+%The GPE introduced in this paper requires to sample finite dimensional distributions of Ornstein Uhlenbeck bridges.
 These results are given here for completeness.
 Consider the following $d$-dimensional SDE:
-\begin{equation}
-\label{eq:sde:ornstein}
-Z_0 = z_0\quad\mbox{and}\quad \rmd Z_t = (\Upsilon Z_t + \vartheta)\rmd t + \rmd W_t\eqsp,
-\end{equation}
-where $W$ is a standard Brownian motion on $\mathbb{R}^d$, $A$ is a $d\times d$ matrix and $b$ a $d$ dimensional vector.
+%\begin{equation}
+%\label{eq:sde:ornstein}
+%Z_0 = z_0\quad\mbox{and}\quad \rmd Z_t = (\Upsilon Z_t + \vartheta)\rmd t + \rmd W_t\eqsp,
+%\end{equation}
+%where $W$ is a standard Brownian motion on $\mathbb{R}^d$, $A$ is a $d\times d$ matrix and $b$ a $d$ dimensional vector.
 In the case of Section~\ref{sec:PaRIS:SDE}, for all $1\le k \le n$, the proposal stochastic differential equation governing $Y_t^k$, $t\in(t_{k-1},t_k)$, is of the form \eqref{eq:sde:ornstein} with
-\[
-\Upsilon = J_{\alpha_{\theta}}(X_{k-1})\;\;\mbox{and}\;\; \vartheta =\alpha_{\theta}(X_{k-1}) - J_{\alpha_{\theta}}(X_{k-1})X_{k-1}\eqsp,
-\]
-$J_{\alpha_{\theta}}$ being the Jacobian of $\alpha_{\theta}$.
+%\[
+%\Upsilon = J_{\alpha_{\theta}}(X_{k-1})\;\;\mbox{and}\;\; \vartheta =\alpha_{\theta}(X_{k-1}) - J_{\alpha_{\theta}}(X_{k-1})X_{k-1}\eqsp,
+%\]
+%$J_{\alpha_{\theta}}$ being the Jacobian of $\alpha_{\theta}$.
 In this case, as $\alpha_{\theta}(x) = \nabla_x A_{\theta}(x)$, $A$ is symmetric and nonsingular.
 By \cite[Section~5.6]{karatzas:shreve:1991}, the unique strong solution of \eqref{eq:sde:ornstein} is:
-\[
-Z_t = \mathrm{e}^{t\Upsilon}z_0 + \int_0^t\mathrm{e}^{(t-s)\Upsilon}\vartheta \rmd s + \int_0^t\mathrm{e}^{(t-s)\Upsilon} \rmd W_s\eqsp.
-\]
-Therefore, the probability density of the conditional distribution of $Z_t$ given $Z_0 = z_0$ is Gaussian with mean $\mu_t(z_0)$ and variance $\Sigma_{t}$ given by:
-\begin{align*}
-\Sigma_{t} &= \int_0^{t} \mathrm{e}^{(t -s)\Upsilon}\mathrm{e}^{(t -s)\Upsilon'}\rmd s\eqsp,\\
-\mu_{t}(z) &= \mathrm{e}^{t \Upsilon}z_0 + \left(\int_0^t\mathrm{e}^{(t-s)\Upsilon} \rmd s\right)\vartheta\eqsp.
-\end{align*}
-As $\Upsilon$ is symetric and nonsingular,
-\begin{align*}
-\Sigma_{t} &= \int_0^{t} \mathrm{e}^{2(t -s)\Upsilon}\rmd s = \Upsilon^{-1}(\mathrm{e}^{2t\Upsilon}-I_d)/2\eqsp,\\
-\mu_{t}(z) &= \mathrm{e}^{t \Upsilon}z + \left(\int_0^t\mathrm{e}^{(t-s)\Upsilon} \rmd s\right)\vartheta = \mathrm{e}^{t \Upsilon}z + \Upsilon^{-1}(\mathrm{e}^{t\Upsilon}-I_d)\vartheta \eqsp.
-\end{align*}
-For each interval $(t_{k-1},t_k)$, $1\le k \le n$, the GPE requires to sample a Ornstein-Uhlenbeck bridge at some ramdom time steps between $X_{k-1}$ at time $t_{k-1}$ and $X_k$ at time $t_k$.
+%\[
+%Z_t = \mathrm{e}^{t\Upsilon}z_0 + \int_0^t\mathrm{e}^{(t-s)\Upsilon}\vartheta \rmd s + \int_0^t\mathrm{e}^{(t-s)\Upsilon} \rmd W_s\eqsp.
+%\]
+%Therefore, the probability density of the conditional distribution of $Z_t$ given $Z_0 = z_0$ is Gaussian with mean $\mu_t(z_0)$ and variance $\Sigma_{t}$ given by:
+%\begin{align*}
+%\Sigma_{t} &= \int_0^{t} \mathrm{e}^{(t -s)\Upsilon}\mathrm{e}^{(t -s)\Upsilon'}\rmd s\eqsp,\\
+%\mu_{t}(z) &= \mathrm{e}^{t \Upsilon}z_0 + \left(\int_0^t\mathrm{e}^{(t-s)\Upsilon} \rmd s\right)\vartheta\eqsp.
+%\end{align*}
+%As $\Upsilon$ is symetric and nonsingular,
+%\begin{align*}
+%\Sigma_{t} &= \int_0^{t} \mathrm{e}^{2(t -s)\Upsilon}\rmd s = \Upsilon^{-1}(\mathrm{e}^{2t\Upsilon}-I_d)/2\eqsp,\\
+%\mu_{t}(z) &= \mathrm{e}^{t \Upsilon}z + \left(\int_0^t\mathrm{e}^{(t-s)\Upsilon} \rmd s\right)\vartheta = \mathrm{e}^{t \Upsilon}z + \Upsilon^{-1}(\mathrm{e}^{t\Upsilon}-I_d)\vartheta \eqsp.
+%\end{align*}
+%For each interval $(t_{k-1},t_k)$, $1\le k \le n$, the GPE requires to sample a Ornstein-Uhlenbeck bridge at some ramdom time steps between $X_{k-1}$ at time $t_{k-1}$ and $X_k$ at time $t_k$.
 Conditional on $Z_s = x$ and $Z_T = y$, the probability density of $Z_t$ is Gaussian with mean $\mu$ and variance $\Sigma_t^{s,T}$ given by:
-\begin{align*}
-\mu &=\Gamma(t,T)\Gamma(s,T)^{-1}\mu_- + \Gamma(s,t)'(\Gamma(s,T)')^{-1}\mu_+\eqsp,\\
-\Sigma_t^{s,T} &=\Gamma(t,T)\Gamma(s,T)^{-1}\Gamma(s,t)\eqsp,\\
-\Gamma(s,t)&=\int_s^{t} \mathrm{e}^{(s -u)\Upsilon}\mathrm{e}^{(t -u)\Upsilon'}\rmd u\eqsp = \Upsilon^{-1}\left(\mathrm{e}^{(t -s)\Upsilon}-\mathrm{e}^{(s -t)\Upsilon}\right)/2\eqsp,\\
-\mu_- &= x - \Upsilon^{-1}(\mathrm{e}^{(s -t)\Upsilon}-I_d)\vartheta\eqsp,\\
-\mu_+ &= y-\Upsilon^{-1}(\mathrm{e}^{(T -t)\Upsilon}-I_d)\vartheta\eqsp.
-\end{align*}
+%\begin{align*}
+%\mu &=\Gamma(t,T)\Gamma(s,T)^{-1}\mu_- + \Gamma(s,t)'(\Gamma(s,T)')^{-1}\mu_+\eqsp,\\
+%\Sigma_t^{s,T} &=\Gamma(t,T)\Gamma(s,T)^{-1}\Gamma(s,t)\eqsp,\\
+%\Gamma(s,t)&=\int_s^{t} \mathrm{e}^{(s -u)\Upsilon}\mathrm{e}^{(t -u)\Upsilon'}\rmd u\eqsp = \Upsilon^{-1}\left(\mathrm{e}^{(t -s)\Upsilon}-\mathrm{e}^{(s -t)\Upsilon}\right)/2\eqsp,\\
+%\mu_- &= x - \Upsilon^{-1}(\mathrm{e}^{(s -t)\Upsilon}-I_d)\vartheta\eqsp,\\
+%\mu_+ &= y-\Upsilon^{-1}(\mathrm{e}^{(T -t)\Upsilon}-I_d)\vartheta\eqsp.
+%\end{align*}
 

### 2017-02-12 ParisEM.tex
minor changes

--- 2017-01-17 ParisEM.tex
@@
 & = \sum_{m\ge 1}h(m)\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\middle| \mathcal{F}_k^N\right]\right)\mathbb{E}\left[h(I_m)\1_{\mathcal{A}^k_{m}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[1-\frac{\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_\ell},\xi_k^{i},\Lambda_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)\\
-&\hspace{5cm}\times\mathbb{E}\left[h(I_m)\frac{\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_m},\xi_k^{i},\Lambda_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \sum_{m\ge 1}\left(\mathbb{E}\left[1-\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)^{m-1}\mathbb{E}\left[h(I_m)\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \mathbb{E}\left[h(I_1)q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{i})\middle| \mathcal{F}_k^N\right]/\mathbb{E}\left[q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{i})\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \sum_{\ell=1}^N \frac{h(\ell)\omega_{k-1}^{\ell}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{\ell},\xi_k^{i})}{\sum_{m=1}^N\omega_{k-1}^{m}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{m},\xi_k^{i})}\eqsp,\\
+& = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[1-\frac{\widehat{q_k}(\xi_{k-1}^{I_\ell},\xi_k^{i};\zeta_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)\\
+&\hspace{5cm}\times\mathbb{E}\left[h(I_m)\frac{\widehat{q_k}(\xi_{k-1}^{I_m},\xi_k^{i};\zeta_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
+& = \sum_{m\ge 1}\left(\mathbb{E}\left[1-\frac{q_k(\xi_{k-1}^{I_1},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)^{m-1}\mathbb{E}\left[h(I_m)\frac{q_k(\xi_{k-1}^{I_m},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
+& = \mathbb{E}\left[h(I_1)q_k(\xi_{k-1}^{I_1},\xi_k^{i})\middle| \mathcal{F}_k^N\right]/\mathbb{E}\left[q_k(\xi_{k-1}^{I_m},\xi_k^{i})\middle| \mathcal{F}_k^N\right]\eqsp,\\
+& = \sum_{\ell=1}^N \frac{h(\ell)\omega_{k-1}^{\ell}q_k(\xi_{k-1}^{\ell},\xi_k^{i})}{\sum_{m=1}^N\omega_{k-1}^{m}q_k(\xi_{k-1}^{m},\xi_k^{i})}\eqsp,\\
 &= \sum_{\ell=1}^N \Lambda_{k-1}^N(i,\ell)h(\ell) \eqsp,
@@
 \begin{proof}[Proof of Lemma \ref{lem:iid}]
-The independence is ensured by the mechanism of SMC methods.\\
-By \eqref{eq:random:weight},
+The independence is ensured by the mechanism of SMC methods. By \eqref{eq:random:weight},
 \[
-\mathbb{E}\left[\tilde{\omega}^i_{k+1}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right] = \mathbb{E}\left[\frac{M^{-1}\sum_{m=1}^M \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}}, \xi^{i}_{k+1};\zeta^m_{k+1})g_{k+1}(\xi^{i}_{k+1})}{\vartheta_{k+1}(\xi^{I^{i}_{k+1}}_{k}) p_{k+1}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right]\eqsp.
+\mathbb{E}\left[\widehat{\omega}^i_{k+1}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right] = \mathbb{E}\left[\frac{M^{-1}\sum_{m=1}^M \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}}, \xi^{i}_{k+1};\zeta^m_{k+1})g_{k+1}(\xi^{i}_{k+1})}{\vartheta_{k+1}(\xi^{I^{i}_{k+1}}_{k}) p_{k+1}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right]\eqsp.
 %&= \mathbb{E}\left[\frac{g_{k+1}(\xi^{1}_{k+1})}{\vartheta_{k+1}(\xi^{I^{1}_{k+1}}_{k}) p_{k+1}(\xi_{k}^{I^{1}_{k+1}},\xi^{\ell}_{k+1})}\tau^{1}_{k+1}\widehat{q}_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{I^{1}_{k+1}},\xi^{1}_{k+1};\Lambda_{k+1})\middle| \mathcal{F}_k^{N}\right]\eqsp,
@@
 \end{multline*}
-Moreover, The p.d.f. of the couple $(\xi_{k+1}^i,I_{k+1}^i)\vert \mathcal{F}_k^N$ is given by
-\begin{align*}
-p(x,j) &= \frac{\omega_k^j\vartheta_{k+1}(\xi_k^j)p_k(\xi_k^j,x)}{\sum_{l=1}^N\int \omega_k^l\vartheta_{k+1}(\xi_k^l)p_k(\xi_k^l,y)\rmd y}=\frac{\omega_k^j\vartheta_{k+1}(\xi_k^j)p_k(\xi_k^j,x)}{\phi_k^N[\vartheta_{k+1}]}
-\end{align*}
+Moreover, conditionally to $\mathcal{F}_k^N$, the probability density function of $(\xi_{k+1}^i,I_{k+1}^i)$ is given by
+\[
+(x,j) \mapsto \frac{\omega_k^j\vartheta_{k+1}(\xi_k^j)p_k(\xi_k^j,x)}{\Omega_k\phi_k^N[\vartheta_{k+1}]}\eqsp.
+\]
 Therefore, this yields:
 \begin{align*}
-\mathbb{E}\left[\tilde{\omega}^i_{k+1}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right]&= \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1} \sum_{j=1}^N\omega_k^j \int \vartheta_{k+1}(\xi^{j}_{k})\frac{\qk(\xi_{k}^{j},x) g_{k+1}(x)}{\vartheta_{k+1}(\xi^{j}_{k}) p_{k}(\xi_{k}^{j},x)}\\
+\mathbb{E}\left[\widehat{\omega}^i_{k+1}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right]&= \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1} \sum_{j=1}^N\frac{\omega_k^j}{\Omega_k} \int \vartheta_{k+1}(\xi^{j}_{k})\frac{\qk(\xi_{k}^{j},x) g_{k+1}(x)}{\vartheta_{k+1}(\xi^{j}_{k}) p_{k}(\xi_{k}^{j},x)}\\
 &\hspace{1cm}\times \sum_{\ell=1}^N\frac{\omega_k^{\ell} \qk (\xi_{k}^{\ell},x)\left(\tau^{\ell}_k + h_{k}(\xi_{k}^{\ell},x)\right)}{\sum_{\ell'=1}^N\omega_k^{\ell'}\qk(\xi_{k}^{\ell'},x)}p_{k}(\xi_{k}^{j},x)\rmd x\eqsp,\\
 &= \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\\
-&~~~~\times\sum_{\ell=1}^N \omega_k^\ell\left[\int \frac{ \sum_{j=1}^N \omega_k^j\qk(\xi_k^j,x) }{ \sum_{\ell'=1}^N\omega_k^{\ell'}\qk(\xi_{k}^{\ell'},x) } g_{k+1}(x)\qk (\xi_{k}^{\ell},x)\left(\tau^{\ell}_k + h_{k}(\xi_{k}^{\ell},x)\right) \rmd x \right]\\
+&~~~~\times\sum_{\ell=1}^N \frac{\omega_k^\ell}{\Omega_k}\left[\int \frac{ \sum_{j=1}^N \omega_k^j\qk(\xi_k^j,x) }{ \sum_{\ell'=1}^N\omega_k^{\ell'}\qk(\xi_{k}^{\ell'},x) } g_{k+1}(x)\qk (\xi_{k}^{\ell},x)\left(\tau^{\ell}_k + h_{k}(\xi_{k}^{\ell},x)\right) \rmd x \right]\\
 & =\left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\int \qk(\cdot,x)g_{k+1}(x)\left\{\tau_k(\cdot) + h_{k}(\cdot,x)\right\}\rmd x\right]\eqsp,

### 2017-02-13 ParisEM.tex
Overall revision

@@
 the computation of the posterior distributions of sequences of hidden states given observations.
-Filtering refers to the estimation of the distributions of the hidden state $X_k$ given the observations $(Y_0,\ldots,Y_k)$, while smoothing stands for the estimation of the distribution of sequence of states $(Z_{k},\ldots,Z_{p})$ given observations $(Y_{0},\ldots,Y_{\ell})$ with $0\le k\le p<\ell \le n$.
+Filtering refers to the estimation of the distributions of the hidden state $X_k$ given the observations $(Y_0,\ldots,Y_k)$, while smoothing stands for the estimation of the distribution of sequence of states $(X_{k},\ldots,X_{p})$ given observations $(Y_{0},\ldots,Y_{\ell})$ with $0\le k\le p<\ell \le n$.
 These posterior distributions are crucial to compute maximum likelihood estimators of the unknown parameters defining the model using the observations $(Y_0,\ldots,Y_n)$ only.
@@
 To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE) introduced in \cite{fearnhead:papaspiliopoulos:roberts:2008}.
-In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities.\\
+In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities.
+
  The only solution to solve the smoothing problem for partially observed SDE using SMC methods has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}.
@@
 
-In this paper, instead of using fixed-lag techniques which introduce a systematic additional bias in the approximation of conditional expectations, we extend the use of  the PaRIS algorithm to partially observed SDE.
-The proposed algorithm allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles in the best cases, and at worse with quadratic complexity.
-The crucial result of the application of PaRIS algorithm to SDE is that the accept reject mechanism ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by positive and bounded unbiased estimates.
-The proposed method results in a generalized random version of the PaRIS algorithm, hereafter named GRand PaRIS algorithm.
-The proposed algorithm apply to general hidden Markov models whose Markovian dynamics is ruled by a stochastic differential equation, belonging to one of the first two domains defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.
+In this paper, we extend the use of  PaRIS algorithm to partially observed SDE. 
+The proposed algorithm allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles.
+The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to SDE is that the accept reject mechanism ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
+The usual FFBS and FFBSi algorithms may not be extended this easily since they both require the computation of weights defined as ratios involving the transition densities, thus replacing these unknown quantities by unbiased estimates does not lead to unbiased estimators of the weights.
+The proposed generalized random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to general hidden Markov models whose Markovian dynamics is ruled by a stochastic differential equation (one of the first two domains defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}) but also to any general state space model where the transition density of the hidden chain may be estimated unbiasedly.
 %We also propose to improve the algorithm of \cite{fearnhead:papaspiliopoulos:roberts:2008} by extending the range of diffusion processes for which SMC methods can be applied.
@@
 See for instance \cite{roberts:tweedie:1996} for conditions under which this diffusion converges exponentially quickly to $\pi$.
-SDE where the drift is given by the gradient of a potential function have been widely used in ecology for instance, see \cite{brillinger:et:al:2011,harris:blackwell:2013,preisler:et:al:2013} and the references therein.\\
-The solution of \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{k=0,\dots,n}$ in $\mathbb{R}^m$. 
+SDE where the drift is given by the gradient of a potential function have been widely used in ecology for instance, see \cite{brillinger:et:al:2011,harris:blackwell:2013,preisler:et:al:2013} and the references therein.
+The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{k=0,\dots,n}$ in $\mathbb{R}^m$.
 For all $0\le k \le n$, the distribution of $Y_k$ given $X_k:= X_{t_k}$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,Y_k) = g_k(X_k)$. 
 The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
-For all $1\le k \le n$, the distribution of $X_{k+1} $ conditional on $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$.\\ 
+For all $1\le k \le n$, the distribution of $X_{k+1} $ conditional on $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$.
+
 Let $0 \leq k \leq k' \leq n$, the joint smoothing distributions of the hidden states are defined, for all measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, by:
@@
 The approximation of \eqref{eq:FFbsm:equality} requires first to approximate the sequence of filtering distributions. 
-Sequential Monte Carlo methods provide an efficient and simple solution to obtain these approximations using sets of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$, $0\le k \le n$.\\
+Sequential Monte Carlo methods provide an efficient and simple solution to obtain these approximations using sets of particles $\{\xi^{\ell}_k\}_{\ell=1}^N$ associated with weights $\{\omega^{\ell}_k\}_{\ell=1}^N$, $0\le k \le n$.
+
 At time $k = 0$, $N$ particles $\{\xi^{\ell}_0\}_{\ell=1}^N$ are sampled independently according to  $\xi^{\ell}_0 \sim \eta_0$, where $\eta_0$ is a probability density with respect to $\mu$. 
@@
 \item sample  $\xi^{\ell}_{k}$ using this chosen particle according to $\xi^{\ell}_{k} \sim p_k(\xi^{I^{\ell}_k}_{k-1},\cdot)$ ; 
-\item $\xi^{\ell}_k$ is associated with:
+\item  associate the particle $\xi^{\ell}_k$ with the importance weight:
 \begin{equation}
@@
 \]
-The PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSi algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions.
+PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSi algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions.
 %The FFBS algorithm has a computational complexity of order $\mathcal{O}(N^{k'-k+1})$ to approximate joint smoothing distributions of the form $\phi_{k:k'|n}$ with $k<k'$ which makes it prohibitive. 
@@
 
-The PaRIS algorithm relies on a fundamental property of $T_k[H_k]$ when $H_k$ for additive functionals.
-Indeed, if $H_k$ satisfies the additive property as in \eqref{def:addfunc}, the following equality holds:
+PaRIS algorithm relies on the following fundamental property of $T_k[H_k]$ when $H_k$ is as in \eqref{def:addfunc}:
+\begin{align*}
+T_k[H_k](X_k) &=\mathbb{E}\left[T_{k-1}[H_{k-1}](X_{k-1}) + h_{k-1}(X_{k-1},X_k)\middle|X_k,Y_{0:k-1} \right]\eqsp,\\
+&= \frac{\int \phi_{k-1}(\rmd x_{k-1})q_{k-1}(x_{k-1},X_k)\left\{T_{k-1}[H_{k-1}](x_{k-1}) + h_{k-1}(x_{k-1},X_k)\right\}}{\int \phi_{k-1}(\rmd x_{k-1})q_{k-1}(x_{k-1},X_k)}\eqsp.
+\end{align*}
+Therefore, \cite{olsson:westerborn:2016} introduces sufficient statistics $\tau^i_k$ (starting with $\tau^i_0 = 0$, $1\le i\le N$), approximating $T_k[H_k](\xi^i_k)$, for $1\le i\le N$ and $0\le k \le n$.
+First, replacing $\phi_{k-1}$ by $\phi^N_{k-1}$ in the last equation leads to the following approximation of $T_k[H_k](\xi^i_k)$:
 \begin{equation}
-T_k[H_k](X_k)=\mathbb{E}\left[T_{k-1}[H_{k-1}](X_{k-1}) + h_{k-1}(X_{k-1},X_k) \right],~1\leq k\leq n.
+\label{eq:Tk}
+T_k^N[H_k](\xi_k^i) = \sum_{j=1}^N \Lambda_{k-1}^N(i,j)\left\{T_{k-1}[H_{k-1}](\xi_{k-1}^j) + h_{k-1}(\xi^j_{k-1},\xi^i_k)\right\}\eqsp, 
 \end{equation}
-Therefore, \cite{olsson:westerborn:2016} introduces sufficient statistics $\tau^i_k$, approximating $T_k[h](\xi^i_k)$, for $1\le i\le N$ and $0\le k \le n$ starting with $\tau^i_0 = 0$ for all $1\le i\le N$.
+where
+\begin{equation}
+\label{eq:Lambda}
+\Lambda_{k}^N(i,\ell) = \frac{\omega^{\ell}_{k} \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})}{\sum_{\ell=1}^N\omega^{\ell}_{k} \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})}\eqsp,\quad 1\le \ell\le N\eqsp.
+\end{equation}
+Then, instead of computing exactly these approximations which would lead to a complexity growing quadratically with $N$ because of the normalizing constant in \eqref{eq:Lambda}, PaRIS algorithm samples particles in the set $\{\xi^j_{k-1}\}_{j=1}^N$ with probabilities $\Lambda_{k}^N(i,\cdot)$ to approximate the expectation \eqref{eq:Tk} and produce $\tau^i_k$.
 Choosing $\tilde{N}\ge 1$, at each time step $0\le k \le {n-1}$ these statistics are updated according to the following steps.
@@
 \item \label{it:PaRIS:filt} Run one step of a particle filter to produce $\{(\xi^{\ell}_k, \omega^{\ell}_k)\}$ for $1\le \ell \le N$.
-\item \label{it:PaRIS:sampleindex} For all $1\le i \le N$, sample independently $J_{k}^{i,\ell}$ in $\{1,\ldots,N\}$ for $1\le \ell \le \widetilde N$ with probabilities $\Lambda_{k}^N(i,\cdot)$, given by
-\[
-\Lambda_{k}^N(i,\ell) = \frac{\omega^{\ell}_{k} \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})}{\sum_{\ell=1}^N\omega^{\ell}_{k} \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})}\eqsp,\quad 1\le \ell\le N\eqsp.
-\]
+\item \label{it:PaRIS:sampleindex} For all $1\le i \le N$, sample independently $J_{k}^{i,\ell}$ in $\{1,\ldots,N\}$ for $1\le \ell \le \widetilde N$ with probabilities $\Lambda_{k}^N(i,\cdot)$, given by \eqref{eq:Lambda}.
 \item \label{it:PaRIS:smooth} Set
@@
 However, there is a significant qualitative difference between the cases $\tilde{N} = 1$ and $\tilde{N} \geq 2$.
-As for the FFBSi algorithm,  when there exists $\sigma_+$ such that $0<\qk <\sigma_+$, the PaRIS algorithm may be implemented with $\mathcal{O}(N)$ complexity using the accept reject mechanism of \cite{douc:garivier:moulines:olsson:2011}.
+As for the FFBSi algorithm,  when there exists $\sigma_+$ such that $0<\qk <\sigma_+$, PaRIS algorithm may be implemented with $\mathcal{O}(N)$ complexity using the accept reject mechanism of \cite{douc:garivier:moulines:olsson:2011}.
 
-In general situations, the PaRIS algorithm cannot be used for stochastic differential equations as $\qk$ is unknown.
-Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio of \cite{douc:garivier:moulines:olsson:2011} is not tractable.\\
+In general situations, PaRIS algorithm cannot be used for stochastic differential equations as $\qk$ is unknown.
+Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio of \cite{douc:garivier:moulines:olsson:2011} is not tractable.
 Following \cite{fearnhead:papaspiliopoulos:roberts:2008,olsson:strojby:2011}, filtering weights can be approximated by replacing $\qk(\xi^{\ell}_{k},\xi_{k+1}^{i})$  by an  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable such that:
@@
 \label{eq:random:weight}
-\widehat{\omega}^{\ell}_k := \frac{\widehat{q}_k(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_k (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
+\widehat{\omega}^{\ell}_k := \frac{\widehat{q}_k(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k;\zeta_k)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_k (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
 \end{equation}
-
-Therefore, to obtain a generalized random version of PaRIS, one need to be able to sample from weight $\Lambda_k^N(i,\ell)$ in the case of SDE based HMM.\\
+Therefore, to obtain a generalized random version of PaRIS, we only need to be able to sample from weight $\Lambda_k^N(i,\ell)$ in the case of SDE based HMM.
 If $\widehat{q}_k$ may be upper bounded almost surely by a constant $\hat{\sigma}_+$, this task is performed using the accept reject mechanism given in Algorithm~\ref{alg:AR:unknownq}.
@@
 Let $0\le k\le n-1$.
-For all $1\le i \le N$ and all $1\le \ell \le \widetilde{N}$, the conditional probability distribution given $\mathcal{F}_{k+1}^N$ of the random variable $J_{k}^{i,\ell}$ produced by Algorithm~\ref{alg:AR:unknownq} is  $\Lambda_{k}^N(i,\cdot)$.
+For all $1\le i \le N$ and all $1\le \ell \le \widetilde{N}$, the conditional probability distribution given $\mathcal{F}_{k+1}^N$ of the random variable $J_{k}^{i,\ell}$ produced by Algorithm~\ref{alg:AR:unknownq} is  $\Lambda_{k-1}^N(i,\cdot)$.
 \end{lemma}
@@
 \end{proof}
-Therefore, having an unbiased positive and bounded estimator is sufficient to perform the Generalized Random PaRIS (GRand PaRIS) algorithm, fully described in Algorithm \ref{alg:Ozaki:PaRIS}.\\
-One can notice that in Algorithm \ref{alg:AR:unknownq}, it is actually sufficient to have a constant $\hat{\sigma}_+^k$ such that, for all $j$, the random variable $\hat{q}_k(\xi_k^j,y,\zeta_k)< \hat{\sigma}_+^k$ almost surely.
-This improvement keeps the linear complexity property of PaRIS algorithm and Algorithm \ref{alg:Ozaki:PaRIS}, and allows to increase the acceptance rate of Algorithm \ref{alg:AR:unknownq}.
-Moreover, the AR algorithm still stands if one can define
-\begin{equation}
-\hat{\sigma}_{+}^{k,i}= \underset{j=1,\dots,N}{\text{sup}}\rho_\Delta(\xi_{k-1}^j,\xi_k^i)\eqsp, \label{eq:quadr:bound}
-\end{equation}
-which is a weaker assumption.
+Note that the assumption that for all $1\le j\le N$, $\sup_y \hat{q}_k(\xi_k^j,y,\zeta_k)< \hat{\sigma}_+^k$ almost surely is actually sufficient to apply Algorithm \ref{alg:AR:unknownq}.
+This improvement keeps the linear complexity property of PaRIS algorithm and Algorithm \ref{alg:Ozaki:PaRIS}, and allows to increase significantly the acceptance rate of Algorithm \ref{alg:AR:unknownq}.
+%Moreover, the AR algorithm still stands if one can define
+%\begin{equation}
+%\hat{\sigma}_{+}^{k,i}= \underset{j=1,\dots,N}{\text{sup}}\rho_\Delta(\xi_{k-1}^j,\xi_k^i)\eqsp, \label{eq:quadr:bound}
+%\end{equation}
+%which is a weaker assumption.
 However, in this case, the resulting algorithm has a quadratic complexity in $N$.
@@
 \subsection*{Bounded estimator of $q_k$}
-For $x, y \in \mathbb{R}^d$, using Girsanov adn Itô's formulas, one can show that the transition density $q_\Delta(x,y):=q(x,y,\Delta)$ related to \eqref{eq:target:sde} satisfies:
+For $x, y \in \mathbb{R}^d$, by Girsanov and Ito's formulas, the transition density $q_k(x,y)$ of \eqref{eq:target:sde} satisfies, with $\Delta_k = t_{k+1}-t_k$,
 \begin{align*}
-q_\Delta(x,y)=\varphi_\Delta(x,y)\exp\left\lbrace A(y)-A(x)\right\rbrace \mathbb{E}_{\mathbb{W}^{x,y,\Delta}}\left[ \exp \left\lbrace - \int_0^\Delta \phi(\mw_s)\rmd s \right\rbrace \right]
+q_k(x,y)=\varphi_{\Delta_k}(x,y)\exp\left\lbrace A(y)-A(x)\right\rbrace \mathbb{E}_{\mathbb{W}^{x,y,\Delta_k}}\left[ \exp \left\lbrace - \int_0^{\Delta_k} \phi(\mw_s)\rmd s \right\rbrace \right]\eqsp,
 \end{align*}
-where $\mathbb{W}^{x,y,\Delta}$ is the law of Brownian bridge starting at $x$ at 0 and hitting $y$ at $\Delta$, $(\mw_t)_{0\leq t \leq \Delta}$ is such a Brownian bridge, $\varphi_\Delta(x,y)$ is the p.d.f.
-of a normal distribution with mean $x$ and variance $\Delta$, evaluated in $y$ and $phi$ is a function from $\mathbb{R}^d$ to $\mathbb{R}$ such that
+where $\mathbb{W}^{x,y,\Delta_k}$ is the law of Brownian bridge starting at $x$ at 0 and hitting $y$ at $\Delta_k$, $(\mw_t)_{0\leq t \leq \Delta_k}$ is such a Brownian bridge, $\varphi_{\Delta_k}(x,y)$ is the p.d.f.
+of a normal distribution with mean $x$ and variance $\Delta_k$, evaluated in $y$ and $\phi:\mathbb{R}^d\to\mathbb{R}$ is defined as:
 \[
-\phi(x) =\parallel \alpha(x)^2 \parallel + \triangle A(x)\eqsp.
+\phi(x) =\|\alpha(x)\|^2  + \triangle A(x)\eqsp,
 \]
-Suppose there exist random variables $\Lo_\mw$ and $\U_\mw$ such that for all $0\leq s \leq \Delta$, $\Lo_\mw \leq \phi(\mw_s)\leq \U_\mw$.
-Let $\kappa$ be a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and $(U_j)_{1\le j\le \kappa}$ be independent uniform random variables on $[0,\Delta]$, and $
+with $\triangle$ the Laplace operator.
+Suppose there exist random variables $\Lo_\mw$ and $\U_\mw$ such that for all $0\leq s \leq \Delta_k$, $\Lo_\mw \leq \phi(\mw_s)\leq \U_\mw$.
+Let $\kappa$ be a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and $(U_j)_{1\le j\le \kappa}$ be independent uniform random variables on $[0,\Delta_k]$, and $
 \zeta_k = \left\{\kappa,\mw,U_1,\ldots,U_\kappa\right\}\eqsp$. 
@@
 \begin{multline}
-\widehat{q}_\Delta(x,y;\zeta_k) = \varphi_\Delta(x,y) \exp \left\{A(y) - A(x)\right\}\\ 
-\times\mathrm{exp}\left\{-\U_\mw\Delta\right\}\frac{\Delta^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\U_\mw-\phi(\mw_{U_j})\right)\eqsp.\label{eq:unbiased:q}
+\widehat{q}_k(x,y;\zeta_k) = \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)\right\}\\ 
+\times\mathrm{exp}\left\{-\U_\mw\Delta\right\}\frac{\Delta_k^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\U_\mw-\phi(\mw_{U_j})\right)\eqsp.\label{eq:unbiased:q}
 \end{multline}
-Interesting choices of $\mu$ are discussed in \cite{fearnhead:papaspiliopoulos:roberts:2008}n and we focus here on the so called GPE-1, when $\mu$ is the distribution of a Poisson law with intensity $(\U_\mw-\Lo_\mw)\Delta$.
+Interesting choices of $\mu$ are discussed in \cite{fearnhead:papaspiliopoulos:roberts:2008} and we focus here on the so called GPE-1, when $\mu$ is a Poisson distribution with intensity $(\U_\mw-\Lo_\mw)\Delta_k$.
 In that case, the estimator \eqref{eq:unbiased:q} becomes:
 \begin{equation}
-\widehat{q}_\Delta(x,y;\zeta_k) = \varphi_\Delta(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta \right\}\prod_{j=1}^{\kappa}\frac{\U_\mw-\phi(\mw_{U_j})}{\U_\mw-\Lo_\mw}\eqsp.\label{eq:GPE1}
+\widehat{q}_{\Delta_k}(x,y;\zeta_k) = \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}\prod_{j=1}^{\kappa}\frac{\U_\mw-\phi(\mw_{U_j})}{\U_\mw-\Lo_\mw}\eqsp.\label{eq:GPE1}
 \end{equation}
-On the r.h.s. of \eqref{eq:GPE1}, the product over $\kappa$ elements is bounded by 1, therefore, a sufficient condition to obtain a bound over $\hat{q}_\Delta$ is that the function:
+On the r.h.s. of \eqref{eq:GPE1}, the product over $\kappa$ elements is bounded by 1, therefore, a sufficient condition to obtain an upper bound  $\hat{q}_k$ is that the function:
 \begin{align*}
-\rho_\Delta:\eqsp\mathbb{R}^d\times \mathbb{R}^d &\mapsto \mathbb{R}\\
-(x,y)&\mapsto \varphi_\Delta(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta \right\}
+\rho_{\Delta_k}:\eqsp\mathbb{R}^d\times \mathbb{R}^d &\mapsto \mathbb{R}\\
+(x,y)&\mapsto \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}
 \end{align*}
-is upper bounded almost surely by $\hat{\sigma}_+$.
-As said above, the GRand PaRIS algorithm only requires a bound $\hat{\sigma}_+^k$ for the function $\rho_\Delta(x,y)$ where $x$ is restricted to its particle support (i.e., at step $k+1$, $x \in \xi_{k}^1,\dots,\xi_{1}^N$.).\\
-In order to obtain a better acceptance rate (at a cost of a quadratic complexity in $N$), one can always compute the bound defined by \eqref{eq:quadr:bound} as long as $\Lo_\mw$ is bounded almost surely.
-Actually, this condition is always satisfied for models being in the domains $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, i.e.
-domains for which the exact algorithms 1 and 2 can be performed.
+is upper bounded almost surely by $\hat{\sigma}_+^k$.
+As said above, the GRand PaRIS algorithm only requires a bound $\hat{\sigma}_+^k$ for the function $\rho_{\Delta_k}$ where $x \in \{\xi_{k}^1,\dots,\xi_{1}^N\}$.
+In order to improve the acceptance rate (at a cost of a quadratic complexity in $N$), we can always compute, for all $1\le i\le N$, the bound
+\begin{equation}
+\hat{\sigma}_{+}^{k,i}= \underset{1 \le j\le N}{\text{sup}}\rho_{\Delta_k}(\xi_{k-1}^j,\xi_k^i)\eqsp, \label{eq:quadr:bound}
+\end{equation}
+as long as $\Lo_\mw$ is bounded almost surely.
+Actually, this condition is always satisfied for modelsin the domains $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, i.e.
+domains for which the exact algorithms EA1 and EA2 can be used.
 \subsection*{Note on the complexity and practical implementation}
@@
 \FORALL{$i \in 1,\dots, N$}
-\STATE Set $\xi_0^i = X_0$, $\tau_0^i = 0$  and  $\omega_0^i = g_0(\xi_0^i)\chi_0(\xi_0^i)/\nu_0(\xi_0^i)$.
+\STATE Sample $\xi_0^i \sim\eta_0$, $\tau_0^i = 0$  and  $\omega_0^i = g_0(\xi_0^i)\chi_0(\xi_0^i)/\eta_0(\xi_0^i)$.
 \ENDFOR
@@
 \begin{proposition}
+\label{prop:exp:deviation}
 Assume that H\ref{assum:boundmodel} and H\ref{assum:boundalgo} hold and that for all $1\le k\le n$, $\mathrm{osc}(h_k)<+\infty$.
@@
 \begin{proof}
-The results is proved by induction.
-At time $k=0$, the result holds using that for all $1\le i \le N$, $\rho_0^i = 0$ and the convention $T_0[h_0] =0$.
-In addition, $\phi_0^N$ is a standard importance sampler estimator of $\phi_0$ with $\omega_0^i\le |\widehat{\omega}_0|_{\infty}$ so that for any bounded function $h$ on $\mathsf{X}$,
-\[
-\mathbb{P}\left(\left|\phi_0^N[h] - \phi_0\left[h\right]\right|\ge \varepsilon\right)\le b_0\exp\left(-c_0N\varepsilon^2\right)\eqsp.
-\]
-Assume the results holds for $k\ge 1$ and write
-\[
-\phi_{k+1}^N[\tau_{k+1}] - \phi_{k+1}\left[T_{k+1}[h_{k+1}]\right] = a_N/b_N\eqsp,
-\]
-where $a_N = N^{-1}\sum_{i=1}^N \widehat{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}[h_{k+1}]\right]\right)$ and $b_N =N^{-1}\sum_{i=1}^N \widehat{\omega}_{k+1}^i$.
-By Lemma~\ref{lem:iid}, the random variables $\{\widehat{\omega}_{k+1}^i\tau_{k+1}^i\}_{i=1}^N$ are independent conditionally on $\mathcal{F}_k^{N}$ and by H\ref{assum:boundalgo},
-\[
-\left|\widehat{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}[h_{k+1}]\right]\right)\right| \le 2|\widehat{\omega}_{k+1}|_{\infty}|H_{k+1}|_{\infty}\eqsp.
-\]
-Therefore, by Hoeffding inequality,
-\[
-\mathbb{P}\left(\left|a_N - \mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right) = \mathbb{E}\left[\mathbb{P}\left(\left|a_N - \mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\middle|\mathcal{F}_k^{N}\right)\right]\le 2\exp\left(-c_kN\varepsilon^2\right)\eqsp.
-\] 
-On the other hand,
-\[
-\mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right] = \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\Upsilon_k\right] \eqsp,
-\]
-where
-\[
-\Upsilon_k(x_k) = \int q_{k}(\cdot,x)g_{k+1}(x)\left(\tau_k(x_k) + h_{k+1}(x_k,x) - \phi_{k+1}\left[T_{k+1}[h_{k+1}]\right]\right)\rmd x\eqsp.
-\]
-By \cite[Lemma~11]{olsson:westerborn:2016}, $\phi_{k}\left[\Upsilon_k\right] = 0$ which implies by the induction assumption that 
-\[
-\mathbb{P}\left(\left|\mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right)\le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
-\]
-Then,
-\[
-\mathbb{P}\left(\left|a_N\right|\ge \varepsilon\right) \le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
-\] 
-Similarly, as $b_N \le |\widehat{\omega}_k|_{\infty}$, by Hoeffding inequality,
-\begin{multline*}
-\mathbb{P}\left(\left|b_N - \mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right) \\
-= \mathbb{E}\left[\mathbb{P}\left(\left|b_N - \mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\middle|\mathcal{F}_k^{N}\right)\right]\le 2\exp\left(-c_kN\varepsilon^2\right)\eqsp.
-\end{multline*}
-Note that
-\[
-\mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right] = \phi^N_{k}\left[\int q_{k}(\cdot,x)g_{k+1}(x)\rmd x\right]\eqsp.
-\]
-By  the induction assumption,
-\[
-\mathbb{P}\left(\left|\mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]-\phi_k\left[\int q_{k}(\cdot,x)g_{k+1}(x)\rmd x\right]\right|\ge \varepsilon\right)\le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
-\]
-The proof is completed using Lemma~\ref{lem:hoeffding:ratio}.
-\end{proof}
-
-\begin{lemma}\label{lem:hoeffding:ratio}
-Assume that $a_N$, $b_N$, and $b$ are random variables defined on the same probability space such that there exist positive constants $\beta$, $B$, $C$, and $M$ satisfying
-\begin{enumerate}[(i)]
-    \item $|a_N/b_N|\leq M$, $\mathbb{P}$-a.s.\ and  $b \geq \beta$, $\mathbb{P}$-a.s.,
-    \item For all $\epsilon>0$ and all $N\geq1$, $\mathbb{P}\left[|b_N-b|>\epsilon \right]\leq B \exp\left(-C N \epsilon^2\right)$,
-    \item For all $\epsilon>0$ and all $N\geq1$, $\mathbb{P} \left[ |a_N|>\epsilon \right]\leq B \exp\left(-C N \left(\epsilon/M\right)^2\right)$.
-\end{enumerate}
-Then,
-$$
-    \mathbb{P}\left\{ \left| \frac{a_N}{b_N} \right| > \epsilon \right\} \leq B \exp{\left(-C N \left(\frac{\epsilon \beta}{2M} \right)^2 \right)} \eqsp.
-$$
-\end{lemma}
-\begin{proof}
-See \cite{douc:garivier:moulines:olsson:2011}.
+See appendix \ref{sec:append:proofs}
 \end{proof}
@@
 \subsection*{The SINE model} 
-The performances of the GRand PaRIS algorithm is first presented on the SINE model, where $(X_t)_{t\geq 0}$ is supposed to be the solution of the SDE: \cite{olsson:strojby:2011}. 
+The performance of the GRand PaRIS algorithm are first highlighted using the SINE model, where $(X_t)_{t\geq 0}$ is supposed to be the solution to: 
 \begin{equation}
-\rmd X_t = \sin \left(X_t-\theta\right)\rmd t + \rmd W_t,~~X_0=X_0. \label{eq:SINE:SDE}
+\rmd X_t = \sin \left(X_t-\theta\right)\rmd t + \rmd W_t,~~X_0=x_0. \label{eq:SINE:SDE}
 \end{equation}
-This simple model has no explicit transition density $q$, however, GPE estimators may be computed by simulating Brownian bridges.
-The process solution to \eqref{eq:SINE:SDE} is observed regularly at times $t_0=0,t_200=200$ through the observation process $(Y_k)_{0\leq k \leq 200}$
+This simple model has no explicit transition density, however GPE estimators may be computed by simulating Brownian bridges.
+The process solution to \eqref{eq:SINE:SDE} is observed regularly at times $t_0=0,\ldots,t_{200}=200$ through the observation process $(Y_k)_{0\leq k \leq 200}$:
 \begin{equation}
-Y_k = X_k + \varepsilon_k\label{eq:obs:SINE},~~\varepsilon\sim \mathcal{N}(0,1)\eqsp,
+Y_k = X_k + \varepsilon_k\label{eq:obs:SINE}\eqsp,
 \end{equation}
-In the example showed on Figure \ref{fig:obs:SINE}, we set $\theta=0$.\\
-\\
+where the $(\varepsilon)_{0\leq k \leq 200}$ are i.i.d. $\mathcal{N}(0,1)$.
+In the example showed on Figure \ref{fig:obs:SINE}, we set $\theta=0$.
+
 In order to evaluate the performance of our algorithm, we compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50).
@@
 Moreover, the E step boils down to the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.
-$\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite{olsson:strojby:2011} (Appendix B) based on 30 independent Monte Carlo simulations.
-The intermediate quantity of the EM algorithm is also estimated with our algorithm 30 times using $N=10000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, see Figure \ref{fig:res:SINE}.\\
-One can see that the GRand PaRIS algorithm outperforms the fixed lag methods (for any lag), as the empirical variance of the estimates is the smallest of all, and the bias is already neglictible for $N=400$, whereas small lags lead to strongly biased estimates for the Fixed lag methods, and unbiased estimates are at the cost of a large variance.
+$\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
+The intermediate quantity of the EM algorithm is also estimated with our algorithm 30 times using $N=10000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, see Figure \ref{fig:res:SINE}.
+
+The GRand PaRIS algorithm outperforms the fixed lag methods, for any value of the lag, as the empirical variance of the estimates is the smallest, and the bias is already negligible for $N=400$, whereas small lags lead to strongly biased estimates for the fixed lag method, and unbiased estimates are at the cost of a large variance.
 %To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated.
@@
 \end{equation}
-In this case, the conditions of the Exact Algorithm~2 defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} are satisfied, as for any $m \in \mathbb{R}$ there exists $\mathsf{U}(m)$ such that for all $x\ge m$, $\psi(x):=\alpha^2(x)+\alpha'(x) \leq \mathsf{U}_m$.
+In this case, the conditions of the Exact Algorithm~2 defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} are satisfied, as for any $m \in \mathbb{R}$ there exists $\mathsf{U}_m$ such that for all $x\ge m$, $\psi(x):=\alpha^2(x)+\alpha'(x) \leq \mathsf{U}_m$.
 Moreover, $\psi$ is lower bounded uniformly by  $\mathsf{L}$.
@@
 The parameters are given by $$\theta =(\kappa=0.1,\sigma=0.1,\gamma=1000,\sigma^2_{obs}=4)\eqsp.$$
-Again, we evaluate the intermediate quantity of the EM algorithm using the same amount of samples as for the SINE model.\\
-The results for the fixed lag technique are similar to the ones presented in \cite{olsson:strojby:2011} on the same model (Figure 1 of their article).
+Again, we evaluate the intermediate quantity of the EM algorithm using the same amount of samples as for the SINE model.
+
+The results for the fixed lag technique are similar to the ones presented in \cite[Figure 1]{olsson:strojby:2011} on the same model.
 For small lags, the variance of the estimates is small, but the estimation is highly biased.
-The bias rapidly decreases as the lag increases, together with a  great variance increase.
-The PaRIS algorithm outperforms the fixed lag smoother as it shows  a similar (vanishing) bias as the fixed lag for the largest lag and a smaller variance than the variance of the fixed lag when the lag is one.
+The bias rapidly decreases as the lag increases, together with a  great increase of variance.
+PaRIS algorithm outperforms the fixed lag smoother as it shows  a similar (vanishing) bias as the fixed lag for the largest lag and a smaller variance than the variance of the fixed lag when the lag is one.
 %It is worth noting here that, as pointed out in \cite{olsson:strojby:2011}, results of the fixed lag technique indicates a quite strong forgetting in the model, and the choice of the lag should take into account this forgetting property.
@@
 \includegraphics[width=0.8\textwidth]{obs_SINE}
-\caption{One realisation of the process $(X_t)_{t\geq 0}$, solution of the SDE \eqref{eq:Lamp:SINE} is represented in white.
-The observations of $Y$ at times $t_0=0,\dots,t_{200}=200$, satisfying equation \eqref{eq:obs:SINE},  are represented in black.}
+\caption{Process $X$ solution to the SDE \eqref{eq:Lamp:SINE} (white) and observations $Y$ satisfying equation \eqref{eq:obs:SINE} (black) at times $t_0=0,\dots,t_{200}=200$.}
 \label{fig:obs:SINE}
@@
 \includegraphics[width=0.8\textwidth]{obs_LG}
-\caption{One realisation of the process $(X_t)_{t\geq 0}$, solution of the SDE \eqref{eq:Lamp:LG} is represented in white.
-The observations of $Y$ at times $t_0=0,\dots,t_{200}=100$, satisfying equation \eqref{eq:obs:LG},  are represented in black.}
+\caption{Process $X$ solution to the SDE \eqref{eq:Lamp:LG} (white) and observations $Y$ satisfying equation \eqref{eq:obs:LG}  (black) at times $t_0=0,\dots,t_{200}=100$.}
 \label{fig:obs:LG}
@@
 \includegraphics[width=0.8\textwidth]{res_Estep_SINE}
-\caption{Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$ using observations of Figure \ref{fig:obs:SINE}, using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm for diffusion process proposed in this work, using 300 replicates.
+\caption{Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$ using observations of Figure \ref{fig:obs:SINE}, using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm for the diffusion process proposed in this work, using 300 replicates.
 The whiskers represent the extent of the 95\% central values.
@@
 \includegraphics[width=0.8\textwidth]{res_Estep_LG}
-\caption{Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$ using observations of Figure \ref{fig:obs:LG}, using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm for diffusion process proposed in this work, using 300 replicates.
+\caption{Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$ using observations of Figure \ref{fig:obs:LG}, using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm for the diffusion process proposed in this work, using 300 replicates.
 The whiskers represent the extent of the 95\% central values.
@@
 \section{Conclusions}
-We presented a new online smoother for partially observed differential equations.
+This paper presents a new online SMC smoother for partially observed differential equations.
 This algorithm relies on an acceptance reject procedure inspired from the recent PaRIS algorithm.
-The main result of the article is that the mechanism of this procedure remains valid when the transition density is approximated by a an unbiased positive estimator.
-The resulting algorithm to approximate smoothing expectations outperforms the existing fixed laged smoother for SDEs of \cite{olsson:strojby:2011}, as it is unbiased, and moreover shows a lower variance on two simulated examples.
-The proposed algorithm can be implemented for the class of models $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} with a linear (or at worse quadratic) complexity in $N$.
+The main result of the article for practical applications is that the mechanism of this procedure remains valid when the transition density is approximated by a an unbiased positive estimator.
+The proposed procedure outperforms the existing fixed lag smoother for SDE of \cite{olsson:strojby:2011}, as it does not introduce intrinsic non vanishing bias,  and numerical simulations highlight a better variance on two examples.
+It can be implemented for the class of models $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} with a linear complexity in $N$.
 % where $\beta\in\mathbb{R}^d$ is chosen by the user.
@@
 \[
-\mathcal{A}^k_{\ell} = \left\{U_\ell<\widehat{q}^{\Delta t_k}(\xi_{k-1}^{I_\ell},\xi_k^{i},\Lambda_{\ell})/\sigma_+\right\}\eqsp.
+\mathcal{A}^k_{\ell} = \left\{U_\ell<\widehat{q}_{k}(\xi_{k-1}^{J_\ell},\xi_k^{i},\zeta^{\ell}_{k})/\sigma_+\right\}\eqsp.
 \]
-
 Let $h$ be a function defined on $\{1,\ldots,N\}$,
 \begin{align*}
-\mathbb{E}\left[h(J^{i,j}_k)\middle| \mathcal{F}_k^N\right] & = \sum_{m\ge 1}\mathbb{E}\left[h(I_m)\1_{\tau=m}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \sum_{m\ge 1}h(m)\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\middle| \mathcal{F}_k^N\right]\right)\mathbb{E}\left[h(I_m)\1_{\mathcal{A}^k_{m}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[1-\frac{\widehat{q_k}(\xi_{k-1}^{I_\ell},\xi_k^{i};\zeta_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)\\
-&\hspace{5cm}\times\mathbb{E}\left[h(I_m)\frac{\widehat{q_k}(\xi_{k-1}^{I_m},\xi_k^{i};\zeta_{\ell})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \sum_{m\ge 1}\left(\mathbb{E}\left[1-\frac{q_k(\xi_{k-1}^{I_1},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\right)^{m-1}\mathbb{E}\left[h(I_m)\frac{q_k(\xi_{k-1}^{I_m},\xi_k^{1})}{\sigma_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \mathbb{E}\left[h(I_1)q_k(\xi_{k-1}^{I_1},\xi_k^{i})\middle| \mathcal{F}_k^N\right]/\mathbb{E}\left[q_k(\xi_{k-1}^{I_m},\xi_k^{i})\middle| \mathcal{F}_k^N\right]\eqsp,\\
+\mathbb{E}\left[h(J^{i,j}_k)\middle| \mathcal{F}_k^N\right] & = \sum_{m\ge 1}\mathbb{E}\left[h(J_m)\1_{\tau=m}\middle| \mathcal{F}_k^N\right]\eqsp,\\
+& = \sum_{m\ge 1}h(m)\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\middle| \mathcal{F}_k^N\right]\right)\mathbb{E}\left[h(J_m)\1_{\mathcal{A}^k_{m}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
+& = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[1-\frac{\widehat{q_k}(\xi_{k-1}^{J_\ell},\xi_k^{i};\zeta_k^{\ell})}{\hat{\sigma}_{+}}\middle| \mathcal{F}_k^N\right]\right)\\
+&\hspace{5cm}\times\mathbb{E}\left[h(I_m)\frac{\widehat{q_k}(\xi_{k-1}^{J_m},\xi_k^{i};\zeta_k^{\ell})}{\hat{\sigma}_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
+& = \sum_{m\ge 1}\left(\mathbb{E}\left[1-\frac{q_k(\xi_{k-1}^{J_1},\xi_k^{1})}{\hat{\sigma}_{+}}\middle| \mathcal{F}_k^N\right]\right)^{m-1}\mathbb{E}\left[h(J_m)\frac{q_k(\xi_{k-1}^{J_m},\xi_k^{1})}{\hat{\sigma}_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
+& = \mathbb{E}\left[h(J_1)q_k(\xi_{k-1}^{J_1},\xi_k^{i})\middle| \mathcal{F}_k^N\right]/\mathbb{E}\left[q_k(\xi_{k-1}^{J_m},\xi_k^{i})\middle| \mathcal{F}_k^N\right]\eqsp,\\
 & = \sum_{\ell=1}^N \frac{h(\ell)\omega_{k-1}^{\ell}q_k(\xi_{k-1}^{\ell},\xi_k^{i})}{\sum_{m=1}^N\omega_{k-1}^{m}q_k(\xi_{k-1}^{m},\xi_k^{i})}\eqsp,\\
@@
 \end{proof}
+
 \begin{proof}[Proof of Lemma \ref{lem:iid}]
@@
 
+\begin{proof}[Proof of Proposition~\ref{prop:exp:deviation}]
+The results is proved by induction.
+At time $k=0$, the result holds using that for all $1\le i \le N$, $\rho_0^i = 0$ and the convention $T_0[h_0] =0$.
+In addition, $\phi_0^N$ is a standard importance sampler estimator of $\phi_0$ with $\omega_0^i\le |\widehat{\omega}_0|_{\infty}$ so that for any bounded function $h$ on $\mathsf{X}$,
+\[
+\mathbb{P}\left(\left|\phi_0^N[h] - \phi_0\left[h\right]\right|\ge \varepsilon\right)\le b_0\exp\left(-c_0N\varepsilon^2\right)\eqsp.
+\]
+Assume the results holds for $k\ge 1$ and write
+\[
+\phi_{k+1}^N[\tau_{k+1}] - \phi_{k+1}\left[T_{k+1}[h_{k+1}]\right] = a_N/b_N\eqsp,
+\]
+where $a_N = N^{-1}\sum_{i=1}^N \widehat{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}[h_{k+1}]\right]\right)$ and $b_N =N^{-1}\sum_{i=1}^N \widehat{\omega}_{k+1}^i$.
+By Lemma~\ref{lem:iid}, the random variables $\{\widehat{\omega}_{k+1}^i\tau_{k+1}^i\}_{i=1}^N$ are independent conditionally on $\mathcal{F}_k^{N}$ and by H\ref{assum:boundalgo},
+\[
+\left|\widehat{\omega}_{k+1}^i \left(\tau_{k+1}^i - \phi_{k+1}\left[T_{k+1}[h_{k+1}]\right]\right)\right| \le 2|\widehat{\omega}_{k+1}|_{\infty}|H_{k+1}|_{\infty}\eqsp.
+\]
+Therefore, by Hoeffding inequality,
+\[
+\mathbb{P}\left(\left|a_N - \mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right) = \mathbb{E}\left[\mathbb{P}\left(\left|a_N - \mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\middle|\mathcal{F}_k^{N}\right)\right]\le 2\exp\left(-c_kN\varepsilon^2\right)\eqsp.
+\] 
+On the other hand,
+\[
+\mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right] = \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\Upsilon_k\right] \eqsp,
+\]
+where
+\[
+\Upsilon_k(x_k) = \int q_{k}(\cdot,x)g_{k+1}(x)\left(\tau_k(x_k) + h_{k+1}(x_k,x) - \phi_{k+1}\left[T_{k+1}[h_{k+1}]\right]\right)\rmd x\eqsp.
+\]
+By \cite[Lemma~11]{olsson:westerborn:2016}, $\phi_{k}\left[\Upsilon_k\right] = 0$ which implies by the induction assumption that 
+\[
+\mathbb{P}\left(\left|\mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right)\le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
+\]
+Then,
+\[
+\mathbb{P}\left(\left|a_N\right|\ge \varepsilon\right) \le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
+\] 
+Similarly, as $b_N \le |\widehat{\omega}_k|_{\infty}$, by Hoeffding inequality,
+\begin{multline*}
+\mathbb{P}\left(\left|b_N - \mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\right) \\
+= \mathbb{E}\left[\mathbb{P}\left(\left|b_N - \mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]\right|\ge \varepsilon\middle|\mathcal{F}_k^{N}\right)\right]\le 2\exp\left(-c_kN\varepsilon^2\right)\eqsp.
+\end{multline*}
+Note that
+\[
+\mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right] = \phi^N_{k}\left[\int q_{k}(\cdot,x)g_{k+1}(x)\rmd x\right]\eqsp.
+\]
+By  the induction assumption,
+\[
+\mathbb{P}\left(\left|\mathbb{E}\left[b_N\middle|\mathcal{F}_k^{N}\right]-\phi_k\left[\int q_{k}(\cdot,x)g_{k+1}(x)\rmd x\right]\right|\ge \varepsilon\right)\le b_k\exp\left(-c_kN\varepsilon^2\right)\eqsp.
+\]
+The proof is completed using Lemma~\ref{lem:hoeffding:ratio}.
+\end{proof}
+
+\begin{lemma}\label{lem:hoeffding:ratio}
+Assume that $a_N$, $b_N$, and $b$ are random variables defined on the same probability space such that there exist positive constants $\beta$, $B$, $C$, and $M$ satisfying
+\begin{enumerate}[(i)]
+    \item $|a_N/b_N|\leq M$, $\mathbb{P}$-a.s.\ and  $b \geq \beta$, $\mathbb{P}$-a.s.,
+    \item For all $\epsilon>0$ and all $N\geq1$, $\mathbb{P}\left[|b_N-b|>\epsilon \right]\leq B \exp\left(-C N \epsilon^2\right)$,
+    \item For all $\epsilon>0$ and all $N\geq1$, $\mathbb{P} \left[ |a_N|>\epsilon \right]\leq B \exp\left(-C N \left(\epsilon/M\right)^2\right)$.
+\end{enumerate}
+Then,
+$$
+    \mathbb{P}\left\{ \left| \frac{a_N}{b_N} \right| > \epsilon \right\} \leq B \exp{\left(-C N \left(\frac{\epsilon \beta}{2M} \right)^2 \right)} \eqsp.
+$$
+\end{lemma}
+\begin{proof}
+See \cite{douc:garivier:moulines:olsson:2011}.
+\end{proof}
+
 \bibliographystyle{plain}

### 2017-02-13 ParisEM.tex
Overall revision #2

@@
 \[
-\mathbb{E}\left[\widehat{\omega}^i_{k+1}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right] = \mathbb{E}\left[\frac{M^{-1}\sum_{m=1}^M \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}}, \xi^{i}_{k+1};\zeta^m_{k+1})g_{k+1}(\xi^{i}_{k+1})}{\vartheta_{k+1}(\xi^{I^{i}_{k+1}}_{k}) p_{k+1}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right]\eqsp.
+\mathbb{E}\left[\widehat{\omega}^i_{k+1}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right] = \mathbb{E}\left[\frac{ \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}}, \xi^{i}_{k+1};\zeta_{k+1})g_{k+1}(\xi^{i}_{k+1})}{\vartheta_{k+1}(\xi^{I^{i}_{k+1}}_{k}) p_{k+1}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right]\eqsp.
 %&= \mathbb{E}\left[\frac{g_{k+1}(\xi^{1}_{k+1})}{\vartheta_{k+1}(\xi^{I^{1}_{k+1}}_{k}) p_{k+1}(\xi_{k}^{I^{1}_{k+1}},\xi^{\ell}_{k+1})}\tau^{1}_{k+1}\widehat{q}_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{I^{1}_{k+1}},\xi^{1}_{k+1};\Lambda_{k+1})\middle| \mathcal{F}_k^{N}\right]\eqsp,
@@
  = \sum_{\ell=1}^N\frac{\omega_k^{\ell} \qk(\xi_{k}^{\ell}, \xi^{i}_{k+1}) \left(\tau^{\ell}_k + h_{k}(\xi_{k}^{\ell},\xi^{i}_{k+1})\right)}{\sum_{\ell'=1}^N\omega_k^{\ell'} \qk(\xi_{k}^{\ell'},\xi^{i}_{k+1})}\eqsp,\\
-&\mathbb{E} \left[ M^{-1}\sum_{m=1}^M \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta^m_{k+1}) \middle| \sigma \left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,I^{i}_{k+1}\}\right)\right]
+&\mathbb{E} \left[\widehat{\qk}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta_{k+1}) \middle| \sigma \left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,I^{i}_{k+1}\}\right)\right]
  = \qk(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})\eqsp.
 \end{align*}
-Since $\tau^{i}_{k+1}$ and $\{\zeta^m_{k+1}\}_{i=1}^M$ are independent conditionally to $\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^l, I_{k+1}^l\}_{l=1}^N\right)$:
+Since $\tau^{i}_{k+1}$ and $\zeta_{k+1}$ are independent conditionally to $\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^{\ell}, I_{k+1}^{\ell}\}_{\ell=1}^N\right)$:
 \begin{multline*}
-\mathbb{E}\left[\tau^{i}_{k+1}M^{-1}\sum_{m=1}^M \widehat{\qk} (\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta^m_{k+1})\middle|\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,I_{k+1}^i\} \right)\right]\\
+\mathbb{E}\left[\tau^{i}_{k+1} \widehat{\qk} (\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta_{k+1})\middle|\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,I_{k+1}^i\} \right)\right]\\
  = q_k(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})\sum_{\ell=1}^N\frac{\omega_k^{\ell} \qk (\xi_{k}^{\ell},\xi^{i}_{k+1})\left(\tau^{\ell}_k + h_{k}(\xi_{k}^{\ell},\xi^{i}_{k+1})\right)}{\sum_{\ell'=1}^N\omega_k^{\ell'} \qk (\xi_{k}^{\ell'},\xi^{i}_{k+1})}\eqsp.
@@
 \]
-Assume the results holds for $k\ge 1$ and write
+Assume the results holds for $k\ge 1$ and that $\vartheta_{k+1} = 1$ for simplicity. Write
 \[
@@
 \[
-\mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right] = \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\Upsilon_k\right] \eqsp,
+%\mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right] = \left(\phi^N_{k}[\vartheta_{k+1}]\right)^{-1}\phi^N_{k}\left[\Upsilon_k\right] \eqsp,
+\mathbb{E}\left[a_N\middle|\mathcal{F}_k^{N}\right] = \phi^N_{k}\left[\Upsilon_k\right] \eqsp,
 \]

### 2017-02-15 ParisEM.tex
Changing results with fair computational time comparison  between PaRIS and FL

@@
-\documentclass[12pt,draft]{article}
+\documentclass[12pt]{article}
 \usepackage[utf8]{inputenc}
@@
 On the r.h.s. of \eqref{eq:GPE1}, the product over $\kappa$ elements is bounded by 1, therefore, a sufficient condition to obtain an upper bound  $\hat{q}_k$ is that the function:
-\begin{align*}
-\rho_{\Delta_k}:\eqsp\mathbb{R}^d\times \mathbb{R}^d &\mapsto \mathbb{R}\\
-(x,y)&\mapsto \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}
-\end{align*}
+\begin{align}
+\rho_{\Delta_k}:\eqsp\mathbb{R}^d\times \mathbb{R}^d &\mapsto \mathbb{R}\nonumber\\
+(x,y)&\mapsto \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}\label{eq:rho:func}
+\end{align}
 is upper bounded almost surely by $\hat{\sigma}_+^k$.
@@
 %The performance of the algorithm is investigated on two known models based on diffusion process.
+We make numerical experiment on two well known SDE models, the sine model and the log-growth model.
+On both cases, the propagation law for the particle filtering is chosen to be a trade-off between the model's dynamics and the observation, namely, we chose:
+$$p_k(x_{k-1},x_k)\propto \tilde{q}_k(x_{k-1}, x_k)g_k(x_{k})$$
+where $\tilde{q}_k(x_{k-1},x_k)$ is the p.d.f.
+of Gaussian distibution with mean $\alpha(x_{k-1})\Delta_k$ and variance $\Delta_kI_d$, corresponding to the Euler approximation of equation \eqref{eq:target:sde}.
+In the cases presented here, the resulting distribution is therefore Gaussian with explicit mean and variance.\\
+In both cases we compare the performance of the GRand PaRIS algorithm with the fixed lag method of \cite{olsson:strojby:2011}.
+To achieve a fair comparison, the GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique is achieved using respectively $N=2000$ (for the sine model) and $N=1400$ (for the log growth model), so that both estimations are obtained in a similar computation time.
 
@@
 In the example showed on Figure \ref{fig:obs:SINE}, we set $\theta=0$.
-
+In that case, the $\rho_{\Delta_k}$ function defined in \eqref{eq:rho:func} can be upper bounded either globaly on $(x,y)$ or only on $y$, the resulting GRand PaRIS algorithm has therefore a linear complexity.\\
 In order to evaluate the performance of our algorithm, we compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50).
 We therefore compute the particle approximation of $\mathcal{Q}(\theta,\theta)$ for a sampled set of observations displayed in Figure~\ref{fig:obs:SINE}.
-This estimation is performed 300 times, using $N=400$ particles, with $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.
+
+This estimation is performed 300 times with $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.
 Moreover, the E step boils down to the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.
@@
 
-The GRand PaRIS algorithm outperforms the fixed lag methods, for any value of the lag, as the empirical variance of the estimates is the smallest, and the bias is already negligible for $N=400$, whereas small lags lead to strongly biased estimates for the fixed lag method, and unbiased estimates are at the cost of a large variance.
+The GRand PaRIS algorithm outperforms the fixed lag methods, for any value of the lag as the bias is already negligible for $N=400$, and with a lower variance than unbiased fixed lag estimates (i.e., in this case, lags larer than 5).
+Small lags lead to strongly biased estimates for the fixed lag method, and unbiased estimates are at the cost of a large variance.
+It is worth noting here that the small bias observed when the lag is equal to five is observation dependent, and no general rule can be made on which lag will lead to a neglictible bias.
 %To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated.
@@
 The parameters are given by $$\theta =(\kappa=0.1,\sigma=0.1,\gamma=1000,\sigma^2_{obs}=4)\eqsp.$$
+In that case, the $\rho_{\Delta_k}$ function defined in \eqref{eq:rho:func} can be upper bounded as a function of $y$ when $x\in \{\xi_k^1,\dots, \xi_k^N\}$, the resulting GRand PaRIS algorithm has therefore a linear complexity.\\
 Again, we evaluate the intermediate quantity of the EM algorithm using the same amount of samples as for the SINE model.
@@
 The bias rapidly decreases as the lag increases, together with a  great increase of variance.
-PaRIS algorithm outperforms the fixed lag smoother as it shows  a similar (vanishing) bias as the fixed lag for the largest lag and a smaller variance than the variance of the fixed lag when the lag is one.
+PaRIS algorithm outperforms the fixed lag smoother as it shows  a similar (vanishing) bias as the fixed lag for the largest lag and a smaller variance than the non biased fixed lags estimates.
 %It is worth noting here that, as pointed out in \cite{olsson:strojby:2011}, results of the fixed lag technique indicates a quite strong forgetting in the model, and the choice of the lag should take into account this forgetting property.

### 2017-02-18 ParisEM.tex
addition of a global numerical study

@@
 \usepackage{graphicx}
+\usepackage{caption}
+\usepackage{subcaption}
 \graphicspath{{/home/gloaguen/Bureau/Travaux_agro/figures_article_oziris/}}
@@
 \begin{equation}
-\rmd X_t = \sin \left(X_t-\theta\right)\rmd t + \rmd W_t,~~X_0=x_0. \label{eq:SINE:SDE}
+\rmd X_t = \sin \left(X_t-\theta\right)\rmd t + \rmd W_t,~~X_0=x_0. \label{eq:Lamp:SINE}
 \end{equation}
 This simple model has no explicit transition density, however GPE estimators may be computed by simulating Brownian bridges.
-The process solution to \eqref{eq:SINE:SDE} is observed regularly at times $t_0=0,\ldots,t_{200}=200$ through the observation process $(Y_k)_{0\leq k \leq 200}$:
+The process solution to \eqref{eq:SINE:SDE} is observed regularly at times $t_0=0,\ldots,t_{100}=50$ through the observation process $(Y_k)_{0\leq k \leq 100}$:
 \begin{equation}
@@
 \end{equation}
-where the $(\varepsilon)_{0\leq k \leq 200}$ are i.i.d. $\mathcal{N}(0,1)$.
+where the $(\varepsilon)_{0\leq k \leq 100}$ are i.i.d. $\mathcal{N}(0,1)$.
 In the example showed on Figure \ref{fig:obs:SINE}, we set $\theta=0$.
@@
 In order to evaluate the performance of our algorithm, we compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50).
-We therefore compute the particle approximation of $\mathcal{Q}(\theta,\theta)$ for a sampled set of observations displayed in Figure~\ref{fig:obs:SINE}.
+The particle approximation of $\mathcal{Q}(\theta,\theta)$ is computed using each algorithm.
+An example of 200 estimates for a sampled set of observations is displayed in Figure~\ref{fig:obs:SINE}.\\
 
@@
 $\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
-The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=10000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, see Figure \ref{fig:res:SINE}.
+The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, see Figure \ref{fig:res:SINE}.\\
+This experiment was reproduced on a hundred different simulated data, in order study the distribution of the empirical absolute relative bias and coefficient of variations for all methods.
+The results are shown on Figure \ref{fig:mult:SINE}.
 
-The GRand PaRIS algorithm outperforms the fixed lag methods, for any value of the lag as the bias is already negligible for $N=400$, and with a lower variance than unbiased fixed lag estimates (i.e., in this case, lags larer than 5).
+The GRand PaRIS algorithm outperforms the fixed lag methods, for any value of the lag as the bias is the lowest already negligible for $N=400$, and with a lower variance than unbiased fixed lag estimates (i.e., in this case, lags larger than 10).
 Small lags lead to strongly biased estimates for the fixed lag method, and unbiased estimates are at the cost of a large variance.
-It is worth noting here that the small bias observed when the lag is equal to five is observation dependent, and no general rule can be made on which lag will lead to a neglictible bias.
+It is worth noting here that the lag for which the bias is small is model dependant, and that the GRand PaRIS algorithm do not require any knowledge of the forgetting properties of the model.
 %To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated.
@@
 The bias rapidly decreases as the lag increases, together with a  great increase of variance.
-PaRIS algorithm outperforms the fixed lag smoother as it shows  a similar (vanishing) bias as the fixed lag for the largest lag and a smaller variance than the non biased fixed lags estimates.
+Again, the GRand PaRIS algorithm outperforms the fixed lag smoother as it shows  a similar (vanishing) bias as the fixed lag for the largest lag and a smaller variance than the non biased fixed lags estimates.
 %It is worth noting here that, as pointed out in \cite{olsson:strojby:2011}, results of the fixed lag technique indicates a quite strong forgetting in the model, and the choice of the lag should take into account this forgetting property.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{obs_SINE}
-\caption{Process $X$ solution to the SDE \eqref{eq:Lamp:SINE} (white) and observations $Y$ satisfying equation \eqref{eq:obs:SINE} (black) at times $t_0=0,\dots,t_{200}=200$.}
-\label{fig:obs:SINE}
-\end{figure}
-\begin{figure}[p]
-\centering
-\includegraphics[width=0.8\textwidth]{obs_LG}
-\caption{Process $X$ solution to the SDE \eqref{eq:Lamp:LG} (white) and observations $Y$ satisfying equation \eqref{eq:obs:LG}  (black) at times $t_0=0,\dots,t_{200}=100$.}
-\label{fig:obs:LG}
-\end{figure}
-\begin{figure}[p]
-\centering
-\includegraphics[width=0.8\textwidth]{res_Estep_SINE}
-\caption{Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$ using observations of Figure \ref{fig:obs:SINE}, using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm for the diffusion process proposed in this work, using 300 replicates.
+\begin{subfigure}{0.49\textwidth}
+\includegraphics[width=\textwidth]{obs_SINE}
+\end{subfigure}
+\begin{subfigure}{0.49\textwidth}
+\includegraphics[width=\textwidth]{res_Estep_SINE}
+\end{subfigure}
+\caption{\textit{Left: }Observations of the process $X$ solution to the SDE \eqref{eq:Lamp:SINE} (black) and observations $Y$ satisfying equation \eqref{eq:obs:SINE} (white) at times $t_0=0,\dots,t_{100}=50$.
+\textit{Right: } Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$  using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm for the diffusion process proposed in this work, using 300 replicates.
 The whiskers represent the extent of the 95\% central values.
@@
 \centering
-\includegraphics[width=0.8\textwidth]{res_Estep_LG}
-\caption{Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$ using observations of Figure \ref{fig:obs:LG}, using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm for the diffusion process proposed in this work, using 300 replicates.
+\begin{subfigure}{0.49\textwidth}
+\includegraphics[width=\textwidth]{res_mult_SINE_bias}
+\end{subfigure}
+\begin{subfigure}{0.49\textwidth}
+\includegraphics[width=\textwidth]{res_mult_SINE_cv}
+\end{subfigure}
+\caption{Evolution of the empirical relative bias for each method in the case of the log-growth model.
+100 replcations of the experience showed on Figure \ref{fig:res:SINE} are made.
+The empirical abolute relative bias is computed for each one.
+The whiskers show the 95\% central values and the dot shows the empirical median.
+All estimations are performed with an equal computation time.}
+\label{fig:mult:SINE}
+\end{figure}
+\begin{figure}[p]
+\centering
+\begin{subfigure}{0.49\textwidth}
+\includegraphics[width=\textwidth]{obs_LG}
+\end{subfigure}
+\begin{subfigure}{0.49\textwidth}
+\includegraphics[width=\textwidth]{res_Estep_LG}
+\end{subfigure}
+\caption{\textit{Left: }Observations of the process $X$ solution to the SDE \eqref{eq:Lamp:LG} (black) and observations $Y$ satisfying equation \eqref{eq:obs:LG} (white) at times $t_0=0,\dots,t_{100}=50$.
+\textit{Right: } Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$  using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm for the diffusion process proposed in this work, using 300 replicates.
 The whiskers represent the extent of the 95\% central values.
@@
 \label{fig:res:LG}
+\end{figure}
+\begin{figure}[p]
+\centering
+\begin{subfigure}{0.49\textwidth}
+\includegraphics[width=\textwidth]{res_mult_LG_bias}
+\end{subfigure}
+\begin{subfigure}{0.49\textwidth}
+\includegraphics[width=\textwidth]{res_mult_LG_cv}
+\end{subfigure}
+\caption{Empirical distribution of the absolute relative bias and te coefficient of variation of each estimation method.
+100 replications of the experience showed on figure \ref{fig:res:LG} are made.
+The whiskers show the 95\% central values and the dot shows the empirical median.
+All estimations are performed with an equal computation time.}
+\label{fig:mult:LG}
 \end{figure}

### 2017-02-21 ParisEM.tex
Merge with Pierre

@@
 the computation of the posterior distributions of sequences of hidden states given observations.
-Filtering refers to the estimation of the distributions of the hidden state $X_k$ given the observations $(Y_0,\ldots,Y_k)$, while smoothing stands for the estimation of the distribution of sequence of states $(X_{k},\ldots,X_{p})$ given observations $(Y_{0},\ldots,Y_{\ell})$ with $0\le k\le p<\ell \le n$.
+The filtering problem refers to the estimation, for each $0\le \le n$,  of the distributions of the hidden state $X_k$ given the observations $(Y_0,\ldots,Y_k)$.
+Smoothing stands for the estimation of the distributions of sequence of states $(X_{k},\ldots,X_{p})$ given observations $(Y_{0},\ldots,Y_{\ell})$ with $0\le k\le p \le \ell \le n$.
 These posterior distributions are crucial to compute maximum likelihood estimators of the unknown parameters defining the model using the observations $(Y_0,\ldots,Y_n)$ only.
@@
 Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$.
+Then, a backward pass is performed to compute new weights (FFBS) or sample trajectories (FFBSi) in order to approximate the smoothing distributions.
 %In the case of partially observed SDE, this forward pass may be performed with the SMC steps combined with the GPE of \cite{fearnhead:papaspiliopoulos:roberts:2008}. 
-Then, the backward pass of the FFBS algorithm keeps all the particles sampled during the forward pass and computes new weights to take into account the information brought by all the observations from time $n$ to time $0$.
+%Then, the backward pass of the FFBS algorithm keeps all the particles sampled during the forward pass and computes new weights to take into account the information brought by all the observations from time $n$ to time $0$.
 Instead of computing new weights, the FFBSi  algorithm samples independently backward in time particle trajectories using the particles and weights produced by the forward pass.
@@
 The paper is organized as follows.
-Section~\ref{sec:rwparis} describes the random weight PaRIS algorithm to approximate smoothed additive functionals using unbiased estimates of the transition density of the hidden states.
-Then, Section~\ref{sec:PaRIS:SDE} details the application of this algorithm when the transition density may be approximated using a GPE. 
+Section~\ref{sec:rwparis} describes the random weight PaRIS algorithm to approximate smoothed additive functionals using unbiased estimates of the transition density of the hidden states and details the application of this algorithm when the transition density may be approximated using a GPE.
 In Section~\ref{sec:convergence}, classical convergence results for SMC smoothers are extended to the setting of this paper and illustrated with numerical experiments in Section~\ref{sec:exp}. 
@@
 A popular choice of $A$  is $A = \log \pi/2$ leading to a Langevin diffusion which may be used to approximate a given target distribution $\pi$ on $\mathbb{R}^d$.
-See for instance \cite{roberts:tweedie:1996} for conditions under which this diffusion converges exponentially quickly to $\pi$.
+See for instance \cite{roberts:tweedie:1996} for conditions under which this diffusion converges  at exponential rate to $\pi$.
 SDE where the drift is given by the gradient of a potential function have been widely used in ecology for instance, see \cite{brillinger:et:al:2011,harris:blackwell:2013,preisler:et:al:2013} and the references therein.
 The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{k=0,\dots,n}$ in $\mathbb{R}^m$.
-For all $0\le k \le n$, the distribution of $Y_k$ given $X_k:= X_{t_k}$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,Y_k) = g_k(X_k)$. 
+For all $0\le k \le n$, the distribution of $Y_k$ given $X_k:= X_{t_k}$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,\cdot) = g_k(X_k)$. 
 The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
@@
 \begin{equation}
-\phi_{0:n|n}[h] = \phi_n[T_n[h]]\eqsp,\;\mbox{where}\; T_n[h](X_n) = \E\left[h(X_{0:n})\vert X_n,~Y_{0:n}\right]\eqsp.\label{eq:FFbsm:equality}
+\phi_{0:n|n}[h] = \phi_n[T_n[h]]\eqsp,\;\mbox{where}\; T_n[h](X_n) = \E\left[h(X_{0:n})\vert X_n,Y_{0:n}\right]\eqsp.\label{eq:FFbsm:equality}
 \end{equation}
@@
 \end{equation}
-Then, instead of computing exactly these approximations which would lead to a complexity growing quadratically with $N$ because of the normalizing constant in \eqref{eq:Lambda}, PaRIS algorithm samples particles in the set $\{\xi^j_{k-1}\}_{j=1}^N$ with probabilities $\Lambda_{k}^N(i,\cdot)$ to approximate the expectation \eqref{eq:Tk} and produce $\tau^i_k$.
+Computing exactly these approximations would lead to a complexity growing quadratically with $N$ because of the normalizing constant in \eqref{eq:Lambda}.
+Therefore, PaRIS algorithm samples particles in the set $\{\xi^j_{k-1}\}_{j=1}^N$ with probabilities $\Lambda_{k}^N(i,\cdot)$ to approximate the expectation \eqref{eq:Tk} and produce $\tau^i_k$.
 Choosing $\tilde{N}\ge 1$, at each time step $0\le k \le {n-1}$ these statistics are updated according to the following steps.
@@
 \[
-\phi_{0:n\vert n}^N = \frac{1}{\Omega_n^N}\sum_{i=1}^N \omega^{i}_n \tau_n^i\eqsp.
+\phi_{0:n\vert n}^N[\tau_n] = \frac{1}{\Omega_n^N}\sum_{i=1}^N \omega^{i}_n \tau_n^i\eqsp.
 \] 
-As proved in \cite{olsson:westerborn:2016}, the algorithm is asymptotically consistent (as $N$ goes to infinity) for any fixed precision parameter $\tilde N$.
+As proved in \cite{olsson:westerborn:2016}, the algorithm is asymptotically consistent (as $N$ goes to infinity) for any precision parameter $\tilde N$.
 However, there is a significant qualitative difference between the cases $\tilde{N} = 1$ and $\tilde{N} \geq 2$.
@@
 where $\mathbb{W}^{x,y,\Delta_k}$ is the law of Brownian bridge starting at $x$ at 0 and hitting $y$ at $\Delta_k$, $(\mw_t)_{0\leq t \leq \Delta_k}$ is such a Brownian bridge, $\varphi_{\Delta_k}(x,y)$ is the p.d.f.
-of a normal distribution with mean $x$ and variance $\Delta_k$, evaluated in $y$ and $\phi:\mathbb{R}^d\to\mathbb{R}$ is defined as:
+of a normal distribution with mean $x$ and variance $\Delta_k$, evaluated at $y$ and $\phi:\mathbb{R}^d\to\mathbb{R}$ is defined as:
 \[
@@
 with $\triangle$ the Laplace operator.
-Suppose there exist random variables $\Lo_\mw$ and $\U_\mw$ such that for all $0\leq s \leq \Delta_k$, $\Lo_\mw \leq \phi(\mw_s)\leq \U_\mw$.
+Assume that there exist random variables $\Lo_\mw$ and $\U_\mw$ such that for all $0\leq s \leq \Delta_k$, $\Lo_\mw \leq \phi(\mw_s)\leq \U_\mw$.
 Let $\kappa$ be a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and $(U_j)_{1\le j\le \kappa}$ be independent uniform random variables on $[0,\Delta_k]$, and $
@@
 \end{multline}
-Interesting choices of $\mu$ are discussed in \cite{fearnhead:papaspiliopoulos:roberts:2008} and we focus here on the so called GPE-1, when $\mu$ is a Poisson distribution with intensity $(\U_\mw-\Lo_\mw)\Delta_k$.
+Interesting choices of $\mu$ are discussed in \cite{fearnhead:papaspiliopoulos:roberts:2008} and we focus here on the so called GPE-1, where $\mu$ is a Poisson distribution with intensity $(\U_\mw-\Lo_\mw)\Delta_k$.
 In that case, the estimator \eqref{eq:unbiased:q} becomes:
 \begin{equation}
-\widehat{q}_{\Delta_k}(x,y;\zeta_k) = \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}\prod_{j=1}^{\kappa}\frac{\U_\mw-\phi(\mw_{U_j})}{\U_\mw-\Lo_\mw}\eqsp.\label{eq:GPE1}
+\widehat{q}_{k}(x,y;\zeta_k) = \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}\prod_{j=1}^{\kappa}\frac{\U_\mw-\phi(\mw_{U_j})}{\U_\mw-\Lo_\mw}\eqsp.\label{eq:GPE1}
 \end{equation}
@@
 as long as $\Lo_\mw$ is bounded almost surely.
-Actually, this condition is always satisfied for modelsin the domains $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, i.e.
+This might increase significantly the acceptance rate of the algorithm, and therefore reduce the number of drawings of the random variable $\zeta_k$, which has a much higher cost than the computation of $\rho$, as it requires simulation of Brownian Bridges.
+It is worth noting here that this latter option can also avoid numerical optimization if no analytical version of $\hat{\sigma}_+^k$ is available.
+In the example of the SINE model presented below, for $N=400$ particles, the algorithm with linear complexity  has a 40\% higher computation time than the algorithm with a quadratic complexity.
+Actually, this condition is always satisfied for models in the domains $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, i.e.
 domains for which the exact algorithms EA1 and EA2 can be used.
-\subsection*{Note on the complexity and practical implementation}
 
-As said above, as long as one can define a $\sigma_+^k$ such that
-$$ \forall\eqsp 0\leq k\leq n-1,\eqsp 1\leq i\leq N,\eqsp y\in\mathbb{R}^d,\eqsp \hat{q}_k(\xi_k^i,y,\zeta_k)<\hat{\sigma}_+^k \text{ a.s.,}$$ then the resulting GRand Paris algorithm has a linear complexity in $N$.\\
-However, it is worth noting here that for a reasonable amount of particles, it might be more efficient in terms of computation time to compute, for all $i$, the bound given by \eqref{eq:quadr:bound}.
+%\subsection*{Note on the complexity and practical implementation}
+%As said above, as long as one can define a $\sigma_+^k$ such that
+%$$ \forall\eqsp 0\leq k\leq n-1,\eqsp 1\leq i\leq N,\eqsp y\in\mathbb{R}^d,\eqsp \hat{q}_k(\xi_k^i,y,\zeta_k)<\hat{\sigma}_+^k \text{ a.s.,}$$ then the resulting GRand Paris algorithm has a linear complexity in $N$.\\
+%However, it is worth noting here that for a reasonable amount of particles, it might be more efficient in terms of computation time to compute, for all $i$, the bound given by \eqref{eq:quadr:bound}.
 Indeed, this bound requires, at each time step $k$, $N^2$ computation of the function $\rho$ given above.
@@
 It is worth noting here that this latter option can also avoid numerical optimization if no analytical version of $\hat{\sigma}_+^k$ is available.\\
-In the example of the SINE model presented below, for $N=400$ particles, the linear complexity algorithm has a 40\% higher computation time than the algorithm with a quadratic complexity.
+%In the example of the SINE model presented below, for $N=400$ particles, the linear complexity algorithm has a 40\% higher computation time than the algorithm with a quadratic complexity.
 In practice, in addition to $n$ and $N$, the computation time will increase with
-\begin{itemize}
-\item $\tilde{N}$ the number of retrospective sampling;
-\item $M$ the number of Monte Carlo samples to obtain $\hat{q}$;
-\item $\Delta$ the acquisition time step;
-\item The expected value of $\U_\mw-\Lo_\mw$.
-\end{itemize}
+%\begin{itemize}
+%\item $\tilde{N}$ the number of retrospective sampling;
+%\item $M$ the number of Monte Carlo samples to obtain $\hat{q}$;
+%\item $\Delta$ the acquisition time step;
+%\item The expected value of $\U_\mw-\Lo_\mw$.
+%\end{itemize}
 
@@
 %The performance of the algorithm is investigated on two known models based on diffusion process.
-We make numerical experiment on two well known SDE models, the sine model and the log-growth model.
-On both cases, the propagation law for the particle filtering is chosen to be a trade-off between the model's dynamics and the observation, namely, we chose:
-$$p_k(x_{k-1},x_k)\propto \tilde{q}_k(x_{k-1}, x_k)g_k(x_{k})$$
+This section investigates the performance of the proposed algorithm with the sine and log-growth models.
+In both cases, the proposal distribution $p_k$ is chosen as the following approximation of the optimal filter (or the fully adapted particle filter in the terminology of \cite{pitt:shephard:1999}):
+$$p_k(x_{k-1},x_k)\propto \tilde{q}_k(x_{k-1}, x_k)g_k(x_{k})\eqsp,$$
 where $\tilde{q}_k(x_{k-1},x_k)$ is the p.d.f.
-of Gaussian distibution with mean $\alpha(x_{k-1})\Delta_k$ and variance $\Delta_kI_d$, corresponding to the Euler approximation of equation \eqref{eq:target:sde}.
-In the cases presented here, the resulting distribution is therefore Gaussian with explicit mean and variance.\\
-In both cases we compare the performance of the GRand PaRIS algorithm with the fixed lag method of \cite{olsson:strojby:2011}.
-To achieve a fair comparison, the GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique is achieved using respectively $N=2000$ (for the sine model) and $N=1400$ (for the log growth model), so that both estimations are obtained in a similar computation time.
+of Gaussian distibution with mean $\alpha(x_{k-1})\Delta_k$ and variance $\Delta_kI_d$, i.e.
+the Euler approximation of equation \eqref{eq:target:sde}.
+As the observation model is linear and Gaussian, the proposal distribution is therefore Gaussian with explicit mean and variance.
+The GRand PaRIS algorithm is compared with the fixed lag method of \cite{olsson:strojby:2011}.
+The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=2000$ for the sine model and $N=1400$ for the log growth model so that both estimations have similar computational times.
 
@@
 \begin{equation}
-\rmd X_t = \sin \left(X_t-\theta\right)\rmd t + \rmd W_t,~~X_0=x_0. \label{eq:Lamp:SINE}
+\rmd X_t = \sin \left(X_t-\theta\right)\rmd t + \rmd W_t,~~X_0=x_0\eqsp. \label{eq:Lamp:SINE}
 \end{equation}
 This simple model has no explicit transition density, however GPE estimators may be computed by simulating Brownian bridges.
-The process solution to \eqref{eq:SINE:SDE} is observed regularly at times $t_0=0,\ldots,t_{100}=50$ through the observation process $(Y_k)_{0\leq k \leq 100}$:
+The process solution to \eqref{eq:Lamp:SINE} is observed regularly at times $t_0=0,\ldots,t_{100}=50$ through the observation process $(Y_k)_{0\leq k \leq 100}$:
 \begin{equation}
@@
 where the $(\varepsilon)_{0\leq k \leq 100}$ are i.i.d. $\mathcal{N}(0,1)$.
-In the example showed on Figure \ref{fig:obs:SINE}, we set $\theta=0$.
-In that case, the $\rho_{\Delta_k}$ function defined in \eqref{eq:rho:func} can be upper bounded either globaly on $(x,y)$ or only on $y$, the resulting GRand PaRIS algorithm has therefore a linear complexity.\\
+In the example displayed on Figure \ref{fig:res:SINE}, we set $\theta=0$.
+In that case, the function $\rho_{\Delta_k}$ defined in \eqref{eq:rho:func} can be upper bounded either on $(x,y)$ or only on $y$, the resulting GRand PaRIS algorithm has therefore a linear complexity.
+
 In order to evaluate the performance of our algorithm, we compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50).
 The particle approximation of $\mathcal{Q}(\theta,\theta)$ is computed using each algorithm.
-An example of 200 estimates for a sampled set of observations is displayed in Figure~\ref{fig:obs:SINE}.\\
+An example of 200 estimates for a sampled set of observations is displayed in Figure~\ref{fig:obs:SINE}.
 
@@
 $\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
-The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, see Figure \ref{fig:res:SINE}.\\
+The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, see Figure \ref{fig:res:SINE}.
 This experiment was reproduced on a hundred different simulated data, in order study the distribution of the empirical absolute relative bias and coefficient of variations for all methods.
@@
 
-The GRand PaRIS algorithm outperforms the fixed lag methods, for any value of the lag as the bias is the lowest already negligible for $N=400$, and with a lower variance than unbiased fixed lag estimates (i.e., in this case, lags larger than 10).
+The GRand PaRIS algorithm outperforms the fixed lag methods for any value of the lag as the bias is the lowest (it is already negligible for $N=400$,) and with a lower variance than fixed lag estimates with negligible bias (i.e., in this case, lags larger than 10).
 Small lags lead to strongly biased estimates for the fixed lag method, and unbiased estimates are at the cost of a large variance.
-It is worth noting here that the lag for which the bias is small is model dependant, and that the GRand PaRIS algorithm do not require any knowledge of the forgetting properties of the model.
+It is worth noting here that the lag for which the bias is small is model dependent.
 %To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated.
@@
 The parameters are given by $$\theta =(\kappa=0.1,\sigma=0.1,\gamma=1000,\sigma^2_{obs}=4)\eqsp.$$
-In that case, the $\rho_{\Delta_k}$ function defined in \eqref{eq:rho:func} can be upper bounded as a function of $y$ when $x\in \{\xi_k^1,\dots, \xi_k^N\}$, the resulting GRand PaRIS algorithm has therefore a linear complexity.\\
-Again, we evaluate the intermediate quantity of the EM algorithm using the same amount of samples as for the SINE model.
+In that case, the $\rho_{\Delta_k}$ function defined in \eqref{eq:rho:func} can be upper bounded as a function of $y$ when $x\in \{\xi_k^1,\dots, \xi_k^N\}$, the resulting GRand PaRIS algorithm has therefore a linear complexity.
+The intermediate quantity of the EM algorithm is evaluated  as for the SINE model.
 
@@
 The bias rapidly decreases as the lag increases, together with a  great increase of variance.
-Again, the GRand PaRIS algorithm outperforms the fixed lag smoother as it shows  a similar (vanishing) bias as the fixed lag for the largest lag and a smaller variance than the non biased fixed lags estimates.
+Again, the GRand PaRIS algorithm outperforms the fixed lag smoother as it shows  a similar (vanishing) bias as the fixed lag for the largest lag and a smaller variance than the fixed lags estimates with negligible bias.
 %It is worth noting here that, as pointed out in \cite{olsson:strojby:2011}, results of the fixed lag technique indicates a quite strong forgetting in the model, and the choice of the lag should take into account this forgetting property.
@@
 \caption{\textit{Left: }Observations of the process $X$ solution to the SDE \eqref{eq:Lamp:SINE} (black) and observations $Y$ satisfying equation \eqref{eq:obs:SINE} (white) at times $t_0=0,\dots,t_{100}=50$.
-\textit{Right: } Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$  using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm for the diffusion process proposed in this work, using 300 replicates.
+\textit{Right: } Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$  using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm using 300 replicates.
 The whiskers represent the extent of the 95\% central values.
 The dot represents the empirical mean over the 300 replicates.
-The dotted line shows the reference value, computed using the unbiased GRand PaRIS algorithm with $N=10000$ particles.}
+The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=10000$ particles.}
 \label{fig:res:SINE}
@@
 \caption{\textit{Left: }Observations of the process $X$ solution to the SDE \eqref{eq:Lamp:LG} (black) and observations $Y$ satisfying equation \eqref{eq:obs:LG} (white) at times $t_0=0,\dots,t_{100}=50$.
-\textit{Right: } Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$  using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm for the diffusion process proposed in this work, using 300 replicates.
+\textit{Right: } Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$  using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm using 300 replicates.
 The whiskers represent the extent of the 95\% central values.
 The dot represents the empirical mean over the 300 replicates.
-The dotted line shows the reference value, computed using the unbiased GRand PaRIS algorithm with $N=10000$ particles.}
+The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=10000$ particles.}
 \label{fig:res:LG}

### 2017-02-21 ParisEM.tex
Final revision #4

@@
 \[
-\phi(x) =\|\alpha(x)\|^2  + \triangle A(x)\eqsp,
+\phi(x) =\left(\|\alpha(x)\|^2  + \triangle A(x)\right)/2\eqsp,
 \]
@@
 In order to evaluate the performance of our algorithm, we compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50).
-The particle approximation of $\mathcal{Q}(\theta,\theta)$ is computed using each algorithm.
-An example of 200 estimates for a sampled set of observations is displayed in Figure~\ref{fig:obs:SINE}.
-
-This estimation is performed 300 times with $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.
-Moreover, the E step boils down to the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.
+The particle approximation of $\mathcal{Q}(\theta,\theta)$ is computed using each algorithm, see Figure~\ref{fig:res:SINE}.
+This estimation is performed 200 times with $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.
+Moreover, the E step requires the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.
 $\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
 The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, see Figure \ref{fig:res:SINE}.
+
+\textcolor{red}{Detailler plus clairement ce que dit la figure 2 ici (donner clairement les estimateurs voire m\^eme les formules pour le biais et la variance relatifs)}
 This experiment was reproduced on a hundred different simulated data, in order study the distribution of the empirical absolute relative bias and coefficient of variations for all methods.
@@
 
-The GRand PaRIS algorithm outperforms the fixed lag methods for any value of the lag as the bias is the lowest (it is already negligible for $N=400$,) and with a lower variance than fixed lag estimates with negligible bias (i.e., in this case, lags larger than 10).
+The GRand PaRIS algorithm outperforms the fixed lag methods for any value of the lag as the bias is the lowest (it is already negligible for $N=400$) and with a lower variance than fixed lag estimates with negligible bias (i.e., in this case, lags larger than 10).
 Small lags lead to strongly biased estimates for the fixed lag method, and unbiased estimates are at the cost of a large variance.
@@
 In that case, the $\rho_{\Delta_k}$ function defined in \eqref{eq:rho:func} can be upper bounded as a function of $y$ when $x\in \{\xi_k^1,\dots, \xi_k^N\}$, the resulting GRand PaRIS algorithm has therefore a linear complexity.
-The intermediate quantity of the EM algorithm is evaluated  as for the SINE model.
+The intermediate quantity of the EM algorithm is evaluated  as for the SINE model, see Figures~\ref{fig:res:LG} and~\ref{fig:mult:LG}.
 
@@
 \end{subfigure}
-\caption{\textit{Left: }Observations of the process $X$ solution to the SDE \eqref{eq:Lamp:SINE} (black) and observations $Y$ satisfying equation \eqref{eq:obs:SINE} (white) at times $t_0=0,\dots,t_{100}=50$.
-\textit{Right: } Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$  using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm using 300 replicates.
+\caption{\textit{Left: }Process $X$ solution to the SDE \eqref{eq:Lamp:SINE} (black) and observations $Y$ satisfying equation \eqref{eq:obs:SINE} (white) at times $t_0=0,\dots,t_{100}=50$.
+\textit{Right: } Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$  using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm using 200 replicates.
 The whiskers represent the extent of the 95\% central values.
-The dot represents the empirical mean over the 300 replicates.
-The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=10000$ particles.}
+The dot represents the empirical mean over the 200 replicates.
+The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=5000$ particles.}
 \label{fig:res:SINE}
@@
 \caption{Evolution of the empirical relative bias for each method in the case of the log-growth model.
-100 replcations of the experience showed on Figure \ref{fig:res:SINE} are made.
+100 replications of the experience showed on Figure \ref{fig:res:SINE} are made.
 The empirical abolute relative bias is computed for each one.
@@
 \end{subfigure}
-\caption{\textit{Left: }Observations of the process $X$ solution to the SDE \eqref{eq:Lamp:LG} (black) and observations $Y$ satisfying equation \eqref{eq:obs:LG} (white) at times $t_0=0,\dots,t_{100}=50$.
-\textit{Right: } Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$  using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm using 300 replicates.
+\caption{\textit{Left: }Process $X$ solution to the SDE \eqref{eq:Lamp:LG} (black) and observations $Y$ satisfying equation \eqref{eq:obs:LG} (white) at times $t_0=0,\dots,t_{100}=50$.
+\textit{Right: } Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$  using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm using 200 replicates.
 The whiskers represent the extent of the 95\% central values.
-The dot represents the empirical mean over the 300 replicates.
-The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=10000$ particles.}
+The dot represents the empirical mean over the 200 replicates.
+The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=5000$ particles.}
 \label{fig:res:LG}
@@
 \end{subfigure}
-\caption{Empirical distribution of the absolute relative bias and te coefficient of variation of each estimation method.
+\caption{Empirical distribution of the absolute relative bias and the coefficient of variation of each estimation method.
 100 replications of the experience showed on figure \ref{fig:res:LG} are made.

### 2017-02-22 ParisEM.tex
clarifying simulation experiments, addition of formulas for used indexes, changes in the captions of the figures

@@
 \newtheorem{proposition}{Proposition}
-
+\newcommand{\hQ}{\widehat{Q}}
 \newcounter{hypA}
@@
 The GRand PaRIS algorithm is compared with the fixed lag method of \cite{olsson:strojby:2011}.
-The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=2000$ for the sine model and $N=1400$ for the log growth model so that both estimations have similar computational times.
+The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=1600$ so that both estimations have similar computational times.
 
@@
 The particle approximation of $\mathcal{Q}(\theta,\theta)$ is computed using each algorithm, see Figure~\ref{fig:res:SINE}.
-This estimation is performed 200 times with $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.
+This estimation is performed 200 times, resulting in estimates $\hQ_1,\dots,\hQ_{200}$, using $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.
 Moreover, the E step requires the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.
 $\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
-The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, see Figure \ref{fig:res:SINE}.
-
-\textcolor{red}{Detailler plus clairement ce que dit la figure 2 ici (donner clairement les estimateurs voire m\^eme les formules pour le biais et la variance relatifs)}
-This experiment was reproduced on a hundred different simulated data, in order study the distribution of the empirical absolute relative bias and coefficient of variations for all methods.
-The results are shown on Figure \ref{fig:mult:SINE}.
+The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, and denoted by $\hQ_\star$.
+Figure \ref{fig:res:SINE} shows this estimates for an example on one simulated data set.\\
+\\
+This same experiment was reproduced on 100 different simulated data sets.
+For each simulation $s$, the empirical absolute relative bias $\mathsf{arb}_s$ and the empirical absolute coefficient of variation $\mathsf{acv}_s$ are computed as
+\begin{align}
+\mathsf{arb}_s &= \frac{\vert m(\hQ^s)-\hQ^s_\star\vert }{\vert \hQ^s_\star\vert }\label{eq:emp:bias}\\
+\mathsf{acv}_s&=\frac{\sigma(\hQ^s)}{\vert m(\hQ^s)\vert }\label{eq:emp:cv}
+\end{align}
+where $m(\hQ^s)$ and $\sigma(\hQ^s)$ are the empirical mean and standard deviation of the sample $Q_1^s,\dots,Q_{200}^s$.
+For each estimation method, the resulting distributions of $\mathsf{arb}_1,\dots,\mathsf{arb}_{100}$ and $\mathsf{acv}_1,\dots,\mathsf{acv}_{100}$  are shown on Figure \ref{fig:mult:SINE}.\\
 
@@
 \end{subfigure}
-\caption{Evolution of the empirical relative bias for each method in the case of the log-growth model.
-100 replications of the experience showed on Figure \ref{fig:res:SINE} are made.
-The empirical abolute relative bias is computed for each one.
-The whiskers show the 95\% central values and the dot shows the empirical median.
-All estimations are performed with an equal computation time.}
+\caption{\textit{Left: }Distribution of the empirical absolute relative bias (given in equation \eqref{eq:emp:bias}) for each method in the case of the SINE model.
+\textit{Right: }Distribution of the empirical absolute coefficient of variation (given in equation \eqref{eq:emp:cv}) for each method for the SINE model.These distributions are obtained using 100 replications of the experience showed on Figure \ref{fig:res:SINE}.
+All estimations are performed in a similar computation time.}
 \label{fig:mult:SINE}
@@
 \end{subfigure}
-\caption{Empirical distribution of the absolute relative bias and the coefficient of variation of each estimation method.
-100 replications of the experience showed on figure \ref{fig:res:LG} are made.
-The whiskers show the 95\% central values and the dot shows the empirical median.
-All estimations are performed with an equal computation time.}
+\caption{\textit{Left: }Distribution of the empirical absolute relative bias (given in equation \eqref{eq:emp:bias}) for each method in the case of the Log-growth model.
+\textit{Right: }Distribution of the empirical absolute coefficient of variation (given in equation \eqref{eq:emp:cv}) for each method for the Log-growth model.These distributions are obtained using 100 replications of the experience showed on Figure \ref{fig:res:LG}.
+All estimations are performed in a similar computation time.}
 \label{fig:mult:LG}

### 2017-02-23 ParisEM.tex
Final revision #5

@@
 The proposed algorithm allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles.
-The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to SDE is that the accept reject mechanism ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
+The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to SDE is that the accept-reject mechanism ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
 The usual FFBS and FFBSi algorithms may not be extended this easily since they both require the computation of weights defined as ratios involving the transition densities, thus replacing these unknown quantities by unbiased estimates does not lead to unbiased estimators of the weights.
@@
 However, there is a significant qualitative difference between the cases $\tilde{N} = 1$ and $\tilde{N} \geq 2$.
-As for the FFBSi algorithm,  when there exists $\sigma_+$ such that $0<\qk <\sigma_+$, PaRIS algorithm may be implemented with $\mathcal{O}(N)$ complexity using the accept reject mechanism of \cite{douc:garivier:moulines:olsson:2011}.
+As for the FFBSi algorithm,  when there exists $\sigma_+$ such that $0<\qk <\sigma_+$, PaRIS algorithm may be implemented with $\mathcal{O}(N)$ complexity using the accept-reject mechanism of \cite{douc:garivier:moulines:olsson:2011}.
 
@@
 \widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)&> 0~~\text{a.s}\eqsp,\\
-\mathbb{E}\left[\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)\middle| \mathcal{F}_{k+1}^N\right] &= \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})\eqsp,
+\mathbb{E}\left[\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)\middle| \mathcal{G}_{k+1}^N\right] &= \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})\eqsp,
 \end{align*}
 where, for all $0\le k \le n$, 
-\[
-\mathcal{F}_{k}^N = \sigma\left\{Y_{0:k};(\xi^{\ell}_{u},\omega^{\ell}_{u},\tau^{\ell}_{u}),~1\le \ell\le N,~0\le u\le k\right\}\eqsp.
-\]
+\begin{align*}
+\mathcal{F}_{k}^N &= \sigma\left\{Y_{0:k};(\xi^{\ell}_{u},\omega^{\ell}_{u},\tau^{\ell}_{u});J_{v}^{\ell,j};~1\le \ell\le N,~0\le u\le k, 1\le j \le \widetilde{N}, 0\le v< k\right\}\eqsp,\\
+\mathcal{G}_{k+1}^N &= \mathcal{F}_{k}^N \vee \sigma\left\{Y_{k+1};(\xi^{\ell}_{k+1},\omega^{\ell}_{k+1});~1\le \ell\le N\right\}\eqsp.
+\end{align*}
 In the case where $\qk$ is unknown, the filtering weights in \eqref{eq:importance:weights} then become:
@@
 Therefore, to obtain a generalized random version of PaRIS, we only need to be able to sample from weight $\Lambda_k^N(i,\ell)$ in the case of SDE based HMM.
-If $\widehat{q}_k$ may be upper bounded almost surely by a constant $\hat{\sigma}_+$, this task is performed using the accept reject mechanism given in Algorithm~\ref{alg:AR:unknownq}.
+If $\widehat{q}_k$ may be upper bounded almost surely by a constant $\hat{\sigma}_+$, this task is performed using the accept-reject mechanism given in Algorithm~\ref{alg:AR:unknownq}.
 Lemma~\ref{lem:AR:unbiased} states that replacing  $\qk$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.\\
@@
 Let $0\le k\le n-1$.
-For all $1\le i \le N$ and all $1\le \ell \le \widetilde{N}$, the conditional probability distribution given $\mathcal{F}_{k+1}^N$ of the random variable $J_{k}^{i,\ell}$ produced by Algorithm~\ref{alg:AR:unknownq} is  $\Lambda_{k-1}^N(i,\cdot)$.
+For all $1\le i \le N$ and all $1\le \ell \le \widetilde{N}$, the conditional probability distribution given $\mathcal{G}_{k+1}^N$ of the random variable $J_{k}^{i,\ell}$ produced by Algorithm~\ref{alg:AR:unknownq} is  $\Lambda_{k-1}^N(i,\cdot)$.
 \end{lemma}
@@
 In the example displayed on Figure \ref{fig:res:SINE}, we set $\theta=0$.
-In that case, the function $\rho_{\Delta_k}$ defined in \eqref{eq:rho:func} can be upper bounded either on $(x,y)$ or only on $y$, the resulting GRand PaRIS algorithm has therefore a linear complexity.
+In that case, the function $\rho_{\Delta_k}$ defined in \eqref{eq:rho:func} can be upper bounded either on $(x,y)$ or only on $y$, the GRand PaRIS algorithm has therefore a linear complexity.
 
@@
 The particle approximation of $\mathcal{Q}(\theta,\theta)$ is computed using each algorithm, see Figure~\ref{fig:res:SINE}.
-This estimation is performed 200 times, resulting in estimates $\hQ_1,\dots,\hQ_{200}$, using $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.
-Moreover, the E step requires the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.
+This estimation is performed 200 times using $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$, and provides the estimates $\hQ_1,\dots,\hQ_{200}$.
+The E step requires the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.
 $\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
 The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, and denoted by $\hQ_\star$.
-Figure \ref{fig:res:SINE} shows this estimates for an example on one simulated data set.\\
-\\
-This same experiment was reproduced on 100 different simulated data sets.
-For each simulation $s$, the empirical absolute relative bias $\mathsf{arb}_s$ and the empirical absolute coefficient of variation $\mathsf{acv}_s$ are computed as
-\begin{align}
-\mathsf{arb}_s &= \frac{\vert m(\hQ^s)-\hQ^s_\star\vert }{\vert \hQ^s_\star\vert }\label{eq:emp:bias}\\
-\mathsf{acv}_s&=\frac{\sigma(\hQ^s)}{\vert m(\hQ^s)\vert }\label{eq:emp:cv}
-\end{align}
-where $m(\hQ^s)$ and $\sigma(\hQ^s)$ are the empirical mean and standard deviation of the sample $Q_1^s,\dots,Q_{200}^s$.
-For each estimation method, the resulting distributions of $\mathsf{arb}_1,\dots,\mathsf{arb}_{100}$ and $\mathsf{acv}_1,\dots,\mathsf{acv}_{100}$  are shown on Figure \ref{fig:mult:SINE}.\\
+This estimate is also displayed in Figure \ref{fig:res:SINE}.
+
+The same experiment was reproduced with 100 different simulated data sets.
+For each data set $s$, the empirical absolute relative bias $\mathsf{arb}_s$ and the empirical absolute coefficient of variation $\mathsf{acv}_s$ are computed as
+\begin{equation}
+\label{eq:emp:bias}
+\mathsf{arb}_s = \frac{\vert m(\hQ^s)-\hQ^s_\star\vert }{\vert \hQ^s_\star\vert }\quad\mbox{and}\quad \mathsf{acv}_s=\frac{\sigma(\hQ^s)}{\vert m(\hQ^s)\vert }\eqsp,
+\end{equation}
+where $m(\hQ^s)$ and $\sigma(\hQ^s)$ are the empirical mean and standard deviation of the estimates $\hQ_1^s,\dots,\hQ_{200}^s$ for each data set.
+The distributions of $\mathsf{arb}_1,\dots,\mathsf{arb}_{100}$ and $\mathsf{acv}_1,\dots,\mathsf{acv}_{100}$  are shown in Figure \ref{fig:mult:SINE} for both estimation methods.
 
@@
 It is worth noting here that the lag for which the bias is small is model dependent.
+
+\begin{figure}[p]
+\centering
+\begin{subfigure}{0.49\textwidth}
+\includegraphics[width=\textwidth]{obs_SINE}
+\end{subfigure}
+\begin{subfigure}{0.49\textwidth}
+\includegraphics[width=\textwidth]{res_Estep_SINE}
+\end{subfigure}
+\caption{{\em SINE model}.
+Process $X$ solution to the SDE (balls) and observations $Y$ (circles) at times $t_0=0,\dots,t_{100}=50$ [left].
+Estimation of the EM intermediate quantity $\mathcal{Q}(\theta,\theta)$  using the fixed lag (FL) technique for 5 different lags, and the GRand PaRIS algorithm using 200 replicates [right].
+The whiskers represent the extent of the 95\% central values.
+The dot represents the empirical mean over the 200 replicates.
+The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=5000$ particles.}
+\label{fig:res:SINE}
+\end{figure}
+\begin{figure}[p]
+\centering
+\begin{subfigure}{0.49\textwidth}
+\includegraphics[width=\textwidth]{res_mult_SINE_bias}
+\end{subfigure}
+\begin{subfigure}{0.49\textwidth}
+\includegraphics[width=\textwidth]{res_mult_SINE_cv}
+\end{subfigure}
+\caption{{\em SINE model}. Distribution of the empirical absolute relative bias [left] and of the empirical absolute coefficient of variation [right] for each method, see \eqref{eq:emp:bias}.}
+\label{fig:mult:SINE}
+\end{figure}
+
 %To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated.
@@
 The parameters are given by $$\theta =(\kappa=0.1,\sigma=0.1,\gamma=1000,\sigma^2_{obs}=4)\eqsp.$$
-In that case, the $\rho_{\Delta_k}$ function defined in \eqref{eq:rho:func} can be upper bounded as a function of $y$ when $x\in \{\xi_k^1,\dots, \xi_k^N\}$, the resulting GRand PaRIS algorithm has therefore a linear complexity.
+In that case, the $\rho_{\Delta_k}$ function defined in \eqref{eq:rho:func} can be upper bounded as a function of $y$ when $x\in \{\xi_k^1,\dots, \xi_k^N\}$, the GRand PaRIS algorithm has therefore a linear complexity.
 The intermediate quantity of the EM algorithm is evaluated  as for the SINE model, see Figures~\ref{fig:res:LG} and~\ref{fig:mult:LG}.
@@
 Again, the GRand PaRIS algorithm outperforms the fixed lag smoother as it shows  a similar (vanishing) bias as the fixed lag for the largest lag and a smaller variance than the fixed lags estimates with negligible bias.
-%It is worth noting here that, as pointed out in \cite{olsson:strojby:2011}, results of the fixed lag technique indicates a quite strong forgetting in the model, and the choice of the lag should take into account this forgetting property.
-%As the PaRIS algorithm is always unbiased, there is no need to consider the forgetting property of a given model in the approximation of the EM intermediate quantity. 
-\begin{figure}[p]
-\centering
-\begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{obs_SINE}
-\end{subfigure}
-\begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{res_Estep_SINE}
-\end{subfigure}
-\caption{\textit{Left: }Process $X$ solution to the SDE \eqref{eq:Lamp:SINE} (black) and observations $Y$ satisfying equation \eqref{eq:obs:SINE} (white) at times $t_0=0,\dots,t_{100}=50$.
-\textit{Right: } Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$  using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm using 200 replicates.
-The whiskers represent the extent of the 95\% central values.
-The dot represents the empirical mean over the 200 replicates.
-The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=5000$ particles.}
-\label{fig:res:SINE}
-\end{figure}
-\begin{figure}[p]
-\centering
-\begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{res_mult_SINE_bias}
-\end{subfigure}
-\begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{res_mult_SINE_cv}
-\end{subfigure}
-\caption{\textit{Left: }Distribution of the empirical absolute relative bias (given in equation \eqref{eq:emp:bias}) for each method in the case of the SINE model.
-\textit{Right: }Distribution of the empirical absolute coefficient of variation (given in equation \eqref{eq:emp:cv}) for each method for the SINE model.These distributions are obtained using 100 replications of the experience showed on Figure \ref{fig:res:SINE}.
-All estimations are performed in a similar computation time.}
-\label{fig:mult:SINE}
-\end{figure}
+
 \begin{figure}[p]
@@
 \end{subfigure}
-\caption{\textit{Left: }Process $X$ solution to the SDE \eqref{eq:Lamp:LG} (black) and observations $Y$ satisfying equation \eqref{eq:obs:LG} (white) at times $t_0=0,\dots,t_{100}=50$.
-\textit{Right: } Estimation of the EM intermediate quantity $\mathcal{Q}(\theta_{true},\theta_{true})$  using the fixed lag (FL) technique for 5 different lags (namely 1, 2, 5,10 ,50), and the GRand PaRIS algorithm using 200 replicates.
+\caption{{\em Log-growth model}.
+Process $X$ solution to the SDE (balls) and observations $Y$ (circles) at times $t_0=0,\dots,t_{100}=50$ [left].
+Estimation of the EM intermediate quantity $\mathcal{Q}(\theta,\theta)$  using the fixed lag (FL) technique for 5 different lags, and the GRand PaRIS algorithm using 200 replicates [right].
 The whiskers represent the extent of the 95\% central values.
@@
 \end{subfigure}
-\caption{\textit{Left: }Distribution of the empirical absolute relative bias (given in equation \eqref{eq:emp:bias}) for each method in the case of the Log-growth model.
-\textit{Right: }Distribution of the empirical absolute coefficient of variation (given in equation \eqref{eq:emp:cv}) for each method for the Log-growth model.These distributions are obtained using 100 replications of the experience showed on Figure \ref{fig:res:LG}.
-All estimations are performed in a similar computation time.}
+\caption{{\em Log-growth model}. Distribution of the empirical absolute relative bias [left] and of the empirical absolute coefficient of variation [right] for each method, see \eqref{eq:emp:bias}.}
 \label{fig:mult:LG}
@@
 This paper presents a new online SMC smoother for partially observed differential equations.
-This algorithm relies on an acceptance reject procedure inspired from the recent PaRIS algorithm.
+This algorithm relies on an accept-reject procedure inspired from the recent PaRIS algorithm.
 The main result of the article for practical applications is that the mechanism of this procedure remains valid when the transition density is approximated by a an unbiased positive estimator.
-The proposed procedure outperforms the existing fixed lag smoother for SDE of \cite{olsson:strojby:2011}, as it does not introduce intrinsic non vanishing bias,  and numerical simulations highlight a better variance on two examples.
+The proposed procedure outperforms the existing fixed lag smoother for SDE of \cite{olsson:strojby:2011}, as it does not introduce an intrinsic and non vanishing bias.
+In addition, numerical simulations highlight a better variance using data from two different models.
 It can be implemented for the class of models $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} with a linear complexity in $N$.
@@
 \[
-\mathcal{A}^k_{\ell} = \left\{U_\ell<\widehat{q}_{k}(\xi_{k-1}^{J_\ell},\xi_k^{i},\zeta^{\ell}_{k})/\sigma_+\right\}\eqsp.
+\mathcal{A}^k_{\ell} = \left\{U_\ell<\widehat{q}_{k}(\xi_{k}^{J_\ell},\xi_{k+1}^{i},\zeta^{\ell}_{k})/\sigma_+\right\}\eqsp.
 \]
@@
 \begin{align*}
-\mathbb{E}\left[h(J^{i,j}_k)\middle| \mathcal{F}_k^N\right] & = \sum_{m\ge 1}\mathbb{E}\left[h(J_m)\1_{\tau=m}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \sum_{m\ge 1}h(m)\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\middle| \mathcal{F}_k^N\right]\right)\mathbb{E}\left[h(J_m)\1_{\mathcal{A}^k_{m}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[1-\frac{\widehat{q_k}(\xi_{k-1}^{J_\ell},\xi_k^{i};\zeta_k^{\ell})}{\hat{\sigma}_{+}}\middle| \mathcal{F}_k^N\right]\right)\\
-&\hspace{5cm}\times\mathbb{E}\left[h(I_m)\frac{\widehat{q_k}(\xi_{k-1}^{J_m},\xi_k^{i};\zeta_k^{\ell})}{\hat{\sigma}_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \sum_{m\ge 1}\left(\mathbb{E}\left[1-\frac{q_k(\xi_{k-1}^{J_1},\xi_k^{1})}{\hat{\sigma}_{+}}\middle| \mathcal{F}_k^N\right]\right)^{m-1}\mathbb{E}\left[h(J_m)\frac{q_k(\xi_{k-1}^{J_m},\xi_k^{1})}{\hat{\sigma}_{+}}\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \mathbb{E}\left[h(J_1)q_k(\xi_{k-1}^{J_1},\xi_k^{i})\middle| \mathcal{F}_k^N\right]/\mathbb{E}\left[q_k(\xi_{k-1}^{J_m},\xi_k^{i})\middle| \mathcal{F}_k^N\right]\eqsp,\\
-& = \sum_{\ell=1}^N \frac{h(\ell)\omega_{k-1}^{\ell}q_k(\xi_{k-1}^{\ell},\xi_k^{i})}{\sum_{m=1}^N\omega_{k-1}^{m}q_k(\xi_{k-1}^{m},\xi_k^{i})}\eqsp,\\
+\mathbb{E}\left[h(J^{i,j}_k)\middle| \mathcal{G}_{k+1}^N\right] & = \sum_{m\ge 1}\mathbb{E}\left[h(J_m)\1_{\tau=m}\middle| \mathcal{G}_{k+1}^N\right]\eqsp,\\
+& = \sum_{m\ge 1}h(m)\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\middle| \mathcal{G}_{k+1}^N\right]\right)\mathbb{E}\left[h(J_m)\1_{\mathcal{A}^k_{m}}\middle| \mathcal{G}_{k+1}^N\right]\eqsp,\\
+& = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[1-\frac{\widehat{q_k}(\xi_{k}^{J_\ell},\xi_{k+1}^{i};\zeta_k^{\ell})}{\hat{\sigma}_{+}}\middle| \mathcal{G}_{k+1}^N\right]\right)\\
+&\hspace{5cm}\times\mathbb{E}\left[h(J_m)\frac{\widehat{q_k}(\xi_{k}^{J_m},\xi_{k+1}^{i};\zeta_k^{m})}{\hat{\sigma}_{+}}\middle| \mathcal{G}_{k+1}^N\right]\eqsp,\\
+& = \sum_{m\ge 1}\left(\mathbb{E}\left[1-\frac{q_k(\xi_{k}^{J_1},\xi_{k+1}^{1})}{\hat{\sigma}_{+}}\middle| \mathcal{G}_{k+1}^N\right]\right)^{m-1}\mathbb{E}\left[h(J_1)\frac{q_k(\xi_{k}^{J_1},\xi_{k+1}^{1})}{\hat{\sigma}_{+}}\middle| \mathcal{G}_{k+1}^N\right]\eqsp,\\
+& = \mathbb{E}\left[h(J_1)q_k(\xi_{k}^{J_1},\xi_{k+1}^{i})\middle| \mathcal{F}_k^N\right]/\mathbb{E}\left[q_k(\xi_{k}^{J_1},\xi_{k+1}^{i})\middle| \mathcal{G}_{k+1}^N\right]\eqsp,\\
+& = \sum_{\ell=1}^N \frac{h(\ell)\omega_{k-1}^{\ell}q_k(\xi_{k}^{\ell},\xi_{k+1}^{i})}{\sum_{m=1}^N\omega_{k-1}^{m}q_k(\xi_{k}^{m},\xi_{k+1}^{i})}\eqsp,\\
 &= \sum_{\ell=1}^N \Lambda_{k-1}^N(i,\ell)h(\ell) \eqsp,
@@
 \[
-\mathbb{E}\left[\widehat{\omega}^i_{k+1}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right] = \mathbb{E}\left[\frac{ \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}}, \xi^{i}_{k+1};\zeta_{k+1})g_{k+1}(\xi^{i}_{k+1})}{\vartheta_{k+1}(\xi^{I^{i}_{k+1}}_{k}) p_{k+1}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right]\eqsp.
-%&= \mathbb{E}\left[\frac{g_{k+1}(\xi^{1}_{k+1})}{\vartheta_{k+1}(\xi^{I^{1}_{k+1}}_{k}) p_{k+1}(\xi_{k}^{I^{1}_{k+1}},\xi^{\ell}_{k+1})}\tau^{1}_{k+1}\widehat{q}_{\theta}^{\Delta t_{k+1}}(\xi_{k}^{I^{1}_{k+1}},\xi^{1}_{k+1};\Lambda_{k+1})\middle| \mathcal{F}_k^{N}\right]\eqsp,
+\mathbb{E}\left[\widehat{\omega}^i_{k+1}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right] = \mathbb{E}\left[\frac{ \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}}, \xi^{i}_{k+1};\zeta_{k})g_{k+1}(\xi^{i}_{k+1})}{\vartheta_{k+1}(\xi^{I^{i}_{k+1}}_{k}) p_{k+1}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right]\eqsp.
 \]
-Note that
+Note that by Lemma~\ref{lem:AR:unbiased},
+%\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,I_{k+1}^i\}\right)
 \begin{align*}
-&\mathbb{E}\left[\tau^{i}_{k+1}\middle|\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,I_{k+1}^i\}\right)\right]
+&\mathbb{E}\left[\tau^{i}_{k+1}\middle|\mathcal{G}_{k+1}^{N}\right]
  = \sum_{\ell=1}^N\frac{\omega_k^{\ell} \qk(\xi_{k}^{\ell}, \xi^{i}_{k+1}) \left(\tau^{\ell}_k + h_{k}(\xi_{k}^{\ell},\xi^{i}_{k+1})\right)}{\sum_{\ell'=1}^N\omega_k^{\ell'} \qk(\xi_{k}^{\ell'},\xi^{i}_{k+1})}\eqsp,\\
-&\mathbb{E} \left[\widehat{\qk}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta_{k+1}) \middle| \sigma \left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,I^{i}_{k+1}\}\right)\right]
+&\mathbb{E} \left[\widehat{\qk}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta_{k}) \middle| \mathcal{G}_{k+1}^{N}\right]
  = \qk(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})\eqsp.
 \end{align*}
-Since $\tau^{i}_{k+1}$ and $\zeta_{k+1}$ are independent conditionally to $\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^{\ell}, I_{k+1}^{\ell}\}_{\ell=1}^N\right)$:
+Since $\tau^{i}_{k+1}$ and $\zeta_{k+1}$ are independent conditionally to $\mathcal{G}_{k+1}^{N}$:
 \begin{multline*}
-\mathbb{E}\left[\tau^{i}_{k+1} \widehat{\qk} (\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta_{k+1})\middle|\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,I_{k+1}^i\} \right)\right]\\
+\mathbb{E}\left[\tau^{i}_{k+1} \widehat{\qk} (\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta_{k+1})\middle|\mathcal{G}_{k+1}^{N}\right]\\
  = q_k(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})\sum_{\ell=1}^N\frac{\omega_k^{\ell} \qk (\xi_{k}^{\ell},\xi^{i}_{k+1})\left(\tau^{\ell}_k + h_{k}(\xi_{k}^{\ell},\xi^{i}_{k+1})\right)}{\sum_{\ell'=1}^N\omega_k^{\ell'} \qk (\xi_{k}^{\ell'},\xi^{i}_{k+1})}\eqsp.

### 2017-02-24 ParisEM.tex
Minor Typo + general strategy for example

--- 2017-02-22 ParisEM.tex
@@
 \usepackage{subcaption}
-\graphicspath{{/home/gloaguen/Bureau/Travaux_agro/figures_article_oziris/}}
+\graphicspath{{figures/}}
 \usepackage{enumerate}
@@
 the computation of the posterior distributions of sequences of hidden states given observations.
-The filtering problem refers to the estimation, for each $0\le \le n$,  of the distributions of the hidden state $X_k$ given the observations $(Y_0,\ldots,Y_k)$.
-Smoothing stands for the estimation of the distributions of sequence of states $(X_{k},\ldots,X_{p})$ given observations $(Y_{0},\ldots,Y_{\ell})$ with $0\le k\le p \le \ell \le n$.
-These posterior distributions are crucial to compute maximum likelihood estimators of the unknown parameters defining the model using the observations $(Y_0,\ldots,Y_n)$ only.
+The filtering problem refers to the estimation, for each $0\le k \le n$,  of the distributions of the hidden state $X_k$ given the observations $(Y_0,\ldots,Y_k)$.
+Smoothing stands for the estimation of the distributions of the sequence of states $(X_{k},\ldots,X_{p})$ given observations $(Y_{0},\ldots,Y_{\ell})$ with $0\le k\le p \le \ell \le n$.
+These posterior distributions are crucial to compute maximum likelihood estimators of the unknown model parameters using the observations $(Y_0,\ldots,Y_n)$ only.
 % $\theta\mapsto \ell_{\theta}^{n}$ given by
@@
 The usual FFBS and FFBSi algorithms may not be extended this easily since they both require the computation of weights defined as ratios involving the transition densities, thus replacing these unknown quantities by unbiased estimates does not lead to unbiased estimators of the weights.
-The proposed generalized random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to general hidden Markov models whose Markovian dynamics is ruled by a stochastic differential equation (one of the first two domains defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}) but also to any general state space model where the transition density of the hidden chain may be estimated unbiasedly.
+The proposed Generalized Random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to general hidden Markov models whose Markovian dynamics is ruled by a stochastic differential equation (one of the first two domains defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}) but also to any general state space model where the transition density of the hidden chain may be estimated unbiasedly.
 %We also propose to improve the algorithm of \cite{fearnhead:papaspiliopoulos:roberts:2008} by extending the range of diffusion processes for which SMC methods can be applied.
@@
 
-The paper is organized as follows.
-Section~\ref{sec:rwparis} describes the random weight PaRIS algorithm to approximate smoothed additive functionals using unbiased estimates of the transition density of the hidden states and details the application of this algorithm when the transition density may be approximated using a GPE.
+Section~\ref{sec:rwparis} describes the GRand PaRIS algorithm  aimed to approximate smoothed additive functionals using unbiased estimates of the transition density of the hidden states and details the application of this algorithm when the transition density may be approximated using a GPE.
 In Section~\ref{sec:convergence}, classical convergence results for SMC smoothers are extended to the setting of this paper and illustrated with numerical experiments in Section~\ref{sec:exp}. 
@@
 
-\section{The generalized random PaRIS algorithm}
+\section{The Generalized Random PaRIS algorithm}
 \label{sec:rwparis}
@@
 It is assumed that $\alpha$ is of the form $\alpha(x) = \nabla_x A(x)$ where $A: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function.
-A popular choice of $A$  is $A = \log \pi/2$ leading to a Langevin diffusion which may be used to approximate a given target distribution $\pi$ on $\mathbb{R}^d$.
+% A popular choice of $A$  is $A(x) = \log \pi(x)/2$ leading to a Langevin diffusion which may be used to approximate a given target distribution $\pi$ on $\mathbb{R}^d$.
 See for instance \cite{roberts:tweedie:1996} for conditions under which this diffusion converges  at exponential rate to $\pi$.
-SDE where the drift is given by the gradient of a potential function have been widely used in ecology for instance, see \cite{brillinger:et:al:2011,harris:blackwell:2013,preisler:et:al:2013} and the references therein.
-The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{k=0,\dots,n}$ in $\mathbb{R}^m$.
+% SDE where the drift is given by the gradient of a potential function have been widely used in ecology for instance, see \cite{brillinger:et:al:2011,harris:blackwell:2013,preisler:et:al:2013} and the references therein.
+The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{k=0,\dots,n}$ in $(\mathbb{R}^m)^{n+1}$. 
 For all $0\le k \le n$, the distribution of $Y_k$ given $X_k:= X_{t_k}$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,\cdot) = g_k(X_k)$. 
@@
 \[
-\phi^N_{k}[h] := \frac{1}{\Omega_t^N} \sum_{\ell=1}^N \omega_k^{\ell} h \left(\xi^{\ell}_k \right)\eqsp,\quad\Omega_k^N:= \sum_{\ell=1}^N \omega_k^{\ell}\eqsp.
+\phi^N_{k}[h] := \frac{1}{\Omega_k^N} \sum_{\ell=1}^N \omega_k^{\ell} h \left(\xi^{\ell}_k \right)\eqsp,\quad\Omega_k^N:= \sum_{\ell=1}^N \omega_k^{\ell}\eqsp.
 \]
@@
 It does not produce an approximation of the smoothing distributions but of the smoothed expectation of a fixed additive functional and thus  may be used to approximate \eqref{def:addfunc}.
-Its crucial property is that it does not require a backward pass, the smoothed expectations is computed on-the-fly with the particle filter and no storage of the particles or weights is needed.
+Its crucial property is that it does not require a backward pass, the smoothed expectation is computed on-the-fly with the particle filter and no storage of the particles or weights is needed.
 
@@
 Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio of \cite{douc:garivier:moulines:olsson:2011} is not tractable.
-Following \cite{fearnhead:papaspiliopoulos:roberts:2008,olsson:strojby:2011}, filtering weights can be approximated by replacing $\qk(\xi^{\ell}_{k},\xi_{k+1}^{i})$  by an  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable such that:
+Following \cite{fearnhead:papaspiliopoulos:roberts:2008,olsson:strojby:2011}, filtering weights can be approximated by replacing $\qk(\xi^{\ell}_{k},\xi_{k+1}^{i})$  by an unbiased  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable such that:
 \begin{align*}
@@
 \]
+
+Practical choices for $\zeta_k$ are presented in the paragraph devoted to the construction of bounded estimators for $q$ and more specifically equation \ref{eq:GPE1} presents the choice made for the implementation of such estimator in our context.
+
 In the case where $\qk$ is unknown, the filtering weights in \eqref{eq:importance:weights} then become:
@@
 As the observation model is linear and Gaussian, the proposal distribution is therefore Gaussian with explicit mean and variance.
+
+In order to evaluate the performance of the GRand PaRIS, the following strategy has been chosen.
+We compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50).
+The particle approximation of $\mathcal{Q}(\theta,\theta)$ for each model  is computed using each algorithm, see Figure~\ref{fig:res:SINE} for the SINE model (and respectively Figure~\ref{fig:res:LG} for the log-growth model).
+This estimation is performed 200 times, resulting in estimates $\hQ_1,\dots,\hQ_{200}$, using $\tilde{N}=2$ particles for the filter part, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.
+Moreover, the E step requires the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.
+$\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
+The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, and denoted by $\hQ_\star$.
+Figure \ref{fig:res:SINE} (resp.
+\ref{fig:res:LG} ) shows this estimates for an example on one simulated data set.\\
 The GRand PaRIS algorithm is compared with the fixed lag method of \cite{olsson:strojby:2011}.
@@
 
-In order to evaluate the performance of our algorithm, we compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50).
-The particle approximation of $\mathcal{Q}(\theta,\theta)$ is computed using each algorithm, see Figure~\ref{fig:res:SINE}.
-This estimation is performed 200 times, resulting in estimates $\hQ_1,\dots,\hQ_{200}$, using $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.
-Moreover, the E step requires the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.
-$\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
-The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, and denoted by $\hQ_\star$.
-Figure \ref{fig:res:SINE} shows this estimates for an example on one simulated data set.\\
-\\
 This same experiment was reproduced on 100 different simulated data sets.

### 2017-02-24 ParisEM.tex
Resolve conflicts

--- 2017-02-23 ParisEM.tex
@@
 The proposed algorithm allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles.
-The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to SDE is that the accept-reject mechanism ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
+The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to SDE is that the accept reject mechanism ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
 The usual FFBS and FFBSi algorithms may not be extended this easily since they both require the computation of weights defined as ratios involving the transition densities, thus replacing these unknown quantities by unbiased estimates does not lead to unbiased estimators of the weights.
-The proposed generalized random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to general hidden Markov models whose Markovian dynamics is ruled by a stochastic differential equation (one of the first two domains defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}) but also to any general state space model where the transition density of the hidden chain may be estimated unbiasedly.
+The proposed Generalized Random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to general hidden Markov models whose Markovian dynamics is ruled by a stochastic differential equation (one of the first two domains defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}) but also to any general state space model where the transition density of the hidden chain may be estimated unbiasedly.
+
 %We also propose to improve the algorithm of \cite{fearnhead:papaspiliopoulos:roberts:2008} by extending the range of diffusion processes for which SMC methods can be applied.
@@
 where, for all $0\le k \le n$,
+
 \begin{align*}
@@
 \end{align*}
+
+Practical choices for $\zeta_k$ are presented in the paragraph devoted to the construction of bounded estimators for $q$ and more specifically equation \ref{eq:GPE1} presents the choice made for the implementation of such estimator in our context.
+
 In the case where $\qk$ is unknown, the filtering weights in \eqref{eq:importance:weights} then become:
@@
 As the observation model is linear and Gaussian, the proposal distribution is therefore Gaussian with explicit mean and variance.
+
+In order to evaluate the performance of the GRand PaRIS, the following strategy has been chosen.
+We compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50).
+The particle approximation of $\mathcal{Q}(\theta,\theta)$ for each model  is computed using each algorithm, see Figure~\ref{fig:res:SINE} for the SINE model (and respectively Figure~\ref{fig:res:LG} for the log-growth model).
+This estimation is performed 200 times, resulting in estimates $\hQ_1,\dots,\hQ_{200}$, using $\tilde{N}=2$ particles for the PaRIS part, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.
+Moreover, the E step requires the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.
+$\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
+The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, and denoted by $\hQ_\star$.
+Figure \ref{fig:res:SINE} (resp.
+\ref{fig:res:LG} ) shows this estimates for an example on one simulated data set.\\
 The GRand PaRIS algorithm is compared with the fixed lag method of \cite{olsson:strojby:2011}.
@@
 
-In order to evaluate the performance of our algorithm, we compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50).
-The particle approximation of $\mathcal{Q}(\theta,\theta)$ is computed using each algorithm, see Figure~\ref{fig:res:SINE}.
-This estimation is performed 200 times using $\tilde{N}=2$, and $M=30$ replications for the Monte Carlo approximation of each $q_k$, and provides the estimates $\hQ_1,\dots,\hQ_{200}$.
-The E step requires the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.
-$\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
-The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, and denoted by $\hQ_\star$.
-This estimate is also displayed in Figure \ref{fig:res:SINE}.
-
-The same experiment was reproduced with 100 different simulated data sets.
-For each data set $s$, the empirical absolute relative bias $\mathsf{arb}_s$ and the empirical absolute coefficient of variation $\mathsf{acv}_s$ are computed as
-\begin{equation}
-\label{eq:emp:bias}
-\mathsf{arb}_s = \frac{\vert m(\hQ^s)-\hQ^s_\star\vert }{\vert \hQ^s_\star\vert }\quad\mbox{and}\quad \mathsf{acv}_s=\frac{\sigma(\hQ^s)}{\vert m(\hQ^s)\vert }\eqsp,
-\end{equation}
-where $m(\hQ^s)$ and $\sigma(\hQ^s)$ are the empirical mean and standard deviation of the estimates $\hQ_1^s,\dots,\hQ_{200}^s$ for each data set.
-The distributions of $\mathsf{arb}_1,\dots,\mathsf{arb}_{100}$ and $\mathsf{acv}_1,\dots,\mathsf{acv}_{100}$  are shown in Figure \ref{fig:mult:SINE} for both estimation methods.
+This same experiment was reproduced on 100 different simulated data sets.
+For each simulation $s$, the empirical absolute relative bias $\mathsf{arb}_s$ and the empirical absolute coefficient of variation $\mathsf{acv}_s$ are computed as
+\begin{align}
+\mathsf{arb}_s &= \frac{\vert m(\hQ^s)-\hQ^s_\star\vert }{\vert \hQ^s_\star\vert }\label{eq:emp:bias}\\
+\mathsf{acv}_s&=\frac{\sigma(\hQ^s)}{\vert m(\hQ^s)\vert }\label{eq:emp:cv}
+\end{align}
+where $m(\hQ^s)$ and $\sigma(\hQ^s)$ are the empirical mean and standard deviation of the sample $Q_1^s,\dots,Q_{200}^s$.
+For each estimation method, the resulting distributions of $\mathsf{arb}_1,\dots,\mathsf{arb}_{100}$ and $\mathsf{acv}_1,\dots,\mathsf{acv}_{100}$  are shown on Figure \ref{fig:mult:SINE}.\\
 

### 2017-02-27 ParisEM.tex
modification of the lemma 1

@@
 \end{equation}
-Therefore, to obtain a generalized random version of PaRIS, we only need to be able to sample from weight $\Lambda_k^N(i,\ell)$ in the case of SDE based HMM.
-If $\widehat{q}_k$ may be upper bounded almost surely by a constant $\hat{\sigma}_+$, this task is performed using the accept-reject mechanism given in Algorithm~\ref{alg:AR:unknownq}.
-Lemma~\ref{lem:AR:unbiased} states that replacing  $\qk$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.\\
+Therefore, to obtain a generalized random version of PaRIS, we only need to be able to sample from weight $\Lambda_k^N(i,\ell)$ in the case of SDE based HMM.\\
+We need to define the following assumption:\\
+\[
+\textbf{A1: }\forall~0\leq k\leq n,~\exists \hat{\sigma}_+ \text{ such that }\forall (x,y) \in\mathbb{R}^d\times  \mathbb{R}^d,~~\widehat{q}_k(x,y,\zeta_k)\leq \hat{\sigma_+}
+\]
+Then, the following lemma provides the missing step of a Generalized Random PaRIS (GRand Paris)
 \begin{lemma}
 \label{lem:AR:unbiased}
-Let $0\le k\le n-1$.
-For all $1\le i \le N$ and all $1\le \ell \le \widetilde{N}$, the conditional probability distribution given $\mathcal{G}_{k+1}^N$ of the random variable $J_{k}^{i,\ell}$ produced by Algorithm~\ref{alg:AR:unknownq} is  $\Lambda_{k-1}^N(i,\cdot)$.
+Let $0\le k\le n-1$ and $1\le i \le N$, and suppose $A_1$ holds may be upper bounded almost surely by a constant $\hat{\sigma}_+$. Let's define the random variable $J_k^i$ as follow:
+\begin{algorithmic}
+\STATE \textsc{Sampling step:} Sample independently $
+\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
+\IF{ $$U \leq \frac{\widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)}{\hat{\sigma}_+},$$}
+\STATE Set $J_k^{i,\ell} = J$.
+\ELSE 
+\STATE Return to \textbf{\sc Sampling Step}
+\ENDIF
+\end{algorithmic}
+Then, the conditional probability distribution given $\mathcal{G}_{k+1}^N$ of $J_{k}^{i,\ell}$ is $\Lambda_{k-1}^N(i,\cdot)$. 
 \end{lemma}
@@
 \end{proof}
-Note that the assumption that for all $1\le j\le N$, $\sup_y \hat{q}_k(\xi_k^j,y,\zeta_k)< \hat{\sigma}_+^k$ almost surely is actually sufficient to apply Algorithm \ref{alg:AR:unknownq}.
+Note that the assumption \textbf{A1} can be relaxed and replaced by one of the two following properties:
+\begin{align*}
+\textbf{A2: }\forall~0\leq k\leq n,~\exists \hat{\sigma}_+ \text{ such that }\forall (j,y) \in [[1,N]]\times  \mathbb{R}^d,~~\widehat{q}_k(\xi^j_k,y,\zeta_k)\leq \hat{\sigma_+}\\
+\textbf{A3: }\forall~0\leq k\leq n,~\exists \hat{\sigma}_+ \text{ such that }\forall (j,y) \in [[1,N]]\times  \mathbb{R}^d,~~\widehat{q}_k(\xi^j_k,y,\zeta_k)\leq \hat{\sigma_+}
+\end{align*}
+
+ for all $1\le j\le N$, $\sup_y \hat{q}_k(\xi_k^j,y,\zeta_k)< \hat{\sigma}_+^k$ almost surely is actually sufficient to apply Algorithm \ref{alg:AR:unknownq}.
 This improvement keeps the linear complexity property of PaRIS algorithm and Algorithm \ref{alg:Ozaki:PaRIS}, and allows to increase significantly the acceptance rate of Algorithm \ref{alg:AR:unknownq}.
@@
 Nevertheless, in practice, this choice can result in a much more efficient acceptance rate, which would reduce the computational time of the algorithm.
-\begin{algorithm}[H]
-\caption{Random weight accept-reject backward sampling}
-\begin{algorithmic}
-\FORALL{$i \in 1,\dots, N$}
-\FORALL{$\ell \in 1,\dots, \widetilde N$} 
-\STATE \textbf{\sc Sampling Step} \\
-Sample independently $
-\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
-\IF{ $$U \leq \frac{\widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)}{\hat{\sigma}_+},$$}
-\STATE Set $J_k^{i,\ell} = J$.
-\ELSE 
-\STATE Return to \textbf{\sc Sampling Step}
-\ENDIF
-\ENDFOR
-\ENDFOR
-\end{algorithmic}
-\label{alg:AR:unknownq}
-\end{algorithm}
 \subsection*{Bounded estimator of $q_k$}

### 2017-02-27 ParisEM.tex
addition of three boudness assumption

--- 2017-02-24 ParisEM.tex
@@
 \item[{\bf H\arabic{hypA}}]}{\end{itemize}}
-
+%%ENvironmment assumption
+\newcounter{defcounter}
+\setcounter{defcounter}{0}
+\newenvironment{assumpt}{%
+\addtocounter{equation}{-1}
+\refstepcounter{defcounter}
+\renewcommand\theequation{\textbf{A}\thedefcounter}
+\begin{equation}}
+{\end{equation}}
 \begin{document}
@@
 %Statistical inference for partially observed stochastic differential equations is a challenging task since some elementary quantities, such as transition probabilities, are not available explicitly.
-Statistical inference for partially observed stochastic differential equations often requires to solve bayesian filtering and smoothing problems, i.e.
-the computation of the posterior distributions of sequences of hidden states given observations.
+Statistical inference for partially observed stochastic differential equations often requires to solve bayesian filtering and smoothing problems, i.e.\ the computation of the posterior distributions of sequences of hidden states given observations.
 The filtering problem refers to the estimation, for each $0\le k \le n$,  of the distributions of the hidden state $X_k$ given the observations $(Y_0,\ldots,Y_k)$.
@@
 Instead of computing new weights, the FFBSi  algorithm samples independently backward in time particle trajectories using the particles and weights produced by the forward pass.
-Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e.
-using the observations as they are received) smoothed expectations of additive functionals.
+Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e.\ using the observations as they are received) smoothed expectations of additive functionals.
 Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.
@@
 \end{equation}
-Therefore, to obtain a generalized random version of PaRIS, we only need to be able to sample from weight $\Lambda_k^N(i,\ell)$ in the case of SDE based HMM.
-If $\widehat{q}_k$ may be upper bounded almost surely by a constant $\hat{\sigma}_+$, this task is performed using the accept-reject mechanism given in Algorithm~\ref{alg:AR:unknownq}.
-Lemma~\ref{lem:AR:unbiased} states that replacing  $\qk$ by its unbiased approximation is a valid alternative to step~\eqref{it:PaRIS:sampleindex}.\\
+Therefore, to obtain a generalized random version of PaRIS, we only need to be able to sample from weight $\Lambda_k^N(i,\ell)$ in the case of SDE based HMM.\\
+We need to define the following assumption:\\
+\begin{assumpt}
+\label{assumpt:global}
+\forall~0\leq k\leq n,~\exists \hat{\sigma}_+ \text{ such that }\forall (x,y) \in\mathbb{R}^d\times  \mathbb{R}^d,~~\widehat{q}_k(x,y,\zeta_k)\leq \hat{\sigma_+}
+\end{assumpt}
+Then, the following lemma provides the missing step of a Generalized Random PaRIS (GRand Paris)
 \begin{lemma}
 \label{lem:AR:unbiased}
-Let $0\le k\le n-1$.
-For all $1\le i \le N$ and all $1\le \ell \le \widetilde{N}$, the conditional probability distribution given $\mathcal{G}_{k+1}^N$ of the random variable $J_{k}^{i,\ell}$ produced by Algorithm~\ref{alg:AR:unknownq} is  $\Lambda_{k-1}^N(i,\cdot)$.
+Let $0\le k\le n-1$ and $1\le i \le N$, and suppose $A_1$ holds may be upper bounded almost surely by a constant $\hat{\sigma}_+$. Let's define the random variable $J_k^i$ as follow:
+\begin{algorithmic}
+\STATE \textsc{Sampling step:} Sample independently $
+\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
+\IF{ $$U \leq \frac{\widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)}{\hat{\sigma}_+},$$}
+\STATE Set $J_k^{i,\ell} = J$.
+\ELSE 
+\STATE Return to \textbf{\sc Sampling Step}
+\ENDIF
+\end{algorithmic}
+Then, the conditional probability distribution given $\mathcal{G}_{k+1}^N$ of $J_{k}^{i,\ell}$ is $\Lambda_{k}^N(i,\cdot)$. 
 \end{lemma}
@@
 \end{proof}
-Note that the assumption that for all $1\le j\le N$, $\sup_y \hat{q}_k(\xi_k^j,y,\zeta_k)< \hat{\sigma}_+^k$ almost surely is actually sufficient to apply Algorithm \ref{alg:AR:unknownq}.
-This improvement keeps the linear complexity property of PaRIS algorithm and Algorithm \ref{alg:Ozaki:PaRIS}, and allows to increase significantly the acceptance rate of Algorithm \ref{alg:AR:unknownq}.
+Note that lemma \ref{lem:AR:unbiased} still holds if assumption \eqref{assumpt:global} is relaxed and replaced by one of the two following assumptions:
+\begin{assumpt}\label{assumpt:ysupport}
+\forall~0\leq k\leq n,~\exists \hat{\sigma}_+ \text{ such that }\forall (j,y) \in [[1,N]]\times  \mathbb{R}^d,~~\widehat{q}_k(\xi^j_k,y,\zeta_k)\leq \hat{\sigma}_+
+\end{assumpt}
+\vspace{-\baselineskip}
+\begin{assumpt}\label{assumpt:local}
+\forall~0\leq k\leq n,~\exists \hat{\sigma}_+ \text{ such that }\forall j \in [[1,N]],~~\widehat{q}_k(\xi^j_k,\xi^j_{k+1},\zeta_k)\leq \hat{\sigma}_+
+\end{assumpt}
+It is worth noting that under assumptions \eqref{assumpt:global} or \eqref{assumpt:ysupport}, the linear complexity property of PaRIS algorithm still holds, whereas if only assumption \eqref{assumpt:local} holds, the resulting algorithm has a quadratic complexity.
 %Moreover, the AR algorithm still stands if one can define 
@@
 \end{equation}
-On the r.h.s. of \eqref{eq:GPE1}, the product over $\kappa$ elements is bounded by 1, therefore, a sufficient condition to obtain an upper bound  $\hat{q}_k$ is that the function:
+On the r.h.s.
+of \eqref{eq:GPE1}, the product over $\kappa$ elements is bounded by 1, therefore, a sufficient condition to satisfy of the assumptions \eqref{assumpt:global}-\eqref{assumpt:local} is that the function:
 \begin{align}
@@
 \end{align}
-is upper bounded almost surely by $\hat{\sigma}_+^k$.
-As said above, the GRand PaRIS algorithm only requires a bound $\hat{\sigma}_+^k$ for the function $\rho_{\Delta_k}$ where $x \in \{\xi_{k}^1,\dots,\xi_{1}^N\}$.
-In order to improve the acceptance rate (at a cost of a quadratic complexity in $N$), we can always compute, for all $1\le i\le N$, the bound
-\begin{equation}
-\hat{\sigma}_{+}^{k,i}= \underset{1 \le j\le N}{\text{sup}}\rho_{\Delta_k}(\xi_{k-1}^j,\xi_k^i)\eqsp, \label{eq:quadr:bound}
-\end{equation}
-as long as $\Lo_\mw$ is bounded almost surely.
-This might increase significantly the acceptance rate of the algorithm, and therefore reduce the number of drawings of the random variable $\zeta_k$, which has a much higher cost than the computation of $\rho$, as it requires simulation of Brownian Bridges.
-It is worth noting here that this latter option can also avoid numerical optimization if no analytical version of $\hat{\sigma}_+^k$ is available.
-In the example of the SINE model presented below, for $N=400$ particles, the algorithm with linear complexity  has a 40\% higher computation time than the algorithm with a quadratic complexity.
-Actually, this condition is always satisfied for models in the domains $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, i.e.
-domains for which the exact algorithms EA1 and EA2 can be used.
-
+is upper bounded almost surely by $\hat{\sigma}_+$.
+In particular, if $Lo_\mw$ is bounded almost surely, \eqref{eq:rho:func} always satisfies assumption \eqref{assumpt:local}, therefore, algorithm \eqref{alg:Ozaki:PaRIS} can be perfomed, in the worse case with a quadratic complexity in $N$.
+Therefore, this condition is always satisfied for models in the domains $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, i.e.\ domains for which the exact algorithms EA1 and EA2 can be used.\\
+In that case, and when \eqref{assumpt:global} or \eqref{assumpt:ysupport} holds, it can be nonetheless of practical interest to choose the $\hat{\sigma}_+$ corresponding to \eqref{assumpt:global}.
+Indeed, this might increase significantly the acceptance rate of the algorithm, and therefore reduce the number of drawings of the random variable $\zeta_k$, which has a much higher cost than the computation of $\rho$, as it requires simulation of Brownian Bridges.
+Moreover, this latter option can also avoid numerical optimization if no analytical version of $\hat{\sigma}_+^k$ is available.
+In practice, we found this option more efficient in terms of computation time when $N$ has moderate values.
 %\subsection*{Note on the complexity and practical implementation}
@@
 where $\tilde{q}_k(x_{k-1},x_k)$ is the p.d.f.
-of Gaussian distibution with mean $\alpha(x_{k-1})\Delta_k$ and variance $\Delta_kI_d$, i.e.
-the Euler approximation of equation \eqref{eq:target:sde}.
+of Gaussian distibution with mean $\alpha(x_{k-1})\Delta_k$ and variance $\Delta_kI_d$, i.e.\ the Euler approximation of equation \eqref{eq:target:sde}.
 As the observation model is linear and Gaussian, the proposal distribution is therefore Gaussian with explicit mean and variance.
@@
 %for some $\xi_{k+1}$ that has no importance in this result.
-We assume $\omega_k$ are normalized weights, i.e.
-$\sum_\ell \omega_k^l=1$.
+We assume $\omega_k$ are normalized weights, i.e.\ $\sum_\ell \omega_k^l=1$.
 The candidate will therefore be sampled from $Y$, a r.v.

### 2017-02-27 ParisEM.tex
Final revision #6

@@
 
-%Statistical inference for partially observed stochastic differential equations is a challenging task since some elementary quantities, such as transition probabilities, are not available explicitly.
-Statistical inference for partially observed stochastic differential equations often requires to solve bayesian filtering and smoothing problems, i.e.\ the computation of the posterior distributions of sequences of hidden states given observations.
+Statistical inference for partially observed state sequences often requires to solve bayesian filtering and smoothing problems, i.e.\ the computation of the posterior distributions of sequences of hidden states given observations.
 The filtering problem refers to the estimation, for each $0\le k \le n$,  of the distributions of the hidden state $X_k$ given the observations $(Y_0,\ldots,Y_k)$.
 Smoothing stands for the estimation of the distributions of the sequence of states $(X_{k},\ldots,X_{p})$ given observations $(Y_{0},\ldots,Y_{\ell})$ with $0\le k\le p \le \ell \le n$.
-These posterior distributions are crucial to compute maximum likelihood estimators of the unknown model parameters using the observations $(Y_0,\ldots,Y_n)$ only.
-% $\theta\mapsto \ell_{\theta}^{n}$ given by
-%\begin{equation*}
-%\ell_{\theta}^{n}(Y_{0:n}) = \log\left(\int p_{\theta}(x_{0:n},Y_{0:n})\,\rmd x_{0:n}\right)\eqsp,
-%\end{equation*}  
-%where $p_{\theta}$ is the complete data likelihood:
-%\begin{equation*}
-%p_{\theta}(x_{0:n},Y_{0:n}) = \chi(x_0)g_{0}(x_0)\prod^{n-1}_{k=0}\qk(x_{k},x_{k+1})g_{k+1}(x_{k+1})\eqsp.
-%\end{equation*}
-For instance, the E-step of the EM algorithm introduced in \cite{dempster:laird:rubin:1977} %iteratively builds a sequence of parameter estimates following two steps.
-%following the two steps:
-%\begin{enumerate}
-%	\item {\bf E-step}: compute $\theta \mapsto Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log p_{\theta}(X_{0:n},Y_{0:n})\middle|Y_{0:n}\right]$, where $\mathbb{E}_{\theta}\left[\cdot\middle|Y_{0:n}\right]$ is the conditional expectation given $Y_{0:n}$ when the parameter value is $\theta$.
-%	\item {\bf M-step}: choose $\theta_{p+1}$ as a maximizer of $\theta \mapsto Q(\theta,\theta_{p})$.
-%\end{enumerate}
-%step 
-boils down to the computation of a conditional expectation of an additive functional of the hidden states given all the observations up to time $n$. 
-%\begin{multline*}
-%Q(\theta,\theta_{p}) = \mathbb{E}_{\theta_p}\left[\log \left(\chi(X_0)g_{0}(X_0)\right)\middle|Y_{0:n}\right] \\
-%+ \sum_{k=0}^{n-1}\mathbb{E}_{\theta_p}\left[\log \left(\qk(X_{k},X_{k+1})g_{k+1}(X_{k+1})\right)\middle|Y_{0:n}\right] \eqsp.
-%\end{multline*}
+These posterior distributions are crucial to compute maximum likelihood estimators of unknown parameters using the observations $(Y_0,\ldots,Y_n)$ only.
+For instance, the E-step of the EM algorithm introduced in \cite{dempster:laird:rubin:1977}  boils down to the computation of a conditional expectation of an additive functional of the hidden states given all the observations up to time $n$.
 Similarly, by Fisher's identity, recursive maximum likelihood estimates may be computed using the gradient of the loglikelihood which can be written as the conditional expectation of an additive functional of the hidden states.
- See \cite[Chapter $10$ and $11$]{cappe:moulines:ryden:2005}, \cite{kantas:doucet:signh:2015,doucet:poyiadjis:singh:2011,lecorff:fort:2013a,lecorff:fort:2013b}
+ See \cite[Chapter $10$ and $11$]{cappe:moulines:ryden:2005}, \cite{kantas:doucet:signh:2015,lecorff:fort:2013a,lecorff:fort:2013b,doucet:poyiadjis:singh:2011}
 for further references on the use of these smoothed expectations of additive functionals  applied to maximum likelihood parameter inference in latent data models.
@@
 In this paper, we propose to use Sequential Monte Carlo (SMC) methods to approximate smoothing distributions with random particles associated with importance weights.
- \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers for state space models by combining importance sampling steps to propagate particles with importance resampling steps to duplicate or discard particles according to their importance weights.
+ \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers for state space models by combining importance sampling steps to propagate particles with resampling steps to duplicate or discard particles according to their importance weights.
 Unfortunately, these methods cannot be applied directly to partially observed stochastic differential equations since some elementary quantities, such as transition densities of the hidden states, are not available explicitly.
@@
 However, even the most recent discretization based approximations of the transition densities induce a systematic bias of particle based approximations of posterior distributions, see for instance \cite{delmoral:jacod:protter:2001}.
-To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE) introduced in \cite{fearnhead:papaspiliopoulos:roberts:2008}.
+To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE).
 In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities.
@@
 Then, a backward pass is performed to compute new weights (FFBS) or sample trajectories (FFBSi) in order to approximate the smoothing distributions.
-%In the case of partially observed SDE, this forward pass may be performed with the SMC steps combined with the GPE of \cite{fearnhead:papaspiliopoulos:roberts:2008}. 
-%Then, the backward pass of the FFBS algorithm keeps all the particles sampled during the forward pass and computes new weights to take into account the information brought by all the observations from time $n$ to time $0$.
-Instead of computing new weights, the FFBSi  algorithm samples independently backward in time particle trajectories using the particles and weights produced by the forward pass.
 Recently, \cite{olsson:westerborn:2016} proposed a new SMC algorithm, the particle-based rapid incremental smoother (PaRIS), to approximate on-the-fly (i.e.\ using the observations as they are received) smoothed expectations of additive functionals.
@@
 The proposed algorithm allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles.
-The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to SDE is that the accept reject mechanism ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
+The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to SDE is that the accept reject mechanism introduced in \cite{douc:garivier:moulines:olsson:2011} ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
 The usual FFBS and FFBSi algorithms may not be extended this easily since they both require the computation of weights defined as ratios involving the transition densities, thus replacing these unknown quantities by unbiased estimates does not lead to unbiased estimators of the weights.
@@
 
-%We also propose to improve the algorithm of \cite{fearnhead:papaspiliopoulos:roberts:2008} by extending the range of diffusion processes for which SMC methods can be applied.
-To do so, we introduce a new class of GPE based on the Ozaki discretization of the target SDE, see \cite{ozaki:1992,shoji:ozaki:1998}.
-These new GPE estimators use an order one Taylor expansion of the drift of the SDE to obtain more efficient estimates than Brownian Bridge based GPE.
-
-Section~\ref{sec:rwparis} describes the GRand PaRIS algorithm  aimed to approximate smoothed additive functionals using unbiased estimates of the transition density of the hidden states and details the application of this algorithm when the transition density may be approximated using a GPE.
+Section~\ref{sec:rwparis} describes the proposed algorithm to approximate smoothed additive functionals using unbiased estimates of the transition density of the hidden states and details the application of this algorithm when the transition density may be approximated using a GPE.
 In Section~\ref{sec:convergence}, classical convergence results for SMC smoothers are extended to the setting of this paper and illustrated with numerical experiments in Section~\ref{sec:exp}. 
@@
 It is assumed that $\alpha$ is of the form $\alpha(x) = \nabla_x A(x)$ where $A: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function.
-% A popular choice of $A$  is $A(x) = \log \pi(x)/2$ leading to a Langevin diffusion which may be used to approximate a given target distribution $\pi$ on $\mathbb{R}^d$.
-See for instance \cite{roberts:tweedie:1996} for conditions under which this diffusion converges  at exponential rate to $\pi$.
-% SDE where the drift is given by the gradient of a potential function have been widely used in ecology for instance, see \cite{brillinger:et:al:2011,harris:blackwell:2013,preisler:et:al:2013} and the references therein.
-The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{k=0,\dots,n}$ in $(\mathbb{R}^m)^{n+1}$. 
+The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{0\le k \le n}$ in $(\mathbb{R}^m)^{n+1}$.
 For all $0\le k \le n$, the distribution of $Y_k$ given $X_k:= X_{t_k}$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,\cdot) = g_k(X_k)$. 
 The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
-For all $1\le k \le n$, the distribution of $X_{k+1} $ conditional on $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$.
+For all $0\le k \le n-1$, the conditional distribution of $X_{k+1} $ given $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$.
 
@@
 PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSi algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions.
-%The FFBS algorithm has a computational complexity of order $\mathcal{O}(N^{k'-k+1})$ to approximate joint smoothing distributions of the form $\phi_{k:k'|n}$ with $k<k'$ which makes it prohibitive. 
-%An important feature of the FFBS algorithm shown in \cite{mongillo:deneve:2008,cappe:2011,delmoral:doucet:singh:2010} is that it can be implemented using only a forward pass when approximating smoothed expectations of additive functions.
-%Instead of computing new weights in the backward pass, the FFBSi  samples backward new trajectories from time $n$ to time $0$ among all the $N^{n+1}$ trajectories which can be chosen in $\{\xi^{\ell}_k\}$, $1\le \ell\le N$ and $0\le k\le n$.
-Contrary to the FFBS algorithm, the FFBSi algorithm samples trajectories and may be used to approximate joint smoothing distributions without increasing its complexity.
-%Nevertheless, the computation of the transition probabilities to select particles in the backward pass makes it a $\mathcal{O}(N^2)$ algorithm.
-In the case where the transition probability $q_k$ is upper bounded, \cite{douc:garivier:moulines:olsson:2011} introduced an acceptance rejection mechanism to implement the FFBSi algorithm with a $\mathcal{O}(N)$ complexity, see also \cite{dubarry:lecorff:2011}.
 It combines both the forward only version of the FFBS algorithm with the sampling mechanism of the FFBSi algorithm.
@@
 Therefore, the computation of the importance weights $\omega_{k}^{\ell}$ and of the acceptance ratio of \cite{douc:garivier:moulines:olsson:2011} is not tractable.
-Following \cite{fearnhead:papaspiliopoulos:roberts:2008,olsson:strojby:2011}, filtering weights can be approximated by replacing $\qk(\xi^{\ell}_{k},\xi_{k+1}^{i})$  by an unbiased  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable such that:
-\begin{align*}
-\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)&> 0~~\text{a.s}\eqsp,\\
-\mathbb{E}\left[\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)\middle| \mathcal{G}_{k+1}^N\right] &= \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})\eqsp,
-\end{align*}
+Following \cite{fearnhead:papaspiliopoulos:roberts:2008,olsson:strojby:2011}, filtering weights can be approximated by replacing $\qk(\xi^{\ell}_{k},\xi_{k+1}^{i})$  by an unbiased  estimator $\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)$, where $\zeta_k$ is a random variable in $\mathbb{R}^q$ such that:
+\[
+\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)> 0~~\text{a.s}\quad\mbox{and}\quad
+\mathbb{E}\left[\widehat{q}_k(\xi^{\ell}_{k},\xi_{k+1}^{i};\zeta_k)\middle| \mathcal{G}_{k+1}^N\right] = \qk(\xi^{\ell}_{k},\xi_{k+1}^{i})\eqsp,
+\]
 where, for all $0\le k \le n$, 
-
 \begin{align*}
@@
 \end{align*}
-
-Practical choices for $\zeta_k$ are presented in the paragraph devoted to the construction of bounded estimators for $q$ and more specifically equation \ref{eq:GPE1} presents the choice made for the implementation of such estimator in our context.
-
+Practical choices for $\zeta_k$ are discussed below, see for instance \eqref{eq:GPE1} which presents the choice made for the implementation of such estimators in our context.
 In the case where $\qk$ is unknown, the filtering weights in \eqref{eq:importance:weights} then become:
@@
 \end{equation}
-Therefore, to obtain a generalized random version of PaRIS, we only need to be able to sample from weight $\Lambda_k^N(i,\ell)$ in the case of SDE based HMM.\\
-We need to define the following assumption:\\
+Therefore, to obtain a generalized random version of PaRIS algorithm, we only need to be able to sample from the discrete probability distribution $\Lambda_k^N(i,\cdot)$ in the case of SDE based HMM.
+Consider the following assumption: for all  $0\leq k\leq n$, there exists a random variable $\hat{\sigma}^k_+$ measurable with respect to $\mathcal{G}_{k+1}^N$ such that,
 \begin{assumpt}
 \label{assumpt:global}
-\forall~0\leq k\leq n,~\exists \hat{\sigma}_+ \text{ such that }\forall (x,y) \in\mathbb{R}^d\times  \mathbb{R}^d,~~\widehat{q}_k(x,y,\zeta_k)\leq \hat{\sigma_+}
+\mathrm{sup}_{x,y,\zeta}\;\widehat{q}_k(x,y;\zeta)\leq \hat{\sigma}^k_+\eqsp.
 \end{assumpt}
-Then, the following lemma provides the missing step of a Generalized Random PaRIS (GRand Paris)
 \begin{lemma}
 \label{lem:AR:unbiased}
-Let $0\le k\le n-1$ and $1\le i \le N$, and suppose $A_1$ holds may be upper bounded almost surely by a constant $\hat{\sigma}_+$. Let's define the random variable $J_k^i$ as follow:
+Assume that $A_1$ holds for some $0\le k\le n-1$.  For all $1\le i \le N$, define the random variable $J_k^i$ as follows:
 \begin{algorithmic}
-\STATE \textsc{Sampling step:} Sample independently $
-\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
-\IF{ $$U \leq \frac{\widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)}{\hat{\sigma}_+},$$}
-\STATE Set $J_k^{i,\ell} = J$.
-\ELSE 
-\STATE Return to \textbf{\sc Sampling Step}
-\ENDIF
+\REPEAT
+\STATE Sample independently $\zeta$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
+\UNTIL{$U \leq \widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta)/\hat{\sigma}^k_+$.}
+\STATE Set $J_k^i = J$.
+%\STATE \textsc{Sampling step:} Sample independently $
+%\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
+%\IF{ $U \leq \widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)/\hat{\sigma}_+$,}
+%\STATE Set $J_k^{i,\ell} = J$.
+%\ELSE 
+%\STATE Return to \textbf{\sc Sampling Step}
+%\ENDIF
 \end{algorithmic}
-Then, the conditional probability distribution given $\mathcal{G}_{k+1}^N$ of $J_{k}^{i,\ell}$ is $\Lambda_{k}^N(i,\cdot)$. 
+Then, the conditional probability distribution given $\mathcal{G}_{k+1}^N$ of $J_{k}^{i}$ is $\Lambda_{k}^N(i,\cdot)$. 
 \end{lemma}
@@
 \end{proof}
-Note that lemma \ref{lem:AR:unbiased} still holds if assumption \eqref{assumpt:global} is relaxed and replaced by one of the two following assumptions:
+Note that Lemma~\ref{lem:AR:unbiased} still holds if assumption \eqref{assumpt:global} is relaxed and replaced by one of the two following assumptions:
 \begin{assumpt}\label{assumpt:ysupport}
-\forall~0\leq k\leq n,~\exists \hat{\sigma}_+ \text{ such that }\forall (j,y) \in [[1,N]]\times  \mathbb{R}^d,~~\widehat{q}_k(\xi^j_k,y,\zeta_k)\leq \hat{\sigma}_+
+\mathrm{sup}_{j,y,\zeta}\;\widehat{q}_k(\xi^j_k,y,\zeta)\leq \hat{\sigma}^k_+\;.
 \end{assumpt}
@@
 \begin{assumpt}\label{assumpt:local}
-\forall~0\leq k\leq n,~\exists \hat{\sigma}_+ \text{ such that }\forall j \in [[1,N]],~~\widehat{q}_k(\xi^j_k,\xi^j_{k+1},\zeta_k)\leq \hat{\sigma}_+
+\mathrm{sup}_{i,j,\zeta}\;\widehat{q}_k(\xi^j_k,\xi^j_{k+1},\zeta)\leq \hat{\sigma}^k_+\;.
 \end{assumpt}
-It is worth noting that under assumptions \eqref{assumpt:global} or \eqref{assumpt:ysupport}, the linear complexity property of PaRIS algorithm still holds, whereas if only assumption \eqref{assumpt:local} holds, the resulting algorithm has a quadratic complexity.
-%Moreover, the AR algorithm still stands if one can define 
-%\begin{equation}
-%\hat{\sigma}_{+}^{k,i}= \underset{j=1,\dots,N}{\text{sup}}\rho_\Delta(\xi_{k-1}^j,\xi_k^i)\eqsp, \label{eq:quadr:bound}
-%\end{equation}
-%which is a weaker assumption.
-However, in this case, the resulting algorithm has a quadratic complexity in $N$.
-Nevertheless, in practice, this choice can result in a much more efficient acceptance rate, which would reduce the computational time of the algorithm.
+It is worth noting that under assumptions \eqref{assumpt:global} or \eqref{assumpt:ysupport}, the linear complexity property of PaRIS algorithm still holds, whereas if only assumption \eqref{assumpt:local} holds, the algorithm has a quadratic complexity.
+
 \subsection*{Bounded estimator of $q_k$}
@@
 \end{align}
-is upper bounded almost surely by $\hat{\sigma}_+$.
-In particular, if $Lo_\mw$ is bounded almost surely, \eqref{eq:rho:func} always satisfies assumption \eqref{assumpt:local}, therefore, algorithm \eqref{alg:Ozaki:PaRIS} can be perfomed, in the worse case with a quadratic complexity in $N$.
-Therefore, this condition is always satisfied for models in the domains $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, i.e.\ domains for which the exact algorithms EA1 and EA2 can be used.\\
-In that case, and when \eqref{assumpt:global} or \eqref{assumpt:ysupport} holds, it can be nonetheless of practical interest to choose the $\hat{\sigma}_+$ corresponding to \eqref{assumpt:global}.
-Indeed, this might increase significantly the acceptance rate of the algorithm, and therefore reduce the number of drawings of the random variable $\zeta_k$, which has a much higher cost than the computation of $\rho$, as it requires simulation of Brownian Bridges.
-Moreover, this latter option can also avoid numerical optimization if no analytical version of $\hat{\sigma}_+^k$ is available.
+is upper bounded almost surely by $\hat{\sigma}^k_+$.
+In particular, if $\Lo_\mw$ is bounded almost surely, \eqref{eq:rho:func} always satisfies assumption \eqref{assumpt:local} and Algorithm~\ref{alg:Ozaki:PaRIS} can be used.
+This condition is always satisfied for models in the domains $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, i.e.\ domains for which the exact algorithms EA1 and EA2 can be used.
+
+When \eqref{assumpt:global} or \eqref{assumpt:ysupport} holds, it can be nonetheless of practical interest to choose the bound $\hat{\sigma}^k_+$ corresponding to \eqref{assumpt:local}.
+Indeed, this might increase significantly the acceptance rate of the algorithm, and therefore reduce the number of drawings of the random variable $\zeta$, which has a much higher cost than the computation of $\rho$, as it requires simulations of Brownian Bridges.
+Moreover, this latter option can also avoid numerical optimization if no analytical expression of $\hat{\sigma}_+^k$ is available.
 In practice, we found this option more efficient in terms of computation time when $N$ has moderate values.
-%\subsection*{Note on the complexity and practical implementation}
-%As said above, as long as one can define a $\sigma_+^k$ such that
-%$$ \forall\eqsp 0\leq k\leq n-1,\eqsp 1\leq i\leq N,\eqsp y\in\mathbb{R}^d,\eqsp \hat{q}_k(\xi_k^i,y,\zeta_k)<\hat{\sigma}_+^k \text{ a.s.,}$$ then the resulting GRand Paris algorithm has a linear complexity in $N$.\\
-%However, it is worth noting here that for a reasonable amount of particles, it might be more efficient in terms of computation time to compute, for all $i$, the bound given by \eqref{eq:quadr:bound}.
-Indeed, this bound requires, at each time step $k$, $N^2$ computation of the function $\rho$ given above.
-However, this might increase significantly the acceptance rate of the AR algorithm, and therefore reduce the number of drawings for the random variable $\zeta_k$, which has a much higher cost than the computation of $\rho$, as it requires simulation of Brownian Bridges.
-It is worth noting here that this latter option can also avoid numerical optimization if no analytical version of $\hat{\sigma}_+^k$ is available.\\
-%In the example of the SINE model presented below, for $N=400$ particles, the linear complexity algorithm has a 40\% higher computation time than the algorithm with a quadratic complexity.
-In practice, in addition to $n$ and $N$, the computation time will increase with
-%\begin{itemize}
-%\item $\tilde{N}$ the number of retrospective sampling;
-%\item $M$ the number of Monte Carlo samples to obtain $\hat{q}$;
-%\item $\Delta$ the acquisition time step;
-%\item The expected value of $\U_\mw-\Lo_\mw$.
-%\end{itemize}
 
-%\section{The OzIRis algorithm}
-%\label{sec:PaRIS:SDE}
-%In \cite{olsson:strojby:2011}, $\qk$ is estimated using GPE based on a representation of $\qk$ as an expectation under the probability measure of a Brownian bridge. 
-%In order to improve the Monte Carlo estimation of $\qk$,  other diffusion bridges may be used as long as they can be simulated efficiently.
-A simple idea is to consider a diffusion obtained as an approximation of the target SDE \eqref{eq:target:sde}.
-In \cite{giesecke:schwenkler:2016}, the authors proposed to replace the Brownian bridge by a Brownian motion with constant drift between two consecutive observations.
-%As an alternative, it is also possible to follow \cite{ozaki:1992,shoji:ozaki:1998} to propose a higher order approximation of the drift term $\alpha$.
-The authors used a linear approximation of the diffusion drift $\alpha$, together with a constant approximation of the volatility over each time interval $(t_{k},t_{k+1})$.
-%
-%For all $0\le k\le n-1$, let $(Z^{k}_t)_{t\ge 0}$ be governed by the following stochastic differential equation for $t\in(t_{k},t_{k+1})$:
-%\begin{equation}
-%\label{eq:SDE:Ozaki}
-%\Oz^k_{t_{k}} = X_{k}\quad\mbox{and}\quad\rmd \Oz^k_t = \widetilde{\alpha}_k(\Oz^k_t)\rmd t + \rmd W_t\eqsp,
-%\end{equation}
-%where $\widetilde{\alpha}_k$ is an instrumental drift. The probability density function of the transition kernel associated with $\Oz^k$ is written $\widetilde{q}_k$.
-%Writing $\mathbb{Q}_k$ (resp.
-$\mathbb{W}_{k}$) the probability measure induced by $(X_s)_{t_{k}\le s\le t_{k+1}}$  (resp.
-$(\Oz_s)_{t_{k}\le s\le t_{k+1}}$) on $(\mathsf{C},\mathcal{C})$, Girsanov theorem yields:
-%\begin{multline*}
-%\frac{\rmd \mathbb{Q}_k}{\rmd \mathbb{W}_{k}}(w) = \mathrm{exp}\left\{H_{k}(\mw_{t_{k+1}}) - H_{k}(\mw_{t_{k}})\right\}\\
-%\times \exp\left\{- \frac{1}{2}\int_{t_{k}}^{t_{k+1}}\left\{\| \alpha_{}(\mw_s)\|^2-\|\widetilde{\alpha}_k(\mw_s)\|^2+\Delta H_k(\mw_s)\right\}\rmd s\right\}\eqsp,
-%\end{multline*}
-%where $\nabla H_k = \alpha -\widetilde{\alpha}_k$ and $\Delta$ is the Laplace operator. Then, the probability measures conditioned on hitting $X_{k+1}$ at time $t_{k+1}$ satisfy:
-%\begin{multline}
-%\label{eq:Girsanov}
-%\frac{\rmd \mathbb{Q}_k^{X_{k+1}}}{\rmd \mathbb{W}_k^{X_{k+1}}}(\mw) = \frac{\widetilde{\qk}(X_{k},X_{k+1})}{\qk (X_{k},X_{k+1}) }\mathrm{exp}\left\{H_{k}(X_{k+1}) - H_{k}(X_{k})\right\} \\
-%\times \mathrm{exp}\left\{- \frac{1}{2}\int_{t_{k}}^{t_{k+1}}\left\{\|\alpha_{}(\mw_s)\|^2-\|\widetilde{\alpha}_{k}(\mw_s)\|^2+\Delta H_{k}(\mw_s)\right\}\rmd s\right\}\eqsp.
-%\end{multline}
-%\subsection{Linear approximation of the drift}
-%In order to improve the performance of the GPE introduced in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,giesecke:schwenkler:2016}, the instrumental drift is chosen as the Ozaki  \cite{ozaki:1992,shoji:ozaki:1998} linear approximation of the target drift:
-%\begin{equation}
-%\label{eq:alpha:linear}
-%\widetilde{\alpha}_k: x\mapsto \Jk x + \alpha(X_{k}) - \Jk X_{k}\eqsp,
-%\end{equation}
-%where $\Jk$ is the Jacobian of $\alpha$ evaluated at $X_k$.
-In this case, the transition kernel associated with $\Oz^k$ is given in Section~\ref{sec:ornstein:bridge} and the function $H_k$ may be chosen as
-%\[
-%H_{k}: x \mapsto A_{}(x) - \frac{1}{2}x'\Jk x- (\alpha(X_{k}) - \Jk X_{k-1})'x\eqsp.
-%\]
-%%Consider the following assumptions on the model:
-%%\begin{enumerate}[(a)]
-%%\item For all $0\le k \le n-1$, $x\mapsto \alpha_{}(x) - \alpha_{}(X_k) - \Jk(x-X_k)$ is bounded.
-%%\item For all $0\le k \le n-1$, $x\mapsto \mathrm{Tr}\left(J_{\alpha_{}}(x) - \Jk\right)$ is bounded, where $\mathrm{Tr}$ is the trace operator.
-%%\end{enumerate}
-%Assume that there exists $\ell_k$ such that:
-%\[
-%\mathrm{inf}_{u\in \mathbb{R}^d} \left\{\| \alpha_{}(u)\|^2-\|\widetilde{\alpha}_{k}(u)\|^2+\Delta H_{k}(u)\right\}/2\ge \ell_k
-%\]
-%and
-%\[
-%\phi_{k}: x\mapsto \left\{\|\alpha_{}(x)\|^2-\|\widetilde{\alpha}_{k}(x)\|^2+\Delta H_{k}(x)\right\}/2-\ell_k\eqsp.
-%\]
-%Assume also that the model is such that there exists $\Lambda_k$ satisfying  $\mathrm{sup}_x\eqsp \phi_{k}(x)\le \Lambda_{k}$.
-These assumptions might seem somehow restrictive but are necessary to define the transition density estimator in the linear case.
-The alternative given by \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,giesecke:schwenkler:2016} when the instrumental drift is constant does not assume any upper bound for $\phi_{k}$ but requires to sample skeletons of a Brownian bridge given its minimum/maximum between $t_k$ and $t_{k+1}$.
-To the best of our knowledge, there is no closed from formula to sample skeletons of an Ornstein-Uhlenbeck bridge given its extremal values on an interval.
-Then,
-%\begin{multline*}
-%\qk(X_{k},X_{k+1}) = \widetilde{\qk}(X_{k},X_{k+1}) \eqsp\mathrm{exp}\left\{H_{k}(X_{k+1}) - H_{k}(X_{k})-\ell\Delta_k\right\}\\
-%\times\mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}\left[\mathrm{exp}\left\{-\int_{0}^{ \Delta_k}\phi_{k}(\mw_s)\rmd s\right\}\right]\eqsp,
-%\end{multline*}
-%where $\mathbb{E}_{\mathbb{W}_{k}^{X_{k+1}}}$ is the expectation under the law of the bridge process defined by \eqref{eq:SDE:Ozaki} and \eqref{eq:alpha:linear}, starting at $X_{k}$ at time $0$ and ending at $X_{k+1}$ at time $\Delta_k$.
-%An unbiased estimator of $\qk(X_{k},X_{k+1})$ is then obtained by estimating the expectation in the previous equation. 
-%Following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, if $\omega \sim\mathbb{W}_{k}^{X_{k}}$, $\kappa$ is a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and if $(U_j)_{1\le j\le \kappa}$ are independent uniform random variables on $[0,t]$,
-%\begin{align*}
-
-%Then, for all $0\le k \le n-1$, $1\le i\le N$, the importance weight $\omega_k^{\ell}$ is replaced by the unbiased estimator:
-%\begin{equation}
-%\label{eq:random:weight}
-%\widehat{\omega}_{k+1}^{i} = \frac{M^{-1}\sum_{m=1}^M \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta_k^m)g_{k+1}(\xi^{i}_{k+1})}{\vartheta_{k+1}(\xi^{I^{i}_{k+1}}_{k}) p_{k} (\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})}\eqsp,
-%\end{equation}
-%where $\{\zeta_k^i\}_{i=1}^M$ are independent copies of $\zeta_k$.
-The resulting Ozaki in PaRIS (OzIRis) algortihm, which provides an approximation of \eqref{def:addfunc}, is summarized in Algorithm~\ref{alg:Ozaki:PaRIS}.
 \begin{algorithm}
@@
 \FORALL{$i \in 1,\dots, N$}
-\STATE Sample $\xi_0^i \sim\eta_0$, $\tau_0^i = 0$  and  $\omega_0^i = g_0(\xi_0^i)\chi_0(\xi_0^i)/\eta_0(\xi_0^i)$.
+\STATE Sample $\xi_0^i \sim\eta_0$, $\tau_0^i = 0$  and  $\widehat \omega_0^i = g_0(\xi_0^i)\chi_0(\xi_0^i)/\eta_0(\xi_0^i)$.
 \ENDFOR
@@
 \FORALL{$\ell \in 1,\dots, \widetilde N$} 
-\STATE \textbf{\sc Sampling Step} \\
-Sample independently $\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to  $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
-\IF{ $$U \leq \frac{\widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)}{\hat{\sigma}^k_+},$$}
-\STATE Set $J_k^{i,\ell} = j$.
-\STATE Update $\tau_{k+1}^i = \tau_{k+1}^i + \left(\tau^{J_k^{i,\ell}}_{k} + h_k(\xi^{J_k^{i,\ell}}_{k},\xi^i_{k+1})\right)/\tilde{N}$;
-\ELSE 
-\STATE Return to \textbf{\sc Sampling Step}
-\ENDIF
+\STATE Sample $J_k^{i,\ell}$ as in Lemma~\ref{lem:AR:unbiased}.
+\STATE Update $\tau_{k+1}^i = \tau_{k+1}^i + (\tau^{J_k^{i,\ell}}_{k} + h_k(\xi^{J_k^{i,\ell}}_{k},\xi^i_{k+1}))/\tilde{N}$.
+%\STATE \textbf{\sc Sampling Step} \\
+%Sample independently $\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to  $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
+%\IF{ $$U \leq \frac{\widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)}{\hat{\sigma}^k_+},$$}
+%\STATE Set $J_k^{i,\ell} = j$.
+%\STATE Update $\tau_{k+1}^i = \tau_{k+1}^i + \left(\tau^{J_k^{i,\ell}}_{k} + h_k(\xi^{J_k^{i,\ell}}_{k},\xi^i_{k+1})\right)/\tilde{N}$;
+%\ELSE 
+%\STATE Return to \textbf{\sc Sampling Step}
+%\ENDIF
 \ENDFOR
@@
 \end{algorithm}
-
-%\subsection{Constant approximation of the drift}
-%\label{sec:constantdrift}
-%In the case where the function $\phi_k$ cannot be upper bounded when $\widetilde{\alpha}_k$ is set as a linear approximation of $\alpha$, \cite{giesecke:schwenkler:2016} suggests to set $\widetilde{\alpha}_k$ as a constant value $\widetilde{\alpha}_k= \rho_k$.
-The GPE of \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} is a special case with $\rho_k = 0$.
-In this case, $H_k$ may be chosen as $H_k: x\mapsto A(x) - \rho_k x$.
-Then, following \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006,olsson:strojby:2011}, it is enough to sample a pair $(L_k,U_k)$ of random variables and a Brownian bridge $\mw\sim \mathbb{W}_k^{X_{k+1}}$ (or a Brownian bridge with a constant drift $\rho_k$) such that, for all $t_k\le t\le t_{k+1}$,
-%\[
-%L_k\le \phi_k(\mw_s)\le U_k\eqsp.
-%\]
-%Sampling the bridge conditionally on its minimum and maximum is possible using Bessel bridges as noted by \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.
-Then, the second term of \eqref{eq:unbiased:q} is replaced by:
-%\[
-%\mathrm{exp}\left\{-U_{k}\Delta_k\right\}\frac{( \Delta_k)^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(U_{k}-\phi_{k}(\mw_{U_j})\right)\eqsp.
-%\]
-%Here again, the choice of $\mu$ is restricted to distributions such that the associated estimator is upper bounded by an explicit constant $\hat{\sigma}_+^k$.
 
@@
 \end{hypA}
-%The following Lemma is proved in \cite[Lemma~11]{olsson:westerborn:2016}.
-%\begin{lemma}
-%\label{lem:smooth:rec}
-%Forward smoothing
-%\end{lemma}
 
@@
 \label{lem:iid}
-For all $0\le k \le n-1$, the random variables $\{\widehat{\omega}_{k+1}^i\tau_{k+1}^i\}_{i=1}^N$ are independent conditionally on $\mathcal{F}_k^{N}$ and%For all bounded measurable function $h$ on $\mathbb{R}^d$,
+For all $0\le k \le n-1$, the random variables $\{\widehat{\omega}_{k+1}^i\tau_{k+1}^i\}_{i=1}^N$ are independent conditionally on $\mathcal{F}_k^{N}$ and
 \[
@@
 \label{sec:exp}
-%This section investigates the performance of the proposed algorithm. The different tuning parameters are:
-%\begin{enumerate}[i)]
-%\item the number of particles $N$ and the precision parameter $\widetilde{N}$;
-%\item the proposal transition density $p_k$ and the adjustment multiplier $\vartheta_k$ used to propagate and weight particles;
-%\item the discrete distribution $\mu$ used to define $\widetilde{q}_{\theta}^{\Delta t_k}$;
-%\item the number of samples to compute $\widetilde{q}_{\theta}^{\Delta t_k}$.
-%\end{enumerate}
-%The performance of the algorithm is investigated on two known models based on diffusion process.
 This section investigates the performance of the proposed algorithm with the sine and log-growth models.
@@
 
-In order to evaluate the performance of the GRand PaRIS, the following strategy has been chosen.
-We compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method, for different values of the lag (namely, 1,2,5,10,50).
+In order to evaluate the performance of the proposed algorithm, the following strategy has been chosen.
+We compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method of \cite{olsson:strojby:2011}, for different values of the lag (namely, 1,2,5,10,50).
 The particle approximation of $\mathcal{Q}(\theta,\theta)$ for each model  is computed using each algorithm, see Figure~\ref{fig:res:SINE} for the SINE model (and respectively Figure~\ref{fig:res:LG} for the log-growth model).
-This estimation is performed 200 times, resulting in estimates $\hQ_1,\dots,\hQ_{200}$, using $\tilde{N}=2$ particles for the PaRIS part, and $M=30$ replications for the Monte Carlo approximation of each $q_k$.
+This estimation is performed 200 times to obtain the estimates $\hQ_1,\dots,\hQ_{200}$, using $\tilde{N}=2$ particles for PaRIS algorithm, and $M=30$ replications for the Monte Carlo approximation $\widehat q_k$ of each $q_k$.
 Moreover, the E step requires the computation of a quantity such as \eqref{def:addfunc} with $h_k= \log g_k + \log q_k$.
@@
 Figure \ref{fig:res:SINE} (resp.
-\ref{fig:res:LG} ) shows this estimates for an example on one simulated data set.\\
-The GRand PaRIS algorithm is compared with the fixed lag method of \cite{olsson:strojby:2011}.
-The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=1600$ so that both estimations have similar computational times.
+\ref{fig:res:LG} ) shows this estimate for an example on one simulated data set.
+The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=1600$ so that both estimations requre similar computational times.
 
@@
 
-%To complete the numerical analysis, the impact of the bridge process used to estimate $q_{\theta}^{\Delta t_k}$ is also illustrated.
-Instead of the Ozaki discretization, the following bridges process are considered:
-%\begin{enumerate}[i)]
-%\item a bridge with constant drift as proposed in \cite{giesecke:schwenkler:2016}.
-%\item a guided bridge process, see \cite{}.
-%\end{enumerate}
-%\paragraph{Choice of $\mu$}
-%A simple choice is to set $\mu$ as a Poisson distribution with mean $\Lambda_{\theta}t$ leading to the first estimator GPE-1:
-%\begin{multline*}
-%q^{t}_{\mathrm{GPE-1},\theta}(\kappa,\omega,(U_j)_{1\le j\le \kappa},x,y) =  \widetilde{q}^{t}_{\theta}(x,y)\eqsp\mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x)+\ell(\theta)t\right\}\Lambda_{\theta}^{-\kappa}\\
-%\times \prod_{j=1}^{\kappa}\left(\Lambda_{\theta}-\phi_{\theta}(\omega_{U_j})\right)\eqsp.
-%\end{multline*}
-
-% An important feature of this estimator is that under the assumptions of this section it may be bounded uniformly in $x,y$, $\sup_{x,y\in\mathbb{R}^d}  q^{t}_{\mathrm{GPE-1},\theta}(\kappa,\omega,(U_j)_{1\le j\le \kappa},x,y) \le M_\theta$ where:
-% \[
-% M_{\theta} = \sup_{x,y\in\mathbb{R}^d} \left(\widetilde{q}^{t}_{\theta}(x,y)\mathrm{exp}\left\{ H_{\theta}(y) - H_{\theta}(x)\right\}\right)\mathrm{exp}\left\{\ell(\theta)t\right\}\eqsp.
-% \]
-%
-%\paragraph{Choice of bridge process}
-%The approximation of $q_{\theta}^{\Delta t_k}$ described in the paper may be compared to the approximation provided by other bridges with affine drift functions.
-Following \cite{}, let $(Y^{\beta}_t)_{t_{k-1}\le t\le t_k}$ be governed by the following stochastic differential equation for $t_{k-1}\le t\le t_k$:
-%\[
-%Y^{\beta}_{t_{k-1}} = X_{k-1}\quad\mbox{and}\quad\rmd Y^{\beta}_t = \beta\rmd t + \rmd W_t\eqsp.
-%\]
-%As suggestd by \cite{}, a guided process between $t_{k-1}$ and $t_k$ may alo be used:
-%\[
-%Y^{\mathsf{guid}}_{t_{k-1}} = X_{k-1}\quad\mbox{and}\quad\rmd Y^{\mathsf{guid}}_t = \frac{X_k - Y^{\mathsf{guid}}_t}{t_k-t}\rmd t + \rmd W_t\eqsp.
-%\]
-%Both processes lead to estimators of $q_{\theta}^{\Delta t_k}$ following the same steps as in Section~\ref{}.
 \subsection*{Log-growth model}
@@
 It can be implemented for the class of models $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} with a linear complexity in $N$.
-% where $\beta\in\mathbb{R}^d$ is chosen by the user.
-Assume that $\alpha_{\theta}$ is of the form $\alpha_{\theta}(x) = \nabla_x A_{\theta}(x)$ where $A_{\theta}: \mathbb{R}^d \to \mathbb{R}$.
-Writing $\mathbb{Q}_{\theta}$ (resp.
-$\mathbb{W}^{\beta}_{\theta}$) the probability measure induced by $(X_s)_{0\le s\le t}$  (resp.
-$(Y_s)_{0\le s\le t}$) on $(\mathsf{C},\mathcal{C})$, Girsanov theorem yields:
-% \[
-% \frac{\rmd \mathbb{Q}_{\theta}}{\rmd \mathbb{W}^{\beta}_{\theta}}(w) = \mathrm{exp}\left\{H_{\theta}(\omega_t) - H_{\theta}(\omega_0) - \frac{1}{2}\int_{0}^{t}\left\{\|\alpha_{\theta}(\mw_s)-\beta\|^2+\Delta H_{\theta}(\mw_s)\right\}\rmd s\right\}\eqsp,
-% \]
-% where $H_{\theta}(x) = A_{\theta}(x) -\beta'x$. Then, the probability measures conditioned on hitting $y$ at time $t$ satisfy:
-% \[
-% \frac{\rmd \mathbb{Q}^{y}_{\theta}}{\rmd \mathbb{W}^{\beta,y}_{\theta}}(w) = \frac{\varphi_t(y-x-t\beta)}{q^{t}_{\theta}(x,y)} \mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x) - \frac{1}{2}\int_{0}^{t}\left\{\|\alpha_{\theta}(\mw_s)-\beta\|^2+\Delta H_{\theta}(\mw_s)\right\}\rmd s\right\}\eqsp.
-% \]
-% Assume also that for all $\theta$, there exists $\ell(\theta)$ such that:
-% \[
-% \mathrm{inf}_{u\in \mathbb{R}^d} \left\{\frac{1}{2}\left(\|\alpha_{\theta}(u)-\beta\|^2+\Delta H_{\theta}(u)\right)\right\}\ge \ell(\theta)\eqsp.
-% \]
-% Then
-% \[
-% q^{t}_{\theta}(x,y) = \varphi_t(y-x-t\beta)\eqsp\mathrm{exp}\left\{H_{\theta}(y) - H_{\theta}(x)+\ell(\theta)t\right\}\mathbb{E}_{\mathbb{W}^{\beta,y}}\left[\mathrm{exp}\left\{-\int_{0}^{t}\phi_{\theta}(\mw_s)\rmd s\right\}\right]\eqsp,
-% \]
-% where 
-% \[
-% \phi_{\theta}(x) = \frac{1}{2}\left\{\|\alpha_{\theta}(x)-\beta\|^2+\Delta H_{\theta}(x)\right\}-\ell(\theta)\eqsp.
-% \]
-% An unbiased estimator of $q^{t}_{\theta}(x,y)$ is then obtained by estimating $\mathbb{E}_{\mathbb{W}^{\beta,y}}\left[\mathrm{exp}\left\{-\int_{0}^{t}\phi_{\theta}(\mw_s)\rmd s\right\}\right]$.
-
-% Let $h$ be a function defined on $\{1,\ldots,N\}$,
-% \begin{align*}
-% \mathbb{E}\left[h(I_{\tau})\right] & = \sum_{m\ge 0}\mathbb{E}\left[h(I_m)\1_{\tau=m}\right]\eqsp,\\
-% & = \sum_{m\ge 0}h(m)\left(\prod_{\ell=0}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\right]\right)\mathbb{E}\left[h(I_m)\1_{\mathcal{A}^k_{m}}\right]\eqsp,\\
-% & = \sum_{m\ge}\left(\prod_{\ell=0}^{m-1}\mathbb{E}\left[1-\frac{q^{\Delta t_k}_{\mathrm{GPE-1},\theta}(\kappa_{\ell},\omega^{\kappa_{\ell}},(U^{\kappa_{\ell}}_j)_{1\le j\le \kappa_{\ell}},\xi_{k-1}^{I_\ell},\xi_k^{1})}{M_{\theta}}\right]\right)\\
-% &\hspace{5cm}\times\mathbb{E}\left[h(I_m)\frac{q^{\Delta t_k}_{\mathrm{GPE-1},\theta}(\kappa_{m},\omega^{\kappa_{m}},(U^{\kappa_{m}}_j)_{1\le j\le \kappa_{m}},\xi_{k-1}^{I_m},\xi_k^{1})}{M_{\theta}}\right]\eqsp,\\
-% & = \sum_{m\ge 0}\left(\mathbb{E}\left[1-\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{1})}{M_{\theta}}\right]\right)^m\mathbb{E}\left[h(I_m)\frac{q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{1})}{M_{\theta}}\right]\eqsp,\\
-% & = \mathbb{E}\left[h(I_1)q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_1},\xi_k^{1})\right]/\mathbb{E}\left[q^{\Delta t_k}_{\theta}(\xi_{k-1}^{I_m},\xi_k^{1})\right]\eqsp,\\
-% & = \sum_{\ell=1}^N \frac{h(\ell)\omega_{k-1}^{\ell}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{\ell},\xi_k^{1})}{\sum_{m=1}^N\omega_{k-1}^{m}q^{\Delta t_k}_{\theta}(\xi_{k-1}^{m},\xi_k^{1})}\eqsp,
-% \end{align*}
-% which concludes the proof.
 
@@
 & = \sum_{m\ge 1}h(m)\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\middle| \mathcal{G}_{k+1}^N\right]\right)\mathbb{E}\left[h(J_m)\1_{\mathcal{A}^k_{m}}\middle| \mathcal{G}_{k+1}^N\right]\eqsp,\\
-& = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[1-\frac{\widehat{q_k}(\xi_{k}^{J_\ell},\xi_{k+1}^{i};\zeta_k^{\ell})}{\hat{\sigma}_{+}}\middle| \mathcal{G}_{k+1}^N\right]\right)\\
-&\hspace{5cm}\times\mathbb{E}\left[h(J_m)\frac{\widehat{q_k}(\xi_{k}^{J_m},\xi_{k+1}^{i};\zeta_k^{m})}{\hat{\sigma}_{+}}\middle| \mathcal{G}_{k+1}^N\right]\eqsp,\\
-& = \sum_{m\ge 1}\left(\mathbb{E}\left[1-\frac{q_k(\xi_{k}^{J_1},\xi_{k+1}^{1})}{\hat{\sigma}_{+}}\middle| \mathcal{G}_{k+1}^N\right]\right)^{m-1}\mathbb{E}\left[h(J_1)\frac{q_k(\xi_{k}^{J_1},\xi_{k+1}^{1})}{\hat{\sigma}_{+}}\middle| \mathcal{G}_{k+1}^N\right]\eqsp,\\
-& = \mathbb{E}\left[h(J_1)q_k(\xi_{k}^{J_1},\xi_{k+1}^{i})\middle| \mathcal{F}_k^N\right]/\mathbb{E}\left[q_k(\xi_{k}^{J_1},\xi_{k+1}^{i})\middle| \mathcal{G}_{k+1}^N\right]\eqsp,\\
+& = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[1-\frac{\widehat{q_k}(\xi_{k}^{J_\ell},\xi_{k+1}^{i};\zeta_k^{\ell})}{\hat{\sigma}^k_{+}}\middle| \mathcal{G}_{k+1}^N\right]\right)\\
+&\hspace{5cm}\times\mathbb{E}\left[h(J_m)\frac{\widehat{q_k}(\xi_{k}^{J_m},\xi_{k+1}^{i};\zeta_k^{m})}{\hat{\sigma}^k_{+}}\middle| \mathcal{G}_{k+1}^N\right]\eqsp,\\
+& = \sum_{m\ge 1}\left(\mathbb{E}\left[1-\frac{q_k(\xi_{k}^{J_1},\xi_{k+1}^{1})}{\hat{\sigma}^k_{+}}\middle| \mathcal{G}_{k+1}^N\right]\right)^{m-1}\mathbb{E}\left[h(J_1)\frac{q_k(\xi_{k}^{J_1},\xi_{k+1}^{1})}{\hat{\sigma}^k_{+}}\middle| \mathcal{G}_{k+1}^N\right]\eqsp,\\
+& = \mathbb{E}\left[h(J_1)q_k(\xi_{k}^{J_1},\xi_{k+1}^{i})\middle| \mathcal{G}_{k+1}^N\right]/\mathbb{E}\left[q_k(\xi_{k}^{J_1},\xi_{k+1}^{i})\middle| \mathcal{G}_{k+1}^N\right]\eqsp,\\
 & = \sum_{\ell=1}^N \frac{h(\ell)\omega_{k-1}^{\ell}q_k(\xi_{k}^{\ell},\xi_{k+1}^{i})}{\sum_{m=1}^N\omega_{k-1}^{m}q_k(\xi_{k}^{m},\xi_{k+1}^{i})}\eqsp,\\
@@
 which concludes the proof.
-
-%The goal is to simulate realisations of  a discrete random variable $X$ having the following target distribution:
-%\begin{equation}
-%\mP(X=\ell)= p_X(\ell)=\frac{\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})}{\sum_{\ell=1}^N\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})},~~\ell\in \{1,\dots,N\}
-%\end{equation}
-%for some $\xi_{k+1}$ that has no importance in this result.
-We assume $\omega_k$ are normalized weights, i.e.\ $\sum_\ell \omega_k^l=1$.
-The candidate will therefore be sampled from $Y$, a r.v.
-with the following distribution:
-%\begin{equation}
-%\mP(Y=\ell)=p_Y(\ell)= \omega_k^\ell,~~\ell\in \{1,\dots,N\}
-%\end{equation}
-%Unfortunately, the $q_k$ function is unknown.
-Howevere, we let's suppose we can define an unbiased estimator $\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)$, where $\zeta$ is a random variable having th p.d.f.
-$p_\zeta$.
-We have therefore:
-%$$\E_{p_\zeta}\left[\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)\right]=q(\xi_k^\ell,\xi_{k+1})$$
-%Moreover, we impose the following property over this estimator, we want that:
-%$$\exists \sigma_+\in \mathbb{R}\text{ such that }\forall x,y,~~\hat{q}_k(x,y,\zeta)\leq \sigma_+$$
-%To simulate a random variable $X\sim p_X$, we propose the following algorithm:
-%\begin{enumerate}
-%\item Sample independently $\ell\sim p_Y$, $z\sim p_\zeta$ and $u \sim \mathcal{U}[0,1]$
-%\item \textbf{If} 
-%$$u \leq \frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},z)}{\sigma_+},$$
-%set $x=\ell$. \textbf{Else}, return to 1).
-%\end{enumerate}
-
-%The lemma \ref{lem:AR:unbiased} states that $x$ is a realisation of a random variable $X\sim p_X$. Indeed
-%\begin{align*}
-%\mP\left(Y=\ell\vert U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\vert Y=l \right) \mathbb{P}(Y=\ell)}{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)}\\
-%&=\frac{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)}{\sigma_+} \right) \omega_k^\ell}{\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)}
-%\intertext{We now note that}
-%\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\E_{p_\zeta}\left[\E\left[\E_{p_Y}\left[\E\left[\mP\left(U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right) \vert Y\right] \right] \vert \zeta\right]\right]\\
-%&=\E_{p_\zeta}\left[\sum_{\ell=1}^N w_k^l\frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)}{\sigma_+} \right]~~(\text{As }\frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\leq 1)\\
-%&=\frac{1}{\sigma_+}\sum_{\ell=1}^N w_k^\ell q_k(\xi_k^\ell,\xi_{k+1})
-%\intertext{In the same way, we have:}
-%\mP\left(U\leq \frac{\hat{q}_k(\xi_k^\ell,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{1}{\sigma_+} q_k(\xi_k^\ell,\xi_{k+1})
-%\intertext{Which gives overall the wanted result}
-%\mP\left(Y=\ell\vert U\leq \frac{\hat{q}_k(\xi_k^Y,\xi_{k+1},\zeta)}{\sigma_+}\right)&=\frac{\omega_k^\ell q_k(\xi_k^\ell,\xi_{k+1})}{\sum_{l=1}^N w_k^\ell q_k(\xi_k^\ell,\xi_{k+1})}=p_X(\ell),~~\ell\in \{1,\dots,N\}
-%\end{align*}
-%\end{proof}
 \end{proof}
@@
 Note that by Lemma~\ref{lem:AR:unbiased},
-%\sigma\left(\mathcal{F}_k^{N},\{\xi_{k+1}^i,I_{k+1}^i\}\right)
 \begin{align*}
@@
 \end{align*}
-Since $\tau^{i}_{k+1}$ and $\zeta_{k+1}$ are independent conditionally to $\mathcal{G}_{k+1}^{N}$:
+Since $\tau^{i}_{k+1}$ and $\zeta_{k}$ are independent conditionally to $\mathcal{G}_{k+1}^{N}$:
 \begin{multline*}
-\mathbb{E}\left[\tau^{i}_{k+1} \widehat{\qk} (\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta_{k+1})\middle|\mathcal{G}_{k+1}^{N}\right]\\
+\mathbb{E}\left[\tau^{i}_{k+1} \widehat{\qk} (\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1};\zeta_{k})\middle|\mathcal{G}_{k+1}^{N}\right]\\
  = q_k(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})\sum_{\ell=1}^N\frac{\omega_k^{\ell} \qk (\xi_{k}^{\ell},\xi^{i}_{k+1})\left(\tau^{\ell}_k + h_{k}(\xi_{k}^{\ell},\xi^{i}_{k+1})\right)}{\sum_{\ell'=1}^N\omega_k^{\ell'} \qk (\xi_{k}^{\ell'},\xi^{i}_{k+1})}\eqsp.
@@
 
-%\section{Multidimensional Ornstein Uhlenbeck bridge}
-%\label{sec:ornstein:bridge}
-%The GPE introduced in this paper requires to sample finite dimensional distributions of Ornstein Uhlenbeck bridges.
-These results are given here for completeness.
-Consider the following $d$-dimensional SDE:
-%\begin{equation}
-%\label{eq:sde:ornstein}
-%Z_0 = z_0\quad\mbox{and}\quad \rmd Z_t = (\Upsilon Z_t + \vartheta)\rmd t + \rmd W_t\eqsp,
-%\end{equation}
-%where $W$ is a standard Brownian motion on $\mathbb{R}^d$, $A$ is a $d\times d$ matrix and $b$ a $d$ dimensional vector.
-In the case of Section~\ref{sec:PaRIS:SDE}, for all $1\le k \le n$, the proposal stochastic differential equation governing $Y_t^k$, $t\in(t_{k-1},t_k)$, is of the form \eqref{eq:sde:ornstein} with
-%\[
-%\Upsilon = J_{\alpha_{\theta}}(X_{k-1})\;\;\mbox{and}\;\; \vartheta =\alpha_{\theta}(X_{k-1}) - J_{\alpha_{\theta}}(X_{k-1})X_{k-1}\eqsp,
-%\]
-%$J_{\alpha_{\theta}}$ being the Jacobian of $\alpha_{\theta}$.
-In this case, as $\alpha_{\theta}(x) = \nabla_x A_{\theta}(x)$, $A$ is symmetric and nonsingular.
-By \cite[Section~5.6]{karatzas:shreve:1991}, the unique strong solution of \eqref{eq:sde:ornstein} is:
-%\[
-%Z_t = \mathrm{e}^{t\Upsilon}z_0 + \int_0^t\mathrm{e}^{(t-s)\Upsilon}\vartheta \rmd s + \int_0^t\mathrm{e}^{(t-s)\Upsilon} \rmd W_s\eqsp.
-%\]
-%Therefore, the probability density of the conditional distribution of $Z_t$ given $Z_0 = z_0$ is Gaussian with mean $\mu_t(z_0)$ and variance $\Sigma_{t}$ given by:
-%\begin{align*}
-%\Sigma_{t} &= \int_0^{t} \mathrm{e}^{(t -s)\Upsilon}\mathrm{e}^{(t -s)\Upsilon'}\rmd s\eqsp,\\
-%\mu_{t}(z) &= \mathrm{e}^{t \Upsilon}z_0 + \left(\int_0^t\mathrm{e}^{(t-s)\Upsilon} \rmd s\right)\vartheta\eqsp.
-%\end{align*}
-%As $\Upsilon$ is symetric and nonsingular,
-%\begin{align*}
-%\Sigma_{t} &= \int_0^{t} \mathrm{e}^{2(t -s)\Upsilon}\rmd s = \Upsilon^{-1}(\mathrm{e}^{2t\Upsilon}-I_d)/2\eqsp,\\
-%\mu_{t}(z) &= \mathrm{e}^{t \Upsilon}z + \left(\int_0^t\mathrm{e}^{(t-s)\Upsilon} \rmd s\right)\vartheta = \mathrm{e}^{t \Upsilon}z + \Upsilon^{-1}(\mathrm{e}^{t\Upsilon}-I_d)\vartheta \eqsp.
-%\end{align*}
-%For each interval $(t_{k-1},t_k)$, $1\le k \le n$, the GPE requires to sample a Ornstein-Uhlenbeck bridge at some ramdom time steps between $X_{k-1}$ at time $t_{k-1}$ and $X_k$ at time $t_k$.
-Conditional on $Z_s = x$ and $Z_T = y$, the probability density of $Z_t$ is Gaussian with mean $\mu$ and variance $\Sigma_t^{s,T}$ given by:
-%\begin{align*}
-%\mu &=\Gamma(t,T)\Gamma(s,T)^{-1}\mu_- + \Gamma(s,t)'(\Gamma(s,T)')^{-1}\mu_+\eqsp,\\
-%\Sigma_t^{s,T} &=\Gamma(t,T)\Gamma(s,T)^{-1}\Gamma(s,t)\eqsp,\\
-%\Gamma(s,t)&=\int_s^{t} \mathrm{e}^{(s -u)\Upsilon}\mathrm{e}^{(t -u)\Upsilon'}\rmd u\eqsp = \Upsilon^{-1}\left(\mathrm{e}^{(t -s)\Upsilon}-\mathrm{e}^{(s -t)\Upsilon}\right)/2\eqsp,\\
-%\mu_- &= x - \Upsilon^{-1}(\mathrm{e}^{(s -t)\Upsilon}-I_d)\vartheta\eqsp,\\
-%\mu_+ &= y-\Upsilon^{-1}(\mathrm{e}^{(T -t)\Upsilon}-I_d)\vartheta\eqsp.
-%\end{align*}
-
 \begin{proof}[Proof of Proposition~\ref{prop:exp:deviation}]
@@
 At time $k=0$, the result holds using that for all $1\le i \le N$, $\rho_0^i = 0$ and the convention $T_0[h_0] =0$.
-In addition, $\phi_0^N$ is a standard importance sampler estimator of $\phi_0$ with $\omega_0^i\le |\widehat{\omega}_0|_{\infty}$ so that for any bounded function $h$ on $\mathsf{X}$,
+In addition, $\phi_0^N$ is a standard importance sampler estimator of $\phi_0$ with $\widehat \omega_0^i\le |\widehat{\omega}_0|_{\infty}$ so that for any bounded function $h$ on $\mathsf{X}$,
 \[

### 2017-08-08 TexSubmission/answer_referees.tex
Answer referee 1

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{amssymb}
\usepackage{makeidx}
\usepackage[english]{babel}
\usepackage{graphicx}
\usepackage{amsfonts,amsmath,amssymb,amsthm}
\usepackage{oldgerm}
\usepackage{mathrsfs}
\usepackage[active]{srcltx}
\usepackage{verbatim}
\usepackage[toc,page]{appendix}
\usepackage{aliascnt}
\usepackage{array}
\usepackage{hyperref}
\usepackage[textwidth=4cm,textsize=footnotesize]{todonotes}
\usepackage{xargs}
\usepackage{cellspace}
\usepackage[Symbolsmallscale]{upgreek}
\usepackage{geometry}
\usepackage{hyperref}
\usepackage{array}
\geometry{top=3.5cm, bottom=3.5cm, left=3.5cm , right=3.5cm}
\usepackage{fancyhdr}
\pagestyle{fancy}

\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\graphicspath{{figures/}}
\usepackage{enumerate}
\usepackage{xcolor}
\usepackage{algorithm,algorithmic}

\newcommand{\x}[2]{x_{#1}^{(#2)}}
\newcommand{\w}[2]{w_{#1}^{(#2)}}
\newcommand{\tw}[2]{\tilde{w}_{#1}^{(#2)}}
\newcommand{\p}[2]{\xi_{#1}^{(#2)}}
\newcommand{\tp}[2]{\tilde{\xi}_{#1}^{(#2)}}
\newcommand{\ta}[2]{\tau_{#1}^{(#2)}}
\newcommand{\rmd}{\mathrm{d}}
\newcommand{\eqsp}{\;}
\newcommand{\1}{\mathrm{1}}
\newcommand{\com}[1]{{\color{gray} // #1}}
\newcommand{\mP}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\qk}{q_{k}}
\newcommand{\acom}[1]{\textit{\color{gray} //#1}}
\newcommand{\Oz}{Z}%Letter for the Ozaki approximation
\newcommand{\Jk}{J_{\alpha}^k}%Command for Jacobian of alpha
\newcommand{\mw}{\mathsf{w}}%For bridge realisations
\newcommand{\U}{\mathsf{U}}
\newcommand{\Lo}{\mathsf{L}}
%\newcommand{\qk}{q^{\Delta t_k}_{\theta}}
\newtheorem{lemma}{Lemma}
\newtheorem{proposition}{Proposition}
\newcommand{\hQ}{\widehat{Q}}
\newcounter{hypA}
\newenvironment{hypA}{\refstepcounter{hypA}\begin{itemize}
\item[{\bf H\arabic{hypA}}]}{\end{itemize}}
%%ENvironmment assumption
\newcounter{defcounter}
\setcounter{defcounter}{0}
\newenvironment{assumpt}{%
\addtocounter{equation}{-1}
\refstepcounter{defcounter}
\renewcommand\theequation{\textbf{A}\thedefcounter}
\begin{equation}}
{\end{equation}}
\begin{document}

\author{Pierre Gloaguen\footnotemark[1] \and Marie-Pierre Etienne\footnotemark[1] \and Sylvain Le {C}orff\footnotemark[2]}

\footnotetext[1]{AgroParistech, UMR MIA 518, F-75231 Paris, France.}
\footnotetext[2]{Laboratoire de Math\'ematiques d'Orsay, Univ. Paris-Sud, CNRS, Universit\'e Paris-Saclay.}

\title{Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations}

\lhead{Gloaguen et al.}
\rhead{Particle smoother for SDE}

\maketitle

The authors are grateful to the associate editor and the anonymous referees for their comments to improve the manuscript. The paper has been modified to take these remarks into account and also carefully proof read to remove remaining typos and minor mistakes. We provide below detailed answers for all comments.

\subsection*{Referee 1}
{\em I have two major comments that I think should be addressed before publishing. First of all the whole presentation is written as given a  fixed observation record $(Y_k)_{1\le k \le n}$, one of the main points of the PaRIS algorithm, which you rely on, is the ability to work with data streams of observations with a  fixed memory usage. Maybe it would be a good idea to include this in the paper.}

\vspace{.3cm}

As pointed by the Referee, one of the best feature of PaRIS algorithm is that it may be implemented "online", processing the observations $(Y_k)_{k\ge 1}$ as they are received, without any storage. This may be seen for instance in steps (i)-(iii) in page 6 which describe the way the intermediate quantities $(\tau_k^i)_{1\le i \le N}$ can be computed at each time step. Each time a new observation $Y_{n+1}$ is received, these quantities can be updated only using $Y_{n+1}$, $(\tau_n^i)_{1\le i \le N}$ and the particle filter at time $n$. This means that storage requirements do not increase. The update is stopped when the last observations is received to compute the approximation of the smoothed additive functional.

This remark was added after steps (i)-(iii) to highlight the online property of the extension of PaRIS algorithm (with no additional storage requirements). The introduction also insists on this property of the algorithm.

In common applications (for instance in the cas of the EM algorithm),  the smoothed additive functional to be computed is known in advance, this is the reason why we chose to stop the updates for a given arbitrary time $n$, but it is now clear that the algorithm may be stopped at any time.

\vspace{1cm}

{\em Secondly in the algorithm presented on page 9, it says to draw $M$ samples from $\zeta_k$ while not mentioning
what to do with these. This should be specified in the manuscript.}

\vspace{.3cm}

The way the unbiased estimation of $q_k(\xi^{I_k^\ell}_{k-1},\xi^\ell_k)$ was computed was indeed not clear. In (7) and at the bottom of page 6 we only stated the existence of a random variable $\zeta_k$ such that $\widehat{q}_k(\xi^{I_k^\ell}_{k-1},\xi^\ell_k;\zeta_k)$ is an unbiased estimation of $q_k(\xi^{I_k^\ell}_{k-1},\xi^\ell_k)$.

At each time step, we then sample independently $(\zeta^m_k)_{1\le m \le M}$ and define the following estimator of $q_k(\xi^{I_k^\ell}_{k-1},\xi^\ell_k)$:
\[
\frac{1}{M}\sum_{m=1}^M \widehat{q}_k(\xi^{I_k^\ell}_{k-1},\xi^\ell_k;\zeta^m_k)\eqsp.
\]
This is now detailed in clearly in \eqref{}.

\vspace{1cm}

{\em On page 4 line 28 $q_k$ is defined as the transition from $X_k$ to $X_{k+1}$, while on page 5 line 30 (equation
4) it is used as the transition from $X_{k-1}$ to $X_k$, this also happens on page 10 line 35.}

\vspace{.3cm}

This was corrected, $q_k$ is the transition density of the law of $X_k$ given $X_{k+1}$.

\vspace{1cm}

{\em On page 15, line 13, I believe that $h(m)$ has sneaked its way into the equation and should be removed.}

\vspace{.3cm}
 $h(m)$ was removed in this equation.

\subsection*{Referee 2}

\end{document}

### 2017-08-08 TexSubmission/ParisEM.tex
Answer referee 1

--- 2017-02-27 ParisEM.tex
@@
 \maketitle
+
+\begin{abstract}
+This paper introduces a new algorithm to approximate smoothed additive functionals for partially observed stochastic differential equations.
+This method relies on the recent procedure introduced in \cite{olsson:westerborn:2016} which allows to compute such approximations online, i.e.
+as the observations are received, and with a computational complexity growing linearly with the number of Monte Carlo samples.
+The algorithm of \cite{olsson:westerborn:2016} cannot be used in the case of partially observed stochastic differential equations since the transition density of the latent data is usually unknown.
+We prove that a similar algorithm may still be defined for partially observed continuous processes by replacing this unknown quantity by an unbiased estimator obtained for instance using general Poisson estimators.
+We prove that this estimator is consistent and its performance are illustrated using data from two models.
+\end{abstract}
+
+{\bf Keywords}: Stochastic differential equations, Smoothing, Sequential Monte Carlo Methods.
 
@@
 We compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method of \cite{olsson:strojby:2011}, for different values of the lag (namely, 1,2,5,10,50).
-The particle approximation of $\mathcal{Q}(\theta,\theta)$ for each model  is computed using each algorithm, see Figure~\ref{fig:res:SINE} for the SINE model (and respectively Figure~\ref{fig:res:LG} for the log-growth model).
+The particle approximation of $\mathcal{Q}(\theta,\theta)$ for each model  is computed using each algorithm, see Figure~\ref{fig:res:SINE:EM} for the SINE model (and respectively Figure~\ref{fig:res:LG:EM} for the log-growth model).
 This estimation is performed 200 times to obtain the estimates $\hQ_1,\dots,\hQ_{200}$, using $\tilde{N}=2$ particles for PaRIS algorithm, and $M=30$ replications for the Monte Carlo approximation $\widehat q_k$ of each $q_k$.
@@
 The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, and denoted by $\hQ_\star$.
-Figure \ref{fig:res:SINE} (resp.
-\ref{fig:res:LG} ) shows this estimate for an example on one simulated data set.
-The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=1600$ so that both estimations requre similar computational times.
+Figure \ref{fig:res:SINE:EM} (resp.
+\ref{fig:res:LG:EM}) shows this estimate for an example on one simulated data set.
+The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=1600$ so that both estimations require similar computational times.
 
@@
 where the $(\varepsilon)_{0\leq k \leq 100}$ are i.i.d. $\mathcal{N}(0,1)$.
-In the example displayed on Figure \ref{fig:res:SINE}, we set $\theta=0$.
+In the example displayed on Figure \ref{fig:res:SINE:obs}, we set $\theta=0$.
 In that case, the function $\rho_{\Delta_k}$ defined in \eqref{eq:rho:func} can be upper bounded either on $(x,y)$ or only on $y$, the GRand PaRIS algorithm has therefore a linear complexity.
@@
 where $m(\hQ^s)$ and $\sigma(\hQ^s)$ are the empirical mean and standard deviation of the sample $Q_1^s,\dots,Q_{200}^s$.
-For each estimation method, the resulting distributions of $\mathsf{arb}_1,\dots,\mathsf{arb}_{100}$ and $\mathsf{acv}_1,\dots,\mathsf{acv}_{100}$  are shown on Figure \ref{fig:mult:SINE}.\\
+For each estimation method, the resulting distributions of $\mathsf{arb}_1,\dots,\mathsf{arb}_{100}$ and $\mathsf{acv}_1,\dots,\mathsf{acv}_{100}$  are shown on Figures \ref{fig:mult:SINE:b} and  \ref{fig:mult:SINE:cv}.
 
@@
 \centering
-\begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{obs_SINE}
-\end{subfigure}
-\begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{res_Estep_SINE}
-\end{subfigure}
-\caption{{\em SINE model}.
-Process $X$ solution to the SDE (balls) and observations $Y$ (circles) at times $t_0=0,\dots,t_{100}=50$ [left].
-Estimation of the EM intermediate quantity $\mathcal{Q}(\theta,\theta)$  using the fixed lag (FL) technique for 5 different lags, and the GRand PaRIS algorithm using 200 replicates [right].
+%\includegraphics[scale = .5]{obs_SINE}
+\caption{{\em SINE model - observations}. Process $X$ solution to the SDE (balls) and observations $Y$ (circles) at times $t_0=0,\dots,t_{100}=50$.}
+\label{fig:res:SINE:obs}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+%\includegraphics[scale=0.5]{res_Estep_SINE}
+\caption{{\em SINE model - EM intermediate quantity}.
+Estimation of the EM intermediate quantity $\mathcal{Q}(\theta,\theta)$  using the fixed lag (FL) technique for 5 different lags, and the GRand PaRIS algorithm using 200 replicates.
 The whiskers represent the extent of the 95\% central values.
@@
 The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=5000$ particles.}
-\label{fig:res:SINE}
+\label{fig:res:SINE:EM}
 \end{figure}
+
 \begin{figure}[p]
 \centering
-\begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{res_mult_SINE_bias}
-\end{subfigure}
-\begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{res_mult_SINE_cv}
-\end{subfigure}
-\caption{{\em SINE model}. Distribution of the empirical absolute relative bias [left] and of the empirical absolute coefficient of variation [right] for each method, see \eqref{eq:emp:bias}.}
-\label{fig:mult:SINE}
+%\includegraphics[scale=0.5]{res_mult_SINE_bias}
+\caption{{\em SINE model - bias}. Distribution of the empirical absolute relative bias.}
+\label{fig:mult:SINE:b}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+%\includegraphics[scale=0.5]{res_mult_SINE_cv}
+\caption{{\em SINE model - variance}. Distribution of the empirical absolute coefficient of variation.}
+\label{fig:mult:SINE:cv}
 \end{figure}
@@
 In that case, the $\rho_{\Delta_k}$ function defined in \eqref{eq:rho:func} can be upper bounded as a function of $y$ when $x\in \{\xi_k^1,\dots, \xi_k^N\}$, the GRand PaRIS algorithm has therefore a linear complexity.
-The intermediate quantity of the EM algorithm is evaluated  as for the SINE model, see Figures~\ref{fig:res:LG} and~\ref{fig:mult:LG}.
+The intermediate quantity of the EM algorithm is evaluated  as for the SINE model, see Figures~\ref{fig:res:LG:EM},~\ref{fig:mult:LG:b} and~\ref{fig:mult:LG:cv}.
 
@@
 \centering
-\begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{obs_LG}
-\end{subfigure}
-\begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{res_Estep_LG}
-\end{subfigure}
-\caption{{\em Log-growth model}.
-Process $X$ solution to the SDE (balls) and observations $Y$ (circles) at times $t_0=0,\dots,t_{100}=50$ [left].
-Estimation of the EM intermediate quantity $\mathcal{Q}(\theta,\theta)$  using the fixed lag (FL) technique for 5 different lags, and the GRand PaRIS algorithm using 200 replicates [right].
+%\includegraphics[scale=0.5]{obs_LG}
+\caption{{\em Log-growth model - observations}. Process $X$ solution to the SDE (balls) and observations $Y$ (circles) at times $t_0=0,\dots,t_{100}=50$.}
+\label{fig:res:LG:obs}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+%\includegraphics[scale=0.5]{res_Estep_LG}
+\caption{{\em Log-growth model - EM intermediate quantity}.
+Estimation of the EM intermediate quantity $\mathcal{Q}(\theta,\theta)$  using the fixed lag (FL) technique for 5 different lags, and the GRand PaRIS algorithm using 200 replicates.
 The whiskers represent the extent of the 95\% central values.
@@
 The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=5000$ particles.}
-\label{fig:res:LG}
+\label{fig:res:LG:EM}
 \end{figure}
+
 \begin{figure}[p]
 \centering
-\begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{res_mult_LG_bias}
-\end{subfigure}
-\begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{res_mult_LG_cv}
-\end{subfigure}
-\caption{{\em Log-growth model}. Distribution of the empirical absolute relative bias [left] and of the empirical absolute coefficient of variation [right] for each method, see \eqref{eq:emp:bias}.}
-\label{fig:mult:LG}
+%\includegraphics[scale=0.5]{res_mult_LG_bias}
+\caption{{\em Log-growth model - biais}. Distribution of the empirical absolute relative bias.}
+\label{fig:mult:LG:b}
 \end{figure}
+
+\begin{figure}[p]
+\centering
+%\includegraphics[scale=0.5]{res_mult_LG_cv}
+\caption{{\em Log-growth model - variance}. Distribution of the empirical absolute coefficient of variation.}
+\label{fig:mult:LG:cv}
+\end{figure}
+
 \section{Conclusions}
@@
 
+\subsection*{Competing interests}
+The authors have no competing interests.
+
+\subsection*{Funding}
+This work has been developed during a one year postdoc funded by Paris-Saclay Center for Data Science.
+
+\subsection*{Authors' contributions}
+All the authors have contributed to the conception of the algorithms, the analysis of the proposed estimator and to the redaction of the manuscript.
+Pierre Gloaguen provided the simulations displayed in the final version.
+
 \appendix
@@
 \[
-\mathcal{A}^k_{\ell} = \left\{U_\ell<\widehat{q}_{k}(\xi_{k}^{J_\ell},\xi_{k+1}^{i},\zeta^{\ell}_{k})/\sigma_+\right\}\eqsp.
+\mathcal{A}^k_{\ell} = \left\{U_\ell<\widehat{q}_{k}(\xi_{k}^{J_\ell},\xi_{k+1}^{i},\zeta^{\ell}_{k})/\hat{\sigma}^k_+\right\}\eqsp.
 \]
@@
 \mathbb{E}\left[h(J^{i,j}_k)\middle| \mathcal{G}_{k+1}^N\right] & = \sum_{m\ge 1}\mathbb{E}\left[h(J_m)\1_{\tau=m}\middle| \mathcal{G}_{k+1}^N\right]\eqsp,\\
-& = \sum_{m\ge 1}h(m)\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\middle| \mathcal{G}_{k+1}^N\right]\right)\mathbb{E}\left[h(J_m)\1_{\mathcal{A}^k_{m}}\middle| \mathcal{G}_{k+1}^N\right]\eqsp,\\
+& = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[\1_{(\mathcal{A}^k_{\ell})^c}\middle| \mathcal{G}_{k+1}^N\right]\right)\mathbb{E}\left[h(J_m)\1_{\mathcal{A}^k_{m}}\middle| \mathcal{G}_{k+1}^N\right]\eqsp,\\
 & = \sum_{m\ge 1}\left(\prod_{\ell=1}^{m-1}\mathbb{E}\left[1-\frac{\widehat{q_k}(\xi_{k}^{J_\ell},\xi_{k+1}^{i};\zeta_k^{\ell})}{\hat{\sigma}^k_{+}}\middle| \mathcal{G}_{k+1}^N\right]\right)\\
@@
 \bibliographystyle{plain}
-\bibliography{./ParisEM_bib}
+\begin{thebibliography}{10}
+
+\bibitem{ait-sahalia:1999}
+Y.~Ait-Sahalia.
+\newblock Transition densities for interest rate and other nonlinear
+  diffusions.
+\newblock {\em Journal of {Finance}}, 54:1361--1395, 1999.
+
+\bibitem{ait-sahalia:2002}
+Y.~Ait-Sahalia.
+\newblock Maximum likelihood estimation of discretely sampled diffusions: a
+  closed-form approximation approach.
+\newblock {\em Econometrica}, 70:223--262, 2002.
+
+\bibitem{ait-sahalia:2008}
+Y.~Ait-Sahalia.
+\newblock Closed-form likelihood expansions for multivariate diffu- sions.
+\newblock {\em The {A}nnals of {S}tatistics}, 36:906--937, 2008.
+
+\bibitem{beskos:papaspiliopoulos:roberts:fearnhead:2006}
+A.~Beskos, O.~Papaspiliopoulos, G.~Roberts, and P.~Fearnhead.
+\newblock Exact and computationally efficient likelihood-based estimation for
+  discretely observed diffusion processes (with discusion).
+\newblock {\em J. Roy. Statist. Soc. Ser. B}, 68(3):333--382, 2006.
+
+\bibitem{cappe:moulines:ryden:2005}
+O.~Capp\'{e}, E.~Moulines, and T.~Ryd\'{e}n.
+\newblock {\em Inference in Hidden {M}arkov Models}.
+\newblock Springer, 2005.
+
+\bibitem{delmoral:jacod:protter:2001}
+P.~Del~Moral, J.~Jacod, and P.~Protter.
+\newblock The {M}onte {C}arlo method for filtering with discrete-time
+  observations.
+\newblock {\em Probability {T}heory and {R}elated {F}ields}, 120:346 -- 368,
+  2001.
+
+\bibitem{dempster:laird:rubin:1977}
+A.~P. Dempster, N.~M. Laird, and D.~B. Rubin.
+\newblock Maximum likelihood from incomplete data via the {EM} algorithm.
+\newblock {\em J. Roy. Statist. Soc. B}, 39(1):1--38 (with discussion), 1977.
+
+\bibitem{douc:garivier:moulines:olsson:2011}
+R.~Douc, A.~Garivier, E.~Moulines, and J.~Olsson.
+\newblock Sequential {M}onte {C}arlo smoothing for general state space hidden
+  {M}arkov models.
+\newblock {\em Ann. Appl. Probab.}, 21(6):2109--2145, 2011.
+
+\bibitem{doucet:godsill:andrieu:2000}
+A.~Doucet, S.~Godsill, and C.~Andrieu.
+\newblock On sequential {M}onte-{C}arlo sampling methods for {B}ayesian
+  filtering.
+\newblock {\em Stat. Comput.}, 10:197--208, 2000.
+
+\bibitem{doucetgodsillandrieu:2000}
+A.~Doucet, S.~Godsill, and C.~Andrieu.
+\newblock On sequential monte-carlo sampling methods for bayesian filtering.
+\newblock {\em Statistics and Computing}, 10:197 -- 208, 2000.
+
+\bibitem{fearnhead:papaspiliopoulos:roberts:2008}
+P.~Fearnhead, O.~Papaspiliopoulos, and G~Roberts.
+\newblock Particle filters for partially observed diffusions.
+\newblock {\em J. Roy. Statist. Soc. Ser. B}, 70(4):755--777, 2008.
+
+\bibitem{godsill:doucet:west:2004}
+S.~J. Godsill, A.~Doucet, and M.~West.
+\newblock {M}onte {C}arlo smoothing for non-linear time series.
+\newblock {\em J. Am. Statist. Assoc.}, 50:438--449, 2004.
+
+\bibitem{gordon:salmond:smith:1993}
+N.~Gordon, D.~Salmond, and A.F. Smith.
+\newblock Novel approach to nonlinear/non-{G}aussian bayesian state estimation.
+\newblock {\em {IEE} Proc. {F}, Radar Signal Process}, 140:107--113, 1993.
+
+\bibitem{huerzeler:kunsch:1998}
+M.~H{\"u}rzeler and H.~R. K{\"u}nsch.
+\newblock {M}onte {C}arlo approximations for general state-space models.
+\newblock {\em J. Comput. Graph. Statist.}, 7:175--193, 1998.
+
+\bibitem{kantas:doucet:signh:2015}
+N.~Kantas, A.~Doucet, S.S. Singh, J.~Maciejowski, and N.~Chopin.
+\newblock On particle methods for parameter estimation in state-space models.
+\newblock {\em Statist. Sci.}, 30(3):328--351, 2015.
+
+\bibitem{kessler:1997}
+M.~Kessler.
+\newblock Estimation of an ergodic diffusion from discrete observations.
+\newblock {\em Scandinavian Journal of Statistics}, 24(2):211--229, 1997.
+
+\bibitem{kessler:lindner:sorensen:2012}
+M.~Kessler, A.~Lindner, and M.~Sorensen.
+\newblock {\em Statistical methods for stochastic differential equations}.
+\newblock CRC Press, 2012.
+
+\bibitem{kitagawa:1996}
+G.~Kitagawa.
+\newblock {M}onte-{C}arlo filter and smoother for non-{G}aussian nonlinear
+  state space models.
+\newblock {\em J. Comput. Graph. Statist.}, 1:1--25, 1996.
+
+\bibitem{lecorff:fort:2013a}
+S.~Le~Corff and G.~Fort.
+\newblock Convergence of a particle-based approximation of the block online
+  {E}xpectation {M}aximization algorithm.
+\newblock {\em ACM Transactions on Modeling and Computer Simulation}, 23(1):2,
+  2013.
+
+\bibitem{lecorff:fort:2013b}
+S.~Le~Corff and G.~Fort.
+\newblock Online {E}xpectation {M}aximization based algorithms for inference in
+  hidden {M}arkov models.
+\newblock {\em Electronic Journal of Statistics}, 7:763--792, 2013.
+
+\bibitem{li:2013}
+C.~Li.
+\newblock Maximum-likelihood estimation for diffusion processes via closed-form
+  density expansions.
+\newblock {\em The {A}nnals of {S}tatistics}, 41(3):1350--1380, 2013.
+
+\bibitem{olsson:cappe:douc:moulines:2008}
+J.~Olsson, O.~Cappe, R.~Douc, and E.~Moulines.
+\newblock Sequential monte carlo smoothing with application to parameter
+  estimation in nonlinear state space models.
+\newblock {\em Bernoulli}, 14(1):155--179, 2008.
+
+\bibitem{olsson:strojby:2011}
+J.~Olsson and J.~Strojby.
+\newblock Particle-based likelihood inference in partially observed diffusion
+  processes using generalised {P}oisson estimators.
+\newblock {\em Electron. J. Statist.}, 5:1090--1122, 2011.
+
+\bibitem{olsson:westerborn:2016}
+J.~Olsson and J.~Westerborn.
+\newblock Efficient particle-based online smoothing in general hidden {M}arkov
+  models: the {PaRIS} algorithm.
+\newblock {\em ArXiv:1412.7550}, 2016.
+
+\bibitem{ozaki:1992}
+T.~Ozaki.
+\newblock A bridge between nonlinear time series models and nonlinear
+  stochastic dynamical systems: a local linearization approach.
+\newblock {\em Statistica Sinica}, 2:1130--135, 1992.
+
+\bibitem{pitt:shephard:1999}
+M.~K. Pitt and N.~Shephard.
+\newblock Filtering via simulation: Auxiliary particle filters.
+\newblock {\em J. Am. Statist. Assoc.}, 94(446):590--599, 1999.
+
+\bibitem{doucet:poyiadjis:singh:2011}
+G.~Poyiadjis, A.~Doucet, and S.S Singh.
+\newblock {Particle approximations of the score and observed information matrix
+  in state space models with application to parameter estimation}.
+\newblock {\em Biometrika}, 98:65--80, 2011.
+
+\bibitem{shoji:1998}
+I.~Shoji and T.~Ozaki.
+\newblock Estimation for nonlinear stochastic differential equations by a local
+  linearization method 1.
+\newblock {\em Stochastic {A}nalysis and {A}pplications}, 16(4):733--752, 1998.
+
+\bibitem{uchida:yoshida:2012}
+Masayuki Uchida and Nakahiro Yoshida.
+\newblock Adaptive estimation of an ergodic diffusion process based on sampled
+  data.
+\newblock {\em Stochastic Processes and their Applications}, 122(8):2885 --
+  2924, 2012.
+
+\end{thebibliography}
+
 \end{document}

### 2017-08-08 TexSubmission/answer_referees.tex
First answers to referee 2

@@
 \]
-This is now detailed in clearly in \eqref{}.
+This is now detailed clearly just after (7).
+
+\vspace{1cm}
+
+{\em Maybe if the time continuous process was denoted by $\widetilde{X}$ or something similar the difference would be clear.}
+
+\vspace{.3cm}
+
+\textcolor{red}{Il est gentil Michel mais introduire une notation pour ca moi ca ne me plait pas. On repond au reste, je suis d'avis de lui dire merci mais non merci.}
+
+\vspace{1cm}
+
+{\em
+On page 3, you talk about not being able to extend the FFBSi algorithm to this class of models, but
+as far as I understand it the same sampling technique should be possible to draw from the approximative
+smoothing distribution.
+}
+
+\vspace{.3cm}
+
+The usual FFBSi cannot be extended easily for the same reasons as the FFBS (it relies on the approximation of the backward kernel which is a ratio as defined in (6) in our paper).
+But it is true that the linear version of the FFBSi may be extended in a similar way as PaRIS algorithm (while it does stil require a backward pass).
+The introduction was modified to make this statement clear.
 
@@
 
+\subsubsection*{Major comments}
+\begin{enumerate}
+\item
+\item {\em When presenting equation (2), the authors should explain the importance of approximating those type of functionals.}
+
+\vspace{.3cm}
+
+The importance of these quantities is now clearly stated after equation (2).
+We added the usual applications to the computation of the Fisher score and the intermediate quantity of the EM algorithm.
+We also highlighted that it was crucial for online inference of latent data models which supports the use of the proposed online smoother.
+
+\item {\em General practical strategies to specify $\vartheta_k$, $p_k$, $\hat{\sigma}_+^k$ in the SDE context should be
+discussed.}
+
+\vspace{.3cm}
+
+Choices that can be made for  $\vartheta_k$ and $p_{k-1}$ are discussed after (4) and the definition of $\phi_k^N$.
+The specific approximation of the fully adapted particle filter is also illustrated in the numerical section.
+\textcolor{red}{Pour le sigma Pierre tu as peut-etre plus d'elements.
+On peut bien sur parler des conditions de l'EA mais je pense qu'une intuition pratique est plus percutante ici.}
+
+\item {\em Conditions (A1)-(A3) should be carefully discussed. How restrictive are they? For
+example, what if $\phi(\omega_s)$ (or $L_\omega$) is not bounded below? Is there any general solution
+in this case?}
+
+\vspace{.3cm}
+
+\item
+\item
+\item
+\item
+\item {\em Can anything be said about the bias of the proposed estimator for finite $N$? Is it
+only due to the form of  $\Lambda_k^N$ ?}
+
+\vspace{.3cm}
+
+\item {\em What can be said about a CLT for the proposed estimator?}
+\textcolor{red}{On peut le faire, pas tres dur, mais ca va allonger les choses et sans message tres clair sur la variance asymptotique a mon avis. Voila une suggestion de reponse...}
+
+\vspace{.3cm}
+
+A CLT was derived for the original version of PaRIS algorithm as the number $N$ of particles grows to infinity.
+A CLT may also derived for our extension as it is a noisy version of the original method.
+However, even if its proof is not too involved it still require additional assumptions and technicalities.
+And the asymptotic variance remains hard to analyze (for instance to choose the parameters in comparison to other smoothing methods).
+Analyzing the variance of particle smoothers (FFBS, FFBSi, PaRIS, two-filer) is at the core of ongoing works and we believe that comparing the variance would be more suitable and interesting than just providing a CLT with no rules to choose between methods.
+We are also working on the estimation of such asymptotic variances which is a complex problem out of the scope of this paper.
+
+\item {\em It is worth mentioning that, for the log-growth model, $X$ is a function of $\sigma$ and what
+are the consequences of that in a context where $\sigma$ is an unknown parameter.}
+
+\vspace{.3cm}
+
+\item {\em I quite like the choice of $Q(\theta;\theta)$ in the examples. It would be very interesting to
+have an implementation of the actual EM algorithm for parameter estimation in the
+examples provided, using GRand Paris on the E step.}
+
+\vspace{.3cm}
+
+The numerical section now contains a complete implementation of the EM algorithm which highlights clearly \textcolor{red}{BIM} the performance of the proposed PaRIS algorithm.  
+\end{enumerate}
+
+\subsubsection*{Minor comments}
+\begin{enumerate}
+\item{\em page 3, line 36: The authors should clearly state that the proposed methodology
+does not rely on time discretisation approximations.}
+
+\vspace{.3cm}
+
+This is now explicitly written in the introduction.
+
+\item{\em page 5, line 21: computing $\longrightarrow$ computed.}
+
+\vspace{.3cm}
+
+This typo was corrected.
+
+\item {\em page 8, line 52: I suggest "... $L_w$ is bounded below almost surely, ..."}
+
+\vspace{.3cm}
+
+This was changed as suggested.
+
+\item {\em page 11, line 10: does the fixed lag algorithm have the same cost for all lag values?
+What is the computational time of one run of GRand PaRIS for the examples presented?}
+
+\vspace{.3cm}
+
+\textcolor{red}{M. Gloaguen, your call !}
+
+\item {\em page 16, line 49: For the proof to be valid, the induction result needs to be established for $k\ge 0$.}
+
+\vspace{.3cm}
+
+\end{enumerate}
+
 \end{document}

### 2017-08-08 TexSubmission/ParisEM.tex
First answers to referee 2

@@
 To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE).
-In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities.
+In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities, i.e.
+no discretization scheme is used.
 
- The only solution to solve the smoothing problem for partially observed SDE using SMC methods has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}.
+ The only solution to solve the smoothing problem for partially observed SDE using SMC methods without any discretization procedure has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}.
  Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias that does not vanish as the number of particles grows to infinity.
@@
 Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.
+One of the best features of PaRIS algorithm is that it may be implemented online, using the observations $(Y_k)_{k\ge 1}$ as they are received, without any increasing storage requirements.
+The total number $n$ of observations is fixed in this paper to describe the algorithm but it could process any additional observation straightforwardly.
 
 In this paper, we extend the use of  PaRIS algorithm to partially observed SDE. 
-The proposed algorithm allows to approximate smoothed expectations of additive functionals online and with a complexity growing only linearly with the number of particles.
+The proposed algorithm allows to approximate smoothed expectations of additive functionals online, with a complexity growing only linearly with the number of particles and without any discretization procedure or Taylor expansion of the transition densities.
 The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to SDE is that the accept reject mechanism introduced in \cite{douc:garivier:moulines:olsson:2011} ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
 The usual FFBS and FFBSi algorithms may not be extended this easily since they both require the computation of weights defined as ratios involving the transition densities, thus replacing these unknown quantities by unbiased estimates does not lead to unbiased estimators of the weights.
+The linear version of the FFBSi algorithm proposed in \cite{douc:garivier:moulines:olsson:2011} could be extended in a similar way as PaRIS algorithm but it would still require a backward pass and would not be an online smoother.
 The proposed Generalized Random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to general hidden Markov models whose Markovian dynamics is ruled by a stochastic differential equation (one of the first two domains defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}) but also to any general state space model where the transition density of the hidden chain may be estimated unbiasedly.
@@
 when the transition density of the hidden states is not available explicitly and where $\{h_k\}_{k=0}^{n-1}$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$.
+Smoothed additive functionals as \eqref{def:addfunc} are crucial for maximum likelihood inference of latent data models.
+These quantities appear naturally when computing the Fisher score in hidden Markov models or the intermediate quantity of the Expectation Maximization algorithm (see Section~\ref{sec:exp}).
+They are also pivotal to design online Expectation Maximization based algorithms which motivates the method introduced in this paper that does not require growing storage and can process observations online.
+
 The algorithm is based on the following link between the filtering and smoothing distributions for additive functionals, see \cite{olsson:westerborn:2016}:
@@
 \]
-Then, for $1\le k \le n$, using $\{(\xi^{\ell}_{k-1},\omega^{\ell}_{k-1})\}_{\ell=1}^N$, the auxiliary particle filter of \cite{pitt:shephard:1999} samples pairs $\{(I^{\ell}_k,\xi^{\ell}_{k})\}_{\ell=1}^N$ of indices and particles using an instrumental transition density $p_k$ on $\mathbb{R}^d\times \mathbb{R}^d$ and an adjustment multiplier function $\vartheta_k$ on $\mathbb{R}^d$.
-Each new particle $\xi^{\ell}_{k}$ and weight $\omega^{\ell}_k$ at time $k$ are computing following these steps:
+Then, for $1\le k \le n$, using $\{(\xi^{\ell}_{k-1},\omega^{\ell}_{k-1})\}_{\ell=1}^N$, the auxiliary particle filter of \cite{pitt:shephard:1999} samples pairs $\{(I^{\ell}_k,\xi^{\ell}_{k})\}_{\ell=1}^N$ of indices and particles using an instrumental transition density $p_{k-1}$ on $\mathbb{R}^d\times \mathbb{R}^d$ and an adjustment multiplier function $\vartheta_k$ on $\mathbb{R}^d$.
+Each new particle $\xi^{\ell}_{k}$ and weight $\omega^{\ell}_k$ at time $k$ are computed following these steps:
 \begin{enumerate}[-]
 \item choose a particle index $I^{\ell}_k$ at time $k-1$ in $\{1,\ldots,N\}$ with probabilities proportional to $\omega_{k-1}^{j} \vartheta_k (\xi^{j}_{k-1})$, for $j$ in $\{1,\ldots,N\}$ ;
-\item sample  $\xi^{\ell}_{k}$ using this chosen particle according to $\xi^{\ell}_{k} \sim p_k(\xi^{I^{\ell}_k}_{k-1},\cdot)$ ; 
+\item sample  $\xi^{\ell}_{k}$ using this chosen particle according to $\xi^{\ell}_{k} \sim p_{k-1}(\xi^{I^{\ell}_k}_{k-1},\cdot)$ ; 
 \item  associate the particle $\xi^{\ell}_k$ with the importance weight:
@@
 \label{eq:importance:weights}
-\omega^{\ell}_k := \frac{\qk(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_k (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
+\omega^{\ell}_k := \frac{q_{k-1}(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_{k-1} (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
 \end{equation}
@@
 \]
+The most simple choice for $p_{k-1}$ and $\vartheta_k$ is the bootstrap filter proposed by \cite{gordon:salmond:smith:1993} which sets $p_{k-1} = q_{k-1}$ and for all $x\in\mathbb{R}^d$, $\vartheta(x) = 1$.
+In the case of partially observed SDE $q_{k-1}$ is unknown but it can be replaced by any approximation to sample the particles as any choice of $p_{k-1}$ can be made.
+The approximation can be obtained using a discretization scheme such as Euler method or a Poisson based approximation as detailed below.
+A more appealing choice is the fully adapted particle filter which sets for all $x,x'\in\mathbb{R}^d$, $p_{k-1}(x,x') \propto q_{k-1}(x,x')g_k(x')$ and for all $x\in\mathbb{R}^d$, $\vartheta(x) = \int q_{k-1}(x,x')g_k(x')\mu(\rmd x')$.
+Here again $q_{k-1}$ has to be replaced by an approximation.
+In Section~\ref{sec:exp}, it is replaced by the Gaussian approximation provided by a Euler scheme which leads to a Gaussian proposal density $p_{k-1}$ as the observation model is linear and Gaussian.
+
 PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSi algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions.
@@
 \] 
+It is clear from steps (i) to (iii) that each time a new observation $Y_{n+1}$ is received, the quantities $(\tau_{n+1}^i)_{1\le i \le N}$ can be updated only using $Y_{n+1}$, $(\tau_n^i)_{1\le i \le N}$ and the particle filter at time $n$.
+This means that storage requirements do not increase when processing additional data.
+
 As proved in \cite{olsson:westerborn:2016}, the algorithm is asymptotically consistent (as $N$ goes to infinity) for any precision parameter $\tilde N$.
@@
 \label{eq:random:weight}
-\widehat{\omega}^{\ell}_k := \frac{\widehat{q}_k(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k;\zeta_k)g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_k (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
+\widehat{\omega}^{\ell}_k := \frac{\widehat{q}_{k-1}(\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k;\zeta_{k-1})g_k(\xi^{\ell}_k)}{\vartheta_k(\xi^{I^{\ell}_k}_{k-1}) p_{k-1} (\xi_{k-1}^{I^{\ell}_k},\xi^{\ell}_k)}\eqsp.
 \end{equation}
+In Algorithm~\ref{alg:Ozaki:PaRIS}, $M$ independent copies $(\zeta^m_{k-1})_{1\le m \le M}$ of $\zeta_{k-1}$ are sampled and the empirical mean of the associated estimates of the transition density are used to compute $\widehat{\omega}^{\ell}_k$ instead of a single realization.
 Therefore, to obtain a generalized random version of PaRIS algorithm, we only need to be able to sample from the discrete probability distribution $\Lambda_k^N(i,\cdot)$ in the case of SDE based HMM.
-Consider the following assumption: for all  $0\leq k\leq n$, there exists a random variable $\hat{\sigma}^k_+$ measurable with respect to $\mathcal{G}_{k+1}^N$ such that,
+Consider the following assumption: for all  $0\leq k\leq n-1$, there exists a random variable $\hat{\sigma}^k_+$ measurable with respect to $\mathcal{G}_{k+1}^N$ such that,
 \begin{assumpt}
@@
 is upper bounded almost surely by $\hat{\sigma}^k_+$.
-In particular, if $\Lo_\mw$ is bounded almost surely, \eqref{eq:rho:func} always satisfies assumption \eqref{assumpt:local} and Algorithm~\ref{alg:Ozaki:PaRIS} can be used.
+In particular, if $\Lo_\mw$ is bounded below almost surely, \eqref{eq:rho:func} always satisfies assumption \eqref{assumpt:local} and Algorithm~\ref{alg:Ozaki:PaRIS} can be used.
 This condition is always satisfied for models in the domains $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, i.e.\ domains for which the exact algorithms EA1 and EA2 can be used.
@@
 \[
-\widehat{\omega}_{0}(x) = \frac{\chi(x)g_0(x)}{\eta_0(x)} \quad\mbox{and for}\; k\ge1\quad\widehat{\omega}_{k}(x,x';z) = \frac{\widehat{\qk}(x,x';z)g_{k+1}(x')}{\vartheta_{k+1}(x) p_{k} (x,x')}\eqsp.
+\widehat{\omega}_{0}(x) = \frac{\chi(x)g_0(x)}{\eta_0(x)} \quad\mbox{and for}\; k\ge1\quad\widehat{\omega}_{k}(x,x';z) = \frac{\widehat{q}_{k-1}(x,x';z)g_{k}(x')}{\vartheta_{k}(x) p_{k-1} (x,x')}\eqsp.
 \]
@@
 In both cases, the proposal distribution $p_k$ is chosen as the following approximation of the optimal filter (or the fully adapted particle filter in the terminology of \cite{pitt:shephard:1999}):
-$$p_k(x_{k-1},x_k)\propto \tilde{q}_k(x_{k-1}, x_k)g_k(x_{k})\eqsp,$$
-where $\tilde{q}_k(x_{k-1},x_k)$ is the p.d.f.
+$$p_{k-1}(x_{k-1},x_k)\propto \tilde{q}_{k-1}(x_{k-1}, x_k)g_k(x_{k})\eqsp,$$
+where $\tilde{q}_{k-1}(x_{k-1},x_k)$ is the p.d.f.
 of Gaussian distibution with mean $\alpha(x_{k-1})\Delta_k$ and variance $\Delta_kI_d$, i.e.\ the Euler approximation of equation \eqref{eq:target:sde}.
@@
 \centering
-%\includegraphics[scale = .5]{obs_SINE}
+\includegraphics[scale = .5]{obs_SINE}
 \caption{{\em SINE model - observations}. Process $X$ solution to the SDE (balls) and observations $Y$ (circles) at times $t_0=0,\dots,t_{100}=50$.}
@@
 \centering
-%\includegraphics[scale=0.5]{res_Estep_SINE}
+\includegraphics[scale=0.5]{res_Estep_SINE}
 \caption{{\em SINE model - EM intermediate quantity}.
@@
 \centering
-%\includegraphics[scale=0.5]{res_mult_SINE_bias}
+\includegraphics[scale=0.5]{res_mult_SINE_bias}
 \caption{{\em SINE model - bias}. Distribution of the empirical absolute relative bias.}
@@
 \centering
-%\includegraphics[scale=0.5]{res_mult_SINE_cv}
+\includegraphics[scale=0.5]{res_mult_SINE_cv}
 \caption{{\em SINE model - variance}. Distribution of the empirical absolute coefficient of variation.}
@@
 \centering
-%\includegraphics[scale=0.5]{obs_LG}
+\includegraphics[scale=0.5]{obs_LG}
 \caption{{\em Log-growth model - observations}. Process $X$ solution to the SDE (balls) and observations $Y$ (circles) at times $t_0=0,\dots,t_{100}=50$.}
@@
 \centering
-%\includegraphics[scale=0.5]{res_Estep_LG}
+\includegraphics[scale=0.5]{res_Estep_LG}
 \caption{{\em Log-growth model - EM intermediate quantity}.
@@
 \centering
-%\includegraphics[scale=0.5]{res_mult_LG_bias}
+\includegraphics[scale=0.5]{res_mult_LG_bias}
 \caption{{\em Log-growth model - biais}. Distribution of the empirical absolute relative bias.}
@@
 \centering
-%\includegraphics[scale=0.5]{res_mult_LG_cv}
+\includegraphics[scale=0.5]{res_mult_LG_cv}
 \caption{{\em Log-growth model - variance}. Distribution of the empirical absolute coefficient of variation.}
@@
 \[
-\mathbb{E}\left[\widehat{\omega}^i_{k+1}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right] = \mathbb{E}\left[\frac{ \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}}, \xi^{i}_{k+1};\zeta_{k})g_{k+1}(\xi^{i}_{k+1})}{\vartheta_{k+1}(\xi^{I^{i}_{k+1}}_{k}) p_{k+1}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right]\eqsp.
+\mathbb{E}\left[\widehat{\omega}^i_{k+1}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right] = \mathbb{E}\left[\frac{ \widehat{\qk}(\xi_{k}^{I^{i}_{k+1}}, \xi^{i}_{k+1};\zeta_{k})g_{k+1}(\xi^{i}_{k+1})}{\vartheta_{k+1}(\xi^{I^{i}_{k+1}}_{k}) p_{k}(\xi_{k}^{I^{i}_{k+1}},\xi^{i}_{k+1})}\tau^{i}_{k+1}\middle| \mathcal{F}_k^{N}\right]\eqsp.
 \]

### 2017-08-08 TexArXiv/ParisEM.tex
Sources for ArXiv

--- 2017-02-27 ParisEM.tex
@@
 
-\title{Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations}
+\title{Online Sequential Monte Carlo smoother for partially observed stochastic differential equations}
 
@@
 \rhead{Particle smoother for SDE}
+\date{}
 
 \maketitle
+
+\begin{abstract}
+This paper introduces a new algorithm to approximate smoothed additive functionals for partially observed stochastic differential equations.
+This method relies on the recent procedure introduced in \cite{olsson:westerborn:2016} which allows to compute such approximations online, i.e.
+as the observations are received, and with a computational complexity growing linearly with the number of Monte Carlo samples.
+The algorithm of \cite{olsson:westerborn:2016} cannot be used in the case of partially observed stochastic differential equations since the transition density of the latent data is usually unknown.
+We prove that a similar algorithm may still be defined for partially observed continuous processes by replacing this unknown quantity by an unbiased estimator obtained for instance using general Poisson estimators.
+We prove that this estimator is consistent and its performance are illustrated using data from two models.
+\end{abstract}
+
+{\bf Keywords}: Stochastic differential equations, Smoothing, Sequential Monte Carlo Methods.
 
@@
 \ref{fig:res:LG} ) shows this estimate for an example on one simulated data set.
-The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=1600$ so that both estimations requre similar computational times.
+The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=1600$ so that both estimations require similar computational times.
 
@@
 \begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{obs_SINE}
+\includegraphics[width=\textwidth]{obs_SINE.png}
 \end{subfigure}
 \begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{res_Estep_SINE}
+\includegraphics[width=\textwidth]{res_Estep_SINE.png}
 \end{subfigure}
@@
 \begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{res_mult_SINE_bias}
+\includegraphics[width=\textwidth]{res_mult_SINE_bias.png}
 \end{subfigure}
 \begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{res_mult_SINE_cv}
+\includegraphics[width=\textwidth]{res_mult_SINE_cv.png}
 \end{subfigure}
-\caption{{\em SINE model}. Distribution of the empirical absolute relative bias [left] and of the empirical absolute coefficient of variation [right] for each method, see \eqref{eq:emp:bias}.}
+\caption{{\em SINE model}. Distribution of the empirical absolute relative bias [left] and of the empirical absolute coefficient of variation [right] for each method.}
 \label{fig:mult:SINE}
@@
 \begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{obs_LG}
+\includegraphics[width=\textwidth]{obs_LG.png}
 \end{subfigure}
 \begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{res_Estep_LG}
+\includegraphics[width=\textwidth]{res_Estep_LG.png}
 \end{subfigure}
@@
 \begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{res_mult_LG_bias}
+\includegraphics[width=\textwidth]{res_mult_LG_bias.png}
 \end{subfigure}
 \begin{subfigure}{0.49\textwidth}
-\includegraphics[width=\textwidth]{res_mult_LG_cv}
+\includegraphics[width=\textwidth]{res_mult_LG_cv.png}
 \end{subfigure}
-\caption{{\em Log-growth model}. Distribution of the empirical absolute relative bias [left] and of the empirical absolute coefficient of variation [right] for each method, see \eqref{eq:emp:bias}.}
+\caption{{\em Log-growth model}. Distribution of the empirical absolute relative bias [left] and of the empirical absolute coefficient of variation [right] for each method.}
 \label{fig:mult:LG}

### 2017-08-09 TexSubmission/answer_referees.tex
Added GEM application

@@
 
-The numerical section now contains a complete implementation of the EM algorithm which highlights clearly \textcolor{red}{BIM} the performance of the proposed PaRIS algorithm.  
+The numerical section now contains a complete implementation of a generalized EM algorithm for the SINE model which highlights clearly the performance of the proposed PaRIS algorithm.  
 \end{enumerate}

### 2017-08-09 TexSubmission/ParisEM.tex
Added GEM application

@@
 \end{equation}
-where the $(\varepsilon)_{0\leq k \leq 100}$ are i.i.d. $\mathcal{N}(0,1)$.
+where the $(\varepsilon_k)_{0\leq k \leq 100}$ are i.i.d. $\mathcal{N}(0,1)$.
 In the example displayed on Figure \ref{fig:res:SINE:obs}, we set $\theta=0$.
@@
 \label{fig:mult:SINE:cv}
+\end{figure}
+
+The performance of our algorithm are also assessed in the case where $\theta$ and the variance $\sigma^2$ of $\varepsilon_1$ are unknown and estimated using a generalized EM algorithm: $n=200$ observations are sampled with $\theta = 0$ and $\sigma^2 = 0.5$ and the Grand PaRIS algorithm is based on $N=200$ particles, $\widetilde{N} = \textcolor{red}{???}$ backward samples and $M = \textcolor{red}{???}$ replicates of the transition density estimator to approximate the importance weights.
+As their is no closed form solution to compute the M step of the EM algorithm and propose new parameter estimates, at each iteration $50$ new parameter estimates are sampled \textcolor{red}{comment ?} close to the current estimate and the intermediate quantity is approximated using our algorithm for each candidate.
+Then, parameter values leading to the greatest value of the approximate intermediate quantities are selected as the new parameter estimates.
+Figures~\ref{fig:SINE:theta} and~\ref{fig:SINE:sigma} illustrate the performance of the estimation for several initializations of $\theta$ (resp.
+$\sigma^2$) uniformly chosen in  $(-\pi,\pi)$ (resp.
+in $(0,6)$) after only a few iterations of the EM procedure.
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=0.5]{figure_theta}
+\caption{{\em SINE model - EM}. Estimation  of $\theta$.}
+\label{fig:SINE:theta}
+\end{figure}
+
+\begin{figure}[p]
+\centering
+\includegraphics[scale=0.5]{figure_sigma2}
+\caption{{\em SINE model - EM}. Estimation of $\sigma^2$.}
+\label{fig:SINE:sigma}
 \end{figure}

### 2017-08-10 TexSubmission/answer_referees.tex
Additional answers

@@
 
+The bias of forward backward based particle smoothers due to $\Lambda_k^N$ (i.e.
+to the particle approximations) was analyzed in \cite{delmoral:doucet:singh:2010}.
+In the case of additive functional as in (2), the bias of the forward only version of the FFBS algorithm is shown to be upper bounded by a term of order $n/N$ for strongly mixing hidden Markov models.
+
 \item {\em What can be said about a CLT for the proposed estimator?}
@@
 \vspace{.3cm}
+
+Even if there is no actual implementation of the EM algorithm in the paper for the log-growth model, we agree that providing some insights on the way it could be used may be very useful.
+We added a paragraph at the end of the log-growth section following the work of [4] explaining how the process can be transformed to estimate $\sigma$ using an EM algorithm although $X$ is a function of $\sigma$.
+\textcolor{red}{Pas si simple, la transformation en deux etapes de Beskos et al ne fonctionne pas comme le processus X n'est pas directement observe...}
 
@@
 
+\bibliographystyle{plain}
+\begin{thebibliography}{10}
+
+\bibitem{delmoral:doucet:singh:2010}
+P.~{D}el {M}oral, {A}.~{D}oucet, and {S}.{S}.~{S}ingh.
+\newblock A backward particle intepretation of {F}eynman-{K}ac formulae.
+\newblock {\em {ESAIM:~M2AN}}, 44:947--975, 2010.
+\end{thebibliography}
+
 \end{document}

### 2017-08-10 TexSubmission/answer_referees.tex
Additional answers

@@
 \begin{enumerate}
-\item
+\item \textcolor{red}{Je pense que c'est le gros morceau... La reponse a cette question recouvre en fait les questions 1/4/5/6/8 donc quasiment tout ce qu'il reste.
+On s'est focalises sur ces EDS essentiellement pour l'utilisation des estimateurs sans biais et leurs applications a des modeles en ecologie.
+C'est facile d'expliquer la partie gradient et d'utiliser Ait-Sahalia pour expliquer que c'est dur en non scalaire.
+Ensuite, je pense que le mieux c'est de bien decrire que les estimateurs de densites peuvent s'obtenir avec tous les EA et dans ce cas detailler les hypotheses proprement pour justifier notre approche.
+On peut aller jusqu'a l'EA3, qui sera quadratique probablement mais remplit la condition A3 donc peut etre a priori utilise.
+Ensuite, je pense que l'on peut mentionner que ce que l'on fait peut s'etendre a d'autres EDS avec une diffusion non unitaire grace aux tres recentes methodes d'importance sampling.
+La team Roberts n'a pas encore publie son algo (le fera-t-elle un jour ?) mais Kohatsu-Higa si et son algor ressemble beaucoup a l'approche Warwickienne.
+On a dans ce cas un estimateur sans biais, et que l'on peut simuler, sous des hypotheses plus generales.
+On ne change pas tout notre papier, on garde une focalisation de type Poisson mais on peut faire un paragraphe que se passe-t-il dans le cas non unitaire...
+PaRIS marche exactement pareil, a chacun d'utiliser son estimateur de densite prefere.
+What ya think ?
+Je change dans cet esprit, vous me direz ce que vous en pensez !}
+
+\vspace{.3cm}
+
+We agree with the Referee that the model was not introduced properly and that the use of generalized Poisson estimators required additional comments.
+This remark led to a major modification of the beginning of Section~2.
+The assumptions mentioned by the Referee are highlighted clearly and much more commented to explain the setting in which GPE can be used.
+
+As some assumptions may be restrictive (specifically the gradient assumption on the drift), we also provide alternative to GPE to estimate transition densities.
+The only requirement to extend PaRIS algorithm is to obtain an unbiased estimator than can be evaluated explicitly to compute the importance weights.
+A recent class of estimators which do not use the gradient assumptions are the continuous importance sampling based methods of \cite{}.
+They introduce other sets of assumptions such as...
+
 \item {\em When presenting equation (2), the authors should explain the importance of approximating those type of functionals.}
@@
 \newblock {\em {ESAIM:~M2AN}}, 44:947--975, 2010.
+
+\bibitem{olsson:westerborn:2016}
+J.~Olsson and J.~Westerborn.
+\newblock Efficient particle-based online smoothing in general hidden {M}arkov
+  models: the {PaRIS} algorithm.
+\newblock {\em ArXiv:1412.7550}, 2016.
+
 \end{thebibliography}

### 2017-08-10 TexSubmission/ParisEM.tex
Additional answers

@@
 \label{sec:rwparis}
-$(X_t)_{t\ge 0}$ is defined as a weak solution to the following SDE in $\mathbb{R}^d$:
+%. These estimators are designed for stochastic processes
+Let $(X_t)_{t\ge 0}$ be defined as a weak solution to the following SDE in $\mathbb{R}^d$:
 \begin{equation}
 \label{eq:target:sde}
-X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha(X_t)\rmd t + \rmd W_t\eqsp,
+X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha(X_t)\rmd t + \Gamma(X_t)\rmd W_t\eqsp,
 \end{equation}
-where $(W_t)_{t\ge 0}$ is a standard Brownian motion.
-It is assumed that $\alpha$ is of the form $\alpha(x) = \nabla_x A(x)$ where $A: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function.
+where $(W_t)_{t\ge 0}$ is a standard Brownian motion on $\mathbb{R}^d$, $\alpha: \mathbb{R}^d\to\mathbb{R}^d$ and $\Gamma: \mathbb{R}^d\to \mathbb{R}^{d\times d}$. 
+In this paper, we focus on unbiased estimator of the unknown transition densities given by generalized Poisson estimators defined in \cite{fearnhead:papaspiliopoulos:roberts:2008}.
+The use of generalized Poisson estimators requires that the diffusion coefficient $\Gamma$ is constant and equals to the identity matrix.
+Note that, when this is not the case, there exists a class of diffusion processes that may be transformed to satisfy this assumption.
+This paper then focuses on reducible SDE for which there exists an invertible and infinitely differentiable function $\eta$ such that the process $\left\{Z_t = \eta(X_t)\right\}_{t\ge 0}$ satisfies the SDE  $z_0 = \eta(x_0)$ and
+\begin{equation}
+\label{eq:diff:lamperti}
+\mathrm{d}Z_t = \beta(Z_t)\mathrm{d}t + \mathrm{d}W_t\eqsp.
+\end{equation}
+By Ito's formula,  it is straightforward to show that, in the case of a reducible diffusion, the Jacobian matrix of $\eta$ satisfies
+\[
+\nabla \eta = \Gamma^{-1}\eqsp,
+\]
+and, in the case $d=1$,
+\[
+\beta:u \mapsto \frac{\alpha(\eta^{-1}(u))}{\Gamma(\eta^{-1}(u))} - \frac{\Gamma'(\eta^{-1}(u))}{2}\eqsp.
+\]
+In the case of a scalar diffusion, this Lamperti transform is given by:
+\[
+\eta:x\mapsto \int_{x_0}^x\Gamma^{-1}(u)\mathrm{d}u\eqsp.
+\]
+In the general case, \cite[Proposition~$1$]{ait-sahalia:2008} shows that when $\Gamma$ is nonsingular, the SDE is reducible if and only if, for all $1\le i,j,k\le d$,
+\begin{equation}
+\label{eq:reducibility}
+\frac{\partial\Gamma^{-1}_{i,j}}{\partial x_k} = \frac{\partial\Gamma^{-1}_{i,k}}{\partial x_j}\eqsp.
+\end{equation}
+In the case of a diagonal matrix $\Gamma$ \eqref{eq:reducibility} is equivalent to assume that $\Gamma$ is such that for each $1\le i\le d$, $\Gamma_{i,i}$ depends on $x_i$ only.
+\cite{ait-sahalia:2008} notes that the reducibility condition \eqref{eq:reducibility} holds also for some nondiagonal matrices $\Gamma$.
+This is true in particular in the case $d=2$ for stochastic volatility models where $\sigma$ is of the form:
+\[
+\Gamma(x) = \begin{pmatrix}a(x_1) & a(x_1)b(x_2)\\0&c(x_2)\end{pmatrix}\eqsp.
+\]
+In the following, it is assumed that the process $(X_t)_{t\ge 0}$ satisfies the SDE \eqref{eq:target:sde} with $\Gamma$ being the identity matrix, i.e.
+we consider a diffusion after the application of the Lamperti transform.
+In addition, designing general Poisson estimators also requires that:
+\begin{enumerate}[i)]
+\item $\alpha$ is of the form $\alpha(x) = \nabla_x A(x)$ where $A: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function ;
+\item the function $x\mapsto \left(\|\alpha(x)\|^2  + \triangle A(x)\right)/2$ is lower bounded where $\triangle$ is the Laplace operator.
+\end{enumerate}
+Assumption i) is the most restrictive \textcolor{red}{quelques exemples} while assumption ii) is satisfied in many situations \textcolor{red}{quelques exemples}.
+They are for instance satisfied for potential based movement models in ecology, see \cite{gloaguen:etienne:lecorff:2017}.
+
+In the case where these assumptions are not fulfilled, an alternative to generalized Poisson estimators is given by continuous importance sampling procedures for SDE, see \cite{}.
+
 The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{0\le k \le n}$ in $(\mathbb{R}^m)^{n+1}$. 
@@
 \[
-\phi(x) =\left(\|\alpha(x)\|^2  + \triangle A(x)\right)/2\eqsp,
+\phi(x) =\left(\|\alpha(x)\|^2  + \triangle A(x)\right)/2\eqsp.
 \]
-with $\triangle$ the Laplace operator.
 Assume that there exist random variables $\Lo_\mw$ and $\U_\mw$ such that for all $0\leq s \leq \Delta_k$, $\Lo_\mw \leq \phi(\mw_s)\leq \U_\mw$.
@@
 Again, the GRand PaRIS algorithm outperforms the fixed lag smoother as it shows  a similar (vanishing) bias as the fixed lag for the largest lag and a smaller variance than the fixed lags estimates with negligible bias.
+
+Note that in this case the Lamperti transform to obtain a diffusion with a unitary diffusion term depends on $\sigma$.
+The process $(X_t)_{t\ge 0}$ is a function of $\sigma$ and is not directly observed if $\sigma$ is unknown, which prevents a direct use of an EM algorithm to estimate $\sigma$.
+Following \cite[Section~8.2]{beskos:papaspiliopoulos:roberts:fearnhead:2006}, this may be overcome with a two-step transformation of the process $(Z_t)_{t\ge 0}$.
+After the Lamperti transform, the process is transformed into a diffusion bridge  is used to obtai
 
@@
 
+\bibitem{gloaguen:etienne:lecorff:2017}
+P.~ Gloaguen, M.-P.~\'Etienne, and S.~Le~Corff.
+\newblock Inference procedures to estimate potential based movement models in ecology.
+\newblock {\em To appear in the {J}ournal of the {R}oyal {S}tatistical {S}ociety: {S}eries {C}}.
+
 \bibitem{godsill:doucet:west:2004}

### 2017-08-18 TexSubmission/ParisEM.tex

@@
  \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers for state space models by combining importance sampling steps to propagate particles with resampling steps to duplicate or discard particles according to their importance weights.
+
 Unfortunately, these methods cannot be applied directly to partially observed stochastic differential equations since some elementary quantities, such as transition densities of the hidden states, are not available explicitly.
@@
      To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE).
-In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities, i.e.
-no discretization scheme is used.
+     In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities, i.e. no discretization scheme is used.
 
  The only solution to solve the smoothing problem for partially observed SDE using SMC methods without any discretization procedure has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}.
- Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias that does not vanish as the number of particles grows to infinity.
+ Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias, due to the fixed lag approximation, that does not vanish as the number of particles grows to infinity.
 In the case of discrete time state space models, approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}.
@@
 
-PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSi algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions.
+The PaRIS algorithm uses the same decomposition as the FFBS algorithm introduced in \cite{doucetgodsillandrieu:2000} and the FFBSi algorithm proposed by \cite{godsill:doucet:west:2004} to approximate smoothing distributions.
  It combines both the forward only version of the FFBS algorithm with the sampling mechanism of the FFBSi algorithm.
@@
 Therefore, to obtain a generalized random version of PaRIS algorithm, we only need to be able to sample from the discrete probability distribution $\Lambda_k^N(i,\cdot)$ in the case of SDE based HMM.
+
 Consider the following assumption: for all  $0\leq k\leq n-1$, there exists a random variable $\hat{\sigma}^k_+$ measurable with respect to $\mathcal{G}_{k+1}^N$ such that,
@@
 \label{lem:AR:unbiased}
-Assume that $A_1$ holds for some $0\le k\le n-1$.  For all $1\le i \le N$, define the random variable $J_k^i$ as follows:
+Assume that \eqref{assumpt:global} holds for some $0\le k\le n-1$.  For all $1\le i \le N$, define the random variable $J_k^i$ as follows:
 \begin{algorithmic}
@@
 \begin{assumpt}\label{assumpt:local}
-\mathrm{sup}_{i,j,\zeta}\;\widehat{q}_k(\xi^j_k,\xi^j_{k+1},\zeta)\leq \hat{\sigma}^k_+\;.
+\mathrm{sup}_{i,j,\zeta}\;\widehat{q}_k(\xi^j_k,\xi^i_{k+1},\zeta)\leq \hat{\sigma}^k_+\;.
 \end{assumpt}
@@
 \STATE Update $\tau_{k+1}^i = \tau_{k+1}^i + (\tau^{J_k^{i,\ell}}_{k} + h_k(\xi^{J_k^{i,\ell}}_{k},\xi^i_{k+1}))/\tilde{N}$.
-%\STATE \textbf{\sc Sampling Step} \\
-%Sample independently $\zeta_k$, $U\sim \mathcal{U}[0,1]$ and $J\in\{1,\ldots,N\}$ with probabilities proportional to  $\{\widehat{\omega}_{k}^1,\dots,\widehat{\omega}_{k}^N\}$.
-%\IF{ $$U \leq \frac{\widehat{\qk}(\xi_{k}^J,\xi_{k+1}^i,\zeta_k)}{\hat{\sigma}^k_+},$$}
-%\STATE Set $J_k^{i,\ell} = j$.
-%\STATE Update $\tau_{k+1}^i = \tau_{k+1}^i + \left(\tau^{J_k^{i,\ell}}_{k} + h_k(\xi^{J_k^{i,\ell}}_{k},\xi^i_{k+1})\right)/\tilde{N}$;
-%\ELSE 
-%\STATE Return to \textbf{\sc Sampling Step}
-%\ENDIF
 \ENDFOR
@@
 $\log q_k$ is not available explicitly and is approximated using the unbiased estimator proposed in \cite[Appendix B]{olsson:strojby:2011} based on 30 independent Monte Carlo simulations.
-The intermediate quantity of the EM algorithm is also estimated                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                                      with our algorithm 30 times using $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, and denoted by $\hQ_\star$.
-Figure \ref{fig:res:SINE:EM} (resp.
-\ref{fig:res:LG:EM}) shows this estimate for an example on one simulated data set.
+In order to obtain a reference value for our study, the intermediate quantity of the EM algorithm is also estimated 30 times using the GRand PaRIS algorithm with $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, and denoted by $\hQ_\star$.
+Figure \ref{fig:res:SINE:EM} (resp. \ref{fig:res:LG:EM}) shows this estimate for an example on one simulated data set. 
 The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=1600$ so that both estimations require similar computational times.
@@
 \begin{equation}
-\rmd X_t = \sin \left(X_t-\theta\right)\rmd t + \rmd W_t,~~X_0=x_0\eqsp. \label{eq:Lamp:SINE}
+\rmd X_t = \sin \left(X_t-\mu\right)\rmd t + \rmd W_t,~~X_0=x_0\eqsp. \label{eq:Lamp:SINE}
 \end{equation}
@@
 The process solution to \eqref{eq:Lamp:SINE} is observed regularly at times $t_0=0,\ldots,t_{100}=50$ through the observation process $(Y_k)_{0\leq k \leq 100}$:
-\begin{equation}
-Y_k = X_k + \varepsilon_k\label{eq:obs:SINE}\eqsp,
-\end{equation}
-where the $(\varepsilon_k)_{0\leq k \leq 100}$ are i.i.d. $\mathcal{N}(0,1)$.
-In the example displayed on Figure \ref{fig:res:SINE:obs}, we set $\theta=0$.
+\begin{equation*}
+Y_k = X_k + \varepsilon_k\eqsp,
+\end{equation*}
+where the $(\varepsilon_k)_{0\leq k \leq 100}$ are i.i.d. $\mathcal{N}(0, \sigma^2_\text{obs})$, the resulting set of model parameters is $\theta = (\mu, \sigma_\text{obs})$.
+In the example displayed on Figure \ref{fig:res:SINE:obs}, we set $\mu=0$ an d $ \sigma_\text{obs} = 1$.
 In that case, the function $\rho_{\Delta_k}$ defined in \eqref{eq:rho:func} can be upper bounded either on $(x,y)$ or only on $y$, the GRand PaRIS algorithm has therefore a linear complexity.
@@
 For each simulation $s$, the empirical absolute relative bias $\mathsf{arb}_s$ and the empirical absolute coefficient of variation $\mathsf{acv}_s$ are computed as
-\begin{align}
-\mathsf{arb}_s &= \frac{\vert m(\hQ^s)-\hQ^s_\star\vert }{\vert \hQ^s_\star\vert }\label{eq:emp:bias}\\
-\mathsf{acv}_s&=\frac{\sigma(\hQ^s)}{\vert m(\hQ^s)\vert }\label{eq:emp:cv}
-\end{align}
+\begin{align*}
+\mathsf{arb}_s &= \frac{\vert m(\hQ^s)-\hQ^s_\star\vert }{\vert \hQ^s_\star\vert }\\
+\mathsf{acv}_s&=\frac{\sigma(\hQ^s)}{\vert m(\hQ^s)\vert }\eqsp,
+\end{align*}
 where $m(\hQ^s)$ and $\sigma(\hQ^s)$ are the empirical mean and standard deviation of the sample $Q_1^s,\dots,Q_{200}^s$.
@@
 The dot represents the empirical mean over the 200 replicates. 
-The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=5000$ particles.}
+The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N = 5000$ particles.}
 \label{fig:res:SINE:EM}
@@
 \end{figure}
-
-The performance of our algorithm are also assessed in the case where $\theta$ and the variance $\sigma^2$ of $\varepsilon_1$ are unknown and estimated using a generalized EM algorithm: $n=200$ observations are sampled with $\theta = 0$ and $\sigma^2 = 0.5$ and the Grand PaRIS algorithm is based on $N=200$ particles, $\widetilde{N} = \textcolor{red}{???}$ backward samples and $M = \textcolor{red}{???}$ replicates of the transition density estimator to approximate the importance weights.
-As their is no closed form solution to compute the M step of the EM algorithm and propose new parameter estimates, at each iteration $50$ new parameter estimates are sampled \textcolor{red}{comment ?} close to the current estimate and the intermediate quantity is approximated using our algorithm for each candidate.
-Then, parameter values leading to the greatest value of the approximate intermediate quantities are selected as the new parameter estimates.
-Figures~\ref{fig:SINE:theta} and~\ref{fig:SINE:sigma} illustrate the performance of the estimation for several initializations of $\theta$ (resp.
-$\sigma^2$) uniformly chosen in  $(-\pi,\pi)$ (resp.
-in $(0,6)$) after only a few iterations of the EM procedure.
+\paragraph{Generalized EM procedure}
+The performance of our algorithm is also assessed in the case where $\theta$ and the variance $\sigma^2_\text{obs}$ are unknown and estimated using a generalized EM algorithm. 
+The study is done using a data set with $n = 200$ observations simulated with $\mu=0$ an d $ \sigma^2_\text{obs} = 1$.
+The Grand PaRIS algorithm is used to perform the E step, with the same settings as before for $N$, $\tilde{N}$ and $M$.
+As their is no closed form solution to compute the M step of the EM algorithm and propose new parameter estimates, we use a Generalized EM procedure: given the current estimation $\theta^{(k)} := (\mu^{(k)}, \sigma_{\text{obs}}^{(k)}$, the function $\mathcal{Q}(\cdot, \theta^{(k)})$ is approximated for $50$ new candidates $\theta_1, \dots, \theta_{50} $ chosen by the user.
+The new estimate is set as
+\[
+\theta^{(k + 1)} = \text{argmax}_i\mathcal{Q}(\theta_i, \theta^{(k)})\eqsp .
+\]
+This procedure has the nice property of using the same particle filter and the same retrospective sampling of Lemma \ref{lem:AR:unbiased} for all candidates, avoiding to repeat this time consuming procedure.
+The number of candidates and the way to choose them is problem dependent and then left to the user. 
+In our case, we sampled candidates using Gaussian distributions around the currend estimate $\theta^{(k)}$, decreasing the variance when $k$ increases.
+Figures~\ref{fig:SINE:theta} and~\ref{fig:SINE:sigma} illustrate the performance of the estimation for 12 different initializations of $\mu$ (resp.
+$\sigma_\text{obs}$) uniformly chosen in  $(-\pi,\pi)$ (resp.
+in $(0,6)$), showing a convergence after only a few iterations of the EM procedure.
 
@@
 \includegraphics[scale=0.5]{figure_theta}
-\caption{{\em SINE model - EM}. Estimation  of $\theta$.}
+\caption{{\em SINE model - EM}. Estimation  of $\mu$.}
 \label{fig:SINE:theta}
@@
 \begin{equation}
-\rmd Z_t = \kappa Z_t\left(1-\frac{Z_t}{\gamma}\right)\rmd t + \sigma Z_t \rmd W_t,~~Z_0=z_0. \label{eq:LG:SDE}
+\rmd Z_t = \kappa Z_t\left(1-\frac{Z_t}{\gamma}\right)\rmd t + \sigma Z_t \rmd W_t,~~Z_0=z_0\eqsp . \label{eq:LG:SDE}
 \end{equation}
@@
 The process solution to \eqref{eq:Lamp:LG} is observed regularly at times $t_0=0,\dots,t_{50}=100$ through the observation process $(Y_k)_{0\le k\le 50}$ defined as:
-\begin{equation}
-Y_k = X_k + \varepsilon_k\label{eq:obs:LG}\eqsp,
-\end{equation}
+\begin{equation*}
+Y_k = X_k + \varepsilon_k\eqsp,
+\end{equation*}
 where the $(\varepsilon_k)_{0\le k \le 50}$ are i.i.d. $ \mathcal{N}(0,\sigma^2_{obs})$.
-The parameters are given by $$\theta =(\kappa=0.1,\sigma=0.1,\gamma=1000,\sigma^2_{obs}=4)\eqsp.$$
+The parameters are given by $$\theta =(\kappa=0.1,\sigma=0.1,\gamma=1000,\sigma_{obs} = 2)\eqsp.$$
 In that case, the $\rho_{\Delta_k}$ function defined in \eqref{eq:rho:func} can be upper bounded as a function of $y$ when $x\in \{\xi_k^1,\dots, \xi_k^N\}$, the GRand PaRIS algorithm has therefore a linear complexity.
@@
 Following \cite[Section~8.2]{beskos:papaspiliopoulos:roberts:fearnhead:2006}, this may be overcome with a two-step transformation of the process $(Z_t)_{t\ge 0}$.
-After the Lamperti transform, the process is transformed into a diffusion bridge  is used to obtai
-
 \begin{figure}[p]
@@
 \includegraphics[scale=0.5]{obs_LG}
-\caption{{\em Log-growth model - observations}. Process $X$ solution to the SDE (balls) and observations $Y$ (circles) at times $t_0=0,\dots,t_{100}=50$.}
+\caption{{\em Log-growth model - observations}.
+ Process $X$ solution to the SDE (balls) and observations $Y$ (circles) at times $t_0=0,\dots,t_{100}=50$.}
 \label{fig:res:LG:obs}
@@
  Estimation of the EM intermediate quantity $\mathcal{Q}(\theta,\theta)$  using the fixed lag (FL) technique for 5 different lags, and the GRand PaRIS algorithm using 200 replicates.
-The whiskers represent the extent of the 95\% central values.
-The dot represents the empirical mean over the 200 replicates.
+  The whiskers represent the extent of the 95\% central values. The dot represents the empirical mean over the 200 replicates.
    The dotted line shows the reference value, computed using the GRand PaRIS algorithm with $N=5000$ particles.}
@@
 \includegraphics[scale=0.5]{res_mult_LG_bias}
-\caption{{\em Log-growth model - biais}. Distribution of the empirical absolute relative bias.}
+\caption{{\em Log-growth model - biais}. 
+Distribution of the empirical absolute relative bias.}
 \label{fig:mult:LG:b}
@@
 \includegraphics[scale=0.5]{res_mult_LG_cv}
-\caption{{\em Log-growth model - variance}. Distribution of the empirical absolute coefficient of variation.}
+\caption{{\em Log-growth model - variance}.
+ Distribution of the empirical absolute coefficient of variation.}
 \label{fig:mult:LG:cv}
@@
 This paper presents a new online SMC smoother for partially observed differential equations.
-This algorithm relies on an accept-reject procedure inspired from the recent PaRIS algorithm.
+ This algorithm relies on an acceptance-rejection procedure inspired from the recent PaRIS algorithm.
  The main result of the article for practical applications is that the mechanism of this procedure remains valid when the transition density is approximated by a an unbiased positive estimator.
-The proposed procedure outperforms the existing fixed lag smoother for SDE of \cite{olsson:strojby:2011}, as it does not introduce an intrinsic and non vanishing bias.
+  The proposed procedure therefore extend the PaRIS algorithm to HMM whose Markov transition density is unknown and can be approximated without bias.
+The GRand PaRIS algorithm outperforms the existing fixed lag smoother for SDE of \cite{olsson:strojby:2011}, as it does not introduce an intrinsic and non vanishing bias.
    In addition, numerical simulations highlight a better variance using data from two different models.

### 2017-08-18 TexSubmission/ParisEM.tex
Modifications in the introduction, focus on POD processes (as in olsson and strojby 2011) and HMMs to distinguish our work. Order in paragraphs has been changed to now be 1) What smoothers exist in classical HMMs, 2) Why it can't work for POD processes, 3) What smoothers exist for POD processes 4) what do we propose, Changes should still be made page 5. I'm not convinced that conditions i) and ii) are well placed there. Furthermore, maybe a bit of investigations on whether the condition on gradient norm and laplacien is always true on R^d should be done

@@
 
-\title{Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations}
+\title{Online Sequential Monte Carlo smoother for partially observed diffusion processes}
 
@@
 \section{Introduction}
-This paper introduces a new algorithm to solve the smoothing problem for partially observed continuous time stochastic processes.
+This paper introduces a new algorithm to solve the smoothing problem for hidden Markov models (HMMs) whose hidden state is supposed to be the solution of a stochastic differential equation (SDE).
+In \cite{olsson:strojby:2011}, these models are referred to as partially observed diffusion (POD) processes.
  In this setting, the hidden state process $(X_t)_{t\ge 0}$ is assumed to be a solution to a stochastic differential equation (SDE) and the only information available is given by noisy observations $(Y_{k})_{0\le k\le n}$ of the states $(X_k)_{0\le k\le  n}$ at some discrete time points $(t_k)_{0\le k\le n}$.
@@
 
-Statistical inference for partially observed state sequences often requires to solve bayesian filtering and smoothing problems, i.e.\ the computation of the posterior distributions of sequences of hidden states given observations.
+Statistical inference for HMMs often requires to solve bayesian filtering and smoothing problems, i.e. the computation of the posterior distributions of sequences of hidden states given observations.
  The filtering problem refers to the estimation, for each $0\le k \le n$,  of the distributions of the hidden state $X_k$ given the observations $(Y_0,\ldots,Y_k)$.
@@
 
-The exact computation of these expectations is usually not possible in the case of partially observed diffusions. 
-In this paper, we propose to use Sequential Monte Carlo (SMC) methods to approximate smoothing distributions with random particles associated with importance weights.
+However, for most HMMs, the exact computation of these expectations is usually not possible, and has to be replace by approximations.
+Sequential Monte Carlo (SMC) methods has been a popular class of algorithms to approximate smoothing distributions with random particles associated with importance weights.
  \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers for state space models by combining importance sampling steps to propagate particles with resampling steps to duplicate or discard particles according to their importance weights.
-
-Unfortunately, these methods cannot be applied directly to partially observed stochastic differential equations since some elementary quantities, such as transition densities of the hidden states, are not available explicitly.
-  Discretization procedures may be used to approximate transition densities, for instance the Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998}, or Gaussian based approximations using Taylor expansions of the posterior mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}.
-   Other approaches based on Hermite polynomials expansion were also introduced by \cite{ait-sahalia:1999,ait-sahalia:2002,ait-sahalia:2008} and extended in several directions recently, see \cite{li:2013} and all the references on the approximation of transition densities therein.
-    However, even the most recent discretization based approximations of the transition densities induce a systematic bias of particle based approximations of posterior distributions, see for instance \cite{delmoral:jacod:protter:2001}.
-     To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE).
-     In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities, i.e. no discretization scheme is used.
-
- The only solution to solve the smoothing problem for partially observed SDE using SMC methods without any discretization procedure has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}.
- Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias, due to the fixed lag approximation, that does not vanish as the number of particles grows to infinity.
-In the case of discrete time state space models, approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}.
+In the case of HMMs, approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}.
 Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$. 
@@
  One of the best features of PaRIS algorithm is that it may be implemented online, using the observations $(Y_k)_{k\ge 1}$ as they are received, without any increasing storage requirements.
- The total number $n$ of observations is fixed in this paper to describe the algorithm but it could process any additional observation straightforwardly.
 
-In this paper, we extend the use of  PaRIS algorithm to partially observed SDE. 
+Unfortunately, these methods cannot be applied directly to POD processes since some elementary quantities, such as transition densities of the hidden states, are not available explicitly.
+  In the context of SDEs,  discretization procedures may be used to approximate transition densities.
+   For instance, the classical Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998}, or Gaussian based approximations using Taylor expansions of the posterior mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}.
+   Other approaches based on Hermite polynomials expansion were also introduced by \cite{ait-sahalia:1999,ait-sahalia:2002,ait-sahalia:2008} and extended in several directions recently, see \cite{li:2013} and all the references on the approximation of transition densities therein.
+    However, even the most recent discretization based approximations of the transition densities induce a systematic bias in the approximation of the transition densities (see for instance \cite{delmoral:jacod:protter:2001}).
+    This bias approximation therefore induces a bias of particle based approximations of posterior distributions in POD processes.
+
+To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE).
+     In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities, i.e. no discretization scheme is used.
+
+ The only solution to solve the smoothing problem for POD processes using SMC methods without any discretization procedure has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}.
+ Using forgetting properties of the hidden chain, the algorithm improves the performance of \cite{fearnhead:papaspiliopoulos:roberts:2008} to approximate smoothing distributions but at the cost of a bias, this time due to the fixed lag approximation, that does not vanish as the number of particles grows to infinity.
+
+In this paper, we propose to use Sequential Monte Carlo (SMC) methods to obtain asymptoticaly unbiased approximations of smoothing expectations in the POD processes, by extending the use of  PaRIS algorithm to POD processes.
 The proposed algorithm allows to approximate smoothed expectations of additive functionals online, with a complexity growing only linearly with the number of particles and without any discretization procedure or Taylor expansion of the transition densities.
- The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to SDE is that the accept reject mechanism introduced in \cite{douc:garivier:moulines:olsson:2011} ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
+ The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to SDE is that the acceptance rejection mechanism introduced in \cite{douc:garivier:moulines:olsson:2011} ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
  The usual FFBS and FFBSi algorithms may not be extended this easily since they both require the computation of weights defined as ratios involving the transition densities, thus replacing these unknown quantities by unbiased estimates does not lead to unbiased estimators of the weights.
  The linear version of the FFBSi algorithm proposed in \cite{douc:garivier:moulines:olsson:2011} could be extended in a similar way as PaRIS algorithm but it would still require a backward pass and would not be an online smoother.
- The proposed Generalized Random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to general hidden Markov models whose Markovian dynamics is ruled by a stochastic differential equation (one of the first two domains defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}) but also to any general state space model where the transition density of the hidden chain may be estimated unbiasedly.
+ The proposed Generalized Random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to POD processes but also to any general state space model where the transition density of the hidden chain may be approximated using a strictly positive and unbiased estimator.
 
@@
 \end{enumerate}
-Assumption i) is the most restrictive \textcolor{red}{quelques exemples} while assumption ii) is satisfied in many situations \textcolor{red}{quelques exemples}.
- They are for instance satisfied for potential based movement models in ecology, see \cite{gloaguen:etienne:lecorff:2017}.
+Assumption i) is somewhat restrictive as it requires $\alpha$ to derive from a scalar potential, however, it has natural applications in many fields, for instance in movement ecology, see \cite{gloaguen:etienne:lecorff:2017}.
+Assumption ii) is technical {\color{red} JAI LIMPRESSION QUE CEST TOUJOURS VRAI SI A(X) EST DEFINI SUR $R^d$}.
 
-In the case where these assumptions are not fulfilled, an alternative to generalized Poisson estimators is given by continuous importance sampling procedures for SDE, see \cite{}.
+{\color{red} In the case where these assumptions are not fulfilled, an alternative to generalized Poisson estimators is given by continuous importance sampling procedures for SDE, see \cite{}.
+REFERENCE A METTRE}
 
@@
 When \eqref{assumpt:global} or \eqref{assumpt:ysupport} holds, it can be nonetheless of practical interest to choose the bound $\hat{\sigma}^k_+$ corresponding to \eqref{assumpt:local}. 
-Indeed, this might increase significantly the acceptance rate of the algorithm, and therefore reduce the number of drawings of the random variable $\zeta$, which has a much higher cost than the computation of $\rho$, as it requires simulations of Brownian Bridges.
+Indeed, this might increase significantly the acceptance rate of the algorithm, and therefore reduce the number of drawings of the random variable $\zeta_k$, which has a much higher cost than the computation of $\rho_{\Delta_k}$, as it requires simulations of Brownian bridges.
 Moreover, this latter option can also avoid numerical optimization if no analytical expression of $\hat{\sigma}_+^k$ is available. 

### 2017-08-18 TexSubmission/ParisEM.tex
harmonization of POD processes appelation

@@
 \lhead{Gloaguen et al.}
-\rhead{Particle smoother for SDE}
+\rhead{Particle smoother for POD processes}
 
@@
 \section{Introduction}
-This paper introduces a new algorithm to solve the smoothing problem for hidden Markov models (HMMs) whose hidden state is supposed to be the solution of a stochastic differential equation (SDE).
+This paper introduces a new algorithm to solve the smoothing problem for hidden Markov models (HMMs) whose hidden state is supposed to be the solution to a stochastic differential equation (SDE).
 In \cite{olsson:strojby:2011}, these models are referred to as partially observed diffusion (POD) processes.
- In this setting, the hidden state process $(X_t)_{t\ge 0}$ is assumed to be a solution to a stochastic differential equation (SDE) and the only information available is given by noisy observations $(Y_{k})_{0\le k\le n}$ of the states $(X_k)_{0\le k\le  n}$ at some discrete time points $(t_k)_{0\le k\le n}$.
+ In this setting, the hidden state process $(X_t)_{t\ge 0}$ is assumed to be a solution to a SDE and the only information available is given by noisy observations $(Y_{k})_{0\le k\le n}$ of the states $(X_k)_{0\le k\le  n}$ at some discrete time points $(t_k)_{0\le k\le n}$.
   The bivariate stochastic process $\{(X_{k},Y_{k})\}_{0\le k\le n}$ is a state space model such that conditional on the state sequence $(X_{k})_{0\le k\le n}$ the observations $(Y_{k})_{0\le k\le n}$ are independent and for all $0\le \ell\le n$ the conditional distribution of $Y_{\ell}$ given $\{X_{k}\}_{0\le k\le n}$ depends on $X_{\ell}$ only.
@@
 The proposed algorithm allows to approximate smoothed expectations of additive functionals online, with a complexity growing only linearly with the number of particles and without any discretization procedure or Taylor expansion of the transition densities.
- The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to SDE is that the acceptance rejection mechanism introduced in \cite{douc:garivier:moulines:olsson:2011} ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
+ The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to POD processes is that the acceptance rejection mechanism introduced in \cite{douc:garivier:moulines:olsson:2011} ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
  The usual FFBS and FFBSi algorithms may not be extended this easily since they both require the computation of weights defined as ratios involving the transition densities, thus replacing these unknown quantities by unbiased estimates does not lead to unbiased estimators of the weights.
@@
 The most simple choice for $p_{k-1}$ and $\vartheta_k$ is the bootstrap filter proposed by \cite{gordon:salmond:smith:1993} which sets $p_{k-1} = q_{k-1}$ and for all $x\in\mathbb{R}^d$, $\vartheta(x) = 1$.
- In the case of partially observed SDE $q_{k-1}$ is unknown but it can be replaced by any approximation to sample the particles as any choice of $p_{k-1}$ can be made.
+ In the case of POD processes $q_{k-1}$ is unknown but it can be replaced by any approximation to sample the particles as any choice of $p_{k-1}$ can be made.
   The approximation can be obtained using a discretization scheme such as Euler method or a Poisson based approximation as detailed below.
@@
 In Algorithm~\ref{alg:Ozaki:PaRIS}, $M$ independent copies $(\zeta^m_{k-1})_{1\le m \le M}$ of $\zeta_{k-1}$ are sampled and the empirical mean of the associated estimates of the transition density are used to compute $\widehat{\omega}^{\ell}_k$ instead of a single realization.
-Therefore, to obtain a generalized random version of PaRIS algorithm, we only need to be able to sample from the discrete probability distribution $\Lambda_k^N(i,\cdot)$ in the case of SDE based HMM.
+Therefore, to obtain a generalized random version of PaRIS algorithm, we only need to be able to sample from the discrete probability distribution $\Lambda_k^N(i,\cdot)$ in the case of POD processes.
 
@@
   The proposed procedure therefore extend the PaRIS algorithm to HMM whose Markov transition density is unknown and can be approximated without bias.
-The GRand PaRIS algorithm outperforms the existing fixed lag smoother for SDE of \cite{olsson:strojby:2011}, as it does not introduce an intrinsic and non vanishing bias.
+The GRand PaRIS algorithm outperforms the existing fixed lag smoother for POD processes of \cite{olsson:strojby:2011}, as it does not introduce an intrinsic and non vanishing bias.
    In addition, numerical simulations highlight a better variance using data from two different models.

### 2017-08-25 TexSubmission/ParisEM.tex
parenthesis

@@
 The Grand PaRIS algorithm is used to perform the E step, with the same settings as before for $N$, $\tilde{N}$ and $M$.
-As their is no closed form solution to compute the M step of the EM algorithm and propose new parameter estimates, we use a Generalized EM procedure: given the current estimation $\theta^{(k)} := (\mu^{(k)}, \sigma_{\text{obs}}^{(k)}$, the function $\mathcal{Q}(\cdot, \theta^{(k)})$ is approximated for $50$ new candidates $\theta_1, \dots, \theta_{50} $ chosen by the user.
+As their is no closed form solution to compute the M step of the EM algorithm and propose new parameter estimates, we use a Generalized EM procedure: given the current estimation $\theta^{(k)} := \left(\mu^{(k)}, \sigma_{\text{obs}}^{(k)} \right)$, the function $\mathcal{Q}(\cdot, \theta^{(k)})$ is approximated for $50$ new candidates $\theta_1, \dots, \theta_{50} $ chosen by the user.
 The new estimate is set as

### 2017-09-06 TexSubmission/ParisEM.tex
Relu jusque page 7, à finir

@@
 where $(W_t)_{t\ge 0}$ is a standard Brownian motion on $\mathbb{R}^d$, $\alpha: \mathbb{R}^d\to\mathbb{R}^d$ and $\Gamma: \mathbb{R}^d\to \mathbb{R}^{d\times d}$. 
-In this paper, we focus on unbiased estimator of the unknown transition densities given by generalized Poisson estimators defined in \cite{fearnhead:papaspiliopoulos:roberts:2008}.
- The use of generalized Poisson estimators requires that the diffusion coefficient $\Gamma$ is constant and equals to the identity matrix.
-  Note that, when this is not the case, there exists a class of diffusion processes that may be transformed to satisfy this assumption. 
+The algorithm proposed in this paper requires an  unbiased estimator of the unknown transition densities and we propose to use  generalized Poisson estimators defined in \cite{fearnhead:papaspiliopoulos:roberts:2008}.
+ The use of generalized Poisson estimators entails that the diffusion coefficient $\Gamma$ is constant and equals to the identity matrix.
+  Note that, when this is not the case, there exists a class of diffusion processes that may be transformed using Lamperti transformation to satisfy this assumption. 
   This paper then focuses on reducible SDE for which there exists an invertible and infinitely differentiable function $\eta$ such that the process $\left\{Z_t = \eta(X_t)\right\}_{t\ge 0}$ satisfies the SDE  $z_0 = \eta(x_0)$ and
@@
 Assumption i) is somewhat restrictive as it requires $\alpha$ to derive from a scalar potential, however, it has natural applications in many fields, for instance in movement ecology, see \cite{gloaguen:etienne:lecorff:2017}.
-Assumption ii) is technical {\color{red} JAI LIMPRESSION QUE CEST TOUJOURS VRAI SI A(X) EST DEFINI SUR $R^d$}.
+Assumption ii) is technical {\color{red} JAI LIMPRESSION QUE CEST TOUJOURS VRAI SI A(X) EST DEFINI SUR $R^d$}.{\color{blue} pkoi  $\triangle A(x)$ serait il borné inférieurement ?}
 
@@
 The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{0\le k \le n}$ in $(\mathbb{R}^m)^{n+1}$. 
-For all $0\le k \le n$, the distribution of $Y_k$ given $X_k:= X_{t_k}$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,\cdot) = g_k(X_k)$. 
+For all $0\le k \le n$, the distribution of $Y_k$ given $X_k$, $X_k$ being a short noation for $X_{t_k}$, has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,\cdot)$.
+For the sake of silmplicity, we will use the classical short notation $g_k(X_k)$ instead of  $g(X_k,\cdot)$.
 The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
@@
 \]
-For all $0\le k\le n$, $\phi_{k} = \phi_{k:k|k}$ denote the filtering distributions. The aim of this section is to detail the extension of PaRIS algorithm to approximate expectations of the form
+For all $0\le k\le n$, $\phi_{k} = \phi_{k:k|k}$ denotes the filtering distributions. The aim of this section is to detail the extension of PaRIS algorithm to approximate expectations of the form
 \begin{equation}
@@
    A more appealing choice is the fully adapted particle filter which sets for all $x,x'\in\mathbb{R}^d$, $p_{k-1}(x,x') \propto q_{k-1}(x,x')g_k(x')$ and for all $x\in\mathbb{R}^d$, $\vartheta(x) = \int q_{k-1}(x,x')g_k(x')\mu(\rmd x')$.
+{\color{blue} C'est si clair que $\vartheta$ est calculable pratiquement ?}
    Here again $q_{k-1}$ has to be replaced by an approximation.

### 2017-09-12 TexSubmission/answer_referees.tex
Small change
End of revision

--- 2017-08-10 TexSubmission/answer_referees.tex
@@
 As pointed by the Referee, one of the best feature of PaRIS algorithm is that it may be implemented "online", processing the observations $(Y_k)_{k\ge 1}$ as they are received, without any storage.
-This may be seen for instance in steps (i)-(iii) in page 6 which describe the way the intermediate quantities $(\tau_k^i)_{1\le i \le N}$ can be computed at each time step.
+This may be seen for instance in steps (i)-(iii) in pages 8-9 which describe the way the intermediate quantities $(\tau_k^i)_{1\le i \le N}$ can be computed at each time step.
 Each time a new observation $Y_{n+1}$ is received, these quantities can be updated only using $Y_{n+1}$, $(\tau_n^i)_{1\le i \le N}$ and the particle filter at time $n$.
@@
 The way the unbiased estimation of $q_k(\xi^{I_k^\ell}_{k-1},\xi^\ell_k)$ was computed was indeed not clear.
-In (7) and at the bottom of page 6 we only stated the existence of a random variable $\zeta_k$ such that $\widehat{q}_k(\xi^{I_k^\ell}_{k-1},\xi^\ell_k;\zeta_k)$ is an unbiased estimation of $q_k(\xi^{I_k^\ell}_{k-1},\xi^\ell_k)$.
+We only stated the existence of a random variable $\zeta_k$ such that $\widehat{q}_k(\xi^{I_k^\ell}_{k-1},\xi^\ell_k;\zeta_k)$ is an unbiased estimation of $q_k(\xi^{I_k^\ell}_{k-1},\xi^\ell_k)$.
 
@@
 \]
-This is now detailed clearly just after (7).
+This is now detailed clearly just after the definition of $\widehat{\omega}_k^\ell$.
 
@@
 
-\textcolor{red}{Il est gentil Michel mais introduire une notation pour ca moi ca ne me plait pas. On repond au reste, je suis d'avis de lui dire merci mais non merci.}
+As noted by the Referee, it is not clear that adding a different notation for the observed points and for the continuous process would improve the readability of the paper.
+As the SMC setting already implies a lot of somehow cumbersome notations, we decided not to add unnecessary notations.
 
@@
 
-The usual FFBSi cannot be extended easily for the same reasons as the FFBS (it relies on the approximation of the backward kernel which is a ratio as defined in (6) in our paper).
+The usual FFBSi cannot be extended easily for the same reasons as the FFBS (it relies on the approximation of the backward kernel which is a ratio as defined  in our paper).
 But it is true that the linear version of the FFBSi may be extended in a similar way as PaRIS algorithm (while it does stil require a backward pass).
@@
 \begin{enumerate}
-\item
+\item {\em 
+The authors should clearly characterise the class of processes for which the proposed
+methodology can be applied. In particular, three issues should be addressed when
+presenting the process in (1). First, condition $\alpha(x) = \nabla_x A(x)$ should be emphasised
+with its validity discussed in the univariate and multivariate cases. Second, it should
+be said that processes with general diffusion coefficients may be considered as long
+as a suitable transformation to lead to a unit coefficient is available. This should
+also be discussed in the univariate and multivariate cases. Third, it could mention
+that $\phi$ (which depends on $\alpha$) should be bounded below in the state space of X. It
+would also be nice to give an insight as to why all these conditions are necessary.
+}
+
+\vspace{.3cm}
+
+We agree with the Referee that the model was not introduced properly and that the use of generalized Poisson estimators required additional comments.
+This remark led to a major modification of Section~2.
+The assumptions mentioned by the Referee are highlighted clearly and much more commented to explain the setting in which GPEs can be used.
+
+As some assumptions may be restrictive (specifically the gradient assumption on the drift), we also provide alternative to GPEs to estimate transition densities.
+The only requirement to extend PaRIS algorithm is to obtain an unbiased estimator than can be evaluated explicitly to compute the importance weights.
+A recent class of estimators which do not use the gradient assumptions are the continuous importance sampling based methods of [14].
+They introduce other sets of assumptions (moment conditions and Holder type assumptions).
+
 \item {\em When presenting equation (2), the authors should explain the importance of approximating those type of functionals.}
@@
 
-\item
-\item
-\item
-\item
+Assumptions (A1)-(A3) are discussed at the end of Section~3.
+As soon as $\phi(\omega_s)$ (or $L_\omega$) is bounded below, (A3) is satisfied  for instance with the GPE based on the  EA3 procedure  which allows to use our algorithm with GPEs to approximate the transition densities.
+When $\phi(\omega_s)$ is not bounded below, the accept reject procedure at the core of the exact algorithms cannot be used which indicates that alternative to GPEs must be used to approximate the transition densities.
+We suggest in Section~2 the use of continuous importance sampling estimators which require other sets of assumptions.
+We did not comment assumptions (A1)-(A3) for these methods as we focus on GPEs in the paper.
+It is important to note that we stress that any unbiased estimator is a candidate to be used in our algorithm and that the way assumptions (A1)-(A3) are checked depends on the specific choice made by the user.
+
+\item {\em The authors should discuss how to obtain $L_w$ and $U_w$, and the efficiency of their
+algorithm in terms of these bounds. In particular, they should reference the layered
+BB construction - Beskos et al. (2008). Also, they can take advantage of the factorisation (into two terms) in (9) to perform the reject-accept step of the algorithm
+in Lemma 1 using two coins - note that the first term does not depend on BB points
+so a fail in the first coin would make the simulation of the BB points unnecessary.}
+
+\vspace{.3cm}
+
+\item {\em It is worth mentioning that (A3) is also satisfied for the EA3 class of diffusions. See
+Beskos et al. (2008).}
+
+\vspace{.3cm}
+
+This is now mentioned at the end of Section~3.
+
+\item {\em Concerning the paragraph starting in page 8 and ending in page 9. Can the authors
+provide some analytical result about when that strategy is better as a function of the
+rejection sampling global accep. prob. and the cost to sample (layered) Brownian
+bridges?}
+
+\vspace{.3cm}
+
+\textcolor{red}{Je propose de ne pas mettre de resultat analytique pour ne pas alourdir la chose (surtout que c'est tres dependant du modele et du GPE utilise).
+On peut par contre donner quelques resultats numeriques, tu as garde les couts de calcul Pierre ?}
+
+\item {\em Item 3 above should include a discussion regarding assumptions H1 and H2.}
+
+\vspace{.3cm}
+
+A few comments were added after these two assumptions.
+Assumption H1 only involves the marginal likelihood $g_k$ of the observations and does not depend on our algorithm.
+In the case where the observations are given as in the practical section, this assumption holds as soon as the variance of the observation is bounded away from zero.
+Assumption H2 depends on the algorithm used to estimate the transition densities and on the tuning parameters of the  SMC filter.
+The most common choice is $\vartheta_k = 1$ so that under H1, the only requirement is to control $\widehat{q}_{k-1}$ and $p_{k-1}$.
+For instance, in the case of the GPE-1, as explained in Section~3, H2 is satisfied if $\phi$ is upper bounded (as for the EA1).
+
 \item {\em Can anything be said about the bias of the proposed estimator for finite $N$? Is it
@@
 In the case of additive functional as in (2), the bias of the forward only version of the FFBS algorithm is shown to be upper bounded by a term of order $n/N$ for strongly mixing hidden Markov models.
+This bias is due to the form of  $\Lambda_k^N$ for the FFBS, FFBSi and PaRIS algorithms.
 
 \item {\em What can be said about a CLT for the proposed estimator?}
-\textcolor{red}{On peut le faire, pas tres dur, mais ca va allonger les choses et sans message tres clair sur la variance asymptotique a mon avis. Voila une suggestion de reponse...}
 
@@
 We added a paragraph at the end of the log-growth section following the work of [4] explaining how the process can be transformed to estimate $\sigma$ using an EM algorithm although $X$ is a function of $\sigma$.
-\textcolor{red}{Pas si simple, la transformation en deux etapes de Beskos et al ne fonctionne pas comme le processus X n'est pas directement observe...}
 

### 2017-09-12 TexSubmission/ParisEM.tex
small typos
End of revision

--- 2017-08-18 TexSubmission/ParisEM.tex
@@
 This paper introduces a new algorithm to approximate smoothed additive functionals for partially observed stochastic differential equations. 
-This method relies on the recent procedure introduced in \cite{olsson:westerborn:2016} which allows to compute such approximations online, i.e.
+This method relies on a new Sequential Monte Carlo method which allows to compute such approximations online, i.e.
 as the observations are received, and with a computational complexity growing linearly with the number of Monte Carlo samples.
-The algorithm of \cite{olsson:westerborn:2016} cannot be used in the case of partially observed stochastic differential equations since the transition density of the latent data is usually unknown. 
-We prove that a similar algorithm may still be defined for partially observed continuous processes by replacing this unknown quantity by an unbiased estimator obtained for instance using general Poisson estimators.
-We prove that this estimator is consistent and its performance are illustrated using data from two models. 
+The original algorithm cannot be used in the case of partially observed stochastic differential equations since the transition density of the latent data is usually unknown.
+We prove that it may be extended to partially observed continuous processes by replacing this unknown quantity by an unbiased estimator obtained for instance using general Poisson estimators. 
+This estimator is consistent and its performance are illustrated using data from two models. 
 \end{abstract}
 
-{\bf Keywords}: Stochastic differential equations, Smoothing, Sequential Monte Carlo Methods.
+{\bf Keywords}: Stochastic differential equations, Smoothing, Sequential Monte Carlo Methods, Online estimation.
 
 \section{Introduction}
-This paper introduces a new algorithm to solve the smoothing problem for hidden Markov models (HMMs) whose hidden state is supposed to be the solution to a stochastic differential equation (SDE).
+This paper introduces a new algorithm to solve the smoothing problem for hidden Markov models (HMMs) whose hidden state is a solution to a stochastic differential equation (SDE).
 In \cite{olsson:strojby:2011}, these models are referred to as partially observed diffusion (POD) processes.
- In this setting, the hidden state process $(X_t)_{t\ge 0}$ is assumed to be a solution to a SDE and the only information available is given by noisy observations $(Y_{k})_{0\le k\le n}$ of the states $(X_k)_{0\le k\le  n}$ at some discrete time points $(t_k)_{0\le k\le n}$.
+The hidden state process $(X_t)_{t\ge 0}$ is assumed to be a solution to a SDE and the only information available is given by noisy observations $(Y_{k})_{0\le k\le n}$ of the states $(X_k)_{0\le k\le  n}$ at some discrete time points $(t_k)_{0\le k\le n}$.
 The bivariate stochastic process $\{(X_{k},Y_{k})\}_{0\le k\le n}$ is a state space model such that conditional on the state sequence $(X_{k})_{0\le k\le n}$ the observations $(Y_{k})_{0\le k\le n}$ are independent and for all $0\le \ell\le n$ the conditional distribution of $Y_{\ell}$ given $\{X_{k}\}_{0\le k\le n}$ depends on $X_{\ell}$ only.
@@
 
-However, for most HMMs, the exact computation of these expectations is usually not possible, and has to be replace by approximations.
-Sequential Monte Carlo (SMC) methods has been a popular class of algorithms to approximate smoothing distributions with random particles associated with importance weights.
+However, in most cases the exact computation of these expectations is usually not possible explicitly.
+Sequential Monte Carlo (SMC) methods are popular algorithms to approximate smoothing distributions with random particles associated with importance weights.
  \cite{gordon:salmond:smith:1993,kitagawa:1996} introduced the first particle filters and smoothers for state space models by combining importance sampling steps to propagate particles with resampling steps to duplicate or discard particles according to their importance weights.
-In the case of HMMs, approximations of the smoothing distributions may also be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}.
+In the case of HMMs, approximations of the smoothing distributions may be obtained using the Forward Filtering Backward Smoothing algorithm (FFBS) and  the Forward Filtering Backward Simulation algorithm (FFBSi) developed respectively in \cite{kitagawa:1996,huerzeler:kunsch:1998,doucet:godsill:andrieu:2000} and \cite{godsill:doucet:west:2004}.
 Both algorithms require first a forward pass which produces a set of particles and weights approximating the sequence of filtering distributions up to time $n$. 
@@
   In the context of SDEs,  discretization procedures may be used to approximate transition densities.
-   For instance, the classical Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998}, or Gaussian based approximations using Taylor expansions of the posterior mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}.
+   For instance, the classical Euler-Maruyama method, the Ozaki discretization which proposes a linear approximation of the drift coefficient between two observations \cite{ozaki:1992,shoji:1998}, or Gaussian based approximations using Taylor expansions of the conditional mean and variance of an observation given the observation at the previous time step, \cite{kessler:1997,kessler:lindner:sorensen:2012,uchida:yoshida:2012}.
    Other approaches based on Hermite polynomials expansion were also introduced by \cite{ait-sahalia:1999,ait-sahalia:2002,ait-sahalia:2008} and extended in several directions recently, see \cite{li:2013} and all the references on the approximation of transition densities therein.
-    However, even the most recent discretization based approximations of the transition densities induce a systematic bias in the approximation of the transition densities (see for instance \cite{delmoral:jacod:protter:2001}).
-    This bias approximation therefore induces a bias of particle based approximations of posterior distributions in POD processes.
+    However, even the most recent discretization based approximations of the transition densities induce a systematic bias in the approximation of the transition densities, see for instance \cite{delmoral:jacod:protter:2001}.
+  %This bias approximation therefore induces a bias of particle based approximations of posterior distributions in POD processes.
 
 To overcome this difficulty, \cite{fearnhead:papaspiliopoulos:roberts:2008} proposed to solve the filtering problem by combining SMC methods with an unbiased estimate of the transition densities based on the generalized Poisson estimator (GPE).
-     In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities, i.e. no discretization scheme is used.
-
+     In this case, only the Monte Carlo error has to be controlled as there is no Taylor expansion to approximate unknown transition densities, i.e.
+no discretization scheme is used.
 The only solution to solve the smoothing problem for POD processes using SMC methods without any discretization procedure has been proposed in \cite{olsson:strojby:2011} and extends the fixed-lag smoother of \cite{olsson:cappe:douc:moulines:2008}.
@@
 
-In this paper, we propose to use Sequential Monte Carlo (SMC) methods to obtain asymptoticaly unbiased approximations of smoothing expectations in the POD processes, by extending the use of  PaRIS algorithm to POD processes.
+In this paper, we propose to use Sequential Monte Carlo (SMC) methods to obtain asymptotically unbiased approximations of smoothing expectations of POD processes by extending PaRIS algorithm.
 The proposed algorithm allows to approximate smoothed expectations of additive functionals online, with a complexity growing only linearly with the number of particles and without any discretization procedure or Taylor expansion of the transition densities.
@@
  The linear version of the FFBSi algorithm proposed in \cite{douc:garivier:moulines:olsson:2011} could be extended in a similar way as PaRIS algorithm but it would still require a backward pass and would not be an online smoother.
- The proposed Generalized Random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to POD processes but also to any general state space model where the transition density of the hidden chain may be approximated using a strictly positive and unbiased estimator.
+ The proposed Generalized Random version of PaRIS algorithm, hereafter named GRand PaRIS algorithm, may be applied to POD processes but also to any general state space model where the transition density of the hidden chain may be approximated using a positive and unbiased estimator.
 
-Section~\ref{sec:rwparis} describes the proposed algorithm to approximate smoothed additive functionals using unbiased estimates of the transition density of the hidden states and details the application of this algorithm when the transition density may be approximated using a GPE.
+Section~\ref{sec:model} describes the model and the smoothing quantities to be estimated.
+Section~\ref{sec:rwparis} provides the algorithm to approximate smoothed additive functionals using unbiased estimates of the transition density of the hidden states.
+This section also details the application of this algorithm when the transition density may be approximated using a GPE.
 In Section~\ref{sec:convergence}, classical convergence results for SMC smoothers are extended to the setting of this paper and illustrated with numerical experiments in Section~\ref{sec:exp}. 
@@
 
-\section{The Generalized Random PaRIS algorithm}
-\label{sec:rwparis}
+\section{Model and framework}
+\label{sec:model}
 %. These estimators are designed for stochastic processes
@@
 where $(W_t)_{t\ge 0}$ is a standard Brownian motion on $\mathbb{R}^d$, $\alpha: \mathbb{R}^d\to\mathbb{R}^d$ and $\Gamma: \mathbb{R}^d\to \mathbb{R}^{d\times d}$.
-In this paper, we focus on unbiased estimator of the unknown transition densities given by generalized Poisson estimators defined in \cite{fearnhead:papaspiliopoulos:roberts:2008}.
- The use of generalized Poisson estimators requires that the diffusion coefficient $\Gamma$ is constant and equals to the identity matrix.
-  Note that, when this is not the case, there exists a class of diffusion processes that may be transformed to satisfy this assumption. 
-  This paper then focuses on reducible SDE for which there exists an invertible and infinitely differentiable function $\eta$ such that the process $\left\{Z_t = \eta(X_t)\right\}_{t\ge 0}$ satisfies the SDE  $z_0 = \eta(x_0)$ and
+The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{0\le k \le n}$ in $(\mathbb{R}^m)^{n+1}$.
+In the following, for all $0 \le k \le n$, the state $X_{t_k}$ at time $k$ is referred to as $X_k$.
+For all $0\le k \le n$, the distribution of $Y_k$ given $X_k$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,\cdot)$.
+For the sake of simplicity, we use the shorthand notation $g_k(X_k)$ for $g(X_k,Y_k)$.
+The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
+For all $0\le k \le n-1$, the conditional distribution of $X_{k+1} $ given $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$.
+
+For all $0 \leq k \leq k' \leq n$, the joint smoothing distributions of the hidden states are defined, for all measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, by:
+\[
+\phi_{k:k'|n}[h] = \mathbb{E}\left[h(X_k,\ldots,X_{k'})\middle|Y_{0:n}\right]
+\]
+and $\phi_{k} = \phi_{k:k|k}$ denotes the filtering distributions. The aim of this paper is to approximate expectations of the form:
+\begin{equation}
+\label{def:addfunc}
+\phi_{0:n\vert n}[H_{n}] = \mathbb{E}\left[H_n(X_{0:n})\middle|Y_{0:n}\right] \text{ where } H_n=\sum_{k=0}^{n-1}h_k(X_k,X_{k+1})\eqsp,
+\end{equation}
+when $\{h_k\}_{k=0}^{n-1}$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$. 
+% the transition density of the hidden states is not available explicitly and where
+Smoothed additive functionals as \eqref{def:addfunc} are crucial for maximum likelihood inference of latent data models. 
+These quantities appear naturally when computing the Fisher score in hidden Markov models or the intermediate quantity of the Expectation Maximization algorithm (see Section~\ref{sec:exp}). 
+They are also pivotal to design online Expectation Maximization based algorithms which motivates the method introduced in this paper that does not require growing storage and can process observations online.
+
+The algorithm proposed in this paper is based on Sequential Monte Carlo methods which offer a flexible framework to approximate such distributions with weighted empirical measures associated with random samples.
+At each time step, the samples are moved randomly in $\mathbb{R}^d$ and  associated with importance weights.
+In general situations, the computation of these importance weights involve the unknown transition density of the process \eqref{eq:target:sde}.
+The solution introduced in Section~\ref{sec:rwparis} requires an unbiased estimator of these unknown transition densities.
+Statistical inference of stochastic differential equations is an active area of research and several solutions have been proposed to design unbiased estimates of transition densities.
+Those estimators require different assumptions on the model \eqref{eq:target:sde}, we provide below several solutions that can be investigated.
+
+\paragraph{General Poisson estimators}
+This paper focuses mainly on GPEs which have been widely used recently and applied in a variety of disciplines.
+These estimators require that the diffusion coefficient $\Gamma$ is constant and equal to the identity matrix, see \cite{fearnhead:papaspiliopoulos:roberts:2008}.
+They may be applied to reducible SDE for which there exists an invertible and infinitely differentiable function $\eta$ such that the process $\left\{Z_t = \eta(X_t)\right\}_{t\ge 0}$ satisfies the SDE  $z_0 = \eta(x_0)$ and
 \begin{equation}
@@
 \]
-In the following, it is assumed that the process $(X_t)_{t\ge 0}$ satisfies the SDE \eqref{eq:target:sde} with $\Gamma$ being the identity matrix, i.e.
+GPEs consider that the process $(X_t)_{t\ge 0}$ satisfies the SDE \eqref{eq:target:sde} with $\Gamma$ being the identity matrix, i.e.
 we consider a diffusion after the application of the Lamperti transform.
- In addition, designing general Poisson estimators also requires that:
+ In addition, designing GPEs also requires that:
 \begin{enumerate}[i)]
@@
 \end{enumerate}
-Assumption i) is somewhat restrictive as it requires $\alpha$ to derive from a scalar potential, however, it has natural applications in many fields, for instance in movement ecology, see \cite{gloaguen:etienne:lecorff:2017}.
-Assumption ii) is technical {\color{red} JAI LIMPRESSION QUE CEST TOUJOURS VRAI SI A(X) EST DEFINI SUR $R^d$}.
+Assumption i) is somewhat restrictive as it requires $\alpha$ to derive from a scalar potential, however it has natural applications in many fields such as movement ecology, see \cite{gloaguen:etienne:lecorff:2017}.
+Assumption ii) is a technical condition which ensures that exact sampling of processes solution to  \eqref{eq:target:sde} using acceptance rejection methods, see for instance \cite{beskos:papaspiliopoulos:roberts:2006,beskos:papaspiliopoulos:roberts:2008,fearnhead:papaspiliopoulos:roberts:2008}.
 
-{\color{red} In the case where these assumptions are not fulfilled, an alternative to generalized Poisson estimators is given by continuous importance sampling procedures for SDE, see \cite{}.
-REFERENCE A METTRE}
+\paragraph{Continuous importance sampling based estimators}
+In the case the previous assumptions are not fulfilled, alternatives to GPEs are given by continuous importance sampling procedures for SDE.
+In \cite{wagner:1989}, for each $0\le k \le n-1$, the transition density between $t_k$ and $t_{k+1}$ is expressed as an infinite expansion obtained using the Kolmogorov backward operator associated with  \eqref{eq:target:sde}.
+This analytical expression of the transition density is not tractable and is estimated by updating random samples at random times using tractable proposal distributions (for instance based on an Euler discretization of the original SDE).
+Then, these samples are associated with random weights to ensure that the proposed estimator is unbiased.
+More recently, \cite{fearnhead:latuszynski:roberts:sermaidis:2017} extended the discrete time importance sampling estimator by introducing updates at random times associated with a renewal process.
+The random samples are weighted using the Kolmogorov forward operator associated with the SDE which relies on the first two order derivatives  of the drift and diffusion coefficients and are therefore tractable.
 
-The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{0\le k \le n}$ in $(\mathbb{R}^m)^{n+1}$. 
-For all $0\le k \le n$, the distribution of $Y_k$ given $X_k:= X_{t_k}$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,\cdot) = g_k(X_k)$. 
-The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
-For all $0\le k \le n-1$, the conditional distribution of $X_{k+1} $ given $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$.
+The unbiasedness of these procedures as long as the controls of the variability of the estimates require moments assumptions and Holder type conditions on the parameters of the SDE \eqref{eq:target:sde}.
+Their efficiency require a fair amount of tuning as they highly depend on the proposal densities used to obtain the Monte Carlo samples and the point processes generating the underlying random times.
 
-Let $0 \leq k \leq k' \leq n$, the joint smoothing distributions of the hidden states are defined, for all measurable function $h$ on $(\mathbb{R}^d)^{k'-k + 1}$, by:
-\[
-\phi_{k:k'|n}[h] = \mathbb{E}\left[h(X_k,\ldots,X_{k'})\middle|Y_{0:n}\right]\eqsp.
-\]
-For all $0\le k\le n$, $\phi_{k} = \phi_{k:k|k}$ denote the filtering distributions. The aim of this section is to detail the extension of PaRIS algorithm to approximate expectations of the form
-\begin{equation}
-\label{def:addfunc}
-\phi_{0:n\vert n}[H_{n}] = \mathbb{E}\left[H_n(X_{0:n})\middle|Y_{0:n}\right] \text{ where } H_n=\sum_{k=0}^{n-1}h_k(X_k,X_{k+1})\eqsp,
-\end{equation}
-when the transition density of the hidden states is not available explicitly and where $\{h_k\}_{k=0}^{n-1}$ are given functions on $\mathbb{R}^d\times \mathbb{R}^d$. 
-Smoothed additive functionals as \eqref{def:addfunc} are crucial for maximum likelihood inference of latent data models. 
-These quantities appear naturally when computing the Fisher score in hidden Markov models or the intermediate quantity of the Expectation Maximization algorithm (see Section~\ref{sec:exp}). 
-They are also pivotal to design online Expectation Maximization based algorithms which motivates the method introduced in this paper that does not require growing storage and can process observations online.
-
+\section{The Generalized Random PaRIS algorithm}
+\label{sec:rwparis}
 The algorithm is based on the following link between the filtering and smoothing distributions for additive functionals, see \cite{olsson:westerborn:2016}:
@@
 
-\subsection*{Bounded estimator of $q_k$}
+\paragraph{Bounded estimator of $q_k$ using GPEs}
 For $x, y \in \mathbb{R}^d$, by Girsanov and Ito's formulas, the transition density $q_k(x,y)$ of \eqref{eq:target:sde} satisfies, with $\Delta_k = t_{k+1}-t_k$,
@@
 When \eqref{assumpt:global} or \eqref{assumpt:ysupport} holds, it can be nonetheless of practical interest to choose the bound $\hat{\sigma}^k_+$ corresponding to \eqref{assumpt:local}. 
-Indeed, this might increase significantly the acceptance rate of the algorithm, and therefore reduce the number of drawings of the random variable $\zeta_k$, which has a much higher cost than the computation of $\rho_{\Delta_k}$, as it requires simulations of Brownian bridges.
-Moreover, this latter option can also avoid numerical optimization if no analytical expression of $\hat{\sigma}_+^k$ is available. 
-In practice, we found this option more efficient in terms of computation time when $N$ has moderate values.
+Indeed, this might increase significantly the acceptance rate of the algorithm, and therefore reduce the number of draws of the random variable $\zeta_k$, which has a much higher cost than the computation of $\rho_{\Delta_k}$, as it requires simulations of Brownian bridges.
+Moreover, this option allows to avoid numerical optimization if no analytical expression of $\hat{\sigma}_+^k$ is available.
+In practice, this seems more efficient in terms of computational time when $N$ has moderate values.
 
@@
 
+Assumption H\ref{assum:boundmodel} only involves the marginal likelihood $g_k$ of the observations and does not depend on the unbiased estimation of the transition density.
+In the case where the observations are given as in the Section~\ref{sec:exp}, this assumption holds as soon as the variance of the observation is bounded away fro zero.
+
 \begin{hypA}
@@
 \end{hypA}
+Assumption H2 depends on the algorithm used to estimate the transition densities and on the tuning parameters of the  SMC filter.
+The most common choice is $\vartheta_k = 1$ so that under H1, the only requirement is to control $\widehat{q}_{k-1}$ and $p_{k-1}$.
+For instance, in the case of the GPE-1, as explained in Section~3, H2 is satisfied if $\phi$ is upper bounded (as for the EA1).
 
@@
 We compare the estimation of the EM intermediate quantity with the one obtained by the fixed lag method of \cite{olsson:strojby:2011}, for different values of the lag (namely, 1,2,5,10,50). 
-The particle approximation of $\mathcal{Q}(\theta,\theta)$ for each model  is computed using each algorithm, see Figure~\ref{fig:res:SINE:EM} for the SINE model (and respectively Figure~\ref{fig:res:LG:EM} for the log-growth model).
+The particle approximation of $\mathcal{Q}(\theta,\theta)$ for each model  is computed using each algorithm, see Figure~\ref{fig:res:SINE:EM} for the SINE model and Figure~\ref{fig:res:LG:EM} for the log-growth model.
 This estimation is performed 200 times to obtain the estimates $\hQ_1,\dots,\hQ_{200}$, using $\tilde{N}=2$ particles for PaRIS algorithm, and $M=30$ replications for the Monte Carlo approximation $\widehat q_k$ of each $q_k$.
@@
 In order to obtain a reference value for our study, the intermediate quantity of the EM algorithm is also estimated 30 times using the GRand PaRIS algorithm with $N=5000$ particles, the reference value is then computed as the arithmetic mean of these 30 estimations, and denoted by $\hQ_\star$.
-Figure \ref{fig:res:SINE:EM} (resp. \ref{fig:res:LG:EM}) shows this estimate for an example on one simulated data set. 
+Figures \ref{fig:res:SINE:EM} and \ref{fig:res:LG:EM} display this estimate for an example with one simulated data set. 
 The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=1600$ so that both estimations require similar computational times.
@@
 \end{equation}
-This simple model has no explicit transition density, however GPE estimators may be computed by simulating Brownian bridges.
+This simple model has no explicit transition density, however GPEs may be computed by simulating Brownian bridges.
 The process solution to \eqref{eq:Lamp:SINE} is observed regularly at times $t_0=0,\ldots,t_{100}=50$ through the observation process $(Y_k)_{0\leq k \leq 100}$:
@@
 where the $(\varepsilon_k)_{0\leq k \leq 100}$ are i.i.d. $\mathcal{N}(0, \sigma^2_\text{obs})$, the resulting set of model parameters is $\theta = (\mu, \sigma_\text{obs})$.
-In the example displayed on Figure \ref{fig:res:SINE:obs}, we set $\mu=0$ an d $ \sigma_\text{obs} = 1$.
+In the example displayed in Figure \ref{fig:res:SINE:obs}, we set $\mu=0$ and $ \sigma_\text{obs} = 1$.
 In that case, the function $\rho_{\Delta_k}$ defined in \eqref{eq:rho:func} can be upper bounded either on $(x,y)$ or only on $y$, the GRand PaRIS algorithm has therefore a linear complexity.
@@
 \end{align*}
-where $m(\hQ^s)$ and $\sigma(\hQ^s)$ are the empirical mean and standard deviation of the sample $Q_1^s,\dots,Q_{200}^s$.
+where $m(\hQ^s)$ and $\sigma(\hQ^s)$ are the empirical mean and standard deviation of the sequence $Q_1^s,\dots,Q_{200}^s$.
 For each estimation method, the resulting distributions of $\mathsf{arb}_1,\dots,\mathsf{arb}_{100}$ and $\mathsf{acv}_1,\dots,\mathsf{acv}_{100}$  are shown on Figures \ref{fig:mult:SINE:b} and  \ref{fig:mult:SINE:cv}.
@@
 The Grand PaRIS algorithm is used to perform the E step, with the same settings as before for $N$, $\tilde{N}$ and $M$.
-As their is no closed form solution to compute the M step of the EM algorithm and propose new parameter estimates, we use a Generalized EM procedure: given the current estimation $\theta^{(k)} := (\mu^{(k)}, \sigma_{\text{obs}}^{(k)}$, the function $\mathcal{Q}(\cdot, \theta^{(k)})$ is approximated for $50$ new candidates $\theta_1, \dots, \theta_{50} $ chosen by the user.
+As there is no closed form solution to compute the M step of the EM algorithm and propose new parameter estimates, we use a Generalized EM procedure: given the current estimation $\theta^{(k)} := \left(\mu^{(k)}, \sigma_{\text{obs}}^{(k)} \right)$, the function $\mathcal{Q}(\cdot, \theta^{(k)})$ is approximated for $50$ new candidates $\theta_1, \dots, \theta_{50} $ chosen by the user.
 The new estimate is set as
@@
 $\sigma_\text{obs}$) uniformly chosen in  $(-\pi,\pi)$ (resp.
-in $(0,6)$), showing a convergence after only a few iterations of the EM procedure.
+in $(0,6)$), illustrating a convergence after only a few iterations of the EM procedure.
 
@@
 \includegraphics[scale=0.5]{figure_sigma2}
-\caption{{\em SINE model - EM}. Estimation of $\sigma^2$.}
+\caption{{\em SINE model - EM}. Estimation of $\sigma_\text{obs}$.}
 \label{fig:SINE:sigma}
@@
 \begin{equation}
-\rmd X_t = \overbrace{\left( \frac{\sigma}{2} -  \frac{\kappa}{\sigma} + \frac{\kappa}{\gamma\sigma}\exp\left(-\sigma X_t\right)\right)}^{:=\alpha(X_t)}\rmd t +\rmd W_t,~~X_0=x_0=\eta(z_0).\label{eq:Lamp:LG}
+\rmd X_t = \overbrace{\left( \frac{\sigma}{2} -  \frac{\kappa}{\sigma} + \frac{\kappa}{\gamma\sigma}\exp\left(-\sigma X_t\right)\right)}^{:=\alpha(X_t)}\rmd t +\rmd W_t,~~X_0=x_0=\eta(z_0)\eqsp.\label{eq:Lamp:LG}
 \end{equation}
@@
 
-The results for the fixed lag technique are similar to the ones presented in \cite[Figure 1]{olsson:strojby:2011} on the same model. 
+The results for the fixed lag technique are similar to the ones presented in \cite[Figure 1]{olsson:strojby:2011} using the same model. 
 For small lags, the variance of the estimates is small, but the estimation is highly biased. 
@@
  The main result of the article for practical applications is that the mechanism of this procedure remains valid when the transition density is approximated by a an unbiased positive estimator.
-  The proposed procedure therefore extend the PaRIS algorithm to HMM whose Markov transition density is unknown and can be approximated without bias.
-The GRand PaRIS algorithm outperforms the existing fixed lag smoother for POD processes of \cite{olsson:strojby:2011}, as it does not introduce an intrinsic and non vanishing bias.
+  The proposed procedure therefore extends the PaRIS algorithm to HMMs whose transition density is unknown and can be unbiasedly approximated.
+The GRand PaRIS algorithm outperforms the existing fixed lag smoother for POD processes of \cite{olsson:strojby:2011}, as it does not introduce any intrinsic and non vanishing bias.
    In addition, numerical simulations highlight a better variance using data from two different models.
@@
 \bibliographystyle{plain}
+%\bibliography{Paris_EM}
 \begin{thebibliography}{10}
@@
 
+\bibitem{beskos:papaspiliopoulos:roberts:2006}
+A.~Beskos,  O.~Papaspiliopoulos, and G.O.~Roberts.
+\newblock Retrospective exact simulation of diffusion sample paths with applications.
+\newblock {\em Bernoulli}, 12(6):1077:1098, 2006.
+
+\bibitem{beskos:papaspiliopoulos:roberts:2008}
+A.~Beskos, O.~Papaspiliopoulos, G.O.~Roberts.
+\newblock A factorisation of diffusion measure and finite sample path constructions.
+\newblock {\em Methodology and {C}omputing in {A}pplied {P}robability}, 10(1):85--104, 2008.
+
 \bibitem{beskos:papaspiliopoulos:roberts:fearnhead:2006}
-A.~Beskos, O.~Papaspiliopoulos, G.~Roberts, and P.~Fearnhead.
+A.~Beskos, O.~Papaspiliopoulos, G.O.~Roberts, and P.~Fearnhead.
 \newblock Exact and computationally efficient likelihood-based estimation for
@@
 \bibitem{fearnhead:papaspiliopoulos:roberts:2008}
-P.~Fearnhead, O.~Papaspiliopoulos, and G~Roberts.
+P.~Fearnhead, O.~Papaspiliopoulos, and G.O.~Roberts.
 \newblock Particle filters for partially observed diffusions.
 \newblock {\em J. Roy. Statist. Soc. Ser. B}, 70(4):755--777, 2008.
+
+\bibitem{fearnhead:latuszynski:roberts:sermaidis:2017}
+P.~Fearnhead, K.~Latuszynski, G.O.~Roberts and G.~Sermaidis.
+\newblock Continuous-time Importance Sampling: {M}onte {C}arlo methods which avoid time-discretisation error.
+\newblock {\em Technical report}, 2017.
 
@@
 \bibitem{pitt:shephard:1999}
-M.~K. Pitt and N.~Shephard.
+M.K. Pitt and N.~Shephard.
 \newblock Filtering via simulation: Auxiliary particle filters.
@@
 \bibitem{doucet:poyiadjis:singh:2011}
-G.~Poyiadjis, A.~Doucet, and S.S Singh.
+G.~Poyiadjis, A.~Doucet, and S.S.~Singh.
 \newblock {Particle approximations of the score and observed information matrix
@@
 \bibitem{uchida:yoshida:2012}
-Masayuki Uchida and Nakahiro Yoshida.
+M.~Uchida and N.~Yoshida.
 \newblock Adaptive estimation of an ergodic diffusion process based on sampled
@@
 
+\bibitem{wagner:1989}
+W.~Wagner.
+\newblock Unbiased {M}onte {C}arlo estimators for functionals of weak solutions of stochastic differential equations.
+\newblock {\em Stochastics and {S}tochastics {R}eports}, 28:1--20, 1989.
+
 \end{thebibliography}

### 2017-09-13 TexSubmission/ParisEM.tex
small changes to precise the necessity of a positive and unbiased estimator. Large changes in the SINE model section to precise the trade off for the choose of A1 bound or A3 bound. Plus answer to reviewers. Two figures to illustrate the choice of bound trade off

@@
 \begin{abstract}
-This paper introduces a new algorithm to approximate smoothed additive functionals for partially observed stochastic differential equations. 
+This paper introduces a new algorithm to approximate smoothed additive functionals for partially diffusion processes. 
 This method relies on a new Sequential Monte Carlo method which allows to compute such approximations online, i.e.
@@
 
-In this paper, we propose to use Sequential Monte Carlo (SMC) methods to obtain asymptotically unbiased approximations of smoothing expectations of POD processes by extending PaRIS algorithm.
+In this paper, we propose to use SMC methods to obtain asymptotically unbiased approximations of smoothing expectations of POD processes by extending the PaRIS algorithm.
 The proposed algorithm allows to approximate smoothed expectations of additive functionals online, with a complexity growing only linearly with the number of particles and without any discretization procedure or Taylor expansion of the transition densities.
- The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of PaRIS algorithm to POD processes is that the acceptance rejection mechanism introduced in \cite{douc:garivier:moulines:olsson:2011} ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
+ The crucial and simple result (Lemma~\ref{lem:AR:unbiased}) of the application of the PaRIS algorithm to POD processes is that the acceptance rejection mechanism introduced in \cite{douc:garivier:moulines:olsson:2011} ensuring the linear complexity of the procedure is still correct when the transition densities are replaced by unbiased estimates.
  The usual FFBS and FFBSi algorithms may not be extended this easily since they both require the computation of weights defined as ratios involving the transition densities, thus replacing these unknown quantities by unbiased estimates does not lead to unbiased estimators of the weights.
@@
 In general situations, the computation of these importance weights involve the unknown transition density of the process \eqref{eq:target:sde}. 
-The solution introduced in Section~\ref{sec:rwparis} requires an unbiased estimator of these unknown transition densities.
+The solution introduced in Section~\ref{sec:rwparis} requires an unbiased estimator of these unknown transition densities. Moreover, this estimator must be almost surely positive and upper bounded.
 Statistical inference of stochastic differential equations is an active area of research and several solutions have been proposed to design unbiased estimates of transition densities. 
@@
 Assumption ii) is a technical condition which ensures that exact sampling of processes solution to  \eqref{eq:target:sde} using acceptance rejection methods, see for instance \cite{beskos:papaspiliopoulos:roberts:2006,beskos:papaspiliopoulos:roberts:2008,fearnhead:papaspiliopoulos:roberts:2008}.
-
+In addition to provide an unbiased estimate of the transition density, the GPE ensure that this estimate is almost surely positive. 
+Moreover, as detailled below, one can define, under certain conditions, a GPE that is almost surely upper bounded.
 \paragraph{Continuous importance sampling based estimators}
@@
 Their efficiency require a fair amount of tuning as they highly depend on the proposal densities used to obtain the Monte Carlo samples and the point processes generating the underlying random times.
-
+In addition to unbiasedness, the proposed algorithm in this work requires that the estimator of the transition density is almost surely positive and upper bounded. 
+It is beyond the scope of this paper to define a continuous importance sampling estimator with such properties, but this could be an interesting perspective if it requires less assumptions than the GPE.
 \section{The Generalized Random PaRIS algorithm}
@@
 \end{equation}
-On the r.h.s.
-of \eqref{eq:GPE1}, the product over $\kappa$ elements is bounded by 1, therefore, a sufficient condition to satisfy of the assumptions \eqref{assumpt:global}-\eqref{assumpt:local} is that the function:
-\begin{align}
+On the r.h.s. of \eqref{eq:GPE1}, the product over $\kappa$ elements is bounded by 1.
+ Therefore, a sufficient condition to satisfy one of the assumptions \eqref{assumpt:global}-\eqref{assumpt:local} is that the function:
+\begin{align*}
 \rho_{\Delta_k}:\eqsp\mathbb{R}^d\times \mathbb{R}^d &\mapsto \mathbb{R}\nonumber\\
-(x,y)&\mapsto \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}\label{eq:rho:func}
-\end{align}
+(x,y)&\mapsto \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}
+\end{align*}
 is upper bounded almost surely by $\hat{\sigma}^k_+$.
-In particular, if $\Lo_\mw$ is bounded below almost surely, \eqref{eq:rho:func} always satisfies assumption \eqref{assumpt:local} and Algorithm~\ref{alg:Ozaki:PaRIS} can be used. 
-This condition is always satisfied for models in the domains $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, i.e.\ domains for which the exact algorithms EA1 and EA2 can be used.
+In particular, if $\Lo_\mw$ is bounded below almost surely, \eqref{eq:GPE1} always satisfies assumption \eqref{assumpt:local} and Algorithm~\ref{alg:Ozaki:PaRIS} can be used. 
+This condition is always satisfied for models in the domains required for the applications of  exact algorithms EA1, EA2 and EA3 defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}.
 
@@
 
-Assumption H\ref{assum:boundmodel} only involves the marginal likelihood $g_k$ of the observations and does not depend on the unbiased estimation of the transition density.
-In the case where the observations are given as in the Section~\ref{sec:exp}, this assumption holds as soon as the variance of the observation is bounded away fro zero.
+Assumption {\bf H\ref{assum:boundmodel}} only involves the marginal likelihood $g_k$ of the observations and does not depend on the unbiased estimation of the transition density. 
+In the case where the observations are given as in the Section~\ref{sec:exp}, this assumption holds as soon as the variance of the observation is bounded away from zero.
 
@@
 \end{hypA}
-Assumption H2 depends on the algorithm used to estimate the transition densities and on the tuning parameters of the  SMC filter.
+Assumption {\bf H\ref{assum:boundalgo}} depends on the algorithm used to estimate the transition densities and on the tuning parameters of the  SMC filter. 
 The most common choice is $\vartheta_k = 1$ so that under H1, the only requirement is to control $\widehat{q}_{k-1}$ and $p_{k-1}$. 
-For instance, in the case of the GPE-1, as explained in Section~3, H2 is satisfied if $\phi$ is upper bounded (as for the EA1).
+For instance, in the case of the GPE-1, as explained in Section~3, {\bf H\ref{assum:boundalgo}} is satisfied if $\phi$ is upper bounded (as for the EA1).
 
@@
 Figures \ref{fig:res:SINE:EM} and \ref{fig:res:LG:EM} display this estimate for an example with one simulated data set. 
-The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=1600$ so that both estimations require similar computational times.
+The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=1600$ so that both estimations require similar computational times, resulting a fair comparison.
+On a personnal computer \footnote{(i7-6600U CPU @ 2.60GHz)}, for the parameters mentionned above, it takes around 25 seconds per E step to be performed.
 
@@
 In the example displayed in Figure \ref{fig:res:SINE:obs}, we set $\mu=0$ and $ \sigma_\text{obs} = 1$.
-In that case, the function $\rho_{\Delta_k}$ defined in \eqref{eq:rho:func} can be upper bounded either on $(x,y)$ or only on $y$, the GRand PaRIS algorithm has therefore a linear complexity.
+
+In the case of the SINE model, the estimator $\widehat{q}_k(\cdot)$ defined by equation \eqref{eq:GPE1} satisfies both \eqref{assumpt:global} and \eqref{assumpt:ysupport}. 
+Therefore, the corresponding bound $\widehat{\sigma}_+^k$ can be obtained using numerical optimization. 
+If that bound is chosen, the GRand PaRIS algorithm has linear complexity in the number of particles. 
+As an alternative, it is worth noting here that the bound $\widehat{\sigma}_+^k$ defined by \eqref{assumpt:local} can also be used.
+This method has a quadratic cost in the number of particles, but provides the optimal bound for the algorithm of Lemma \ref{lem:AR:unbiased}, and therefore reduces the expected time before acceptance, specially when the time step $\Delta_k$ is large.
+Therefore, depending on the number of particles wanted and the time step $\Delta_k$, one has to choose whether to use the  $\widehat{\sigma}_+^k$ defined by \eqref{assumpt:global}, \eqref{assumpt:ysupport} or \eqref{assumpt:local}.
+In the configuration presented here, both bounds resulted in an equivalent computation time.
 
@@
 Figures~\ref{fig:SINE:theta} and~\ref{fig:SINE:sigma} illustrate the performance of the estimation for 12 different initializations of $\mu$ (resp.
-$\sigma_\text{obs}$) uniformly chosen in  $(-\pi,\pi)$ (resp.
-in $(0,6)$), illustrating a convergence after only a few iterations of the EM procedure.
+$\sigma_\text{obs}$) uniformly chosen in  $]-\pi,\pi[$ (resp.
+in $]0,6[$), illustrating a convergence after only a few iterations of the EM procedure.
 
@@
 The parameters are given by $$\theta =(\kappa=0.1,\sigma=0.1,\gamma=1000,\sigma_{obs} = 2)\eqsp.$$
-In that case, the $\rho_{\Delta_k}$ function defined in \eqref{eq:rho:func} can be upper bounded as a function of $y$ when $x\in \{\xi_k^1,\dots, \xi_k^N\}$, the GRand PaRIS algorithm has therefore a linear complexity.
+In the case of the Log-growth model, the estimator $\widehat{q}_k(\cdot)$ defined by equation \eqref{eq:GPE1} satisfies \eqref{assumpt:ysupport}, leading to a GRand PaRIS algorithm with linear complexity in the number of particles.
+However, the remarks about the bound $\widehat{\sigma}_+^k$
+ made for the SINE model above still hold in this case.  
 The intermediate quantity of the EM algorithm is evaluated  as for the SINE model, see Figures~\ref{fig:res:LG:EM},~\ref{fig:mult:LG:b} and~\ref{fig:mult:LG:cv}.
@@
 P.~ Gloaguen, M.-P.~\'Etienne, and S.~Le~Corff.
-\newblock Inference procedures to estimate potential based movement models in ecology.
+\newblock Stochastic differential equation based on a multimodal potential to model movement data in ecology\textit{•}.
 \newblock {\em To appear in the {J}ournal of the {R}oyal {S}tatistical {S}ociety: {S}eries {C}}.
@@
   models: the {PaRIS} algorithm.
-\newblock {\em ArXiv:1412.7550}, 2016.
+\newblock {\em Bernoulli}, 3:1951--1996, 2017 .
 

### 2017-09-14 TexSubmission/answer_referees.tex

@@
 \renewcommand\theequation{\textbf{A}\thedefcounter}
+\usepackage{float}
 \begin{equation}}
@@
 The specific approximation of the fully adapted particle filter is also illustrated in the numerical section.
-\textcolor{red}{Pour le sigma Pierre tu as peut-etre plus d'elements.
-On peut bien sur parler des conditions de l'EA mais je pense qu'une intuition pratique est plus percutante ici.}
+The $\widehat{\sigma}_+^k$ is an upperbound that must be computed depending on which of the assumptions (A1)-(A3) is satisfied, it is not, in this sens a user's choice.
+However, as discussed at the end of page 8, it can be better, even if (A1) is satistified, to choose the bound defined by (A3).
+The first one would (in general) require numerical optimization and provide a general bound (potentially far from the one obtained with the set of particles).
+The other would require $N^2$ computations ($N$ being the number of particles), but provides the optimal bound for the acceptance reject algorithm of Lemma 1.
+In addition to the existing paragraph, this is more explicitely discussed now in the numerical section, specially for the SINE model.
 
@@
 
-\textcolor{red}{Je propose de ne pas mettre de resultat analytique pour ne pas alourdir la chose (surtout que c'est tres dependant du modele et du GPE utilise).
-On peut par contre donner quelques resultats numeriques, tu as garde les couts de calcul Pierre ?}
+This important problem is model dependent. Indeed, if assumption $(A1)$ is satisfied, the user has to
+For each $k$, to perform the rejection sampling, the user has to either
+\begin{enumerate}
+\item Find $\widehat{\sigma}_+^k$ by numerical optimization in the (A1) and (A2) case, and set for every particle $i$, $\widehat{\sigma}_+^{k, i } = \widehat{\sigma}_+^k$.
+\item Or, for each particle $i$, find a $\widehat{\sigma}_+^{k, i } = \mathrm{sup}_{j,\zeta}\;\widehat{q}_k(\xi^j_k,\xi^i_{k+1},\zeta)$, in the (A3) case.
+\end{enumerate}
+For each $i$, the corresponding bound will be used to perform Lemma 1.\\
+The first bound is uniform and do not depend on the particles sampled, therefore can be used for every particle $i$.
+However, the resulting bound  can be "large"(with regards to the simulated set of particles) for the AR algorithm of Lemma 1.
+The second bound requires $N$ computations per particle (therefore $N^2$ computations). However, it is clear that this second bound gives the optimal bound for Lemma 1.\\
+Actually, the first bound is more likely to be "too large" when the time step is large.
+Furthermore, in that case, a rejection in Lemma 1 is more costly as it would require sampling another Brownian bridge skeleton (which would be large due to the large time step).\\
+Overall, it seems difficult to have analytical result about which $\widehat{\sigma}_+^k$ to choose, as it depends both on the number of particles and the sampling time step, and that the bound given by (A3) is a random variable.\\
+However, we did an empirical study on the SINE model, that satisfies A1 and A3. Results are presented in Figure \ref{Figure}.
+
+ As expected from what is said above, for a large time step, even for a large number of particles, the bound given by A3, despite its quadratic cost, leads to less computationnal time than the uniform bound A1.
+ On the other hand, for a small time step, the uniform bound of A1 is good enough, and it's better to choose this bound instead of the optimal bound A3.
+\begin{figure}[H]
+\centering
+\begin{tabular}{cc}
+$\Delta = 0.5$ & $\Delta = 2$\\
+\includegraphics[width = 0.4\textwidth]{Times_Delta05}&
+\includegraphics[width = 0.4\textwidth]{Times_Delta2}
+\end{tabular}
+\caption{Empirical distribution of required time (in logarithmic scale of milliseconds) for the acceptation of a candidate in the acceptance rejection algorithm of Lemma 1, depending on which bound is chosen.
+A1 is when the uniform bound is chosen, (see the paper), A3 is when the particle based bound is chosen.
+One can see that the best choice depends on the number of particles (as expected) but also on the time step.
+A3 gives the optimal bound, but a larger computationnal cost.}\label{Figure}
+\end{figure}
+This is now mentionned (without this empirical study) in the numerical experiments section (in the SINE model subsection).
+We believe that it might be confusing to insist too much on this point by adding this study.
 
@@
 
-\textcolor{red}{M. Gloaguen, your call !}
+The only cost difference for two lags $\delta_1 < \delta_2$ is the memory space to keep tracks of particles between times $t+\delta_1$ and $t+\delta_2$ (basically, a $N*(\delta_2 - \delta_1)$ matrix).
+Therefore the computationnal time is identical for two different lags.\\
+On a personnal laptop  (i7-6600U CPU @ 2.60GHz), for the parameters given in the paper, it took around 25 seconds for each E step.
+As mentionned in the paper, all the comparisons are made for a same amount of computation time (that is why there is more particles used in the fixed lag techniques).
+This is now clarified in the paper.
 

### 2017-09-14 TexSubmission/ParisEM.tex

@@
    In addition, numerical simulations highlight a better variance using data from two different models.
-   It can be implemented for the class of models $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} with a linear complexity in $N$.
+   It can be implemented for the class of models for which exact algorithms of \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} are valids, with a linear complexity in $N$ in the best cases, or at worse in $N^2$.
 

### 2017-09-14 TexSubmission/answer_referees.tex
Last modifications

--- 2017-09-12 TexSubmission/answer_referees.tex
@@
 The specific approximation of the fully adapted particle filter is also illustrated in the numerical section.
-\textcolor{red}{Pour le sigma Pierre tu as peut-etre plus d'elements.
-On peut bien sur parler des conditions de l'EA mais je pense qu'une intuition pratique est plus percutante ici.}
+The $\widehat{\sigma}_+^k$ is an upper bound that must be computed depending on which of the assumptions (A1)-(A3) is satisfied, its choice  is not in this sens completely left to the  user.
+However, as discussed at the end of page 8, it can be better, even if (A1) is satisfied, to choose the bound defined by (A3).
+The first one would (in general) require numerical optimization and provide a general bound (potentially far from the one obtained with the set of particles).
+The other would require $N^2$ computations ($N$ being the number of particles), but provides the optimal bound for the acceptance reject algorithm of Lemma 1.
+In addition to the existing paragraph, this is more explicitely discussed now in the numerical section, specially for the SINE model.
 
@@
 As soon as $\phi(\omega_s)$ (or $L_\omega$) is bounded below, (A3) is satisfied  for instance with the GPE based on the  EA3 procedure  which allows to use our algorithm with GPEs to approximate the transition densities.
-When $\phi(\omega_s)$ is not bounded below, the accept reject procedure at the core of the exact algorithms cannot be used which indicates that alternative to GPEs must be used to approximate the transition densities.
+When $\phi(\omega_s)$ is not bounded below, this may indicate that alternative to GPEs should be used to approximate the transition densities.
 We suggest in Section~2 the use of continuous importance sampling estimators which require other sets of assumptions.
@@
 \vspace{.3cm}
+
+The performance of the estimator of the transition density depends on the choice of $\Lo_\mw$ and $\U_\mw$ which is specific to the SDE.
+In the case of the models analyzed in the numerical section these bounds are discussed in [13] for the SINE model and in [27] for the log-growth model.
+We added a remark to stress the fact that  in the case where $\phi$ is not upper bounded, [5] proposed the EA3 algorithm.
+This {\em layered Brownian bridge} construction first samples random variables to determine in which {\em layer} the Brownian bridge lies before simulating the bridge conditional on the event that it belongs to the layer.
+By continuity of $\phi$, $\Lo_\mw$ and  $\U_\mw$ can be computed easily.
+
+We agree with the Referee that the specific form of the GPE estimate allows to improve (slightly) its computational cost.
+Nonetheless, as discussed after Lemma~1 and in the numerical section, the cost of the AR algorithm depends mainly on the choice of the upper bound for $\widehat{q}_k$ (see answer to item 7.
+with numerical experiments).
 
@@
 
-\textcolor{red}{Je propose de ne pas mettre de resultat analytique pour ne pas alourdir la chose (surtout que c'est tres dependant du modele et du GPE utilise).
-On peut par contre donner quelques resultats numeriques, tu as garde les couts de calcul Pierre ?}
+This important problem is model dependent. Indeed, if assumption $(A1)$ is satisfied, to perform the rejection sampling, the user has to either
+\begin{enumerate}
+\item find $\widehat{\sigma}_+^k$ by numerical optimization in the (A1) and (A2) case, and set for every particle $i$, the upper bound to $\widehat{\sigma}_+^k$ ;
+\item or, for each particle $i$, find a $\widehat{\sigma}_+^{k, i } = \mathrm{sup}_{j,\zeta}\;\widehat{q}_k(\xi^j_k,\xi^i_{k+1},\zeta)$, in the (A3) case.
+\end{enumerate}
+For each $i$, the corresponding bound is then used to apply Lemma 1.
+The first bound is uniform and does not depend on the particles, therefore can be used for every particle $i$.
+However, the resulting bound  can be "large"(with respect to the simulated set of particles) for the AR algorithm of Lemma 1.
+The second bound requires $N$ computations per particle (therefore $N^2$ computations). However, it is clear that this second bound gives the optimal bound for Lemma 1.
+
+Actually, the first bound is more likely to be "too large" when the time step is large.
+Furthermore, in that case, a rejection in Lemma 1 is more costly as it would require sampling another Brownian bridge skeleton (which would be large due to the large time step).
+Overall, it seems difficult to provide an analytical result about which upper bound to choose.
+However, we did an empirical study on the SINE model, that satisfies A1 and A3.
+Results are presented in Figure \ref{Figure}.
+
+ As expected from what is said above, for a large time step, even for a large number of particles, the bound given by A3, despite its quadratic cost, leads to less computational time than the uniform bound A1.
+ On the other hand, for a small time step, the uniform bound of A1 is good enough, and it's better to choose this bound instead of the optimal bound A3.
+\begin{figure}[H]
+\centering
+\begin{tabular}{cc}
+$\Delta = 0.5$ & $\Delta = 2$\\
+\includegraphics[width = 0.4\textwidth]{Times_Delta05}&
+\includegraphics[width = 0.4\textwidth]{Times_Delta2}
+\end{tabular}
+\caption{Empirical distribution of required time (in logarithmic scale of milliseconds) for the acceptation of a candidate in the acceptance rejection algorithm of Lemma 1, depending on which bound is chosen.
+A1 is when the uniform bound is chosen and A3 is when the particle based bound is chosen.
+The best choice depends on the number of particles (as expected) but also on the time step.}\label{Figure}
+\end{figure}
+This is now mentioned (without this empirical study) in the numerical experiments section (in the SINE model subsection).
 
@@
 
-\textcolor{red}{M. Gloaguen, your call !}
+The only cost difference for two lags $\delta_1 < \delta_2$ is the memory space to keep tracks of particles between times $t+\delta_1$ and $t+\delta_2$ (basically, a $N*(\delta_2 - \delta_1)$ matrix).
+Therefore the computational time is identical for two different lags.
+On a personal laptop  (i7-6600U CPU @ 2.60GHz), for the parameters given in the paper, it took around 25 seconds for each E step.
+As mentioned in the paper, all the comparisons are made for a same computational time (that is why there is more particles used in the fixed lag techniques).
+This is now clarified in the paper.
 
@@
 \vspace{.3cm}
+
+This was corrected.
 

### 2017-09-14 TexSubmission/ParisEM.tex
Last modifications

@@
 \begin{abstract}
-This paper introduces a new algorithm to approximate smoothed additive functionals for partially diffusion processes. 
+This paper introduces a new algorithm to approximate smoothed additive functionals for partially observed diffusion processes. 
 This method relies on a new Sequential Monte Carlo method which allows to compute such approximations online, i.e.
@@
 We prove that it may be extended to partially observed continuous processes by replacing this unknown quantity by an unbiased estimator obtained for instance using general Poisson estimators. 
-This estimator is consistent and its performance are illustrated using data from two models. 
+This estimator is proved to be consistent and its performance are illustrated using data from two models. 
 \end{abstract}
@@
 This paper introduces a new algorithm to solve the smoothing problem for hidden Markov models (HMMs) whose hidden state is a solution to a stochastic differential equation (SDE).
-In \cite{olsson:strojby:2011}, these models are referred to as partially observed diffusion (POD) processes.
-The hidden state process $(X_t)_{t\ge 0}$ is assumed to be a solution to a SDE and the only information available is given by noisy observations $(Y_{k})_{0\le k\le n}$ of the states $(X_k)_{0\le k\le  n}$ at some discrete time points $(t_k)_{0\le k\le n}$.
+These models are referred to as partially observed diffusion (POD) processes in \cite{olsson:strojby:2011}.
+The hidden state process $(X_t)_{t\ge 0}$ is assumed to be a solution to a SDE and the only information available is given by noisy observations $(Y_{k})_{0\le k\le n}$ of the states $(X_k)_{0\le k\le  n}$ (where $X_k$ stands for $X_{t_k}$) at some discrete time points $(t_k)_{0\le k\le n}$.
 The bivariate stochastic process $\{(X_{k},Y_{k})\}_{0\le k\le n}$ is a state space model such that conditional on the state sequence $(X_{k})_{0\le k\le n}$ the observations $(Y_{k})_{0\le k\le n}$ are independent and for all $0\le \ell\le n$ the conditional distribution of $Y_{\ell}$ given $\{X_{k}\}_{0\le k\le n}$ depends on $X_{\ell}$ only.
@@
 Unlike the FFBS algorithm, the complexity of this algorithm grows only linearly with the number of particles $N$ and contrary to the FFBSi algorithm, no backward pass is required.
- One of the best features of PaRIS algorithm is that it may be implemented online, using the observations $(Y_k)_{k\ge 1}$ as they are received, without any increasing storage requirements.
+ One of the best features of PaRIS algorithm is that it may be implemented online, using the observations $(Y_k)_{k\ge 0}$ as they are received, without any increasing storage requirements.
 
@@
 
-In this paper, we propose to use SMC methods to obtain asymptotically unbiased approximations of smoothing expectations of POD processes by extending the PaRIS algorithm.
+In this paper, we propose to use SMC methods to obtain consistent approximations of smoothing expectations of POD processes by extending the PaRIS algorithm.
 The proposed algorithm allows to approximate smoothed expectations of additive functionals online, with a complexity growing only linearly with the number of particles and without any discretization procedure or Taylor expansion of the transition densities.
@@
 Section~\ref{sec:rwparis} provides the algorithm to approximate smoothed additive functionals using unbiased estimates of the transition density of the hidden states.
-This section also details the application of this algorithm when the transition density may be approximated using a GPE.
+This section also details the application of this algorithm when the transition density are approximated using a GPE.
 In Section~\ref{sec:convergence}, classical convergence results for SMC smoothers are extended to the setting of this paper and illustrated with numerical experiments in Section~\ref{sec:exp}. 
@@
 For all $0\le k \le n$, the distribution of $Y_k$ given $X_k$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,\cdot)$.
-For the sake of simplicity, we use the shorthand notation $g_k(X_k)$ for $g(X_k,Y_k)$.
+For the sake of simplicity, the shorthand notation $g_k(X_k)$ for $g(X_k,Y_k)$ is used.
 The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
@@
 The solution introduced in Section~\ref{sec:rwparis} requires an unbiased estimator of these unknown transition densities. Moreover, this estimator must be almost surely positive and upper bounded.
-Statistical inference of stochastic differential equations is an active area of research and several solutions have been proposed to design unbiased estimates of transition densities. 
+Statistical inference of stochastic differential equations is an active area of research and several solutions have been proposed to design unbiased estimates of these transition densities. 
 Those estimators require different assumptions on the model \eqref{eq:target:sde}, we provide below several solutions that can be investigated.
@@
 In addition to provide an unbiased estimate of the transition density, the GPE ensure that this estimate is almost surely positive.
-Moreover, as detailled below, one can define, under certain conditions, a GPE that is almost surely upper bounded.
+Moreover, as detailed below, under additional conditions, a GPE that is almost surely upper bounded can be defined.
 \paragraph{Continuous importance sampling based estimators}
-In the case the previous assumptions are not fulfilled, alternatives to GPEs are given by continuous importance sampling procedures for SDE. 
+In the case the previous assumptions are not fulfilled, in particular assumption i),  alternatives to GPEs are given by continuous importance sampling procedures for SDE. 
 In \cite{wagner:1989}, for each $0\le k \le n-1$, the transition density between $t_k$ and $t_{k+1}$ is expressed as an infinite expansion obtained using the Kolmogorov backward operator associated with  \eqref{eq:target:sde}.
-This analytical expression of the transition density is not tractable and is estimated by updating random samples at random times using tractable proposal distributions (for instance based on an Euler discretization of the original SDE).
+This analytical expression of the transition density is not tractable and is estimated by updating random samples at random times between $t_k$ and $t_{k+1}$ using tractable proposal distributions (for instance based on an Euler discretization of the original SDE).
 Then, these samples are associated with random weights to ensure that the proposed estimator is unbiased.
 More recently, \cite{fearnhead:latuszynski:roberts:sermaidis:2017} extended the discrete time importance sampling estimator by introducing updates at random times associated with a renewal process.
-The random samples are weighted using the Kolmogorov forward operator associated with the SDE which relies on the first two order derivatives  of the drift and diffusion coefficients and are therefore tractable.
-
-The unbiasedness of these procedures as long as the controls of the variability of the estimates require moments assumptions and Holder type conditions on the parameters of the SDE \eqref{eq:target:sde}.
+The random samples are weighted using the Kolmogorov forward operator associated with the SDE which relies on the first two order derivatives  of the drift and diffusion coefficients (and is therefore tractable).
+
+The unbiasedness of these procedures and  the controls of the variability of the estimates require moments assumptions and Holder type conditions on the parameters of the SDE \eqref{eq:target:sde}. 
 Their efficiency require a fair amount of tuning as they highly depend on the proposal densities used to obtain the Monte Carlo samples and the point processes generating the underlying random times.
 In addition to unbiasedness, the proposed algorithm in this work requires that the estimator of the transition density is almost surely positive and upper bounded.
-It is beyond the scope of this paper to define a continuous importance sampling estimator with such properties, but this could be an interesting perspective if it requires less assumptions than the GPE.
+This implies additional assumptions on the SDE depending on the chosen estimate and could lead to interesting perspectives.
+
 \section{The Generalized Random PaRIS algorithm}
@@
 \end{proof}
-Note that Lemma~\ref{lem:AR:unbiased} still holds if assumption \eqref{assumpt:global} is relaxed and replaced by one of the two following assumptions:
+Note that Lemma~\ref{lem:AR:unbiased} still holds if assumption \eqref{assumpt:global} is relaxed and replaced by:
 \begin{assumpt}\label{assumpt:ysupport}
@@
 \end{assumpt}
-\vspace{-\baselineskip}
+It is worth noting that under assumptions \eqref{assumpt:global} or \eqref{assumpt:ysupport}, the linear complexity property of PaRIS algorithm is ensured.
+The following assumption can also be considered.
+For all $1\le i \le N$,
 \begin{assumpt}\label{assumpt:local}
-\mathrm{sup}_{i,j,\zeta}\;\widehat{q}_k(\xi^j_k,\xi^i_{k+1},\zeta)\leq \hat{\sigma}^k_+\;.
+\mathrm{sup}_{j,\zeta}\;\widehat{q}_k(\xi^j_k,\xi^i_{k+1},\zeta)\leq \hat{\sigma}^{k,i}_+\;.
 \end{assumpt}
-It is worth noting that under assumptions \eqref{assumpt:global} or \eqref{assumpt:ysupport}, the linear complexity property of PaRIS algorithm still holds, whereas if only assumption \eqref{assumpt:local} holds, the algorithm has a quadratic complexity.
+If only assumption \eqref{assumpt:local} holds, the algorithm has a quadratic complexity.
+The bound of \eqref{assumpt:global} is uniform (it does not depend on the particles) and can be used for every particle $1\le i\le N$.
+However, this  bound  can be large (with respect to the simulated set of particles) for the algorithm of Lemma~\ref{lem:AR:unbiased}.
+The bound of \eqref{assumpt:local} requires $N$ computations per particle (therefore $N^2$ computations).
+However, it is clear that this second bound is sharper that the one of  \eqref{assumpt:global} for the acceptance rejection procedure and may lead to a computationally more efficient algorithm.
 
@@
 Assume that there exist random variables $\Lo_\mw$ and $\U_\mw$ such that for all $0\leq s \leq \Delta_k$, $\Lo_\mw \leq \phi(\mw_s)\leq \U_\mw$.
+The performance of the estimator depends on the choice of $\Lo_\mw$ and $\U_\mw$ which is specific to the SDE.
+In the case of the models analyzed in Section~\ref{sec:exp} these bounds are discussed in \cite{fearnhead:papaspiliopoulos:roberts:2008} for the SINE model and in \cite{olsson:strojby:2011} for the log-growth model.
+Note that in the case where $\phi$ is not upper bounded, \cite{beskos:papaspiliopoulos:roberts:2008} proposed the EA3 algorithm.
+This {\em layered Brownian bridge} construction first samples random variables to determine in which {\em layer} the Brownian bridge lies before simulating the bridge conditional on the event that it belongs to the layer.
+By continuity of $\phi$, $\Lo_\mw$ and  $\U_\mw$ can be computed easily.
+
 Let $\kappa$ be a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and $(U_j)_{1\le j\le \kappa}$ be independent uniform random variables on $[0,\Delta_k]$, and $
@@
 
-When \eqref{assumpt:global} or \eqref{assumpt:ysupport} holds, it can be nonetheless of practical interest to choose the bound $\hat{\sigma}^k_+$ corresponding to \eqref{assumpt:local}. 
+When \eqref{assumpt:global} or \eqref{assumpt:ysupport} holds, it can be nonetheless of practical interest to choose the bounds $\hat{\sigma}^{k,i}_+$, $1\le i \le N$, corresponding to \eqref{assumpt:local}.
 Indeed, this might increase significantly the acceptance rate of the algorithm, and therefore reduce the number of draws of the random variable $\zeta_k$, which has a much higher cost than the computation of $\rho_{\Delta_k}$, as it requires simulations of Brownian bridges.
@@
 The GRand Paris algorithm is performed using $N=400$ particles in both cases, the fixed lag technique using $N=1600$ so that both estimations require similar computational times, resulting a fair comparison.
-On a personnal computer \footnote{(i7-6600U CPU @ 2.60GHz)}, for the parameters mentionned above, it takes around 25 seconds per E step to be performed.
+On a personal computer\footnote{(i7-6600U CPU @ 2.60GHz)}, for the parameters mentioned above, it takes around 25 seconds to perform each E step.
 
@@
 
-In the case of the SINE model, the estimator $\widehat{q}_k(\cdot)$ defined by equation \eqref{eq:GPE1} satisfies both \eqref{assumpt:global} and \eqref{assumpt:ysupport}. 
-Therefore, the corresponding bound $\widehat{\sigma}_+^k$ can be obtained using numerical optimization. 
+In the case of the SINE model, the estimator $\widehat{q}_k$ defined by equation \eqref{eq:GPE1} satisfies both \eqref{assumpt:global} and \eqref{assumpt:ysupport}.
+The corresponding bound $\widehat{\sigma}_+^k$ can be obtained using numerical optimization.
 If that bound is chosen, the GRand PaRIS algorithm has linear complexity in the number of particles. 
-As an alternative, it is worth noting here that the bound $\widehat{\sigma}_+^k$ defined by \eqref{assumpt:local} can also be used.
-This method has a quadratic cost in the number of particles, but provides the optimal bound for the algorithm of Lemma \ref{lem:AR:unbiased}, and therefore reduces the expected time before acceptance, specially when the time step $\Delta_k$ is large.
-Therefore, depending on the number of particles wanted and the time step $\Delta_k$, one has to choose whether to use the  $\widehat{\sigma}_+^k$ defined by \eqref{assumpt:global}, \eqref{assumpt:ysupport} or \eqref{assumpt:local}.
-In the configuration presented here, both bounds resulted in an equivalent computation time.
+As an alternative, it is worth noting here that the bounds $\widehat{\sigma}_+^{k,i}$, $1\le i \le N$, defined by \eqref{assumpt:local} can also be used.
+This method has a quadratic cost in the number of particles but provides the optimal bound for the algorithm of Lemma \ref{lem:AR:unbiased}.
+This may reduce significantly the expected time before acceptance, in particular when the time step $\Delta_k$ is large.
+In this experiment e configuration presented here, both bounds resulted in an equivalent computational time.
 
@@
 \]
-Assume the results holds for $k\ge 1$ and that $\vartheta_{k+1} = 1$ for simplicity. Write
+Assume the results holds for $k\ge 0$ and that $\vartheta_{k+1} = 1$ for simplicity. Write
 \[
@@
 P.~ Gloaguen, M.-P.~\'Etienne, and S.~Le~Corff.
-\newblock Stochastic differential equation based on a multimodal potential to model movement data in ecology\textit{•}.
+\newblock Stochastic differential equation based on a multimodal potential to model movement data in ecology.
 \newblock {\em To appear in the {J}ournal of the {R}oyal {S}tatistical {S}ociety: {S}eries {C}}.

### 2017-09-14 TexSubmission/ParisEM.tex
small typo

@@
    In addition, numerical simulations highlight a better variance using data from two different models.
-   It can be implemented for the class of models for which exact algorithms of \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} are valids, with a linear complexity in $N$ in the best cases, or at worse in $N^2$.
+   It can be implemented for the class of models for which exact algorithms of \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006} are valid, with a linear complexity in $N$ in the best cases, or at worse in $N^2$.
 
